### 2023-02-03 products/manuscript/MANU.qmd
project part 1

---
title: "Manuscript Template for a Data Analysis Project"
subtitle: "Exploring Factors influencing HIV Death-to-Case ratio in the United States and Puerto Rico -- 2020"
author: Hayley Hemme
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis_template_references.bib
csl: ../apa.csl
---

The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(tidyverse)
library(readxl)
library(lubridate)
library(stringr)
```

# Summary/Abstract
_Write a summary of your project._

{{< pagebreak >}}

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

## Description of data and data source
The source of

## Questions/Hypotheses to be addressed
Question: Can HIV death rates be predicted by social determinants of health like poverty and/ or by HIV related outcomes (new diagnoses, prevalence PrEP use, etc?)

H0 : There is no correlation between HIV-related deaths and SDOH or HIV-related outcomes.
Ha : There is a correlation between HIV-related deaths and SDOH or HIV-related outcomes.

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a]

{{< pagebreak >}}

# Methods

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._
Loading the datasets from OASIS for deaths and population (Hispanic, NH White, NH Black) for persons ages 13-59 from 2018 to 2020.
```{r warning=FALSE}
setwd("../../")
his_deaths <- read_xlsx("data/raw_data/hivdeaths_his_1820.xlsx", skip = 1)
nhw_deaths <- read_xlsx("data/raw_data/hivdeaths_nhw_1820.xlsx", skip = 1)
nhb_deaths <- read_xlsx("data/raw_data/hivdeaths_nhb_1820.xlsx", skip = 1)

his_pop <- read_xlsx("data/raw_data/pop_his_1820.xlsx", skip = 1)
nhw_pop <- read_xlsx("data/raw_data/pop_nhw_1820.xlsx", skip = 1)
nhb_pop <- read_xlsx("data/raw_data/pop_nhb_1820.xlsx", skip = 1)
```

Loading dataset from AIDSVu for social determinants of health (2020) and new HIV diagnoses from 2018 to 2020.
```{r warning=FALSE}
setwd("../../")
sdoh <- read_xlsx("data/raw_data/AIDSVu_County_SDOH_2020.xlsx", skip = 3)
sdoh <- sdoh %>%  filter(State == "Georgia")

dx18 <- read_xlsx("data/raw_data/AIDSVu_County_NewDX_2018.xlsx", skip = 3)
dx18 <- dx18 %>%  filter(State == "Georgia") %>% 
  select(!contains(c("Rate Stability", "Percent", "Case")))

dx19 <- read_xlsx("data/raw_data/AIDSVu_County_NewDX_2019.xlsx", skip = 3)
dx19 <- dx19 %>%  filter(State == "Georgia")%>% 
  select(!contains(c("Rate Stability", "Percent", "Case")))

dx20 <- read_xlsx("data/raw_data/AIDSVu_County_NewDX_2020.xlsx", skip = 3)
dx20 <- dx20 %>%  filter(State == "Georgia") %>% 
  select(!contains(c("Rate Stability", "Percent", "Case")))
```

Let's load data from Georgia Data on Poverty rates in Georgia by race/ ethnicity for the period 2016-2020.
```{r warning = false}
setwd("../../")
pov <- read_xlsx("data/raw_data/economics_22.xlsx", sheet = "poverty")

glimpse(pov)
```

We'll just be using the data from the "2016-2020 Persons Below Poverty Level" by race/ethnicity. We'll also change the text formatting of County to match the other data frames.
```{r}
pov <- pov %>%  select(1, starts_with("2016-2020 Persons Below Poverty Level"))

pov$County <- str_to_title(pov$County)
```

Let's do some cleaning on the OASIS data. We'll remove extra columns and rows, as well as rows with "Selected ages/sexes total", pivot Year and Deaths into long format, and create the new variable for race/ethnicity.
```{r}
his_deaths <- his_deaths[c(-1,-2387:-2414),-7]
his_deaths <- his_deaths %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>%
  pivot_longer(c(4:6), names_to = "Year", values_to = "HIV_Deaths") %>% 
  mutate("Race_Ethnicity" = "Hispanic")

nhw_deaths <- nhw_deaths[c(-1,-2387:-2414),-7]
nhw_deaths <- nhw_deaths %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>%
  pivot_longer(c(4:6), names_to = "Year", values_to = "HIV_Deaths") %>% 
  mutate("Race_Ethnicity" = "NH White")

nhb_deaths <- nhb_deaths[c(-1,-2387:-2414),-7]
nhb_deaths <- nhb_deaths %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>%
  pivot_longer(c(4:6), names_to = "Year", values_to = "HIV_Deaths") %>% 
  mutate("Race_Ethnicity" = "NH Black")
```

Next, we'll merge the datasets into one dataframe, and remove the original datasets from the environment.
```{r}
deaths <- full_join(his_deaths, nhw_deaths)
deaths <- full_join(deaths, nhb_deaths)

rm(his_deaths, nhw_deaths, nhb_deaths)
```

Next, let's clean up the dataset for population using a similar method.
```{r}
his_pop <- his_pop[c(-1,-2387:-2412),-7]
his_pop <- his_pop %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>% 
  pivot_longer(c(4:6), names_to = "Year", values_to = "Population") %>% 
  mutate("Race_Ethnicity" = "Hispanic")

nhw_pop <- nhw_pop[c(-1,-2387:-2412),-7]
nhw_pop <- nhw_pop %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>% 
  pivot_longer(c(4:6), names_to = "Year", values_to = "Population") %>% 
  mutate("Race_Ethnicity" = "NH White")

nhb_pop <- nhb_pop[c(-1,-2387:-2412),-7]
nhb_pop <- nhb_pop %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>% 
  pivot_longer(c(4:6), names_to = "Year", values_to = "Population") %>% 
  mutate("Race_Ethnicity" = "NH Black")
```

Merging datasets and cleaning the original from the environment.
```{r}
pop <- full_join(his_pop, nhw_pop)
pop <- full_join(pop, nhb_pop)

rm(his_pop, nhw_pop, nhb_pop)
```

Next, let's merge the two OASIS dataframes.
```{r}
df <- full_join(deaths, pop, by = c("Geography", "Sex", "Age", "Year", "Race_Ethnicity"))
```

Next, let's create a new variable for Public Health District, change some variable to factors, and rename the Age variable and Geography variable.
```{r}
df <- df %>% rename(County = Geography) %>% 
  mutate(District = case_when(County %in% c("Bartow", "Catoosa", "Chattooga", "Dade", "Floyd", "Gordon", "Haralson", "Paulding", "Polk", "Walker") ~ "1-1",
 County %in% c("Cherokee", "Fannin", "Gilmer", "Murray", "Pickens", "Whitfield") ~ "1-2",
 County %in% c("Banks", "Dawson", "Forsyth", "Franklin", "Habersham", "Hall", "Hart", "Lumpkin", "Rabun", "Stephens", "Towns", "Union", "White") ~ "2",
 County %in% c("Cobb", "Douglas") ~ "3-1",
 County %in% c("Fulton") ~ "3-2",
 County %in% c("Clayton") ~ "3-3",
 County %in% c("Gwinnett", "Newton", "Rockdale") ~ "3-4",
 County %in% c("DeKalb") ~ "3-5",
 County %in% c("Butts", "Carroll", "Coweta", "Fayette", "Heard", "Henry", "Lamar", "Meriwether", "Pike", "Spalding", "Troup", "Upson") ~ "4",
 County %in% c("Bleckley", "Dodge", "Johnson", "Laurens", "Montgomery", "Pulaski", "Telfair", "Treutlen", "Wheeler", "Wilcox") ~ "5-1",
 County %in% c("Baldwin", "Bibb", "Crawford", "Hancock", "Houston", "Jasper", "Jones", "Monroe", "Peach", "Putnam", "Twiggs", "Washington", "Wilkinson") ~ "5-2",
 County %in% c("Burke", "Columbia", "Emanuel", "Glascock", "Jefferson", "Jenkins", "Lincoln", "McDuffie", "Richmond", "Screven", "Taliaferro", "Warren", "Wilkes") ~ "6",
 County %in% c("Chattahoochee", "Clay", "Crisp", "Dooly", "Harris", "Macon", "Muscogee", "Marion", "Quitman", "Randolph", "Schley", "Stewart", "Sumter", "Talbot", "Taylor", "Webster") ~ "7",
 County %in% c("Ben Hill", "Berrien", "Brooks", "Cook", "Echols", "Irwin", "Lanier", "Lowndes", "Tift", "Turner") ~ "8-1",
 County %in% c("Baker", "Calhoun", "Colquitt", "Dougherty", "Decatur", "Early", "Grady", "Lee", "Miller", "Mitchell", "Seminole", "Terrell", "Thomas", "Worth") ~ "8-2",
 County %in% c("Bryan", "Camden", "Chatham", "Effingham", "Glynn", "Liberty", "Long", "McIntosh") ~ "9-1",
 County %in% c("Appling", "Atkinson", "Bacon", "Brantley", "Bulloch", "Candler", "Charlton", "Clinch", "Coffee", "Evans", "Jeff Davis", "Pierce", "Tattnall", "Toombs", "Ware", "Wayne") ~ "9-2",
 County %in% c("Barrow", "Clarke", "Elbert", "Greene", "Jackson", "Madison", "Morgan", "Oconee", "Oglethorpe", "Walton") ~ "10")) %>% 
   mutate(District = as.factor(District)) %>% 
   mutate(Sex = as.factor(Sex)) %>% 
   mutate(Age = as.factor(Age)) %>% 
   mutate(County = as.factor(County)) %>% 
   mutate(Race_Ethnicity = as.factor(Race_Ethnicity)) %>% 
   rename(Age_Group = Age) %>% 
   mutate(Age_Group = as.factor(Age_Group))
```

Let's clean up the Age_Group variable by removing text
```{r}
df$Age_Group <- str_replace(df$Age_Group, " Adolescence", "")
df$Age_Group <- str_replace(df$Age_Group, " Early Adulthood", "")
df$Age_Group <- str_replace(df$Age_Group, " Young Adulthood", "")
df$Age_Group <- str_replace(df$Age_Group, " Middle Adulthood", "")
```

```{r}
glimpse(df)
```
Things are looking good so far!

Now let's move onto the AIDSVu data. First, we'll merge all new HIV diagnoses from 2018-2020 into one dataframe
```{r}
dx <- full_join(dx18, dx19)
dx <- full_join(dx, dx20)
```

Let's remove columns for race/ethnicities other than Hispanic, NH White, and NH Black, as well as some other variables we won't be using in this analysis. We will also rename county and create the district variable. 
```{r}
dx <- dx %>% select(!contains(c("Asian", "American", "Multiracial", "Hawaiian", "MSM", "2013", "GEO ID", "State")))%>% 
  rename(County = "County Name")

dx$Geography <- str_remove(dx$County, " County")

dx <-  dx %>% mutate(District = case_when(County %in% c("Bartow", "Catoosa", "Chattooga", "Dade", "Floyd", "Gordon", "Haralson", "Paulding", "Polk", "Walker") ~ "1-1",
 County %in% c("Cherokee", "Fannin", "Gilmer", "Murray", "Pickens", "Whitfield") ~ "1-2",
 County %in% c("Banks", "Dawson", "Forsyth", "Franklin", "Habersham", "Hall", "Hart", "Lumpkin", "Rabun", "Stephens", "Towns", "Union", "White") ~ "2",
 County %in% c("Cobb", "Douglas") ~ "3-1",
 County %in% c("Fulton") ~ "3-2",
 County %in% c("Clayton") ~ "3-3",
 County %in% c("Gwinnett", "Newton", "Rockdale") ~ "3-4",
 County %in% c("DeKalb") ~ "3-5",
 County %in% c("Butts", "Carroll", "Coweta", "Fayette", "Heard", "Henry", "Lamar", "Meriwether", "Pike", "Spalding", "Troup", "Upson") ~ "4",
 County %in% c("Bleckley", "Dodge", "Johnson", "Laurens", "Montgomery", "Pulaski", "Telfair", "Treutlen", "Wheeler", "Wilcox") ~ "5-1",
 County %in% c("Baldwin", "Bibb", "Crawford", "Hancock", "Houston", "Jasper", "Jones", "Monroe", "Peach", "Putnam", "Twiggs", "Washington", "Wilkinson") ~ "5-2",
 County %in% c("Burke", "Columbia", "Emanuel", "Glascock", "Jefferson", "Jenkins", "Lincoln", "McDuffie", "Richmond", "Screven", "Taliaferro", "Warren", "Wilkes") ~ "6",
 County %in% c("Chattahoochee", "Clay", "Crisp", "Dooly", "Harris", "Macon", "Muscogee", "Marion", "Quitman", "Randolph", "Schley", "Stewart", "Sumter", "Talbot", "Taylor", "Webster") ~ "7",
 County %in% c("Ben Hill", "Berrien", "Brooks", "Cook", "Echols", "Irwin", "Lanier", "Lowndes", "Tift", "Turner") ~ "8-1",
 County %in% c("Baker", "Calhoun", "Colquitt", "Dougherty", "Decatur", "Early", "Grady", "Lee", "Miller", "Mitchell", "Seminole", "Terrell", "Thomas", "Worth") ~ "8-2",
 County %in% c("Bryan", "Camden", "Chatham", "Effingham", "Glynn", "Liberty", "Long", "McIntosh") ~ "9-1",
 County %in% c("Appling", "Atkinson", "Bacon", "Brantley", "Bulloch", "Candler", "Charlton", "Clinch", "Coffee", "Evans", "Jeff Davis", "Pierce", "Tattnall", "Toombs", "Ware", "Wayne") ~ "9-2",
 County %in% c("Barrow", "Clarke", "Elbert", "Greene", "Jackson", "Madison", "Morgan", "Oconee", "Oglethorpe", "Walton") ~ "10")) %>% 
  mutate(District = as.factor(District))
```

Things are looking pretty good so frame, so let's save these dataframes.
```{r}
saveRDS(df, "../../data/df.rds")
saveRDS(dx, "../../data/dx.rds")
saveRDS(sdoh, "../../data/sdoh.rds")
saveRDS(pov, "../../data/pov.rds")
```

Clear the environment and start fresh.
```{r}

```
## Statistical analysis
Explore central tendency across demographic groups. Fit a poisson regression model predicting death-to-case ratio from percentage with x SDOH variables and percentage of x demographic group.

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by sex."
#| echo: FALSE
knitr::include_graphics(here("results","height_weight_stratified.png"))
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

{{< pagebreak >}}

# References

### 2023-02-03 products/manuscript/Manuscript.qmd
project part 1

--- 2023-02-03 products/manuscript/MANU.qmd
@@
 library(lubridate)
-library(stringr)
 ```
@@
 ## Description of data and data source
-The source of
+The data used in this analysis was retrieved from AIDSVu, a publicly available data source presented by Emory University's Rollins School of Public Health and partners, whose mission is to make HIV data widely available and accessible.
+AIDSVu compiled the data used in this analysis from the CDC.
+These datasets contain data collected at the state-level from 2020 on: new HIV diagnoses, HIV-related deaths, counts/ percentages of those with late HIV diagnoses, viral suppression, PrEP coverage, HIV prevalence, and state testing, and social determinants of health.
+These datasets contain demographic-level information for sex, race/ethnicity, transmission category, among others.
 
 ## Questions/Hypotheses to be addressed
-Question: Can HIV death rates be predicted by social determinants of health like poverty and/ or by HIV related outcomes (new diagnoses, prevalence PrEP use, etc?)
+Question: Can HIV mortality-prevalence ratio at the state-level in the United States be predicted by social determinant of health (ex.
+poverty rates), percentage of people living with HIV (PLWH) who are virally suppressed, and/ or percentage of late diagnoses/ percentage of minority HIV deaths?
 
-H0 : There is no correlation between HIV-related deaths and SDOH or HIV-related outcomes.
-Ha : There is a correlation between HIV-related deaths and SDOH or HIV-related outcomes.
+https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7842853/
+
+where mortality-prevalence ratio = (number of deaths/number of prevalant cases) / prevelance rate
+
+H0 : Death-to-case ratio is not correlated with differences in demographic factors, social determinants of health, testing prevalence, or factors influencing HIV outcomes, such as late diagnoses, viral suppression, and PrEP coverage.
+
+HA : Death-to-case ratio is  correlated with differences in demographic factors, social determinants of health, testing prevalence, or factors influencing HIV outcomes, such as late diagnoses, viral suppression, and PrEP coverage.
+
+Special interest will be given to those disproportionately impacted by HIV, including Non-Hispanic Black populations and the MSM community.
 
@@
 All materials, including files that contain code, should be commented well so everyone can follow along._
-Loading the datasets from OASIS for deaths and population (Hispanic, NH White, NH Black) for persons ages 13-59 from 2018 to 2020.
-```{r warning=FALSE}
-setwd("../../")
-his_deaths <- read_xlsx("data/raw_data/hivdeaths_his_1820.xlsx", skip = 1)
-nhw_deaths <- read_xlsx("data/raw_data/hivdeaths_nhw_1820.xlsx", skip = 1)
-nhb_deaths <- read_xlsx("data/raw_data/hivdeaths_nhb_1820.xlsx", skip = 1)
-
-his_pop <- read_xlsx("data/raw_data/pop_his_1820.xlsx", skip = 1)
-nhw_pop <- read_xlsx("data/raw_data/pop_nhw_1820.xlsx", skip = 1)
-nhb_pop <- read_xlsx("data/raw_data/pop_nhb_1820.xlsx", skip = 1)
+Loading the datasets.
+```{r}
+deaths <- read_xlsx("../../data/raw_data/AIDSVu_State_Death_2020.xlsx", skip = 3)
+dx <- read_xlsx("../../data/raw_data/AIDSVu_State_NewDX_2020.xlsx", skip = 3)
 ```
 
-Loading dataset from AIDSVu for social determinants of health (2020) and new HIV diagnoses from 2018 to 2020.
-```{r warning=FALSE}
-setwd("../../")
-sdoh <- read_xlsx("data/raw_data/AIDSVu_County_SDOH_2020.xlsx", skip = 3)
-sdoh <- sdoh %>%  filter(State == "Georgia")
-
-dx18 <- read_xlsx("data/raw_data/AIDSVu_County_NewDX_2018.xlsx", skip = 3)
-dx18 <- dx18 %>%  filter(State == "Georgia") %>% 
+##Selecting only states/ districts in the South geographic region, as defined by the CDC.
+```{r}
+deaths <- deaths %>% filter(State %in% c("Alabama", "Arkansas", "Delaware", "Washington, D.C.", "Florida", "Georgia", "Kentucky", "Louisiana", "Maryland", "Mississippi", "North Carolina", "Oklahoma", "South Carolina", "Tennessee", "Texas", "Virginia", "West Virginia")) %>%
   select(!contains(c("Rate Stability", "Percent", "Case")))
 
-dx19 <- read_xlsx("data/raw_data/AIDSVu_County_NewDX_2019.xlsx", skip = 3)
-dx19 <- dx19 %>%  filter(State == "Georgia")%>% 
-  select(!contains(c("Rate Stability", "Percent", "Case")))
+sex_deaths_sub <- deaths %>% select(1:7)
+race_deaths_sub <- deaths  %>% select(c(1:5, 8:14))
+age_deaths_sub <- deaths  %>% select(c(1:5, 15:19))
 
-dx20 <- read_xlsx("data/raw_data/AIDSVu_County_NewDX_2020.xlsx", skip = 3)
-dx20 <- dx20 %>%  filter(State == "Georgia") %>% 
-  select(!contains(c("Rate Stability", "Percent", "Case")))
-```
+sex_deaths_sub <- sex_deaths_sub %>% pivot_longer(6:7, names_to = "Sex", values_to = "Death Rate by Sex")
+sex_deaths_sub$Sex <- gsub(" Death Rate", "", sex_deaths_sub$Sex)
+sex_deaths_sub <- sex_deaths_sub %>% mutate_if(is.character, as.factor) %>% 
+  mutate(`GEO ID` = as.factor(`GEO ID`))
 
-Let's load data from Georgia Data on Poverty rates in Georgia by race/ ethnicity for the period 2016-2020.
-```{r warning = false}
-setwd("../../")
-pov <- read_xlsx("data/raw_data/economics_22.xlsx", sheet = "poverty")
+race_deaths_sub <- race_deaths_sub  %>% pivot_longer(6:12, names_to = "Race/Ethnicity", values_to = "Death Rate by Race/Ethnicity")
+race_deaths_sub$`Race/Ethnicity` <- gsub(" Death Rate", "",race_deaths_sub$`Race/Ethnicity`)
+race_deaths_sub <- race_deaths_sub %>% mutate_if(is.character, as.factor) %>% 
+  mutate(`GEO ID` = as.factor(`GEO ID`))
 
-glimpse(pov)
-```
+age_deaths_sub <- age_deaths_sub %>% pivot_longer(6:10, names_to = "Age_Group", values_to = "Death Rate by Age Group")
+age_deaths_sub$Age_Group <- gsub("Age ", "", age_deaths_sub$Age_Group)
+age_deaths_sub$Age_Group <- gsub(" Death Rate", "", age_deaths_sub$Age_Group)
 
-We'll just be using the data from the "2016-2020 Persons Below Poverty Level" by race/ethnicity. We'll also change the text formatting of County to match the other data frames.
-```{r}
-pov <- pov %>%  select(1, starts_with("2016-2020 Persons Below Poverty Level"))
-
-pov$County <- str_to_title(pov$County)
-```
-
-Let's do some cleaning on the OASIS data.
-We'll remove extra columns and rows, as well as rows with "Selected ages/sexes total", pivot Year and Deaths into long format, and create the new variable for race/ethnicity.
-```{r}
-his_deaths <- his_deaths[c(-1,-2387:-2414),-7]
-his_deaths <- his_deaths %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>%
-  pivot_longer(c(4:6), names_to = "Year", values_to = "HIV_Deaths") %>% 
-  mutate("Race_Ethnicity" = "Hispanic")
-
-nhw_deaths <- nhw_deaths[c(-1,-2387:-2414),-7]
-nhw_deaths <- nhw_deaths %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>%
-  pivot_longer(c(4:6), names_to = "Year", values_to = "HIV_Deaths") %>% 
-  mutate("Race_Ethnicity" = "NH White")
-
-nhb_deaths <- nhb_deaths[c(-1,-2387:-2414),-7]
-nhb_deaths <- nhb_deaths %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>%
-  pivot_longer(c(4:6), names_to = "Year", values_to = "HIV_Deaths") %>% 
-  mutate("Race_Ethnicity" = "NH Black")
-```
-
-Next, we'll merge the datasets into one dataframe, and remove the original datasets from the environment.
-```{r}
-deaths <- full_join(his_deaths, nhw_deaths)
-deaths <- full_join(deaths, nhb_deaths)
-
-rm(his_deaths, nhw_deaths, nhb_deaths)
-```
-
-Next, let's clean up the dataset for population using a similar method.
-```{r}
-his_pop <- his_pop[c(-1,-2387:-2412),-7]
-his_pop <- his_pop %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>% 
-  pivot_longer(c(4:6), names_to = "Year", values_to = "Population") %>% 
-  mutate("Race_Ethnicity" = "Hispanic")
-
-nhw_pop <- nhw_pop[c(-1,-2387:-2412),-7]
-nhw_pop <- nhw_pop %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>% 
-  pivot_longer(c(4:6), names_to = "Year", values_to = "Population") %>% 
-  mutate("Race_Ethnicity" = "NH White")
-
-nhb_pop <- nhb_pop[c(-1,-2387:-2412),-7]
-nhb_pop <- nhb_pop %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>% 
-  pivot_longer(c(4:6), names_to = "Year", values_to = "Population") %>% 
-  mutate("Race_Ethnicity" = "NH Black")
-```
-
-Merging datasets and cleaning the original from the environment.
-```{r}
-pop <- full_join(his_pop, nhw_pop)
-pop <- full_join(pop, nhb_pop)
-
-rm(his_pop, nhw_pop, nhb_pop)
-```
-
-Next, let's merge the two OASIS dataframes.
-```{r}
-df <- full_join(deaths, pop, by = c("Geography", "Sex", "Age", "Year", "Race_Ethnicity"))
-```
-
-Next, let's create a new variable for Public Health District, change some variable to factors, and rename the Age variable and Geography variable.
-```{r}
-df <- df %>% rename(County = Geography) %>% 
-  mutate(District = case_when(County %in% c("Bartow", "Catoosa", "Chattooga", "Dade", "Floyd", "Gordon", "Haralson", "Paulding", "Polk", "Walker") ~ "1-1",
- County %in% c("Cherokee", "Fannin", "Gilmer", "Murray", "Pickens", "Whitfield") ~ "1-2",
- County %in% c("Banks", "Dawson", "Forsyth", "Franklin", "Habersham", "Hall", "Hart", "Lumpkin", "Rabun", "Stephens", "Towns", "Union", "White") ~ "2",
- County %in% c("Cobb", "Douglas") ~ "3-1",
- County %in% c("Fulton") ~ "3-2",
- County %in% c("Clayton") ~ "3-3",
- County %in% c("Gwinnett", "Newton", "Rockdale") ~ "3-4",
- County %in% c("DeKalb") ~ "3-5",
- County %in% c("Butts", "Carroll", "Coweta", "Fayette", "Heard", "Henry", "Lamar", "Meriwether", "Pike", "Spalding", "Troup", "Upson") ~ "4",
- County %in% c("Bleckley", "Dodge", "Johnson", "Laurens", "Montgomery", "Pulaski", "Telfair", "Treutlen", "Wheeler", "Wilcox") ~ "5-1",
- County %in% c("Baldwin", "Bibb", "Crawford", "Hancock", "Houston", "Jasper", "Jones", "Monroe", "Peach", "Putnam", "Twiggs", "Washington", "Wilkinson") ~ "5-2",
- County %in% c("Burke", "Columbia", "Emanuel", "Glascock", "Jefferson", "Jenkins", "Lincoln", "McDuffie", "Richmond", "Screven", "Taliaferro", "Warren", "Wilkes") ~ "6",
- County %in% c("Chattahoochee", "Clay", "Crisp", "Dooly", "Harris", "Macon", "Muscogee", "Marion", "Quitman", "Randolph", "Schley", "Stewart", "Sumter", "Talbot", "Taylor", "Webster") ~ "7",
- County %in% c("Ben Hill", "Berrien", "Brooks", "Cook", "Echols", "Irwin", "Lanier", "Lowndes", "Tift", "Turner") ~ "8-1",
- County %in% c("Baker", "Calhoun", "Colquitt", "Dougherty", "Decatur", "Early", "Grady", "Lee", "Miller", "Mitchell", "Seminole", "Terrell", "Thomas", "Worth") ~ "8-2",
- County %in% c("Bryan", "Camden", "Chatham", "Effingham", "Glynn", "Liberty", "Long", "McIntosh") ~ "9-1",
- County %in% c("Appling", "Atkinson", "Bacon", "Brantley", "Bulloch", "Candler", "Charlton", "Clinch", "Coffee", "Evans", "Jeff Davis", "Pierce", "Tattnall", "Toombs", "Ware", "Wayne") ~ "9-2",
- County %in% c("Barrow", "Clarke", "Elbert", "Greene", "Jackson", "Madison", "Morgan", "Oconee", "Oglethorpe", "Walton") ~ "10")) %>% 
-   mutate(District = as.factor(District)) %>% 
-   mutate(Sex = as.factor(Sex)) %>% 
-   mutate(Age = as.factor(Age)) %>% 
-   mutate(County = as.factor(County)) %>% 
-   mutate(Race_Ethnicity = as.factor(Race_Ethnicity)) %>% 
-   rename(Age_Group = Age) %>% 
-   mutate(Age_Group = as.factor(Age_Group))
-```
-
-Let's clean up the Age_Group variable by removing text
-```{r}
-df$Age_Group <- str_replace(df$Age_Group, " Adolescence", "")
-df$Age_Group <- str_replace(df$Age_Group, " Early Adulthood", "")
-df$Age_Group <- str_replace(df$Age_Group, " Young Adulthood", "")
-df$Age_Group <- str_replace(df$Age_Group, " Middle Adulthood", "")
+age_deaths_sub <- age_deaths_sub %>% mutate_if(is.character, as.factor) %>% 
+  mutate(`GEO ID` = as.factor(`GEO ID`))
 ```
@@
 ```{r}
-glimpse(df)
-```
-Things are looking good so far!
+dx <- dx %>% filter(State %in% c("Alabama", "Arkansas", "Delaware", "Washington, D.C.", "Florida", "Georgia", "Kentucky", "Louisiana", "Maryland", "Mississippi", "North Carolina", "Oklahoma", "South Carolina", "Tennessee", "Texas", "Virginia", "West Virginia")) %>%
+  select(!contains(c("Rate Stability", "Percent", "Case")))
 
-Now let's move onto the AIDSVu data. First, we'll merge all new HIV diagnoses from 2018-2020 into one dataframe
-```{r}
-dx <- full_join(dx18, dx19)
-dx <- full_join(dx, dx20)
+sex_dx_sub <- dx %>% select(1:7)
+race_dx_sub <- dx %>% select(c(1:5, 8:14))
+age_dx_sub <- dx %>% select(c(1:5, 15:19))
+
+sex_dx_sub <- sex_dx_sub %>% pivot_longer(6:7, names_to = "Sex", values_to = "New Diagnoses Rate by Sex")
+sex_dx_sub$Sex <- gsub("New Diagnoses ", "", sex_dx_sub$Sex)
+sex_dx_sub$Sex <- gsub(" Rate", "", sex_dx_sub$Sex)
+sex_dx_sub <- sex_dx_sub %>% mutate_if(is.character, as.factor) %>% 
+  mutate(`GEO ID` = as.factor(`GEO ID`))
+
+race_dx_sub <- race_dx_sub  %>% pivot_longer(6:12, names_to = "Race/Ethnicity", values_to = "New Diagnoses Rate by Race/Ethnicity")
+race_dx_sub$`Race/Ethnicity` <- gsub("New Diagnoses ", "", race_dx_sub$`Race/Ethnicity`)
+race_dx_sub$`Race/Ethnicity` <- gsub(" Rate", "", race_dx_sub$`Race/Ethnicity`)
+race_dx_sub <- race_dx_sub %>% mutate_if(is.character, as.factor) %>% 
+  mutate(`GEO ID` = as.factor(`GEO ID`))
+
+age_dx_sub <- age_dx_sub %>% pivot_longer(6:10, names_to = "Age_Group", values_to = "New Diagnoses Rate by Age Group")
+age_dx_sub$Age_Group <- gsub("New Diagnoses Age ", "", age_dx_sub$Age_Group)
+age_dx_sub$Age_Group <- gsub(" Rate", "", age_dx_sub$Age_Group)
+age_dx_sub <- age_dx_sub %>% mutate_if(is.character, as.factor) %>% 
+  mutate(`GEO ID` = as.factor(`GEO ID`))
 ```
 
-Let's remove columns for race/ethnicities other than Hispanic, NH White, and NH Black, as well as some other variables we won't be using in this analysis.
-We will also rename county and create the district variable.
+Calculate mortality to incidence ratio (MIR) "The mortality-to-incidence ratio (MIR) can be computed from readily accessible, public-use data on cancer incidence and mortality, and a high MIR value is an indicator of poor survival relative to incidence."
+
+The IMR (incidence:mortality ratio) is defined as the ratio of the number of people who become HIV infected per year to the number of people among those already infected who die (from any cause) per year.
+When this number is greater than one, the size of the population of PLHIV grows; when it is less than one, the size of that population shrinks
+https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6201869/#!po=16.6667
+
+https://www.kff.org/global-health-policy/issue-brief/assessing-global-hiv-targets-in-pepfar-countries-a-dashboard/
 ```{r}
-dx <- dx %>% select(!contains(c("Asian", "American", "Multiracial", "Hawaiian", "MSM", "2013", "GEO ID", "State")))%>% 
-  rename(County = "County Name")
+sex_df <- full_join(sex_deaths_sub, sex_dx_sub)
 
-dx$Geography <- str_remove(dx$County, " County")
+sex_df  <- sex_df %>% group_by(State, Sex) %>%
+  mutate(d2c_ratio_sex = sum(`Death Rate by Sex`/`New Diagnoses Rate by Sex`)) %>% 
+  mutate(d2c_ratio_sex = round(d2c_ratio_sex, 1))
 
-dx <-  dx %>% mutate(District = case_when(County %in% c("Bartow", "Catoosa", "Chattooga", "Dade", "Floyd", "Gordon", "Haralson", "Paulding", "Polk", "Walker") ~ "1-1",
- County %in% c("Cherokee", "Fannin", "Gilmer", "Murray", "Pickens", "Whitfield") ~ "1-2",
- County %in% c("Banks", "Dawson", "Forsyth", "Franklin", "Habersham", "Hall", "Hart", "Lumpkin", "Rabun", "Stephens", "Towns", "Union", "White") ~ "2",
- County %in% c("Cobb", "Douglas") ~ "3-1",
- County %in% c("Fulton") ~ "3-2",
- County %in% c("Clayton") ~ "3-3",
- County %in% c("Gwinnett", "Newton", "Rockdale") ~ "3-4",
- County %in% c("DeKalb") ~ "3-5",
- County %in% c("Butts", "Carroll", "Coweta", "Fayette", "Heard", "Henry", "Lamar", "Meriwether", "Pike", "Spalding", "Troup", "Upson") ~ "4",
- County %in% c("Bleckley", "Dodge", "Johnson", "Laurens", "Montgomery", "Pulaski", "Telfair", "Treutlen", "Wheeler", "Wilcox") ~ "5-1",
- County %in% c("Baldwin", "Bibb", "Crawford", "Hancock", "Houston", "Jasper", "Jones", "Monroe", "Peach", "Putnam", "Twiggs", "Washington", "Wilkinson") ~ "5-2",
- County %in% c("Burke", "Columbia", "Emanuel", "Glascock", "Jefferson", "Jenkins", "Lincoln", "McDuffie", "Richmond", "Screven", "Taliaferro", "Warren", "Wilkes") ~ "6",
- County %in% c("Chattahoochee", "Clay", "Crisp", "Dooly", "Harris", "Macon", "Muscogee", "Marion", "Quitman", "Randolph", "Schley", "Stewart", "Sumter", "Talbot", "Taylor", "Webster") ~ "7",
- County %in% c("Ben Hill", "Berrien", "Brooks", "Cook", "Echols", "Irwin", "Lanier", "Lowndes", "Tift", "Turner") ~ "8-1",
- County %in% c("Baker", "Calhoun", "Colquitt", "Dougherty", "Decatur", "Early", "Grady", "Lee", "Miller", "Mitchell", "Seminole", "Terrell", "Thomas", "Worth") ~ "8-2",
- County %in% c("Bryan", "Camden", "Chatham", "Effingham", "Glynn", "Liberty", "Long", "McIntosh") ~ "9-1",
- County %in% c("Appling", "Atkinson", "Bacon", "Brantley", "Bulloch", "Candler", "Charlton", "Clinch", "Coffee", "Evans", "Jeff Davis", "Pierce", "Tattnall", "Toombs", "Ware", "Wayne") ~ "9-2",
- County %in% c("Barrow", "Clarke", "Elbert", "Greene", "Jackson", "Madison", "Morgan", "Oconee", "Oglethorpe", "Walton") ~ "10")) %>% 
-  mutate(District = as.factor(District))
+race_df <- full_join(race_deaths_sub, race_dx_sub)
+
+race_df <- race_df %>% group_by(State, `Race/Ethnicity`) %>%
+  mutate(d2c_ratio_race = sum(`Death Rate by Race/Ethnicity`/`New Diagnoses Rate by Race/Ethnicity`)) %>% 
+  mutate(d2c_ratio_race = round(d2c_ratio_race, 1))
+
+age_df <- full_join(age_deaths_sub, age_dx_sub)
+
+age_df <- age_df %>% group_by(State, Age_Group) %>%
+  mutate(d2c_ratio_age = sum(`Death Rate by Age Group`/`New Diagnoses Rate by Age Group`)) %>% 
+  mutate(d2c_ratio_age = round(d2c_ratio_age, 1))
+
+saveRDS(sex_df, "../../data/sex_df.rds")
+saveRDS(race_df, "../../data/race_df.rds")
+saveRDS(age_df, "../../data/age_df.rds")
 ```
 
-Things are looking pretty good so frame, so let's save these dataframes.
 ```{r}
-saveRDS(df, "../../data/df.rds")
-saveRDS(dx, "../../data/dx.rds")
-saveRDS(sdoh, "../../data/sdoh.rds")
-saveRDS(pov, "../../data/pov.rds")
+rm(list=ls())
+
+sex_df <- readRDS("../../data/sex_df.rds")
+race_df <- readRDS("../../data/race_df.rds")
+age_df <- readRDS("../../data/age_df.rds")
+
 ```
 
-Clear the environment and start fresh.
 ```{r}
 
+```
+
+```{r}
+df <- full_join(deaths, new_dx)
+df <- full_join(df, prep)
+df<- full_join(df, prev)
+
+df <- df %>% select(!contains("Rate Stability"))
+
+saveRDS(df, "../../data/raw_data/raw_df.rds")
+
+rm(list=ls())
+rm(list=ls())
+
+prep <- read_xlsx("../../data/raw_data/AIDSVu_State_PrEP_2020.xlsx", skip = 3)
+prep <- prep %>% select(!contains("Rate Stability"))
+
+df <- readRDS("../../data/raw_data/raw_df.rds")
+
+late_dx <- read_xlsx("../../data/raw_data/AIDSVu_State_LateDiagnoses_2020.xlsx", skip = 3)
+sdoh <- read_xlsx("../../data/raw_data/AIDSVu_State_SDOH_2020.xlsx", skip = 3)
+v_sup <- read_xlsx("../../data/raw_data/AIDSVu_State_ViralSuppression_2020.xlsx", skip = 3)
+testing  <- read_xlsx("../../data/raw_data/AIDSVu_State_Testing_2020.xlsx", skip = 3)
 ```

### 2023-02-03 159_hiv_sdoh.R

@@ -1,0 +1,27 @@
+library(tidyverse)
+library(readxl)
+=
+Data on HIV deaths from OASIS. 2018-2020 By age group, by race/ethnicity
+
+his_deaths <- read_xlsx("data/raw_data/hivdeaths_his_1820.xlsx", skip = 1)
+nhw_deaths <- read_xlsx("data/raw_data/hivdeaths_nhw_1820.xlsx", skip = 1)
+nhb_deaths <- read_xlsx("data/raw_data/hivdeaths_nhb_1820.xlsx", skip = 1)
+
+his_pop <- read_xlsx("data/raw_data/pop_his_1820.xlsx", skip = 1)
+nhw_pop <- read_xlsx("data/raw_data/pop_nhw_1820.xlsx", skip = 1)
+nhb_pop <- read_xlsx("data/raw_data/pop_nhb_1820.xlsx")
+
+sdoh <- read_xlsx("data/raw_data/AIDSVu_County_SDOH_2020.xlsx", skip = 3)
+sdoh <- sdoh %>%  filter(State == "Georgia")
+
+dx18 <- read_xlsx("data/raw_data/AIDSVu_County_NewDX_2018.xlsx", skip = 3)
+dx18 <- dx18 %>%  filter(State == "Georgia") %>% 
+  select(!contains(c("Rate Stability", "Percent", "Case")))
+
+dx19 <- read_xlsx("data/raw_data/AIDSVu_County_NewDX_2019.xlsx", skip = 3)
+dx19 <- dx19 %>%  filter(State == "Georgia")%>% 
+  select(!contains(c("Rate Stability", "Percent", "Case")))
+
+dx20 <- read_xlsx("data/raw_data/AIDSVu_County_NewDX_2020.xlsx", skip = 3)
+dx20 <- dx20 %>%  filter(State == "Georgia") %>% 
+  select(!contains(c("Rate Stability", "Percent", "Case")))

### 2023-02-03 products/manuscript/MANU.qmd
commit

@@
 title: "Manuscript Template for a Data Analysis Project"
-subtitle: "Exploring Factors influencing HIV Death-to-Case ratio in the United States and Puerto Rico -- 2020"
+subtitle: "Exploring Factors influencing HIV-related Mortality Rates in the Georgia -- 2018 - 2020"
 author: Hayley Hemme

### 2023-02-03 products/manuscript/MANU.qmd
commit

@@
 ## Description of data and data source
-The source of
+The data used in this project was obtained from multiple publicly available sources.
+Data pertaining to counts of HIV-related deaths in for Hispanics, non-Hispanic Whites, and non-Hispanic Blacks and populations for these groups was obtained from OASIS, the Georgia Department of Public Health's online data repository.
+This analyse will currently focus on the year 2018-2020, but may be expanded to 2016-2020 moving forward.
+The data on these years will be obtained and processed identically to ways later described.
+Date on new HIV diagnoses and SDOH, particularly related to HIV were obtained from AIDSVu.
+Data on poverty rates was obtained from Georgia Data, a data source provided by the University of Georgia.
+All data described was aggregated at the county level.
 

### 2023-02-03 products/manuscript/MANU.qmd
commit

--- 2023-02-03 products/manuscript/MANU.qmd
@@
 
-The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
+The structure below is one possible setup for a data analysis project (including the course project).
+For a manuscript, adjust as needed.
 You don't need to have exactly these sections, but the content covering those sections should be addressed.
@@
 # Summary/Abstract
-_Write a summary of your project._
+
+*Write a summary of your project.*
 
@@
 ## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
+
+*Provide enough background on your topic that others can understand the why and how of your analysis*
 
 ## Description of data and data source
-The source of
+
+The data used in this project was obtained from multiple publicly available sources.
+Data pertaining to counts of HIV-related deaths in for Hispanics, non-Hispanic Whites, and non-Hispanic Blacks and populations for these groups was obtained from [OASIS](https://oasis.state.ga.us/), the Georgia Department of Public Health's online data repository.
+This analyse will currently focus on the year 2018-2020, but may be expanded to 2016-2020 moving forward.
+The data on these years will be obtained and processed identically to ways later described.
+Date on new HIV diagnoses and SDOH, particularly related to HIV were obtained from [AIDSVu](https://aidsvu.org/resources/#/datasets).
+Data on poverty rates was obtained from [Georgia Data](https://georgiadata.org/), a data source provided by the the Carl Vinson Institute of Government at the University of Georgia.
+All data described was aggregated at the county level.
 
 ## Questions/Hypotheses to be addressed
+
 Question: Can HIV death rates be predicted by social determinants of health like poverty and/ or by HIV related outcomes (new diagnoses, prevalence PrEP use, etc?)
 
-H0 : There is no correlation between HIV-related deaths and SDOH or HIV-related outcomes.
-Ha : There is a correlation between HIV-related deaths and SDOH or HIV-related outcomes.
+H0 : There is no correlation between HIV-related deaths and SDOH or HIV-related outcomes. Ha : There is a correlation between HIV-related deaths and SDOH or HIV-related outcomes.
 
@@
 
-_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
+*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*
 
 ## Data aquisition
-_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
+
+*As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next.*
 
 ## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
+
+*Write code that reads in the file and cleans it so it's ready for analysis.
 Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
@@
 as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
-Loading the datasets from OASIS for deaths and population (Hispanic, NH White, NH Black) for persons ages 13-59 from 2018 to 2020.
+All materials, including files that contain code, should be commented well so everyone can follow along.* Loading the datasets from OASIS for deaths and population (Hispanic, NH White, NH Black) for persons ages 13-59 from 2018 to 2020.
+
 ```{r warning=FALSE}
@@
 Loading dataset from AIDSVu for social determinants of health (2020) and new HIV diagnoses from 2018 to 2020.
+
 ```{r warning=FALSE}
@@
 Let's load data from Georgia Data on Poverty rates in Georgia by race/ ethnicity for the period 2016-2020.
+
 ```{r warning = false}
@@
 We'll just be using the data from the "2016-2020 Persons Below Poverty Level" by race/ethnicity. We'll also change the text formatting of County to match the other data frames.
+
 ```{r}
@@
 We'll remove extra columns and rows, as well as rows with "Selected ages/sexes total", pivot Year and Deaths into long format, and create the new variable for race/ethnicity.
+
 ```{r}
@@
 Next, we'll merge the datasets into one dataframe, and remove the original datasets from the environment.
+
 ```{r}
@@
 Next, let's clean up the dataset for population using a similar method.
+
 ```{r}
@@
 Merging datasets and cleaning the original from the environment.
+
 ```{r}
@@
 Next, let's merge the two OASIS dataframes.
+
 ```{r}
@@
 Next, let's create a new variable for Public Health District, change some variable to factors, and rename the Age variable and Geography variable.
+
 ```{r}
@@
 Let's clean up the Age_Group variable by removing text
+
 ```{r}
@@
 ```
+
 Things are looking good so far!
@@
 Now let's move onto the AIDSVu data. First, we'll merge all new HIV diagnoses from 2018-2020 into one dataframe
+
 ```{r}
@@
 We will also rename county and create the district variable.
+
 ```{r}
@@
 Things are looking pretty good so frame, so let's save these dataframes.
+
 ```{r}
@@
 Clear the environment and start fresh.
+
 ```{r}
@@
 ```
+
 ## Statistical analysis
+
 Explore central tendency across demographic groups. Fit a poisson regression model predicting death-to-case ratio from percentage with x SDOH variables and percentage of x demographic group.
@@
 
-_Use a combination of text/tables/figures to explore and describe your data.
+*Use a combination of text/tables/figures to explore and describe your data.
 Show the most important descriptive results here.
 Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project._
+Even more can be in the R and Quarto files that are part of your project.*
 
@@
 
-_To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
+*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
 simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
+Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
 
@@
 
-_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
+*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
 This might again be code that is best placed in one or several separate R scripts that need to be well documented.
 You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here._
+Then you load them here.*
 
@@
 ## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
+
+*Summarize what you did, what you found and what it means.*
 
 ## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
+
+*Discuss what you perceive as strengths and limitations of your analysis.*
 
 ## Conclusions
-_What are the main take-home messages?_
 
-_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_
+*What are the main take-home messages?*
+
+*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
 

### 2023-02-09 products/manuscript/MANU.qmd
used here package

@@
 library(stringr)
+here::here()
+
 ```
@@
 ```{r warning=FALSE}
-setwd("../../")
-his_deaths <- read_xlsx("data/raw_data/hivdeaths_his_1820.xlsx", skip = 1)
-nhw_deaths <- read_xlsx("data/raw_data/hivdeaths_nhw_1820.xlsx", skip = 1)
-nhb_deaths <- read_xlsx("data/raw_data/hivdeaths_nhb_1820.xlsx", skip = 1)
+his_deaths <- read_xlsx(here("data/raw_data/hivdeaths_his_1820.xlsx"), skip = 1)
+nhw_deaths <- read_xlsx(here("data/raw_data/hivdeaths_nhw_1820.xlsx"), skip = 1)
+nhb_deaths <- read_xlsx(here("data/raw_data/hivdeaths_nhb_1820.xlsx"), skip = 1)
 
-his_pop <- read_xlsx("data/raw_data/pop_his_1820.xlsx", skip = 1)
-nhw_pop <- read_xlsx("data/raw_data/pop_nhw_1820.xlsx", skip = 1)
-nhb_pop <- read_xlsx("data/raw_data/pop_nhb_1820.xlsx", skip = 1)
+his_pop <- read_xlsx(here("data/raw_data/pop_his_1820.xlsx"), skip = 1)
+nhw_pop <- read_xlsx(here("data/raw_data/pop_nhw_1820.xlsx"), skip = 1)
+nhb_pop <- read_xlsx(here("data/raw_data/pop_nhb_1820.xlsx"), skip = 1)
 ```
@@
 ```{r warning=FALSE}
-setwd("../../")
-sdoh <- read_xlsx("data/raw_data/AIDSVu_County_SDOH_2020.xlsx", skip = 3)
+sdoh <- read_xlsx(here("data/raw_data/AIDSVu_County_SDOH_2020.xlsx"), skip = 3)
 sdoh <- sdoh %>%  filter(State == "Georgia")
 
-dx18 <- read_xlsx("data/raw_data/AIDSVu_County_NewDX_2018.xlsx", skip = 3)
+dx18 <- read_xlsx(here("data/raw_data/AIDSVu_County_NewDX_2018.xlsx"), skip = 3)
 dx18 <- dx18 %>%  filter(State == "Georgia") %>% 
@@
 
-dx19 <- read_xlsx("data/raw_data/AIDSVu_County_NewDX_2019.xlsx", skip = 3)
+dx19 <- read_xlsx(here("data/raw_data/AIDSVu_County_NewDX_2019.xlsx"), skip = 3)
 dx19 <- dx19 %>%  filter(State == "Georgia")%>% 
@@
 
-dx20 <- read_xlsx("data/raw_data/AIDSVu_County_NewDX_2020.xlsx", skip = 3)
+dx20 <- read_xlsx(here("data/raw_data/AIDSVu_County_NewDX_2020.xlsx"), skip = 3)
 dx20 <- dx20 %>%  filter(State == "Georgia") %>% 
@@
 setwd("../../")
-pov <- read_xlsx("data/raw_data/economics_22.xlsx", sheet = "poverty")
+pov <- read_xlsx(here("data/raw_data/economics_22.xlsx"), sheet = "poverty")
 
@@
 
-Things are looking pretty good so frame, so let's save these dataframes.
+Things are looking pretty good so far, so let's save these dataframes.
 
 ```{r}
-saveRDS(df, "../../data/df.rds")
-saveRDS(dx, "../../data/dx.rds")
-saveRDS(sdoh, "../../data/sdoh.rds")
-saveRDS(pov, "../../data/pov.rds")
+saveRDS(df, here("data/processed_data/df.rds"))
+saveRDS(dx, here("data/processed_data/dx.rds"))
+saveRDS(sdoh, here("data/processed_data/sdoh.rds"))
+saveRDS(pov, here("data/processed_data/pov.rds"))
 ```

### 2023-02-09 products/manuscript/MANU.qmd
removed (../../) and code chunks causing issues with rendering

@@
 
-```{r warning = false}
-setwd("../../")
+```{r warning = FALSE}
 pov <- read_xlsx(here("data/raw_data/economics_22.xlsx"), sheet = "poverty")
@@
 
-```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
-#| echo: FALSE
-resulttable=readRDS("../../results/summarytable.rds")
-knitr::kable(resulttable)
-```
-
 ## Basic statistical analysis
@@
 ```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by sex."
-#| echo: FALSE
-knitr::include_graphics(here("results","height_weight_stratified.png"))
+
 ```
@@
 ```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-resulttable2 = readRDS(here("results","resulttable2.rds"))
-knitr::kable(resulttable2)
+
 ```

### 2023-02-24 code/analysis_code/archieved/2_exploratory_analysis.Rmd
switching directions a bit after running into issues with lots of zeros at district/county-level.  Will be using state-level data from year 2020 for the United States.

```{r message=FALSE}
library(here)
library(GGally)
library(skimr)
library(tidyverse)
```

```{r}
df <- readRDS(here("data/processed_data/dataframe.rds"))
df_sdoh20 <- readRDS(here("data/processed_data/df20.rds"))
```

Let's collapse age groups to just look at 20-44 due to issue with very small populations among some groups. let also look at hiv deaths over entire time period.
```{r}
df_sub <- df %>% 
  filter(Age_Group %in% c("20-29", "30-44"))

df_sub <- df_sub %>%
  group_by(District, Race_Ethnicity, Sex) %>% 
  summarise(HIV_Deaths = sum(HIV_Deaths),
            Population = sum(Population),
            Uninsured = mean(`% Uninsured`),
            Poverty_Rate = mean(Poverty_Rate_1620),
            n_PCP = sum(`# Primary Care Physicians`),
            mean_PCP_rate = mean(`PCP Rate`)) 
```

Next, let's find the expected number of HIV deaths. First, we'll calculate lambda.
```{r}
lambda <- sum(df_sub$HIV_Deaths)/sum(df_sub$Population)

df_sub <- df_sub %>%
  mutate(Expected_Deaths = lambda * Population)
```

Then, we'll calculate rates for observed HIV deaths and expected HIV deaths.
```{r}
df_sub <-  df_sub %>% 
  mutate(HIV_Rate = ((HIV_Deaths + 0.5)/Population) *1e5)

df_sub <-  df_sub %>% 
  mutate(Expected_Rate = ((Expected_Deaths + 0.5)/Population) *1e5)
```

Next, let's find the standardized mortality ratio.
```{r}
df_sub <- df_sub %>%
  mutate(SMR = HIV_Rate/Expected_Rate)
```

Next, I looked at the distribution of SMR
```{r}
df_sub %>% 
  ggplot() +
  geom_histogram(aes(x = SMR), bins = 10)

summary(df_sub$SMR)
```
The histogram revealed a very right-skewed distribution.

Let's take a look at HIV rate and SMR more closely by running some model diagnostics in ggfortify.
```{r}
model <- lm(df_sub$HIV_Rate ~ 1)
model_SMR <- lm(df_sub$SMR ~ 1)

library(ggfortify)

autoplot(model)
autoplot(model_SMR)
```

Let's see if any transformations can normalize this distribution. I decided to start with log(x +1) since the distribution was so skewed.
```{r}
df_sub %>% 
  mutate(logSMR = log(SMR + 1)) %>% 
  ggplot() +
  geom_histogram(aes(x = logSMR), bins = 20)
```
But based on visual assessment, it appears that this approach was not enough to normalized the distribution! After some quick Googling, it appears some potential options are using a mixture model or Box-Cox transformation.

I'll try the Box-Cox transformation first. chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/https://cran.r-project.org/web/packages/spmoran/vignettes/vignette_spmoran_nongaussian.pdf
```{r}
library(MASS)
b <-  boxcox(lm(df_sub$SMR ~ 1))

lambda2 <- b$x[which.max(b$y)]

df_sub <- df_sub %>%
  ungroup() %>%
  mutate(exact_x = (df_sub$SMR^lambda2 - 1) / lambda2)

exact_model <- lm(df_sub$exact_x ~ 1)
plot(exact_model)

df_sub %>% 
  ggplot() +
  geom_histogram(aes(x = exact_x), bins = 20)

shapiro.test(df_sub$exact_x)
```
With the p = 0.25, we do not find statistically significant evidence to reject normality. Great!

Zero-inflated model. We can try this on the count data..
```{r}
library(pscl)

m1 <- zeroinfl(HIV_Deaths ~ Uninsured | 1,  offset = Population,  data = df_sub, dist = "poisson")
m2 <- zeroinfl(HIV_Deaths ~ . | 1,  offset = Population,  data = df_sub, dist = "negbin")

summary(m1)
summary(m2)
```
Admittedly, I am bit unsure of what route to take next. I am considering 
I am considering pivoting to HIV prevalence data???
https://bmcinfectdis.biomedcentral.com/articles/10.1186/1471-2334-14-285

```{r}
corr_cols <- 
ggpairs(corr_cols)

```

Let's look at HIV-related death rates by race/ethnicity.
```{r message=FALSE}
df_sub %>% 
  ggplot() +
  geom_histogram(aes(x = exact_x), bins = 20) + facet_wrap (. ~ Race_Ethnicity)

df_sub %>%
  group_by(Race_Ethnicity) %>%
  skim(exact_x)
```

Then by sex...
```{r}
df_sub %>% 
  drop_na(HIV_Rate) %>%
  ggplot() +
  geom_histogram(aes(x = HIV_Rate), bins = 20) + facet_wrap (. ~ Sex)

df_sub %>%
  group_by(Sex) %>%
  skim(HIV_Rate)
```

And by race/ ethnicity by sex.
```{r message=FALSE, warning=FALSE}
df_sub %>% 
  drop_na(HIV_Rate) %>%
  ggplot() +
  geom_histogram(aes(x = HIV_Rate), bins = 20) + facet_wrap (Sex ~ Race_Ethnicity)

df_sub %>%
  group_by(Race_Ethnicity, Sex) %>%
  skim(HIV_Rate)
```

Next, let's get a sense of how things look by District. 
```{r}
df_sub %>% 
  drop_na(HIV_Rate) %>%
  ggplot() +
  geom_histogram(aes(x = HIV_Rate), bins = 20) + facet_wrap (. ~ District)

df_sub %>%
  group_by(District) %>%
  skim(HIV_Rate)
```

```{r}
df_sub %>% 
  drop_na(HIV_Rate) %>%
  ggplot() +
geom_histogram(aes(x = HIV_Rate), bins = 20) + facet_wrap (District ~ Sex)

df_sub %>%
  group_by(District) %>%
  skim(HIV_Rate)
```

```{r}
df_sub <- df_sub %>% mutate(logHIV_Rate = log10(HIV_Rate + 1))
```

```{r}
df_sub %>% 
  drop_na(HIV_Rate) %>% 
  ggplot() +
  geom_histogram(aes(x = HIV_Rate), bins = 20)

df_sub %>% 
  drop_na(HIV_Rate) %>%
  ggplot() +
  geom_histogram(aes(x = logHIV_Rate))
```

```{r}
x_test <- df %>%
  drop_na(HIV_Rate) %>%
  filter(HIV_Rate > 0)

shapiro.test(x_test$HIV_Rate)
```

```{r}

```

### 2023-02-24 code/analysis_code/archieved/EDA.Rmd
switching directions a bit after running into issues with lots of zeros at district/county-level.  Will be using state-level data from year 2020 for the United States.

```{r}
library(here)
library(GGally)
library(tidyverse)
```

```{r}
df <- readRDS(here("data/processed_data/dataframe.rds"))
df_sdoh20 <- readRDS(here("data/processed_data/df20.rds"))
```

Creating variable for HIV death rate
```{r}
df <- df %>% mutate(HIV_Rate = HIV_Deaths/Population*1e5) %>% 
  mutate(HIV_Rate = as.numeric(HIV_Rate))
df <- df %>% mutate(logHIV_Rate = log10(HIV_Rate + 1))
```

```{r}
df_DS <- df %>% group_by(Year, District, Race_Ethnicity, Sex, Age_Group) %>% 
  summarise(HIV_Deaths_DS = sum(HIV_Deaths),
            Population_DS = sum(Population),
            Uninsured_DS = mean(`% Uninsured`),
            Poverty_Rate_DS = mean(Poverty_Rate_1620)) %>% 
  mutate(HIV_Rate_DS = HIV_Deaths_DS/Population_DS*1e5)
```

```{r}
df %>% 
  group_by(District) %>% 
  drop_na(HIV_Rate) %>% 
  ggplot() +
  geom_histogram(aes(x = HIV_Rate), bins = 20)

df %>% 
  group_by(District) %>% 
  drop_na(HIV_Rate) %>% 
  ggplot() +
  geom_histogram(aes(x = logHIV_Rate))

df %>% 
  group_by(District) %>% 
  drop_na(HIV_Rate) %>%
  filter(HIV_Rate > 0) %>% 
  ggplot() +
  geom_histogram(aes(x = logHIV_Rate))
```

```{r}
x_test <- df %>%
  drop_na(HIV_Rate) %>%
  filter(HIV_Rate > 0)

shapiro.test(x_test$HIV_Rate)
```

```{r}

```

### 2023-02-24 code/analysis_code/archieved/more wrangling.Rmd
switching directions a bit after running into issues with lots of zeros at district/county-level.  Will be using state-level data from year 2020 for the United States.

```{r}
library(here)
library(tidyverse)
```

```{r}
dths <- readRDS(here("data/processed_data/df.rds"))
dx <- readRDS(here("data/processed_data/dx.rds"))
sdoh <- readRDS(here("data/processed_data/sdoh.rds"))
pov <- readRDS(here("data/processed_data/pov.rds"))
c_hlth <- readRDS(here("data/processed_data/c_hlth.rds"))
```

Let's look into the diagnoses data a bit more. 
```{r}
glimpse(dx)
range(dx$`New Diagnoses Rate`)
```

Since -1 is not a possible value a rate can take, I referred back to AIDSVu.org and found that: '-1: Data are not shown to protect privacy because a small number of cases and/or a small population size for reasons listed in the Data Suppression and Rate Stability section.' A this moment, let's change observations with '-1' to NA.

```{r}
dx <- dx %>% mutate(`New Diagnoses Rate` = if_else(`New Diagnoses Rate` < 0, NA, `New Diagnoses Rate`))

pov <- pov %>% mutate(Race_Ethnicity = as.factor(Race_Ethnicity))

sdoh <- sdoh %>% select(!`Median Household Income`)
```

```{r}
df <- list(dths, dx, c_hlth, pov)

df <- reduce(df, full_join)

df20 <- df %>% filter(Year == 2020)

df20 <- full_join(df20, sdoh)
```

Let's save these dataframes.
```{r}
saveRDS(df20, "data/processed_data/df20.rds")
saveRDS(df, "data/processed_data/dataframe.rds")
```

### 2023-02-24 code/processing_code/1_cleaning_2020.Rmd
switching directions a bit after running into issues with lots of zeros at district/county-level.  Will be using state-level data from year 2020 for the United States.

```{r, echo=FALSE, message=FALSE}
library(knitr)
library(readxl)
library(lubridate)
library(stringr)
library(tidyverse)
library(here)
```

```{r}
deaths <- read_xlsx(here("data/raw_data/AIDSVu_State_Death_2020.xlsx"), skip = 3)
late_dx <- read_xlsx(here("data/raw_data/AIDSVu_State_LateDiagnoses_2020.xlsx"), skip = 3)
us_race <- read_xlsx(here("data/raw_data/DECENNIALPL2020.P2-2023-02-24T124235.xlsx"), sheet = 2)
us_sex <-read_csv(here("data/raw_data/raw_data.csv"), skip = 2)

new_dx <- read_xlsx(here("data/raw_data/AIDSVu_State_NewDX_2020.xlsx"), skip = 3)
prev  <- read_xlsx(here("data/raw_data/AIDSVu_State_Prev_2020.xlsx"), skip = 3)
rec_care  <- read_xlsx(here("data/raw_data/AIDSVu_State_ReceiptofCare_2020.xlsx"), skip = 3)
sdoh  <- read_xlsx(here("data/raw_data/AIDSVu_State_SDOH_2020.xlsx"), skip = 3)
v_sup  <- read_xlsx(here("data/raw_data/AIDSVu_State_ViralSuppression_2020.xlsx"), skip = 3)
``

let's start by separating `deaths` by demographic characteristic . While this data provides information on demographic characteristics such as sex, age group, and race/ethnicity, it does not provide data on all of these terms together. We'll separating by sex and race/ethnicity. This analysis will be limited to Non-Hispanic Black, Non-Hispanic White, and Hispanic persons.
```{r}
deaths_sex <- deaths %>% select(1:4, contains(c("male", "MSM")))

deaths_race <-   deaths %>% select(1:4, contains(c("Black", "White", "Hispanic",  "Multiple")))

summary(deaths_race)
```
Some rates are -9 due to there not being data for these observations. Let's convert these to NA
```{r}
death_race <-deaths_race %>% 
  mutate(across(ends_with("Death Rate"), ~ case_when( .x < 0 ~ NA, TRUE ~ .x)))
```

Next let's move on to the census data. I'll start by re-formatting the data from wide format for States to long, and format race/ethnicity into wide.
```{r}
glimpse(us_race)

race <- us_race %>%  
  pivot_longer(2:54 , names_to = "Location", values_to = "Population" ) %>%  
  pivot_wider(names_from = Label, values_from= Population)
```

```{r}
race <- race %>% 
  select(c(Location, starts_with(c("Total:", "Black or African American alone", "Hispanic", "White alone", "Population of two or more races:")))) %>% 
  rename(c(Hispanic = `Hispanic or Latino`,
         White = `White alone`,
         Black = `Black or African American alone`,
         Multiple = `Population of two or more races:`))
```

Then, I'll use inner join to merge the dataset contain percentage make up by sex for each state. 
```{r}
pop <- inner_join(us_sex, race) %>%
  select(!Total) 
```

I need to remove all commas from the dataframe and convert the columns to numeric. 
```{r}
pop <- pop %>% 
  mutate_all(~ str_remove_all(., ",")) %>% 
  mutate(across(2:8, ~ as.numeric(.)))
```

Then, I'll calculate the counts of males and females in each location.
```{r}
pop <- pop %>% 
  mutate(Male = Male * `Total:`) %>% 
  mutate(Female = Female * `Total:`)
```

```{r}
pop_sex <- pop %>%  select(1:3)

sex_df <- inner_join(deaths_sex, pop_sex, by = join_by("State" == Location))
```

```{r}
sex_df %>% mutate(m_rate = (`Male Death Cases` * Male)*1e5)

```

https://datacenter.kidscount.org/data/tables/101-child-population-by-age-group#detailed/2/2-53/false/574/62,63,64,6,4693/419

We need to estimate the population older than 13. We can get pretty close with the data above.

```{r}
kid_pop <- read_xlsx(here("data/Child population by age group.xlsx"))
kid_pop <- kid_pop %>%  filter(TimeFrame == 2020)
```

 Most recent known address was used in all analyses except for new HIV diagnoses, late diagnoses, and linkage to care, which are calculated based on residence at diagnosis. Denominators used to calculate rates for all state and county populations were obtained by CDC from the U.S. Census Bureau’s census for each respective year. Population denominators are restricted to people aged 13 and older. There are no county-level maps for the District of Columbia because there are no counties in DC.

HIV surveillance data are displayed as case counts, rates, and/or proportions (represented as % cases). Rates were calculated per 100,000 population to permit data standardization and comparison. For rate ratios by race/ethnicity displayed on state profile pages, White race is the referent group.

### 2023-02-24 code/processing_code/archieved/1_cleaning.Rmd
switching directions a bit after running into issues with lots of zeros at district/county-level.  Will be using state-level data from year 2020 for the United States.

```{r, echo=FALSE, message=FALSE}
library(knitr)
library(readxl)
library(lubridate)
library(stringr)
library(tidyverse)
library(here)
```

OASIS for deaths and population (Hispanic, NH White, NH Black) for persons ages 13-59 from 2018 to 2020.

```{r warning=FALSE}
his_deaths <- read_xlsx(here("data/raw_data/hivdeaths_his_1820.xlsx"), skip = 1)
nhw_deaths <- read_xlsx(here("data/raw_data/hivdeaths_nhw_1820.xlsx"), skip = 1)
nhb_deaths <- read_xlsx(here("data/raw_data/hivdeaths_nhb_1820.xlsx"), skip = 1)

his_pop <- read_xlsx(here("data/raw_data/pop_his_1820.xlsx"), skip = 1)
nhw_pop <- read_xlsx(here("data/raw_data/pop_nhw_1820.xlsx"), skip = 1)
nhb_pop <- read_xlsx(here("data/raw_data/pop_nhb_1820.xlsx"), skip = 1)
```

Loading dataset from AIDSVu for social determinants of health (2020) and new HIV diagnoses from 2018 to 2020.
```{r warning=FALSE}
sdoh <- read_xlsx(here("data/raw_data/AIDSVu_County_SDOH_2020.xlsx"), skip = 3)

dx18 <- read_xlsx(here("data/raw_data/AIDSVu_County_NewDX_2018.xlsx"), skip = 3)
dx18 <- dx18 %>%  filter(State == "Georgia") %>% 
  select(!contains(c("Rate Stability", "Percent", "Case")))

dx19 <- read_xlsx(here("data/raw_data/AIDSVu_County_NewDX_2019.xlsx"), skip = 3)
dx19 <- dx19 %>%  filter(State == "Georgia")%>% 
  select(!contains(c("Rate Stability", "Percent", "Case")))

dx20 <- read_xlsx(here("data/raw_data/AIDSVu_County_NewDX_2020.xlsx"), skip = 3)
dx20 <- dx20 %>%  filter(State == "Georgia") %>% 
  select(!contains(c("Rate Stability", "Percent", "Case")))
```

Let's clean the `sdoh`. I'll start by filtering out all observations except those from Georgia, then I'll do a bit of reformatting.
```{r}
glimpse(sdoh)
sdoh <- sdoh %>%  filter(State == "Georgia")
sdoh$County <- str_remove(sdoh$County, " County")

sdoh <- sdoh %>% select(4:12) %>% 
  mutate(Year = "2020") %>% 
  mutate(Year = as.numeric(Year)) %>% 
  mutate(County = as.factor(County))

sdoh <- sdoh %>% select(!`Median Household Income`)
```

Let's load data from Georgia Data on Poverty rates in Georgia by race/ ethnicity for the period 2016-2020.

```{r warning = FALSE}
pov <- read_xlsx(here("data/raw_data/economics_22.xlsx"), sheet = "poverty") 
pov <- pov[c(1:159),]

glimpse(pov)
```

We'll just be using the data from the "2016-2020 Persons Below Poverty Level" by race/ethnicity. We'll also change the text formatting of County to match the other data frames.

```{r}
pov <- pov %>% select(1, starts_with("2016-2020 Persons Below Poverty Level"))

pov$County <- str_to_title(pov$County)
pov$County <- str_replace(pov$County, "Mcintosh", "McIntosh")
pov$County <- str_replace(pov$County, "Mcduffie", "McDuffie")
pov$County <- str_replace(pov$County, "Dekalb", "DeKalb")

pov <- pov %>%
  mutate(`2016-2020 Persons Below Poverty Level, Hispanic, Percent` = as.numeric(`2016-2020 Persons Below Poverty Level, Hispanic, Percent`)) %>% 
  mutate(across(where(is.numeric), function(x) x*100)) %>% 
  mutate_if(is.character, as.factor) %>% 
  pivot_longer(c(2:5), names_to = "Race_Ethnicity", values_to = "Poverty_Rate_1620")

pov$Race_Ethnicity <- str_remove(pov$Race_Ethnicity, "2016-2020 Persons Below Poverty Level, ")
pov$Race_Ethnicity <- str_remove(pov$Race_Ethnicity, ", Percent")
pov$Race_Ethnicity <- str_replace(pov$Race_Ethnicity, "Percent", "All")

pov <- pov %>% 
  filter(Race_Ethnicity %in% c("Black", "Hispanic", "White")) %>% 
  mutate(Race_Ethnicity = as.factor(Race_Ethnicity))

glimpse(pov)
```

Let's do some cleaning on the OASIS data. We'll remove extra columns and rows, as well as rows with "Selected ages/sexes total", pivot Year and Deaths into long format, and create the new variable for race/ethnicity.

```{r}
his_deaths <- his_deaths[c(-1,-2387:-2414),-7]
his_deaths <- his_deaths %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>%
  pivot_longer(c(4:6), names_to = "Year", values_to = "HIV_Deaths") %>% 
  mutate("Race_Ethnicity" = "Hispanic")

nhw_deaths <- nhw_deaths[c(-1,-2387:-2414),-7]
nhw_deaths <- nhw_deaths %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>%
  pivot_longer(c(4:6), names_to = "Year", values_to = "HIV_Deaths") %>% 
  mutate("Race_Ethnicity" = "White")

nhb_deaths <- nhb_deaths[c(-1,-2387:-2414),-7]
nhb_deaths <- nhb_deaths %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>%
  pivot_longer(c(4:6), names_to = "Year", values_to = "HIV_Deaths") %>% 
  mutate("Race_Ethnicity" = "Black")
```

Next, we'll merge the datasets into one dataframe, and remove the original datasets from the environment.

```{r}
deaths <- list(his_deaths, nhb_deaths, nhw_deaths)

deaths <- reduce(deaths, full_join)
rm(his_deaths, nhw_deaths, nhb_deaths)
```

Next, let's clean up the dataset for population using a similar method.

```{r}
his_pop <- his_pop[c(-1,-2387:-2412),-7]
his_pop <- his_pop %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>% 
  pivot_longer(c(4:6), names_to = "Year", values_to = "Population") %>% 
  mutate("Race_Ethnicity" = "Hispanic")

nhw_pop <- nhw_pop[c(-1,-2387:-2412),-7]
nhw_pop <- nhw_pop %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>% 
  pivot_longer(c(4:6), names_to = "Year", values_to = "Population") %>% 
  mutate("Race_Ethnicity" = "White")

nhb_pop <- nhb_pop[c(-1,-2387:-2412),-7]
nhb_pop <- nhb_pop %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>% 
  pivot_longer(c(4:6), names_to = "Year", values_to = "Population") %>% 
  mutate("Race_Ethnicity" = "Black")
```

Merging datasets and cleaning the original from the environment.

```{r}
pop <- list(his_pop, nhb_pop, nhw_pop)
pop <- reduce(pop, full_join)

pop$Population <- as.numeric(gsub(",","",pop$Population))
rm(his_pop, nhw_pop, nhb_pop)
```

Next, let's merge the two OASIS dataframes.

```{r}
df <- full_join(deaths, pop, by = c("Geography", "Sex", "Age", "Year", "Race_Ethnicity"))
```

Next, let's create a new variable for Public Health District, change some variable to factors, and rename the Age variable and Geography variable.

```{r}
df <- df %>% rename(County = Geography) %>% 
  mutate(District = case_when(County %in% c("Bartow", "Catoosa", "Chattooga", "Dade", "Floyd", "Gordon", "Haralson", "Paulding", "Polk", "Walker") ~ "1-1",
 County %in% c("Cherokee", "Fannin", "Gilmer", "Murray", "Pickens", "Whitfield") ~ "1-2",
 County %in% c("Banks", "Dawson", "Forsyth", "Franklin", "Habersham", "Hall", "Hart", "Lumpkin", "Rabun", "Stephens", "Towns", "Union", "White") ~ "2",
 County %in% c("Cobb", "Douglas") ~ "3-1",
 County %in% c("Fulton") ~ "3-2",
 County %in% c("Clayton") ~ "3-3",
 County %in% c("Gwinnett", "Newton", "Rockdale") ~ "3-4",
 County %in% c("DeKalb") ~ "3-5",
 County %in% c("Butts", "Carroll", "Coweta", "Fayette", "Heard", "Henry", "Lamar", "Meriwether", "Pike", "Spalding", "Troup", "Upson") ~ "4",
 County %in% c("Bleckley", "Dodge", "Johnson", "Laurens", "Montgomery", "Pulaski", "Telfair", "Treutlen", "Wheeler", "Wilcox") ~ "5-1",
 County %in% c("Baldwin", "Bibb", "Crawford", "Hancock", "Houston", "Jasper", "Jones", "Monroe", "Peach", "Putnam", "Twiggs", "Washington", "Wilkinson") ~ "5-2",
 County %in% c("Burke", "Columbia", "Emanuel", "Glascock", "Jefferson", "Jenkins", "Lincoln", "McDuffie", "Richmond", "Screven", "Taliaferro", "Warren", "Wilkes") ~ "6",
 County %in% c("Chattahoochee", "Clay", "Crisp", "Dooly", "Harris", "Macon", "Muscogee", "Marion", "Quitman", "Randolph", "Schley", "Stewart", "Sumter", "Talbot", "Taylor", "Webster") ~ "7",
 County %in% c("Ben Hill", "Berrien", "Brooks", "Cook", "Echols", "Irwin", "Lanier", "Lowndes", "Tift", "Turner") ~ "8-1",
 County %in% c("Baker", "Calhoun", "Colquitt", "Dougherty", "Decatur", "Early", "Grady", "Lee", "Miller", "Mitchell", "Seminole", "Terrell", "Thomas", "Worth") ~ "8-2",
 County %in% c("Bryan", "Camden", "Chatham", "Effingham", "Glynn", "Liberty", "Long", "McIntosh") ~ "9-1",
 County %in% c("Appling", "Atkinson", "Bacon", "Brantley", "Bulloch", "Candler", "Charlton", "Clinch", "Coffee", "Evans", "Jeff Davis", "Pierce", "Tattnall", "Toombs", "Ware", "Wayne") ~ "9-2",
 County %in% c("Barrow", "Clarke", "Elbert", "Greene", "Jackson", "Madison", "Morgan", "Oconee", "Oglethorpe", "Walton") ~ "10")) %>% 
   mutate(District = as.factor(District)) %>% 
   mutate(Sex = as.factor(Sex)) %>% 
   mutate(County = as.factor(County)) %>% 
   mutate(Race_Ethnicity = as.factor(Race_Ethnicity)) %>% 
   rename(Age_Group = Age) %>% 
   mutate(Age_Group = as.factor(Age_Group))
```

Let's clean up the Age_Group variable by removing text and make it a factor with levels. Changing from class from character to numeric for Year, HIV_Deaths, and Population.

```{r}
df$Age_Group <- str_replace(df$Age_Group, " Adolescence", "")
df$Age_Group <- str_replace(df$Age_Group, " Early Adulthood", "")
df$Age_Group <- str_replace(df$Age_Group, " Young Adulthood", "")
df$Age_Group <- str_replace(df$Age_Group, " Middle Adulthood", "")

df <- df %>% mutate(Age_Group = as.factor(Age_Group),
                   Age_Group = factor(Age_Group,
    level = c("13-19", "20-29", "30-44", "45-59"))) %>% 
  mutate_if(is.character, as.numeric) %>% 
  relocate(where(is.numeric), .after = last_col()) %>% 
  relocate(`District`, .after = `County`) %>% 
  relocate(`Age_Group`, .after = `Race_Ethnicity`)
glimpse(df)
```

```{r}
glimpse(df)
```
Things are looking good so far!

Now let's move onto the AIDSVu data. First, we'll merge all new HIV diagnoses from 2018-2020 into one dataframe

```{r}
dx <- list(dx18, dx19, dx20)
dx <- reduce(dx, full_join)
rm(dx18, dx19, dx20)
```

Let's remove columns for race/ethnicities other than Hispanic, NH White, and NH Black, as well as some other variables we won't be using in this analysis. We will also rename county and create the district variable.

```{r}
dx <- dx %>% select(!contains(c("Asian", "American", "Multiracial", "Hawaiian", "MSM", "2013", "GEO ID", "State")))%>% 
  rename(County = "County Name")

dx$County <- str_remove(dx$County, " County")

dx <-  dx %>% mutate(District = case_when(County %in% c("Bartow", "Catoosa", "Chattooga", "Dade", "Floyd", "Gordon", "Haralson", "Paulding", "Polk", "Walker") ~ "1-1",
 County %in% c("Cherokee", "Fannin", "Gilmer", "Murray", "Pickens", "Whitfield") ~ "1-2",
 County %in% c("Banks", "Dawson", "Forsyth", "Franklin", "Habersham", "Hall", "Hart", "Lumpkin", "Rabun", "Stephens", "Towns", "Union", "White") ~ "2",
 County %in% c("Cobb", "Douglas") ~ "3-1",
 County %in% c("Fulton") ~ "3-2",
 County %in% c("Clayton") ~ "3-3",
 County %in% c("Gwinnett", "Newton", "Rockdale") ~ "3-4",
 County %in% c("DeKalb") ~ "3-5",
 County %in% c("Butts", "Carroll", "Coweta", "Fayette", "Heard", "Henry", "Lamar", "Meriwether", "Pike", "Spalding", "Troup", "Upson") ~ "4",
 County %in% c("Bleckley", "Dodge", "Johnson", "Laurens", "Montgomery", "Pulaski", "Telfair", "Treutlen", "Wheeler", "Wilcox") ~ "5-1",
 County %in% c("Baldwin", "Bibb", "Crawford", "Hancock", "Houston", "Jasper", "Jones", "Monroe", "Peach", "Putnam", "Twiggs", "Washington", "Wilkinson") ~ "5-2",
 County %in% c("Burke", "Columbia", "Emanuel", "Glascock", "Jefferson", "Jenkins", "Lincoln", "McDuffie", "Richmond", "Screven", "Taliaferro", "Warren", "Wilkes") ~ "6",
 County %in% c("Chattahoochee", "Clay", "Crisp", "Dooly", "Harris", "Macon", "Muscogee", "Marion", "Quitman", "Randolph", "Schley", "Stewart", "Sumter", "Talbot", "Taylor", "Webster") ~ "7",
 County %in% c("Ben Hill", "Berrien", "Brooks", "Cook", "Echols", "Irwin", "Lanier", "Lowndes", "Tift", "Turner") ~ "8-1",
 County %in% c("Baker", "Calhoun", "Colquitt", "Dougherty", "Decatur", "Early", "Grady", "Lee", "Miller", "Mitchell", "Seminole", "Terrell", "Thomas", "Worth") ~ "8-2",
 County %in% c("Bryan", "Camden", "Chatham", "Effingham", "Glynn", "Liberty", "Long", "McIntosh") ~ "9-1",
 County %in% c("Appling", "Atkinson", "Bacon", "Brantley", "Bulloch", "Candler", "Charlton", "Clinch", "Coffee", "Evans", "Jeff Davis", "Pierce", "Tattnall", "Toombs", "Ware", "Wayne") ~ "9-2",
 County %in% c("Barrow", "Clarke", "Elbert", "Greene", "Jackson", "Madison", "Morgan", "Oconee", "Oglethorpe", "Walton") ~ "10")) %>% 
  mutate(District = as.factor(District)) %>% 
  mutate(County = as.factor(County))

range(dx$`New Diagnoses Rate`)
```
Since -1 is not a possible value a rate can take, I referred back to AIDSVu.org and found that: '-1: Data are not shown to protect privacy because a small number of cases and/or a small population size for reasons listed in the Data Suppression and Rate Stability section.' A this moment, let's change observations with '-1' to NA. 
```{r}
dx <- dx %>% select(County, District, Year, `New Diagnoses Rate`)%>% 
  mutate(`New Diagnoses Rate` = if_else(`New Diagnoses Rate` < 0, NA, `New Diagnoses Rate`))

glimpse(dx)
```

Here, I'm loading data from County Health rankings for years 2018 - 2020
```{r}
cnty_hlth18A <- read_xls(path = here("data/raw_data/2018 County Health Rankings Georgia Data - v3.xls"), sheet = 4, skip = 1)
cnty_hlth18B <- read_xls(path = here("data/raw_data/2018 County Health Rankings Georgia Data - v3.xls"), sheet = 5, skip = 1)
cnty_hlth18A <- cnty_hlth18A %>%  select(c(3, 67:75))
cnty_hlth18B <- cnty_hlth18B %>%  select(c(3, 62:64))
c_hlt18 <- full_join(cnty_hlth18A, cnty_hlth18B)

c_hlt18 <- c_hlt18 %>% mutate(Year = as.numeric(2018)) %>%
  select(1:3,7:9,11, 14) %>% 
  rename(`Median Household Income` = `Household Income`)
c_hlt18 <-  c_hlt18[-1,]

cnty_hlth19A <- read_xls(path = here("data/raw_data/2019 County Health Rankings Georgia Data - v1_0.xls"), sheet = 4, skip = 1)
cnty_hlth19B <- read_xls(path = here("data/raw_data/2019 County Health Rankings Georgia Data - v1_0.xls"), sheet = 5, skip = 1)
cnty_hlth19A <- cnty_hlth19A %>%  select(c(3, 67:75))
cnty_hlth19B <- cnty_hlth19B %>%  select(c(3, 66:68))
c_hlt19 <- full_join(cnty_hlth19A, cnty_hlth19B)

c_hlt19 <- c_hlt19 %>% mutate(Year = as.numeric(2019)) %>%
  select(1:3,7:9,11, 14)  %>% 
  rename(`Median Household Income` = `Household Income`)
c_hlt19 <-  c_hlt19[-1,]

cnty_hlth20A <- read_xlsx(path = here("data/raw_data/2020 County Health Rankings Georgia Data - v1_0.xlsx"), sheet = 4, skip = 1)
cnty_hlth20B <- read_xlsx(path = here("data/raw_data/2020 County Health Rankings Georgia Data - v1_0.xlsx"), sheet = 5, skip = 1)
cnty_hlth20A <- cnty_hlth20A %>%  select(c(3, 104:112))
cnty_hlth20B <- cnty_hlth20B %>%  select(c(3, 158:160))

c_hlt20 <- full_join(cnty_hlth20A, cnty_hlth20B)
c_hlt20 <- c_hlt20 %>% mutate(Year = as.numeric(2020)) %>%
  select(1:3,7:9,11, 14)    %>% 
  rename(`PCP Rate` = `Primary Care Physicians Rate`) %>% 
  rename(`PCP Ratio` = `Primary Care Physicians Ratio`)
c_hlt20 <-  c_hlt20[-1,]
```

```{r}
c_hlth <- list(c_hlt18,c_hlt19, c_hlt20)

c_hlth  <- reduce(c_hlth, full_join)
c_hlth <- c_hlth %>% mutate(County = as.factor(County))
```

Things are looking pretty good so far, so let's save these dataframes.
```{r}
saveRDS(df, here("data/processed_data/df.rds"))
saveRDS(dx, here("data/processed_data/dx.rds"))
saveRDS(sdoh, here("data/processed_data/sdoh.rds"))
saveRDS(pov, here("data/processed_data/pov.rds"))
saveRDS(c_hlth, here("data/processed_data/c_hlth.rds"))
```

Now, let's merge the datasets - dths, dx, c_hlth, pov. Since `sdoh` contains only data from 2020, we'll create another subset containing just observations from 2020 and merge with `sdoh`.
```{r}
df1 <- list(df, dx, c_hlth, pov)

df1 <- reduce(df1, full_join)

df20 <- df %>% filter(Year == 2020)

df20 <- full_join(df20, sdoh)
```

Let's save these dataframes.
```{r}
saveRDS(df20, here("data/processed_data/df20.rds"))
saveRDS(df1, here("data/processed_data/dataframe.rds"))
```

### 2023-02-24 products/manuscript/MANU.qmd
switching directions a bit after running into issues with lots of zeros at district/county-level.  Will be using state-level data from year 2020 for the United States.

@@
 # load a few R packages
-library(here)
 library(knitr)
-library(tidyverse)
 library(readxl)
@@
 library(stringr)
-here::here()
-
+library(tidyverse)
+library(here)
 ```
@@
 
-The data used in this project was obtained from multiple publicly available sources.
-Data pertaining to counts of HIV-related deaths in for Hispanics, non-Hispanic Whites, and non-Hispanic Blacks and populations for these groups was obtained from [OASIS](https://oasis.state.ga.us/), the Georgia Department of Public Health's online data repository.
-This analyse will currently focus on the year 2018-2020, but may be expanded to 2016-2020 moving forward.
-The data on these years will be obtained and processed identically to ways later described.
-Date on new HIV diagnoses and SDOH, particularly related to HIV were obtained from [AIDSVu](https://aidsvu.org/resources/#/datasets).
-Data on poverty rates was obtained from [Georgia Data](https://georgiadata.org/), a data source provided by the the Carl Vinson Institute of Government at the University of Georgia.
-All data described was aggregated at the county level.
+The data used in this project was obtained from multiple publicly available source.
+Data pertaining to HIV deaths, new diagnoses, HIV-related outcomes, and social determinants of health relating to HIV were obtained from AIDSVu.
+[AIDSVu](https://aidsvu.org/resources/#/datasets).
+Population data for on persons of race/ethnicities included in the analysis and total population was obtained from the [U.S.
+Census Bureau](https://data.census.gov/).
+Percentage population make-up by sex was obtained from [KFF](https://www.kff.org/other/state-indicator/) .
+Percentage distribution of persons under 14 was obtained from Kids Count [Kids Count](https://datacenter.kidscount.org/).
+All data described was aggregated at the the state-level.
 
@@
 
-Question: Can HIV death rates be predicted by social determinants of health like poverty and/ or by HIV related outcomes (new diagnoses, prevalence PrEP use, etc?)
+Question: Can the HIV death rate from 2020 in the United States be modeled by social determinants of health like poverty and/ or by HIV related outcomes (new diagnoses, prevalence PrEP use, etc?) Explain in separate analysis: Race/ ethnicity (Black, Hispanic, White, Multiple) and Sex (Male, Female).
+Can a spatial model explain variation in HIV death-rates at the state-level and/or regional-level after accounting for variables associated with sociodemographic, SDOH, and HIV-related outcomes?
 
-H0 : There is no correlation between HIV-related deaths and SDOH or HIV-related outcomes. Ha : There is a correlation between HIV-related deaths and SDOH or HIV-related outcomes.
-
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key.
-Then you can include like this:
+H0 : There is not a spatial correlation HIV-related deaths and SDOH or HIV-related outcomes.
+Ha : There is a spatial correlation between HIV-related deaths and SDOH or HIV-related outcomes.
 
@@
 as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along.* Loading the datasets from OASIS for deaths and population (Hispanic, NH White, NH Black) for persons ages 13-59 from 2018 to 2020.
-
-```{r warning=FALSE}
-his_deaths <- read_xlsx(here("data/raw_data/hivdeaths_his_1820.xlsx"), skip = 1)
-nhw_deaths <- read_xlsx(here("data/raw_data/hivdeaths_nhw_1820.xlsx"), skip = 1)
-nhb_deaths <- read_xlsx(here("data/raw_data/hivdeaths_nhb_1820.xlsx"), skip = 1)
-
-his_pop <- read_xlsx(here("data/raw_data/pop_his_1820.xlsx"), skip = 1)
-nhw_pop <- read_xlsx(here("data/raw_data/pop_nhw_1820.xlsx"), skip = 1)
-nhb_pop <- read_xlsx(here("data/raw_data/pop_nhb_1820.xlsx"), skip = 1)
-```
-
-Loading dataset from AIDSVu for social determinants of health (2020) and new HIV diagnoses from 2018 to 2020.
-
-```{r warning=FALSE}
-sdoh <- read_xlsx(here("data/raw_data/AIDSVu_County_SDOH_2020.xlsx"), skip = 3)
-sdoh <- sdoh %>%  filter(State == "Georgia")
-
-dx18 <- read_xlsx(here("data/raw_data/AIDSVu_County_NewDX_2018.xlsx"), skip = 3)
-dx18 <- dx18 %>%  filter(State == "Georgia") %>% 
-  select(!contains(c("Rate Stability", "Percent", "Case")))
-
-dx19 <- read_xlsx(here("data/raw_data/AIDSVu_County_NewDX_2019.xlsx"), skip = 3)
-dx19 <- dx19 %>%  filter(State == "Georgia")%>% 
-  select(!contains(c("Rate Stability", "Percent", "Case")))
-
-dx20 <- read_xlsx(here("data/raw_data/AIDSVu_County_NewDX_2020.xlsx"), skip = 3)
-dx20 <- dx20 %>%  filter(State == "Georgia") %>% 
-  select(!contains(c("Rate Stability", "Percent", "Case")))
-```
-
-Let's load data from Georgia Data on Poverty rates in Georgia by race/ ethnicity for the period 2016-2020.
-
-```{r warning = FALSE}
-pov <- read_xlsx(here("data/raw_data/economics_22.xlsx"), sheet = "poverty")
-
-glimpse(pov)
-```
-
-We'll just be using the data from the "2016-2020 Persons Below Poverty Level" by race/ethnicity. We'll also change the text formatting of County to match the other data frames.
-
-```{r}
-pov <- pov %>%  select(1, starts_with("2016-2020 Persons Below Poverty Level"))
-
-pov$County <- str_to_title(pov$County)
-```
-
-Let's do some cleaning on the OASIS data.
-We'll remove extra columns and rows, as well as rows with "Selected ages/sexes total", pivot Year and Deaths into long format, and create the new variable for race/ethnicity.
-
-```{r}
-his_deaths <- his_deaths[c(-1,-2387:-2414),-7]
-his_deaths <- his_deaths %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>%
-  pivot_longer(c(4:6), names_to = "Year", values_to = "HIV_Deaths") %>% 
-  mutate("Race_Ethnicity" = "Hispanic")
-
-nhw_deaths <- nhw_deaths[c(-1,-2387:-2414),-7]
-nhw_deaths <- nhw_deaths %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>%
-  pivot_longer(c(4:6), names_to = "Year", values_to = "HIV_Deaths") %>% 
-  mutate("Race_Ethnicity" = "NH White")
-
-nhb_deaths <- nhb_deaths[c(-1,-2387:-2414),-7]
-nhb_deaths <- nhb_deaths %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>%
-  pivot_longer(c(4:6), names_to = "Year", values_to = "HIV_Deaths") %>% 
-  mutate("Race_Ethnicity" = "NH Black")
-```
-
-Next, we'll merge the datasets into one dataframe, and remove the original datasets from the environment.
-
-```{r}
-deaths <- full_join(his_deaths, nhw_deaths)
-deaths <- full_join(deaths, nhb_deaths)
-
-rm(his_deaths, nhw_deaths, nhb_deaths)
-```
-
-Next, let's clean up the dataset for population using a similar method.
-
-```{r}
-his_pop <- his_pop[c(-1,-2387:-2412),-7]
-his_pop <- his_pop %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>% 
-  pivot_longer(c(4:6), names_to = "Year", values_to = "Population") %>% 
-  mutate("Race_Ethnicity" = "Hispanic")
-
-nhw_pop <- nhw_pop[c(-1,-2387:-2412),-7]
-nhw_pop <- nhw_pop %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>% 
-  pivot_longer(c(4:6), names_to = "Year", values_to = "Population") %>% 
-  mutate("Race_Ethnicity" = "NH White")
-
-nhb_pop <- nhb_pop[c(-1,-2387:-2412),-7]
-nhb_pop <- nhb_pop %>% filter(!grepl("Selected", Age)) %>% filter(!grepl("Selected", Sex)) %>% 
-  pivot_longer(c(4:6), names_to = "Year", values_to = "Population") %>% 
-  mutate("Race_Ethnicity" = "NH Black")
-```
-
-Merging datasets and cleaning the original from the environment.
-
-```{r}
-pop <- full_join(his_pop, nhw_pop)
-pop <- full_join(pop, nhb_pop)
-
-rm(his_pop, nhw_pop, nhb_pop)
-```
-
-Next, let's merge the two OASIS dataframes.
-
-```{r}
-df <- full_join(deaths, pop, by = c("Geography", "Sex", "Age", "Year", "Race_Ethnicity"))
-```
-
-Next, let's create a new variable for Public Health District, change some variable to factors, and rename the Age variable and Geography variable.
-
-```{r}
-df <- df %>% rename(County = Geography) %>% 
-  mutate(District = case_when(County %in% c("Bartow", "Catoosa", "Chattooga", "Dade", "Floyd", "Gordon", "Haralson", "Paulding", "Polk", "Walker") ~ "1-1",
- County %in% c("Cherokee", "Fannin", "Gilmer", "Murray", "Pickens", "Whitfield") ~ "1-2",
- County %in% c("Banks", "Dawson", "Forsyth", "Franklin", "Habersham", "Hall", "Hart", "Lumpkin", "Rabun", "Stephens", "Towns", "Union", "White") ~ "2",
- County %in% c("Cobb", "Douglas") ~ "3-1",
- County %in% c("Fulton") ~ "3-2",
- County %in% c("Clayton") ~ "3-3",
- County %in% c("Gwinnett", "Newton", "Rockdale") ~ "3-4",
- County %in% c("DeKalb") ~ "3-5",
- County %in% c("Butts", "Carroll", "Coweta", "Fayette", "Heard", "Henry", "Lamar", "Meriwether", "Pike", "Spalding", "Troup", "Upson") ~ "4",
- County %in% c("Bleckley", "Dodge", "Johnson", "Laurens", "Montgomery", "Pulaski", "Telfair", "Treutlen", "Wheeler", "Wilcox") ~ "5-1",
- County %in% c("Baldwin", "Bibb", "Crawford", "Hancock", "Houston", "Jasper", "Jones", "Monroe", "Peach", "Putnam", "Twiggs", "Washington", "Wilkinson") ~ "5-2",
- County %in% c("Burke", "Columbia", "Emanuel", "Glascock", "Jefferson", "Jenkins", "Lincoln", "McDuffie", "Richmond", "Screven", "Taliaferro", "Warren", "Wilkes") ~ "6",
- County %in% c("Chattahoochee", "Clay", "Crisp", "Dooly", "Harris", "Macon", "Muscogee", "Marion", "Quitman", "Randolph", "Schley", "Stewart", "Sumter", "Talbot", "Taylor", "Webster") ~ "7",
- County %in% c("Ben Hill", "Berrien", "Brooks", "Cook", "Echols", "Irwin", "Lanier", "Lowndes", "Tift", "Turner") ~ "8-1",
- County %in% c("Baker", "Calhoun", "Colquitt", "Dougherty", "Decatur", "Early", "Grady", "Lee", "Miller", "Mitchell", "Seminole", "Terrell", "Thomas", "Worth") ~ "8-2",
- County %in% c("Bryan", "Camden", "Chatham", "Effingham", "Glynn", "Liberty", "Long", "McIntosh") ~ "9-1",
- County %in% c("Appling", "Atkinson", "Bacon", "Brantley", "Bulloch", "Candler", "Charlton", "Clinch", "Coffee", "Evans", "Jeff Davis", "Pierce", "Tattnall", "Toombs", "Ware", "Wayne") ~ "9-2",
- County %in% c("Barrow", "Clarke", "Elbert", "Greene", "Jackson", "Madison", "Morgan", "Oconee", "Oglethorpe", "Walton") ~ "10")) %>% 
-   mutate(District = as.factor(District)) %>% 
-   mutate(Sex = as.factor(Sex)) %>% 
-   mutate(Age = as.factor(Age)) %>% 
-   mutate(County = as.factor(County)) %>% 
-   mutate(Race_Ethnicity = as.factor(Race_Ethnicity)) %>% 
-   rename(Age_Group = Age) %>% 
-   mutate(Age_Group = as.factor(Age_Group))
-```
-
-Let's clean up the Age_Group variable by removing text
-
-```{r}
-df$Age_Group <- str_replace(df$Age_Group, " Adolescence", "")
-df$Age_Group <- str_replace(df$Age_Group, " Early Adulthood", "")
-df$Age_Group <- str_replace(df$Age_Group, " Young Adulthood", "")
-df$Age_Group <- str_replace(df$Age_Group, " Middle Adulthood", "")
-```
-
-```{r}
-glimpse(df)
-```
-
-Things are looking good so far!
-
-Now let's move onto the AIDSVu data. First, we'll merge all new HIV diagnoses from 2018-2020 into one dataframe
-
-```{r}
-dx <- full_join(dx18, dx19)
-dx <- full_join(dx, dx20)
-```
-
-Let's remove columns for race/ethnicities other than Hispanic, NH White, and NH Black, as well as some other variables we won't be using in this analysis.
-We will also rename county and create the district variable.
-
-```{r}
-dx <- dx %>% select(!contains(c("Asian", "American", "Multiracial", "Hawaiian", "MSM", "2013", "GEO ID", "State")))%>% 
-  rename(County = "County Name")
-
-dx$Geography <- str_remove(dx$County, " County")
-
-dx <-  dx %>% mutate(District = case_when(County %in% c("Bartow", "Catoosa", "Chattooga", "Dade", "Floyd", "Gordon", "Haralson", "Paulding", "Polk", "Walker") ~ "1-1",
- County %in% c("Cherokee", "Fannin", "Gilmer", "Murray", "Pickens", "Whitfield") ~ "1-2",
- County %in% c("Banks", "Dawson", "Forsyth", "Franklin", "Habersham", "Hall", "Hart", "Lumpkin", "Rabun", "Stephens", "Towns", "Union", "White") ~ "2",
- County %in% c("Cobb", "Douglas") ~ "3-1",
- County %in% c("Fulton") ~ "3-2",
- County %in% c("Clayton") ~ "3-3",
- County %in% c("Gwinnett", "Newton", "Rockdale") ~ "3-4",
- County %in% c("DeKalb") ~ "3-5",
- County %in% c("Butts", "Carroll", "Coweta", "Fayette", "Heard", "Henry", "Lamar", "Meriwether", "Pike", "Spalding", "Troup", "Upson") ~ "4",
- County %in% c("Bleckley", "Dodge", "Johnson", "Laurens", "Montgomery", "Pulaski", "Telfair", "Treutlen", "Wheeler", "Wilcox") ~ "5-1",
- County %in% c("Baldwin", "Bibb", "Crawford", "Hancock", "Houston", "Jasper", "Jones", "Monroe", "Peach", "Putnam", "Twiggs", "Washington", "Wilkinson") ~ "5-2",
- County %in% c("Burke", "Columbia", "Emanuel", "Glascock", "Jefferson", "Jenkins", "Lincoln", "McDuffie", "Richmond", "Screven", "Taliaferro", "Warren", "Wilkes") ~ "6",
- County %in% c("Chattahoochee", "Clay", "Crisp", "Dooly", "Harris", "Macon", "Muscogee", "Marion", "Quitman", "Randolph", "Schley", "Stewart", "Sumter", "Talbot", "Taylor", "Webster") ~ "7",
- County %in% c("Ben Hill", "Berrien", "Brooks", "Cook", "Echols", "Irwin", "Lanier", "Lowndes", "Tift", "Turner") ~ "8-1",
- County %in% c("Baker", "Calhoun", "Colquitt", "Dougherty", "Decatur", "Early", "Grady", "Lee", "Miller", "Mitchell", "Seminole", "Terrell", "Thomas", "Worth") ~ "8-2",
- County %in% c("Bryan", "Camden", "Chatham", "Effingham", "Glynn", "Liberty", "Long", "McIntosh") ~ "9-1",
- County %in% c("Appling", "Atkinson", "Bacon", "Brantley", "Bulloch", "Candler", "Charlton", "Clinch", "Coffee", "Evans", "Jeff Davis", "Pierce", "Tattnall", "Toombs", "Ware", "Wayne") ~ "9-2",
- County %in% c("Barrow", "Clarke", "Elbert", "Greene", "Jackson", "Madison", "Morgan", "Oconee", "Oglethorpe", "Walton") ~ "10")) %>% 
-  mutate(District = as.factor(District))
-```
-
-Things are looking pretty good so far, so let's save these dataframes.
-
-```{r}
-saveRDS(df, here("data/processed_data/df.rds"))
-saveRDS(dx, here("data/processed_data/dx.rds"))
-saveRDS(sdoh, here("data/processed_data/sdoh.rds"))
-saveRDS(pov, here("data/processed_data/pov.rds"))
-```
+All materials, including files that contain code, should be commented well so everyone can follow along.* Loading the datasets from
 
@@
 
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-
 ## Basic statistical analysis
@@
 Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
-
-```{r}
-
-```
 
@@
 Then you load them here.*
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
-
-```{r}
-
-```
 
@@
 
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
 Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.

### 2023-02-24 159_hiv_sdoh.R

@@ -1,27 +1,0 @@
-library(tidyverse)
-library(readxl)
-=
-Data on HIV deaths from OASIS. 2018-2020 By age group, by race/ethnicity
-
-his_deaths <- read_xlsx("data/raw_data/hivdeaths_his_1820.xlsx", skip = 1)
-nhw_deaths <- read_xlsx("data/raw_data/hivdeaths_nhw_1820.xlsx", skip = 1)
-nhb_deaths <- read_xlsx("data/raw_data/hivdeaths_nhb_1820.xlsx", skip = 1)
-
-his_pop <- read_xlsx("data/raw_data/pop_his_1820.xlsx", skip = 1)
-nhw_pop <- read_xlsx("data/raw_data/pop_nhw_1820.xlsx", skip = 1)
-nhb_pop <- read_xlsx("data/raw_data/pop_nhb_1820.xlsx")
-
-sdoh <- read_xlsx("data/raw_data/AIDSVu_County_SDOH_2020.xlsx", skip = 3)
-sdoh <- sdoh %>%  filter(State == "Georgia")
-
-dx18 <- read_xlsx("data/raw_data/AIDSVu_County_NewDX_2018.xlsx", skip = 3)
-dx18 <- dx18 %>%  filter(State == "Georgia") %>% 
-  select(!contains(c("Rate Stability", "Percent", "Case")))
-
-dx19 <- read_xlsx("data/raw_data/AIDSVu_County_NewDX_2019.xlsx", skip = 3)
-dx19 <- dx19 %>%  filter(State == "Georgia")%>% 
-  select(!contains(c("Rate Stability", "Percent", "Case")))
-
-dx20 <- read_xlsx("data/raw_data/AIDSVu_County_NewDX_2020.xlsx", skip = 3)
-dx20 <- dx20 %>%  filter(State == "Georgia") %>% 
-  select(!contains(c("Rate Stability", "Percent", "Case")))

### 2023-02-24 code/processing_code/1_cleaning_2020.Rmd
more cleaning

@@
 deaths <- read_xlsx(here("data/raw_data/AIDSVu_State_Death_2020.xlsx"), skip = 3)
-late_dx <- read_xlsx(here("data/raw_data/AIDSVu_State_LateDiagnoses_2020.xlsx"), skip = 3)
-us_race <- read_xlsx(here("data/raw_data/DECENNIALPL2020.P2-2023-02-24T124235.xlsx"), sheet = 2)
-us_sex <-read_csv(here("data/raw_data/raw_data.csv"), skip = 2)
+#late_dx <- read_xlsx(here("data/raw_data/AIDSVu_State_LateDiagnoses_2020.xlsx"), skip = 3)
+us_race_pop <- read_xlsx(here("data/raw_data/DECENNIALPL2020.P2-2023-02-24T124235.xlsx"), sheet = 2)
+us_sex_pop <-read_csv(here("data/raw_data/raw_data.csv"), skip = 2)
 
-new_dx <- read_xlsx(here("data/raw_data/AIDSVu_State_NewDX_2020.xlsx"), skip = 3)
-prev  <- read_xlsx(here("data/raw_data/AIDSVu_State_Prev_2020.xlsx"), skip = 3)
-rec_care  <- read_xlsx(here("data/raw_data/AIDSVu_State_ReceiptofCare_2020.xlsx"), skip = 3)
-sdoh  <- read_xlsx(here("data/raw_data/AIDSVu_State_SDOH_2020.xlsx"), skip = 3)
-v_sup  <- read_xlsx(here("data/raw_data/AIDSVu_State_ViralSuppression_2020.xlsx"), skip = 3)
-``
+#new_dx <- read_xlsx(here("data/raw_data/AIDSVu_State_NewDX_2020.xlsx"), skip = 3)
+#prev  <- read_xlsx(here("data/raw_data/AIDSVu_State_Prev_2020.xlsx"), skip = 3)
+#rec_care  <- read_xlsx(here("data/raw_data/AIDSVu_State_ReceiptofCare_2020.xlsx"), skip = 3)
+#sdoh  <- read_xlsx(here("data/raw_data/AIDSVu_State_SDOH_2020.xlsx"), skip = 3)
+#v_sup  <- read_xlsx(here("data/raw_data/AIDSVu_State_ViralSuppression_2020.xlsx"), skip = 3)
+```
 
-let's start by separating `deaths` by demographic characteristic .
+Let's start by separating `deaths` by demographic characteristic .
 While this data provides information on demographic characteristics such as sex, age group, and race/ethnicity, it does not provide data on all of these terms together.
@@
 ```
-Some rates are -9 due to there not being data for these observations. Let's convert these to NA
+
+Some rates are -9 due to missingness. Let's convert these to NA
 ```{r}
-death_race <-deaths_race %>% 
+death_race <- deaths_race %>% 
   mutate(across(ends_with("Death Rate"), ~ case_when( .x < 0 ~ NA, TRUE ~ .x)))
@@
 
-Next let's move on to the census data. I'll start by re-formatting the data from wide format for States to long, and format race/ethnicity into wide.
+Next, let's move on to the census data. I'll start by re-formatting the data from wide format for States to long, and format race/ethnicity into wide.
 ```{r}
-glimpse(us_race)
+glimpse(us_race_pop)
 
-race <- us_race %>%  
+race_pop <- us_race_pop %>%  
   pivot_longer(2:54 , names_to = "Location", values_to = "Population" ) %>%  
@@
 ```{r}
-race <- race %>% 
+race_pop <- race_pop %>% 
   select(c(Location, starts_with(c("Total:", "Black or African American alone", "Hispanic", "White alone", "Population of two or more races:")))) %>% 
@@
 
-Then, I'll use inner join to merge the dataset contain percentage make up by sex for each state. 
-```{r}
-pop <- inner_join(us_sex, race) %>%
-  select(!Total) 
-```
-
 I need to remove all commas from the dataframe and convert the columns to numeric. 
 ```{r}
-pop <- pop %>% 
+race_pop <- race_pop %>% 
   mutate_all(~ str_remove_all(., ",")) %>% 
-  mutate(across(2:8, ~ as.numeric(.)))
+  mutate(across(2:6, ~ as.numeric(.)))
+
 ```
@@
 ```{r}
-pop <- pop %>% 
-  mutate(Male = Male * `Total:`) %>% 
-  mutate(Female = Female * `Total:`)
+total_pop <- race_pop %>% select(1:2)
+pop_sex <- inner_join(total_pop, us_sex_pop)
+```
+
+Next, we need to estimate the population older than 12. I will try to estimate this using data from Kids Count.
+
+```{r}
+kid_pop <- read_xlsx(here("data/raw_data/Child population by age group.xlsx"))
+kid_pop <- kid_pop %>% 
+  filter(TimeFrame == 2020) %>% 
+  filter(`Age group` %in% c("0 to 4", "5 to 11", "12 to 14")) %>% 
+  mutate(Data = as.numeric(Data))
+
+pop_less13 <- kid_pop %>% 
+  group_by(Location) %>% 
+  mutate(`Age group` = ifelse(`Age group` == '12 to 14', '12', `Age group`)) %>%
+  group_by(`Age group`) %>%
+  mutate(`Data` = ifelse(`Age group` == '12', `Data` * 0.333, `Data`)) %>%
+  group_by(Location) %>% 
+  summarise(Less13 = sum(Data))
+
+new_pop <- full_join(pop_sex, pop_less13, by = join_by(Location))
+
+new_pop <- new_pop %>% 
+  mutate(Male_Pop = Male * `Total:`,
+  Female_Pop = Female * `Total:`,
+  Female_kid_pop = Female * Less13,
+  Male_kid_pop = Male * Less13)
+
+df_pop <- new_pop %>% 
+  group_by(Location) %>% 
+  summarize(df_pop_male = sum(Male_Pop - Male_kid_pop),
+            df_pop_female = sum(Female_Pop - Female_kid_pop))
+
 ```
@@
 ```{r}
-pop_sex <- pop %>%  select(1:3)
-
-sex_df <- inner_join(deaths_sex, pop_sex, by = join_by("State" == Location))
+sex_df <- inner_join(deaths_sex, df_pop, by = join_by(State == Location))
 ```
@@
 ```{r}
-sex_df %>% mutate(m_rate = (`Male Death Cases` * Male)*1e5)
-
+sex_df %>% mutate(M_Rate = `Male Death Cases`/df_pop_male*1e5)
 ```
 
-https://datacenter.kidscount.org/data/tables/101-child-population-by-age-group#detailed/2/2-53/false/574/62,63,64,6,4693/419
-
-We need to estimate the population older than 13. We can get pretty close with the data above.
-
-```{r}
-kid_pop <- read_xlsx(here("data/Child population by age group.xlsx"))
-kid_pop <- kid_pop %>%  filter(TimeFrame == 2020)
-```
-
- Most recent known address was used in all analyses except for new HIV diagnoses, late diagnoses, and linkage to care, which are calculated based on residence at diagnosis.
-Denominators used to calculate rates for all state and county populations were obtained by CDC from the U.S.
-Census Bureau’s census for each respective year.
-Population denominators are restricted to people aged 13 and older.
-There are no county-level maps for the District of Columbia because there are no counties in DC.
-
-HIV surveillance data are displayed as case counts, rates, and/or proportions (represented as % cases).
-Rates were calculated per 100,000 population to permit data standardization and comparison.
-For rate ratios by race/ethnicity displayed on state profile pages, White race is the referent group.
+Next find expected rate vs observed rate

### 2023-02-25 code/processing_code/1_cleaning_2020.Rmd
more cleaning

@@
 us_sex_pop <-read_csv(here("data/raw_data/raw_data.csv"), skip = 2)
+kid_pop <- read_xlsx(here("data/raw_data/Child population by age group.xlsx"))
 
@@
 ```{r}
-deaths_sex <- deaths %>% select(1:4, contains(c("male", "MSM")))
+deaths <- deaths %>% 
+  filter(!grepl("DC", "State Abbreviation")) %>%
+  filter(!grepl("PR", "State Abbreviation"))
 
-deaths_race <-   deaths %>% select(1:4, contains(c("Black", "White", "Hispanic",  "Multiple")))
+deaths_sex <- deaths %>%
+  select(1:4, contains("male"))
+
+deaths_race <-   deaths %>%
+  select(1:4, contains(c("Black", "White", "Hispanic",  "Multiple")))
 
@@
 ```{r}
-death_race <- deaths_race %>% 
+deaths_race <- deaths_race %>% 
   mutate(across(ends_with("Death Rate"), ~ case_when( .x < 0 ~ NA, TRUE ~ .x)))
@@
   pivot_longer(2:54 , names_to = "Location", values_to = "Population" ) %>%  
-  pivot_wider(names_from = Label, values_from= Population)
+  pivot_wider(names_from = Label, values_from = Population)
 ```
@@
   mutate(across(2:6, ~ as.numeric(.)))
-
 ```
@@
 ```{r}
-total_pop <- race_pop %>% select(1:2)
+total_pop <- race_pop %>%
+  select(1:2)
+
 pop_sex <- inner_join(total_pop, us_sex_pop)
@@
 ```{r}
-kid_pop <- read_xlsx(here("data/raw_data/Child population by age group.xlsx"))
 kid_pop <- kid_pop %>% 
@@
   summarise(Less13 = sum(Data))
+```
 
+```{r}
 new_pop <- full_join(pop_sex, pop_less13, by = join_by(Location))
@@
 
+new_pop <- new_pop %>%
+  filter(!grepl("District", Location)) %>%
+  filter(!grepl("United", Location)) %>% 
+  filter(!grepl("Puerto Rico", Location))
+
 df_pop <- new_pop %>% 
@@
             df_pop_female = sum(Female_Pop - Female_kid_pop))
-
 ```
@@
 ```{r}
-sex_df <- inner_join(deaths_sex, df_pop, by = join_by(State == Location))
+sex_df <- full_join(deaths_sex, df_pop, by = join_by(State == Location))
+```
+
+Let's find the HIV death rate by sex
+```{r}
+sex_df <- sex_df %>% 
+  mutate(M_Rate = `Male Death Cases`/df_pop_male*1e5,
+         F_Rate = `Female Death Cases`/df_pop_female*1e5)
+```
+
+Next, let's find the expected number of HIV deaths. First, we'll calculate lambda.
+```{r}
+lambda_male <- sex_df %>% drop_na(`Male Death Cases`) %>% 
+  drop_na(df_pop_male) %>% 
+  summarize(lambda_male = sum(`Male Death Cases`)/sum(df_pop_male)) %>%
+  as.numeric()
+
+lambda_female <- sex_df %>% drop_na(`Female Death Cases`) %>% 
+  drop_na(df_pop_female) %>% 
+  summarize(lambda_female = sum(`Female Death Cases`)/sum(df_pop_female)) %>%
+  as.numeric()
+
+sex_df <- sex_df %>%
+  mutate(Expected_Male_Cases = df_pop_male * lambda_male,
+         Expected_Female_Cases = df_pop_female *  lambda_female)
 ```
@@
 ```{r}
-sex_df %>% mutate(M_Rate = `Male Death Cases`/df_pop_male*1e5)
+sex_df <- sex_df %>% 
+  mutate(OE_Ratio_male = `Male Death Cases` / Expected_Male_Cases,
+         OE_Ratio_female = `Female Death Cases` / Expected_Female_Cases)
 ```
-
-Next find expected rate vs observed rate

### 2023-02-25 3_cleaning_race_df.Rmd
more cleaning

```{r, echo=FALSE, message=FALSE}
library(knitr)
library(tidyverse)
library(here)
```

```{r}
race_df <- readRDS(here("data/processed_data/race_df.rds"))
```
Let's find the HIV death rate by race
```{r}
race_df <- race_df %>% 
  mutate(black_death_rate = round(`Black Death Cases`/Black_pop*1e5),
         hispanic_death_rate = round(`Hispanic Death Cases`/Hispanic_pop*1e5),
         white_death_rate = round(`White Death Cases`/White_pop*1e5),
         multiple_death_rate = round(`Multiple Race Death Cases`/Multiple_pop*1e5))
```

Next, let's find the expected number of HIV deaths. First, we'll calculate lambda.
```{r}
lambda_black <- race_df %>%
  drop_na(`Black Death Cases`) %>% 
  summarize(lambda_black = sum(`Black Death Cases`)/sum(Black_pop)) %>%
  as.numeric()

lambda_hispanic <- race_df %>%
  drop_na(`Hispanic Death Cases`) %>% 
  summarize(lambda_hispanic = sum(`Hispanic Death Cases`)/sum(Hispanic_pop)) %>%
  as.numeric()

lambda_white <- race_df %>%
  drop_na(`White Death Cases`) %>% 
  summarize(lambda_white = sum(`White Death Cases`)/sum(White_pop)) %>%
  as.numeric()

lambda_multiple <- race_df %>%
  drop_na(`Multiple Race Death Cases`) %>% 
  summarize(lambda_multipe = sum(`Multiple Race Death Cases`)/sum(Multiple_pop)) %>%
  as.numeric()

race_df <- race_df %>%
  mutate(black_exp_deaths = round(Black_pop * lambda_black),
        white_exp_deaths = round(White_pop * lambda_white),
        hispanic_exp_deaths = round(Hispanic_pop *  lambda_hispanic),
        multiple_exp_deaths = round(Multiple_pop *  lambda_multiple))
```

```{r}
race_df <- race_df %>% 
  mutate(ratio_OE_black = `Black Death Cases` / black_exp_deaths,
         ratio_OE_hispanic = `Hispanic Death Cases` / hispanic_exp_deaths,
         ratio_OE_white = `White Death Cases` / white_exp_deaths,
         ratio_OE_multiple = `Multiple Race Death Cases` / multiple_exp_deaths)
```

### 2023-02-25 code/processing_code/1_cleaning_2020.Rmd
more cleaning

@@
 deaths <- read_xlsx(here("data/raw_data/AIDSVu_State_Death_2020.xlsx"), skip = 3)
-#late_dx <- read_xlsx(here("data/raw_data/AIDSVu_State_LateDiagnoses_2020.xlsx"), skip = 3)
 us_race_pop <- read_xlsx(here("data/raw_data/DECENNIALPL2020.P2-2023-02-24T124235.xlsx"), sheet = 2)
@@
 kid_pop <- read_xlsx(here("data/raw_data/Child population by age group.xlsx"))
+kid_race <- read_xlsx(here("data/raw_data/Child population by race and ethnicity.xlsx"))
 
-#new_dx <- read_xlsx(here("data/raw_data/AIDSVu_State_NewDX_2020.xlsx"), skip = 3)
-#prev  <- read_xlsx(here("data/raw_data/AIDSVu_State_Prev_2020.xlsx"), skip = 3)
-#rec_care  <- read_xlsx(here("data/raw_data/AIDSVu_State_ReceiptofCare_2020.xlsx"), skip = 3)
-#sdoh  <- read_xlsx(here("data/raw_data/AIDSVu_State_SDOH_2020.xlsx"), skip = 3)
-#v_sup  <- read_xlsx(here("data/raw_data/AIDSVu_State_ViralSuppression_2020.xlsx"), skip = 3)
+new_dx <- read_xlsx(here("data/raw_data/AIDSVu_State_NewDX_2020.xlsx"), skip = 3)
+late_dx <- read_xlsx(here("data/raw_data/AIDSVu_State_LateDiagnoses_2020.xlsx"), skip = 3)
+prev  <- read_xlsx(here("data/raw_data/AIDSVu_State_Prev_2020.xlsx"), skip = 3)
+rec_care  <- read_xlsx(here("data/raw_data/AIDSVu_State_ReceiptofCare_2020.xlsx"), skip = 3)
+sdoh  <- read_xlsx(here("data/raw_data/AIDSVu_State_SDOH_2020.xlsx"), skip = 3)
+v_sup  <- read_xlsx(here("data/raw_data/AIDSVu_State_ViralSuppression_2020.xlsx"), skip = 3)
 ```
@@
 deaths <- deaths %>% 
-  filter(!grepl("DC", "State Abbreviation")) %>%
-  filter(!grepl("PR", "State Abbreviation"))
+  filter(!grepl("DC", `State Abbreviation`)) %>%
+  filter(!grepl("PR", `State Abbreviation`))
 
 deaths_sex <- deaths %>%
-  select(1:4, contains("male"))
+  select(1:4, contains("male"), -contains("and"), -ends_with("Stability"))
 
 deaths_race <-   deaths %>%
-  select(1:4, contains(c("Black", "White", "Hispanic",  "Multiple")))
+  select(1:4, contains(c("Black", "White", "Hispanic",  "Multiple")), -ends_with("Stability"))
 
 summary(deaths_race)
+```
+```{r}
+new_dx <- new_dx %>% 
+  filter(!grepl("DC", `State Abbreviation`)) %>%
+  filter(!grepl("PR", `State Abbreviation`)) %>% 
+  select(1:3, contains(c("male", "Black", "White", "Hispanic")), -contains(c("and", "Stability")))
+
+sex_dx <- new_dx %>% 
+  select(1:7)
+
+new_dx <- new_dx %>% 
+  select(1:3,8:13)
 ```
@@
   pivot_wider(names_from = Label, values_from = Population)
+
+race_pop <- race_pop %>%
+  filter(!grepl("Puerto Rico", Location)) %>% 
+  filter(!grepl("District", Location)) 
 ```
@@
   mutate(across(2:6, ~ as.numeric(.)))
+
+total_pop <- race_pop %>%  select(1:2)
 ```
@@
 ```{r}
-total_pop <- race_pop %>%
-  select(1:2)
-
 pop_sex <- inner_join(total_pop, us_sex_pop)
@@
 
-Next, we need to estimate the population older than 12. I will try to estimate this using data from Kids Count.
-
+Next, we need to estimate the population older than 12. Since I cannot find a source explicitly population older than 12, I will estimate population younger than 13 this using data from Kids Count. 
 ```{r}
@@
 new_pop <- new_pop %>% 
-  mutate(Male_Pop = Male * `Total:`,
-  Female_Pop = Female * `Total:`,
-  Female_kid_pop = Female * Less13,
-  Male_kid_pop = Male * Less13)
+  mutate(Male_Pop = round(Male * `Total:`),
+  Female_Pop = round(Female * `Total:`),
+  Female_kid_pop = round(Female * Less13),
+  Male_kid_pop = round(Male * Less13))
 
@@
   group_by(Location) %>% 
-  summarize(df_pop_male = sum(Male_Pop - Male_kid_pop),
-            df_pop_female = sum(Female_Pop - Female_kid_pop))
+  summarize(pop_male = sum(Male_Pop - Male_kid_pop),
+            pop_female = sum(Female_Pop - Female_kid_pop))
 ```
@@
 sex_df <- full_join(deaths_sex, df_pop, by = join_by(State == Location))
-```
 
-Let's find the HIV death rate by sex
-```{r}
-sex_df <- sex_df %>% 
-  mutate(M_Rate = `Male Death Cases`/df_pop_male*1e5,
-         F_Rate = `Female Death Cases`/df_pop_female*1e5)
-```
-
-Next, let's find the expected number of HIV deaths. First, we'll calculate lambda.
-```{r}
-lambda_male <- sex_df %>% drop_na(`Male Death Cases`) %>% 
-  drop_na(df_pop_male) %>% 
-  summarize(lambda_male = sum(`Male Death Cases`)/sum(df_pop_male)) %>%
-  as.numeric()
-
-lambda_female <- sex_df %>% drop_na(`Female Death Cases`) %>% 
-  drop_na(df_pop_female) %>% 
-  summarize(lambda_female = sum(`Female Death Cases`)/sum(df_pop_female)) %>%
-  as.numeric()
-
-sex_df <- sex_df %>%
-  mutate(Expected_Male_Cases = df_pop_male * lambda_male,
-         Expected_Female_Cases = df_pop_female *  lambda_female)
+sex_df <- full_join(sex_df, sex_dx)
 ```
@@
 ```{r}
-sex_df <- sex_df %>% 
-  mutate(OE_Ratio_male = `Male Death Cases` / Expected_Male_Cases,
-         OE_Ratio_female = `Female Death Cases` / Expected_Female_Cases)
+glimpse(kid_race)
+
+kid_race <- kid_race %>% 
+  filter(TimeFrame == 2020,
+         DataFormat == "Percent") %>% 
+  mutate(Data = as.numeric(Data)) %>% 
+  filter(Race %in% c("Hispanic or Latino", "Non-Hispanic Black alone", "Non-Hispanic White alone", "Non-Hispanic Two or More Race Groups"))
+
+kid_race <- kid_race %>%
+  pivot_wider(Location, names_from = Race, values_from = Data) %>% 
+  rename(c(Hispanic_k = `Hispanic or Latino`,
+         White_k = `Non-Hispanic White alone`,
+         Black_k = `Non-Hispanic Black alone`,
+         Multiple_k = `Non-Hispanic Two or More Race Groups`))
+
+kid_race <- inner_join(kid_race, pop_less13)
 ```
+
+Let's estimate the number of children under 13 by race
+```{r}
+kid_race <- kid_race %>%
+  mutate(across(2:5, ~ round( . * Less13)))
+
+race_pop <- inner_join(race_pop, kid_race)
+
+race_pop <- race_pop %>% 
+  group_by(Location) %>% 
+  summarize(Black_pop = sum(Black - Black_k),
+            Hispanic_pop = sum(Hispanic - Hispanic_k),
+            White_pop = sum(White - White_k),
+            Multiple_pop = sum(Multiple - Multiple_k))
+```
+
+```{r}
+race_df <- list(deaths_race, late_dx, new_dx)
+
+race_df <- reduce(race_df, inner_join)
+
+race_df <- inner_join(race_df, race_pop, by = join_by(State == Location))
+```
+
+```{r}
+rec_care <- rec_care %>%
+  filter(!grepl("DC", `State Abbreviation`)) %>%
+  filter(!grepl("PR", `State Abbreviation`)) %>% 
+  select(1:3, contains("Percent of")) %>% 
+  select(1:3, contains(c("Hispanic", "Black", "White", "Multiple","male")))
+
+sex_df$`Male_rec_care_percent`<- rec_care$`Percent of Males Living with HIV who Received HIV Care`
+sex_df$`Female_rec_care_percent`<- rec_care$`Percent of Females Living with HIV who Received HIV Care`
+
+rec_care <- rec_care %>%
+  select(1:3, contains(c("Hispanic", "Black", "White", "Multiple")))
+
+race_df <- inner_join(race_df, rec_care)
+```
+```{r}
+v_sup <- v_sup %>%
+  filter(!grepl("DC", `State Abbreviation`)) %>%
+  filter(!grepl("PR", `State Abbreviation`)) %>% 
+  select(1:3, contains("Percent of")) %>% 
+  select(1:3, contains(c("Hispanic", "Black", "White", "Multiple","male")))
+
+sex_df$`Male_viral_sup_percent` <- v_sup$`Percent of Males Living with HIV who were Virally Suppressed`
+sex_df$`Female_viral_sup_percent` <- v_sup$`Percent of Females Living with HIV who were Virally Suppressed`
+
+v_sup <- v_sup %>%
+  select(1:3, contains(c("Hispanic", "Black", "White", "Multiple")))
+race_df <- inner_join(race_df, v_sup)
+```
+
+```{r}
+prev <- prev %>% 
+  filter(!grepl("DC", `State Abbreviation`)) %>%
+  filter(!grepl("PR", `State Abbreviation`))
+
+prev_sex <- prev %>%
+  select(1:4, contains("male"), -contains("and"), -ends_with("Stability"))
+
+sex_df <- full_join(sex_df, prev_sex)
+
+prev_race <-   prev %>%
+  select(1:4, contains(c("Black", "White", "Hispanic",  "Multiple")), -contains("and"), -ends_with("Stability"))
+
+race_df <- full_join(race_df, prev_race)
+```
+
+```{r}
+saveRDS(sex_df, here("data/processed_data/sex_df.rds"))
+saveRDS(race_df, here("data/processed_data/race_df.rds"))
+```
+
+```{r}
+rm(deaths, df_pop, kid_pop, pop_sex, pop_less13, us_sex_pop, us_race_pop, new_pop, deaths_sex, race_pop, kid_race, total_pop, deaths_race, late_dx, new_dx, sex_dx, prev, prev_sex, prev_race, v_sup, rec_care)
+```

### 2023-02-25 code/processing_code/2_cleaning_sex_df.Rmd
more cleaning

```{r, echo=FALSE, message=FALSE}
library(knitr)
library(tidyverse)
library(here)
```

```{r}
sex_df <- readRDS(here("data/processed_data/sex_df.rds"))
```
Let's find the HIV death rate by sex
```{r}
sex_df <- sex_df %>% 
  mutate(M_Rate = round(`Male Death Cases`/pop_male*1e5),
         F_Rate = round(`Female Death Cases`/pop_female*1e5))
```

Next, let's find the expected number of HIV deaths. First, we'll calculate lambda.
```{r}
lambda_male <- sex_df %>% drop_na(`Male Death Cases`) %>% 
  drop_na(pop_male) %>% 
  summarize(lambda_male = sum(`Male Death Cases`)/sum(pop_male)) %>%
  as.numeric()

lambda_female <- sex_df %>% drop_na(`Female Death Cases`) %>% 
  drop_na(pop_female) %>% 
  summarize(lambda_female = sum(`Female Death Cases`)/sum(pop_female)) %>%
  as.numeric()

sex_df <- sex_df %>%
  mutate(male_exp_deaths = round(pop_male * lambda_male),
         female_exp_deaths = round(pop_female *  lambda_female))
```

```{r}
sex_df <- sex_df %>% 
  mutate(ratio_OE_male = `Male Death Cases` / male_exp_deaths,
         ratio_OE_female = `Female Death Cases` / female_exp_deaths)
```

### 2023-02-25 3_cleaning_race_df.Rmd
completed cleaning

@@
 ```
+
+```{r}
+black_df <- race_df %>% select(1:3, contains("Black"))
+
+black_df <- black_df %>% 
+  mutate(Race = "Black") %>% 
+ rename("Death Rate (AIDSVu)" = `Black Death Rate`,
+        "Population" = Black_pop,
+        "Death Cases" = "Black Death Cases",
+        "New Diagnoses Rate" = "New Diagnoses Black Rate",
+        "New Diagnoses Cases" = "New Diagnoses Black Cases",
+        "Percentage Diagnosed Late" = `Percent of Black Persons Diagnosed Late`,
+        "Number of Persons Diagnosed Late" = `Number of Black Persons Diagnosed Late`,
+        "Percentage Received Care" = `Percent of Black Persons Living with HIV who Received HIV Care`,
+        "Percentage Virally Suppressed" = `Percent of Black Persons Living with HIV who were Virally Suppressed`,
+        "HIV Prevalence Rate" = "Black Rate",
+        "Prevalent HIV Cases" = "Black Cases",
+        "Expected Death Cases" = black_exp_deaths,
+        "Ratio Observed/Expected" = ratio_OE_black,
+        "Death Rate (estimated)" = black_death_rate)
+
+hispanic_df <- race_df %>% select(1:3, contains("Hispanic"))
+
+hispanic_df <- hispanic_df %>% 
+  mutate(Race = "Hispanic") %>% 
+ rename("Death Rate (AIDSVu)" = `Hispanic Death Rate`,
+        "Population" = Hispanic_pop,
+        "Death Cases" = "Hispanic Death Cases",
+        "New Diagnoses Rate" = "New Diagnoses Hispanic Rate",
+        "New Diagnoses Cases" = "New Diagnoses Hispanic Cases",
+        "Percentage Diagnosed Late" = `Percent of Hispanic Persons Diagnosed Late`,
+        "Number of Persons Diagnosed Late" = `Number of Hispanic Persons Diagnosed Late`,
+        "Percentage Received Care" = `Percent of Hispanic Persons Living with HIV who Received HIV Care`,
+        "Percentage Virally Suppressed" = `Percent of Hispanic Persons Living with HIV who were Virally Suppressed`,
+        "HIV Prevalence Rate" = "Hispanic Rate",
+        "Prevalent HIV Cases" = "Hispanic Cases",
+        "Expected Death Cases" = hispanic_exp_deaths,
+        "Ratio Observed/Expected" = ratio_OE_hispanic,
+        "Death Rate (estimated)" = hispanic_death_rate)
+
+white_df <- race_df %>% select(1:3, contains("White"))
+
+white_df <- white_df %>% 
+  mutate(Race = "White") %>% 
+ rename("Death Rate (AIDSVu)" = `White Death Rate`,
+        "Population" = White_pop,
+        "Death Cases" = "White Death Cases",
+        "New Diagnoses Rate" = "New Diagnoses White Rate",
+        "New Diagnoses Cases" = "New Diagnoses White Cases",
+        "Percentage Diagnosed Late" = `Percent of White Persons Diagnosed Late`,
+        "Number of Persons Diagnosed Late" = `Number of White Persons Diagnosed Late`,
+        "Percentage Received Care" = `Percent of White Persons Living with HIV who Received HIV Care`,
+        "Percentage Virally Suppressed" = `Percent of White Persons Living with HIV who were Virally Suppressed`,
+        "HIV Prevalence Rate" = "White Rate",
+        "Prevalent HIV Cases" = "White Cases",
+        "Expected Death Cases" = white_exp_deaths,
+        "Ratio Observed/Expected" = ratio_OE_white,
+        "Death Rate (estimated)" = white_death_rate)
+
+multiple_df <- race_df %>% select(1:3, contains("Multiple"))
+
+multiple_df <- multiple_df %>% 
+  mutate(Race = "Multiple") %>% 
+ rename("Death Rate (AIDSVu)" = "Multiple Race Death Rate",
+        "Population" = Multiple_pop,
+        "Death Cases" = "Multiple Race Death Cases",
+        "New Diagnoses Rate" = "New Diagnoses Multiple Race Rate",
+        "New Diagnoses Cases" = "New Diagnoses Multiple Race Cases",
+        "Percentage Received Care" = `Percent of Multiple Race Persons Living with HIV who Received HIV Care`,
+        "Percentage Virally Suppressed" = `Percent of Multiple Race Persons Living with HIV who were Virally Suppressed`,
+        "HIV Prevalence Rate" = "Multiple Races Rate",
+        "Prevalent HIV Cases" = "Multiple Races Cases",
+        "Expected Death Cases" = multiple_exp_deaths,
+        "Ratio Observed/Expected" = ratio_OE_multiple,
+        "Death Rate (estimated)" = multiple_death_rate)
+
+race_df <- list(black_df, hispanic_df, white_df, multiple_df)
+
+race_df <- reduce(race_df, full_join)
+```
+
+```{r}
+saveRDS(race_df, here("data/processed_data/race_df_cleaned.rds"))
+```

### 2023-02-25 code/processing_code/1_cleaning_2020.Rmd
completed cleaning

@@
   filter(!grepl("PR", `State Abbreviation`)) %>% 
-  select(1:3, contains(c("male", "Black", "White", "Hispanic")), -contains(c("and", "Stability")))
+  select(1:3, contains(c("male", "Black", "White", "Hispanic", "Multiple")), -contains(c("and", "Stability")))
 
@@
 new_dx <- new_dx %>% 
-  select(1:3,8:13)
+  select(-contains(c("male", "Year")))
 ```

### 2023-02-25 code/processing_code/2_cleaning_sex_df.Rmd
completed cleaning

@@
 sex_df <- sex_df %>% 
-  mutate(M_Rate = round(`Male Death Cases`/pop_male*1e5),
-         F_Rate = round(`Female Death Cases`/pop_female*1e5))
+  mutate(Male_Rate = round(`Male Death Cases`/pop_male*1e5),
+         Female_Rate = round(`Female Death Cases`/pop_female*1e5)) %>% 
+  select(!`Year`)
 ```
@@
 ```
+
+```{r}
+female_df <- sex_df %>% select(1:3, contains("female")) 
+male_df <- sex_df %>% select(1:3, !contains("female"))
+
+female_df <- female_df %>% 
+  mutate(Sex = "Female") %>% 
+ rename("Death Rate (AIDSVu)" = `Female Death Rate`,
+        "Population" = pop_female,
+        "Death Cases" = "Female Death Cases",
+        "New Diagnoses Rate" = "New Diagnoses Female Rate",
+        "New Diagnoses Cases" = "New Diagnoses Female Cases",
+        "Percentage Received Care" = "Female_rec_care_percent",
+        "Percentage Virally Suppressed" = "Female_viral_sup_percent",
+        "HIV Prevalence Rate" = "Female Rate",
+        "Prevalent HIV Cases" = "Female Cases",
+        "Expected Death Cases" = female_exp_deaths,
+        "Ratio Observed/Expected" = ratio_OE_female,
+        "Death Rate (estimated)" = Female_Rate)
+
+male_df <- male_df %>% 
+  mutate(Sex = "Male") %>% 
+ rename("Death Rate (AIDSVu)" = `Male Death Rate`,
+        "Population" = pop_male,
+        "Death Cases" = "Male Death Cases",
+        "New Diagnoses Rate" = "New Diagnoses Male Rate",
+        "New Diagnoses Cases" = "New Diagnoses Male Cases",
+        "Percentage Received Care" = "Male_rec_care_percent",
+        "Percentage Virally Suppressed" = "Male_viral_sup_percent",
+        "HIV Prevalence Rate" = "Male Rate",
+        "Prevalent HIV Cases" = "Male Cases",
+        "Expected Death Cases" = male_exp_deaths,
+        "Ratio Observed/Expected" = ratio_OE_male,
+        "Death Rate (estimated)" = Male_Rate) %>% 
+  select(!`State Rate`)
+
+sex_df <- full_join(female_df, male_df)
+
+```
+
+```{r}
+saveRDS(sex_df, here("data/processed_data/sex_df_cleaned.rds"))
+```

### 2023-02-26 code/analysis_code/1_EDA_sex.Rmd
eda

```{r message=FALSE}
library(here)
library(GGally)
library(skimr)
library(ggfortify)
library(car)
library(bestNormalize)
library(MASS)
library(tidyverse)
library(readxl)
```
Loading data
```{r}
sex_df <- readRDS(here("data/processed_data/sex_df_cleaned.rds"))
```

Creating `Expected Death Rate` that will be used to calculate the standardized mortality ratio. Filter for only states in the contiguous United States.
```{r}
sex_df <- sex_df %>% 
 mutate(`Expected Death Rate` = (`Expected Death Cases`/Population)*1e5,
        `Standardized Mortality Ratio` = `Death Rate (estimated)`/(`Expected Death Rate`)) %>%
  filter(!grepl("Alaska", State),
         !grepl("Hawaii", State))
```
Carrying out some diagnostic plots on key variables. 
```{r}
autoplot(lm(sex_df$`Death Rate (AIDSVu)` ~ 1))
autoplot(lm(sex_df$`Death Rate (estimated)` ~ 1))
autoplot(lm(sex_df$`Ratio Observed/Expected` ~ 1))
autoplot(lm(sex_df$`Standardized Mortality Ratio` ~ 1))
```

```{r}
summary(sex_df$`Death Rate (AIDSVu)`)
summary(sex_df$`Death Rate (estimated)`)
summary(sex_df$`Ratio Observed/Expected`)
summary(sex_df$`Standardized Mortality Ratio`)
```
It appears that our data does not follow a normal distribution. Let's start by obtaining a histogram.
```{r}
sex_df %>% 
  ggplot() +
  geom_histogram(aes(x = `Death Rate (estimated)`, fill = Sex), bins = 10)
shapiro.test(sex_df$`Death Rate (estimated)`)

sex_df %>% 
  ggplot() +
  geom_histogram(aes(x = `Ratio Observed/Expected`, fill = Sex), bins = 10)
shapiro.test(sex_df$`Ratio Observed/Expected`)

sex_df %>% 
  ggplot() +
  geom_histogram(aes(x = `Standardized Mortality Ratio`, fill = Sex), bins = 10)
shapiro.test(sex_df$`Standardized Mortality Ratio`)
```

Visual assessment and the sharpiro test are in agreement that the data is not normally distributed. Let's see if log (x +1) transformation can help resolve this issue.

```{r}
sex_df %>% 
  ggplot() +
  geom_histogram(aes(x = log(`Death Rate (estimated)` + 1), fill = Sex) , bins = 10)
shapiro.test(log(sex_df$`Death Rate (estimated)` + 1))

sex_df %>% 
  ggplot() +
  geom_histogram(aes(x = log(`Ratio Observed/Expected` + 1), fill = Sex) , bins = 10)
shapiro.test(log(sex_df$`Ratio Observed/Expected` + 1))

sex_df %>% 
  ggplot() +
  geom_histogram(aes(x = log(`Standardized Mortality Ratio` +1), fill = Sex) , bins = 10)
shapiro.test(log(sex_df$`Standardized Mortality Ratio` + 1))
```

While this is much better, we found sufficient evidence to reject that the distribution is normal with p values < 0.05. Let's see if another transformation method can achieve better results. Other options include power transformations, like Box Cox transformation or Yeo Johnson transformation, and mixture models, such as the zero-inflated model (Poisson or negative binomial).

I'll try the Box-Cox transformation first. A limitation of this is that Box Cox cannot values of zero (or negative numbers). We will have to subset observations greater than zero to continue with the transformation. We will use boxcox() to find lambda. 
`library(MASS)`
```{r}
sex_df_DR <- sex_df %>% 
  filter(`Death Rate (estimated)` > 0)
b_DR <-  boxcox(lm(sex_df_DR$`Death Rate (estimated)` ~ 1))
lambda_DR <- b_DR$x[which.max(b_DR$y)]

sex_df_ROE <- sex_df %>% 
  filter(`Ratio Observed/Expected` > 0)

b_ROE <-  boxcox(lm(sex_df_ROE$`Ratio Observed/Expected` ~ 1))
lambda_ROE <- b_ROE$x[which.max(b_ROE$y)]

sex_df_SMR <- sex_df %>% 
  filter(`Standardized Mortality Ratio`  > 0)
b_SMR <-  boxcox(lm(sex_df_SMR$`Standardized Mortality Ratio` ~ 1))
lambda_SMR <- b_SMR$x[which.max(b_SMR$y)]

sex_df_SMR <- sex_df %>% 
  filter(`Standardized Mortality Ratio`  > 0)

```

We'll create a function to carry out the box cox transformation [cite](https://stats.stackexchange.com/questions/1444/how-should-i-transform-non-negative-data-including-zeros)
```{r}
box_cox <- function(x, lambda) {

    eps <- 0.00001
    if (abs(lambda) < eps)
        log(x)
    else
        (x ^ lambda - 1) / lambda

}
```

Box Cox transformation of DR and results of shapiro test.
```{r}
sex_df_DR %>%
  mutate(BOX = box_cox(`Death Rate (estimated)`, lambda_DR)) %>%
  ggplot() + geom_histogram(aes(x= BOX, fill = Sex), bins = 10)

sex_df_DR <- sex_df_DR %>%
  mutate(BOX = box_cox(`Death Rate (estimated)`, lambda_DR))

sex_df_DR %>%
  pull(BOX) %>% 
  shapiro.test()

sex_df_DR %>%
  filter(Sex == "Female") %>% 
  pull(BOX) %>% 
  shapiro.test()

sex_df_DR %>%
  filter(Sex == "Male") %>% 
  pull(BOX) %>% 
  shapiro.test()

par(mfrow=c(1,3))
plot(lm(sex_df_DR$BOX ~ 1))
```
Unsuccessful.

Box Cox transformation of ROE and results of shapiro test.
```{r}
sex_df_ROE %>%
  mutate(BOX = box_cox(`Ratio Observed/Expected`, lambda_ROE)) %>%
  ggplot() + geom_histogram(aes(x= BOX, fill = Sex), bins = 10)

sex_df_ROE <- sex_df_ROE %>%
  mutate(BOX = box_cox(`Ratio Observed/Expected`, lambda_ROE))

sex_df_ROE %>%
  pull(BOX) %>% 
  shapiro.test()

sex_df_ROE %>%
  filter(Sex == "Female") %>% 
  pull(BOX) %>% 
  shapiro.test()

sex_df_ROE %>%
  filter(Sex == "Male") %>% 
  pull(BOX) %>% 
  shapiro.test()

par(mfrow=c(1,3))
plot(lm(sex_df_ROE$BOX ~ 1))
```
The Box Cox transformation was able to normalize the `Ratio Observed/Expected`. We will use the Box Cox transformation method for analyses with `Ratio Observed/Expected`.

Box Cox transformation of SMR and results of shapiro test.
```{r}
sex_df_SMR %>%
  mutate(BOX = box_cox(`Standardized Mortality Ratio`, lambda_SMR)) %>%
  ggplot() + geom_histogram(aes(x= BOX, fill = Sex), bins =10)

sex_df_SMR <- sex_df_SMR %>%
  mutate(BOX = box_cox(`Standardized Mortality Ratio`, lambda_SMR))

sex_df_SMR %>%
  filter(Sex == "Female") %>% 
  pull(BOX) %>% 
  shapiro.test()

sex_df_SMR %>%
  filter(Sex == "Male") %>% 
  pull(BOX) %>% 
  shapiro.test()

par(mfrow=c(1,3))
plot(lm(sex_df_SMR$BOX ~ 1))
```
The Box Cox transformation was not able to normalize the `Standardized Mortality Ratio`. We will move on the Yeo Johnson transformation

We will perform Yeo Johnson transformation on `Death Rate (estimated)` and `Standardized Mortality Ratio`.We will use `yeojohnson` from `BestNormalize` package to extract lambda and `yeo_johnson` from `car` to compute the transformation for each observation.
```{r}
yj_DR <- yeojohnson(sex_df_SMR$`Death Rate (estimated)`)
yj_DR <- yj_DR$lambda

sex_df <- sex_df %>%
  mutate(YJ = yeo_johnson(`Death Rate (estimated)`, yj_DR))

sex_df %>%
  ggplot() + geom_histogram(aes(x= YJ, fill = Sex), bins =10)

sex_df %>%
  pull(YJ) %>% 
  shapiro.test()

sex_df %>%
  filter(Sex == "Female") %>% 
  pull(YJ) %>% 
  shapiro.test()

sex_df %>%
  filter(Sex == "Male") %>% 
  pull(YJ) %>% 
  shapiro.test()

par(mfrow=c(1,3))
plot(lm(sex_df$YJ ~ 1))
```
Unable to normalized the distribution `Death Rate (estimated)` with the Yeo Johnson transformation.

```{r}
yj_SMR <- yeojohnson(sex_df_SMR$`Standardized Mortality Ratio`)
yj_SMR <- yj_SMR$lambda

sex_df_SMR <- sex_df_SMR%>%
  mutate(YJ = yeo_johnson(`Standardized Mortality Ratio`, yj_SMR))

sex_df_SMR %>%
  ggplot() + geom_histogram(aes(x= YJ, fill = Sex), bins =10)

sex_df_SMR %>%
  pull(YJ) %>% 
  shapiro.test()

sex_df_SMR %>%
  filter(Sex == "Female") %>% 
  pull(YJ) %>% 
  shapiro.test()

sex_df_SMR %>%
  filter(Sex == "Male") %>% 
  pull(YJ) %>% 
  shapiro.test()

par(mfrow=c(1,3))
plot(lm(sex_df_SMR$YJ ~ 1))
```
The Yeo Johnson transformation was only able to normalize the SMR for male, but not females or for both sexes combined.

Plotting correlation matrix
```{r message=FALSE, warning=FALSE}
X <- sex_df[,c(9:10, 16:25)]
ggpairs(X)
```

```{r}
saveRDS(sex_df_ROE, here("data/process_data/sex_new.rds"))
```

### 2023-02-26 code/analysis_code/2_EDA_sex.Rmd
eda

```{r}
library(here)
library(tidyverse)
library(spdep)
```
https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state
List-of-neighboring-states-for-each-US-state
```{r}
sex_df <- readRDS(here("data/processed_data/sex_new.rds"))
us_nabs <- read_csv(here("data/raw_data/neighbors-states.csv"))

us_nabs <- us_nabs %>%
  filter(!grepl("AK", StateCode),
         !grepl("HI", StateCode), 
         !grepl("DC", StateCode))
```

```{r}
male_df <- sex_df %>%
  filter(!grepl("Alaska", State),
         !grepl("Hawaii", State),
         Sex == "Male")

h_male <- matrix(0, nrow(male_df), nrow(male_df))
h_male[cbind(match(us_nabs$StateCode, male_df$`State Abbreviation`), match(us_nabs$NeighborStateCode, male_df$`State Abbreviation`))]<-1
nablist_male <- mat2listw(h_male)
```

```{r}
par(mfrow=c(1,2))
moran.test(male_df$BOX, nablist_male, zero.policy = TRUE, randomisation = FALSE)
moran.plot(male_df$BOX, nablist_male, zero.policy = TRUE)

lisa_male <-localmoran(male_df$`Ratio Observed/Expected`,  nablist_male)
summary(lisa_male[,1])
```

Since the data frame for female is missing observations (due to 0) for Vermont and New Hampshire, we need to remove these from the neighbor list.
```{r}
female_df <- sex_df %>%
  filter(!grepl("Alaska", State),
         !grepl("Hawaii", State),
         Sex == "Female")
us_nabs_f <- us_nabs %>%
  filter(!grepl("VT", StateCode),
         !grepl("NH", StateCode))

h_female <- matrix(0, nrow(female_df), nrow(female_df))
h_female[cbind(match(us_nabs_f$StateCode, female_df$`State Abbreviation`), match(us_nabs_f$NeighborStateCode, female_df$`State Abbreviation`))]<-1
nablist_female <- mat2listw(h_female)

par(mfrow=c(1,2))
moran.test(female_df$BOX, nablist_female, zero.policy = TRUE, randomisation = FALSE)
moran.plot(female_df$BOX, nablist_female, zero.policy = TRUE)

lisa_female <-localmoran(female_df$`Ratio Observed/Expected`,  nablist_female)
summary(lisa_female[,1])
```

### 2023-02-26 code/processing_code/3_cleaning_race_df.Rmd
eda

--- 2023-02-25 3_cleaning_race_df.Rmd
@@
 ```{r}
-black_df <- race_df %>% select(1:3, contains("Black"))
+black_df <- race_df %>% select(1:3, contains("Black"), 50:62)
 
@@
 
-hispanic_df <- race_df %>% select(1:3, contains("Hispanic"))
+hispanic_df <- race_df %>% select(1:3, contains("Hispanic"), 50:62)
 
@@
 
-white_df <- race_df %>% select(1:3, contains("White"))
+white_df <- race_df %>% select(1:3, contains("White"), 50:62)
 
@@
 
-multiple_df <- race_df %>% select(1:3, contains("Multiple"))
+multiple_df <- race_df %>% select(1:3, contains("Multiple"), 50:62)
 

### 2023-02-27 code/processing_code/4_cleaning_us_nabs.Rmd
cleaning for reproducibility

This is the list of neighboring states that was obtained from a user on [GitHub](https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state). After some 'investigation' (troubleshooting code), this file is not exactly what I need to create the neighbor list. What is achieved by the following code took me a few seconds to do in Excel with copy and paste. But for the sake of reproducibility, I decided it's well-worth it some spend a little bit of time do this in R.

```{r}
library(reshape2)
library(tidyverse)
library(here)

us_nabs <- read_csv(here("data/raw_data/neighbors-states.csv"))

us_nabs_copy <- us_nabs
```

```{r}
nabs <- full_join(us_nabs, us_nabs_copy, by = join_by(StateCode == NeighborStateCode, NeighborStateCode == StateCode))
```

```{r}
saveRDS(nabs, here("data/processed_data/us_nabs.rds"))
```

### 2023-02-27 code/analysis_code/2_sex.Rmd
comments

--- 2023-02-26 code/analysis_code/2_EDA_sex.Rmd
@@
 sex_df <- readRDS(here("data/processed_data/sex_new.rds"))
-us_nabs <- read_csv(here("data/raw_data/neighbors-states.csv"))
+us_nabs <- readRDS(here("data/processed_data/us_nabs.rds"))
 

### 2023-02-27 code/processing_code/1_cleaning_2020.Rmd
eda
comments

@@
+Cleaning Part 1
+
+Load necessary libraries
 ```{r, echo=FALSE, message=FALSE}
@@
 
+Load the files to be cleaned.
 ```{r}
@@
 We'll separating by sex and race/ethnicity.
-This analysis will be limited to Non-Hispanic Black, Non-Hispanic White, and Hispanic persons.
+This analysis will be limited to Non-Hispanic Black, Non-Hispanic White, Hispanic persons, and persons of Multiple racial backgrounds.
 ```{r}
@@
 ```
+Some rates are -9 due to missingness. Let's convert these to NA
+```{r}
+deaths_race <- deaths_race %>% 
+  mutate(across(ends_with("Death Rate"), ~ case_when( .x < 0 ~ NA, TRUE ~ .x)))
+```
+
+Next, let's clean the dataframe for New diagnoses. We'll select for variables of interest and subset.
 ```{r}
@@
 
-Some rates are -9 due to missingness. Let's convert these to NA
-```{r}
-deaths_race <- deaths_race %>% 
-  mutate(across(ends_with("Death Rate"), ~ case_when( .x < 0 ~ NA, TRUE ~ .x)))
-```
-
-Next, let's move on to the census data. I'll start by re-formatting the data from wide format for States to long, and format race/ethnicity into wide.
+Moving on to the census data. I'll start by re-formatting the data from wide format for States to long, and format race/ethnicity into wide.
 ```{r}
@@
 
+I'll select only the columns of interest, and rename some of them so they are a bit easier to work with.
 ```{r}
@@
 
-total_pop <- race_pop %>%  select(1:2)
+total_pop <- race_pop %>%
+  select(1:2)
 ```
 
-Then, I'll calculate the counts of males and females in each location.
-```{r}
-pop_sex <- inner_join(total_pop, us_sex_pop)
-```
-
-Next, we need to estimate the population older than 12. Since I cannot find a source explicitly population older than 12, I will estimate population younger than 13 this using data from Kids Count. 
+I also need to estimate the population older than 12. Since I cannot find a source explicitly population older than 12, I will estimate population younger than 13 this using data from Kids Count. 
 ```{r}
@@
 
+Then, I'll calculate the counts of males and females in each location using percentage estimates from the KFF and removing the population under 13 years old.
 ```{r}
-new_pop <- full_join(pop_sex, pop_less13, by = join_by(Location))
+new_pop <- list(total_pop, pop_less13, us_sex_pop)
+
+new_pop <- reduce(new_pop, inner_join, by = join_by(Location))
 
@@
 
+Merging dataframes. 
 ```{r}
@@
 ```
-
+We need to repeat the steps above, but with race and ethnicity.
+We'll load the file containing information on the racial/ethnic make up of the population under 18.
+This contains a lot of data, so we'll be sure to only select the data we need.
 ```{r}
@@
 
-Let's estimate the number of children under 13 by race
+Let's estimate the number of children under 13 by race.
+We'll use mutate and across to provide these calculations for each columns for race/ ethnicity.
+Then, we'll merge and subtract the population under 13 years old from the analysis.
 ```{r}
@@
 
+Merging datasets. 
 ```{r}
@@
 
+Cleaning the file `Received Care`. I do some cleaning and merge appropriate dataframes, and use some base R to add a couple columns to `sex_df`.
 ```{r}
@@
 
-sex_df$`Male_rec_care_percent`<- rec_care$`Percent of Males Living with HIV who Received HIV Care`
-sex_df$`Female_rec_care_percent`<- rec_care$`Percent of Females Living with HIV who Received HIV Care`
+sex_df$Male_rec_care_percent <- rec_care$`Percent of Males Living with HIV who Received HIV Care`
+sex_df$Female_rec_care_percent <- rec_care$`Percent of Females Living with HIV who Received HIV Care`
 
@@
 ```
+Repeat for file on `Viral Suppression`
 ```{r}
@@
 
+Cleaning Prevalence file
 ```{r}
@@
 
+Some quick cleaning of the social determinant sof health file, and merging this with our two results dataframes `sex_df` and `race_df`.
+```{r}
+sdoh <- sdoh %>% 
+  filter(!grepl("DC", `State Abbreviation`)) %>% 
+   filter(!grepl("PR", `State Abbreviation`))
+
+sex_df <- full_join(sex_df, sdoh)
+
+race_df <- full_join(race_df,sdoh)
+```
+
+We have some more cleaning to, but let's save these files for now. 
 ```{r}
@@
 ```{r}
-rm(deaths, df_pop, kid_pop, pop_sex, pop_less13, us_sex_pop, us_race_pop, new_pop, deaths_sex, race_pop, kid_race, total_pop, deaths_race, late_dx, new_dx, sex_dx, prev, prev_sex, prev_race, v_sup, rec_care)
+rm(deaths, df_pop, kid_pop, pop_sex, pop_less13, us_sex_pop, us_race_pop, new_pop, deaths_sex, race_pop, kid_race, total_pop, deaths_race, late_dx, new_dx, sex_dx, prev, prev_sex, prev_race, v_sup, rec_care, sdoh)
 ```

### 2023-02-27 code/processing_code/2_cleaning_sex_df.Rmd
eda
comments

@@
 ```
-Let's find the HIV death rate by sex
+
+Let's find the HIV death rate by sex. We use round to get x to the closest whole number.
 ```{r}
@@
 
-Next, let's find the expected number of HIV deaths. First, we'll calculate lambda.
+Next, let's find the expected number of HIV deaths. First, we'll calculate lambda, which we will then multiply by population to get our expected number of deaths.
 ```{r}
-lambda_male <- sex_df %>% drop_na(`Male Death Cases`) %>% 
+lambda_male <- sex_df %>%
+  drop_na(`Male Death Cases`) %>% 
   drop_na(pop_male) %>% 
@@
 
-lambda_female <- sex_df %>% drop_na(`Female Death Cases`) %>% 
+lambda_female <- sex_df %>%
+  drop_na(`Female Death Cases`) %>% 
   drop_na(pop_female) %>% 
@@
 
+Creating variable for ratio of observed deaths over expected deaths.
 ```{r}
@@
 
+Let's tidy this all up. We'll separate the dataframe briefly by sex, then re-join!
 ```{r}
-female_df <- sex_df %>% select(1:3, contains("female")) 
+female_df <- sex_df %>% select(1:3, contains("female"), 23:35) 
 male_df <- sex_df %>% select(1:3, !contains("female"))
@@
 ```
-
+Save the file. Next, we'll clean up the file for race/ethnicity.
 ```{r}

### 2023-02-27 code/analysis_code/1_EDA_sex.Rmd
comments
changes

@@
 library(ggfortify)
-library(car)
-library(bestNormalize)
 library(MASS)
@@
 ```
-Unsuccessful.
+Unsuccessful at normalizing distribution of Death Rate, and overall worse results than log transformation, except for the males.
 
@@
 ```
-The Box Cox transformation was not able to normalize the `Standardized Mortality Ratio`. We will move on the Yeo Johnson transformation
+The Box Cox transformation was not technically able to normalize the `Standardized Mortality Ratio`. We will move on the Yeo Johnson transformation
 
-We will perform Yeo Johnson transformation on `Death Rate (estimated)` and `Standardized Mortality Ratio`.We will use `yeojohnson` from `BestNormalize` package to extract lambda and `yeo_johnson` from `car` to compute the transformation for each observation.
+We will perform Yeo Johnson transformation on `Death Rate (estimated)` and `Standardized Mortality Ratio`.We will use `yeojohnson` from `BestNormalize` package to extract lambda and `yeo_johnson` to compute the transformation for each observation.
+Code for `yeo_johnson` obtained from [GitHub](https://stats.stackexchange.com/questions/1444/how-should-i-transform-non-negative-data-including-zeros)
 ```{r}
-yj_DR <- yeojohnson(sex_df_SMR$`Death Rate (estimated)`)
+library(bestNormalize)
+
+yeo_johnson <- function(x, lambda) {
+
+    eps <- .000001
+    not_neg <- which(x >= 0)
+    is_neg  <- which(x < 0)
+
+    not_neg_trans <- function(x, lambda) {
+        if (abs(lambda) < eps) log(x + 1)
+        else ((x + 1) ^ lambda - 1) / lambda
+    }
+
+    neg_trans <- function(x, lambda) {
+        if (abs(lambda - 2) < eps) - log(-x + 1)
+        else - ((-x + 1) ^ (2 - lambda) - 1) / (2 - lambda)
+    }
+
+    x[not_neg] <- not_neg_trans(x[not_neg], lambda)
+
+    x[is_neg] <- neg_trans(x[is_neg], lambda)
+
+    return(x)
+
+}
+
+```
+
+```{r}
+yj_DR <- yeojohnson(sex_df_DR$`Death Rate (estimated)`)
 yj_DR <- yj_DR$lambda
@@
 ```
-Unable to normalized the distribution `Death Rate (estimated)` with the Yeo Johnson transformation.
+Unable to normalized the distribution `Death Rate (estimated)` with the Yeo Johnson transformation, but improved for overall and males compared to log transformation (not for females).
 
@@
 ```
-
+Let's save the dataframe for ROE since this one was able to be normalized after Box Cox transformation.
 ```{r}
-saveRDS(sex_df_ROE, here("data/process_data/sex_new.rds"))
+saveRDS(sex_df_ROE, here("data/processed_data/sex_new.rds"))
 ```

### 2023-03-18 code/analysis_code/Race/p1_EDA_race.Rmd
more EDA and modeling

```{r message=FALSE, warning=FALSE}
library(here)
library(spdep)
library(GGally)
library(skimr)
library(ggfortify)
library(MASS)
library(ggcorrplot)
library("FactoMineR")
library(tidyverse)
```
https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state
List-of-neighboring-states-for-each-US-state
```{r}
race_df <- readRDS(here("data/processed_data/race_df_cleaned.rds"))
us_nabs <- readRDS(here("data/processed_data/us_nabs.rds"))

#Subsetting to only contiguous states.
race_df <- race_df %>%
  filter(!grepl("Alaska", State),
         !grepl("Hawaii", State))

us_nabs <- us_nabs %>%
  filter(!grepl("AK", StateCode),
         !grepl("HI", StateCode), 
         !grepl("DC", StateCode))
```

```{r}
autoplot(lm(race_df$`Death Rate (AIDSVu)` ~ 1))
autoplot(lm(race_df$`Death Rate (estimated)` ~ 1))
autoplot(lm(race_df$`Ratio Observed/Expected` ~ 1))
```

```{r}
race_df %>% 
  ggplot() +
  geom_histogram(aes(x = `Death Rate (estimated)`, fill = Race), bins = 10)
shapiro.test(race_df$`Death Rate (estimated)`)

race_df %>% 
  ggplot() +
  geom_histogram(aes(x = `Ratio Observed/Expected`, fill = Race), bins = 10)
shapiro.test(race_df$`Ratio Observed/Expected`)
```
Both histograms show a positively skewed distribution. ROE appears to be the slightest bit more normal.

```{r}
race_df %>% 
  ggplot() +
  geom_histogram(aes(x = log(`Death Rate (estimated)` + 1), fill = Race) , bins = 10)
shapiro.test(log(race_df$`Death Rate (estimated)` + 1))

race_df %>% 
  ggplot() +
  geom_histogram(aes(x = log(`Ratio Observed/Expected` + 1), fill = Race) , bins = 10)
shapiro.test(log(race_df$`Ratio Observed/Expected` + 1))
```
The log transformation of Death Rate better fit, but there is still sufficient evidence to reject normality.

```{r}
race_df_DR <- race_df %>% 
  filter(`Death Rate (estimated)` > 0)
b_DR <-  boxcox(lm(race_df_DR$`Death Rate (estimated)` ~ 1))
lambda_DR <- b_DR$x[which.max(b_DR$y)]

race_df_ROE <- race_df %>% 
  filter(`Ratio Observed/Expected` > 0)

b_ROE <-  boxcox(lm(race_df_ROE$`Ratio Observed/Expected` ~ 1))
lambda_ROE <- b_ROE$x[which.max(b_ROE$y)]
```

We'll create a function to carry out the box cox transformation [cite](https://stats.stackexchange.com/questions/1444/how-should-i-transform-non-negative-data-including-zeros)
```{r}
box_cox <- function(x, lambda) {

    eps <- 0.00001
    if (abs(lambda) < eps)
        log(x)
    else
        (x ^ lambda - 1) / lambda

}
```

Box Cox transformation of DR and results of shapiro test.
```{r}
race_df_DR %>%
  mutate(BOX = box_cox(`Death Rate (estimated)`, lambda_DR)) %>%
  ggplot() + geom_histogram(aes(x= BOX, fill = Race), bins = 10) + facet_wrap(. ~ Race)

race_df_DR <- race_df_DR %>%
  mutate(BOX = box_cox(`Death Rate (estimated)`, lambda_DR))

race_df_DR %>%
  pull(BOX) %>% 
  shapiro.test()
##Reject normality

race_df_DR %>%
  filter(Race == "Black") %>% 
  pull(BOX) %>% 
  shapiro.test()
##Fail to reject normality

race_df_DR %>%
  filter(Race == "White") %>% 
  pull(BOX) %>% 
  shapiro.test()
##Reject normality

race_df_DR %>%
  filter(Race == "Hispanic") %>% 
  pull(BOX) %>% 
  shapiro.test()
##Fail to reject normality

race_df_DR %>%
  filter(Race == "Multiple") %>% 
  pull(BOX) %>% 
  shapiro.test()
##Fail to reject normality

par(mfrow=c(1,3))
plot(lm(race_df_DR$BOX ~ 1))
```
Successful transformation of death rate among of Black, Hispanic, and Multiple racial backgrounds persons.

Box Cox transformation of ROE and results of Shapiro test.
```{r}
race_df_ROE %>%
  mutate(BOX = box_cox(`Ratio Observed/Expected`, lambda_ROE)) %>%
  ggplot() + geom_histogram(aes(x= BOX, fill = Race), bins = 10) + facet_wrap(. ~ Race)

race_df_ROE <- race_df_ROE %>%
  mutate(BOX = box_cox(`Ratio Observed/Expected`, lambda_ROE))

race_df_ROE %>%
  pull(BOX) %>% 
  shapiro.test()

race_df_ROE %>%
  filter(Race == "Black") %>% 
  pull(BOX) %>% 
  shapiro.test()

race_df_ROE %>%
  filter(Race == "White") %>% 
  pull(BOX) %>% 
  shapiro.test()

race_df_ROE %>%
  filter(Race == "Hispanic") %>% 
  pull(BOX) %>% 
  shapiro.test()

race_df_ROE %>%
  filter(Race == "Multiple") %>% 
  pull(BOX) %>% 
  shapiro.test()

par(mfrow=c(1,3))
plot(lm(race_df_ROE$BOX ~ 1))
```

Let's save the dataframe for ROE since after the Box Cox transformation for all races was better with ROE than DR.
```{r}
saveRDS(race_df_ROE, here("data/processed_data/race_new.rds"))
```

### 2023-03-18 code/analysis_code/Race/p2_PCA_race.Rmd
more EDA and modeling

```{r message=FALSE, warning=FALSE}
library(ggcorrplot)
library(FactoMineR)
library(factoextra)
library(spatialreg)
library(MASS)
library(here)
library(tidyverse)

df <- readRDS(here("data/processed_data/race_new.rds"))
```

```{r}
pca_function <- function(race) {   
  # filter race_df by the selected race/ethnicity
  df <- df[df$Race == race,]

num_race <- df[,c(9:10, 16:25)]

#normalization
norm_race <- scale(num_race)
print(head(norm_race))

#setting up correlation matrix
corr_matrix_race <- cor(norm_race)
print(ggcorrplot(corr_matrix_race))

#applying pca
pca_race <- princomp(corr_matrix_race)
print(summary(pca_race))
#pca loadings
print(pca_race$loadings[, 1:2])

#Scree plot
print(fviz_eig(pca_race, addlabels = TRUE))
# Biplot of the attributes
print(fviz_pca_var(pca_race, col.var = "darkblue"))

#Contribution of each variable 
print(fviz_cos2(pca_race, choice = "var", axes = 1:2))

# Biplot of the attributes
print(fviz_pca_var(pca_race, col.var = "cos2",
             gradient.cols = c("darkblue", "orange", "green"),
             repel = TRUE))
# Variables in green are a good representation of the variable on that component.
}
```

```{r}
pca_function("Black")
```

```{r}
pca_function("Hispanic")
```

```{r}
pca_function("Multiple")
```

```{r}
pca_function("White")
```

### 2023-03-18 code/analysis_code/Race/p3_model_race.Rmd
more EDA and modeling

```{r message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(spdep)
```

```{r}
df <- readRDS(here("data/processed_data/race_new.rds"))
us_nabs <- readRDS(here("data/processed_data/us_nabs.rds"))

#Subset to contiguous states
us_nabs <- us_nabs %>%
  filter(!grepl("AK", StateCode),
         !grepl("HI", StateCode), 
         !grepl("DC", StateCode))

df <- df %>% 
  filter(!grepl("Alaska", State),
         !grepl("Hawaii", State)) %>% 
  mutate(log_rate = log(`Death Rate (estimated)`+ 1))
```

Setting up functions for models
```{r}
  #poisson
pois_function <- function(race) {
  pois_race <- glm(`Death Cases` ~ `Median Household Income` +
                       `Percent High School Education` +
                       offset(log(`Population`)),
                     data = df[df$Race == race,],
                     family = poisson)
  print(summary(pois_race))
}

  #Negative binomial
nb_function  <- function(race) {
  neg_bin_race <- glm.nb(`Death Cases` ~ `Median Household Income` +
                              `Percent High School Education` + 
                              offset(log(`Population`)),
                            data = df[df$Race == race,])
  print(summary(neg_bin_race))
}

  #Fitting CAR model (requires neighbors)
car_function <- function(race) {   
  # filter race_df by the selected race/ethnicity
  df <- df[df$Race == race,]

  h_race_df <- matrix(0, nrow(df), nrow(df))
  h_race_df[cbind(match(us_nabs$StateCode, df$`State Abbreviation`),
                  match(us_nabs$NeighborStateCode, df$`State Abbreviation`))] <- 1

  nablist_race_df <- mat2listw(h_race_df, style = "W")

  car_race <- spautolm(log_rate ~ `Percent Living in Poverty` + 
                             `Percent High School Education`,
                           data = df,
                           listw = nablist_race_df,
                           family = "CAR")
  print(summary(car_race))
}
```

Fitting Poisson models
```{r}
pois_function("Black")
```

```{r}
pois_function("Hispanic")
```

```{r}
pois_function("Multiple")
```

```{r}
pois_function("White")
```

Fitting Negative binomial models
```{r}
nb_function("Black")
```

```{r}
nb_function("Hispanic")
```

```{r}
nb_function("Multiple")
```

```{r}
nb_function("White")
```
The NB model accounting for overdispersion fit better than the poisson model.

Fitting CAR models
```{r}
car_function("Black")
```

```{r}
car_function("Hispanic")
```

```{r}
car_function("Multiple")
```

```{r}
car_function("White")
```
The model accounting for spatial variation provide the best fit.

Creating function that subsets by specified race, and compute/ plots Moran's I and lisa.
```{r}
spacy_function <- function(race) {
  # filter race_df by the selected race/ethnicity
  df <- df[df$Race == race,]

  h_race_df <- matrix(0, nrow(df), nrow(df))
  h_race_df[cbind(match(us_nabs$StateCode, df$`State Abbreviation`),
  match(us_nabs$NeighborStateCode, df$`State Abbreviation`))] <- 1

  nablist_race_df <- mat2listw(h_race_df, style = "W")

 # plot Moran's I
  par(mfrow=c(1,2))
  print(moran.test(df$BOX, nablist_race_df, zero.policy = TRUE, randomisation = FALSE))
  moran.plot(df$BOX, nablist_race_df, zero.policy = TRUE)

  # calculate and print LISA statistics
  lisa_race <- localmoran(df$`Ratio Observed/Expected`, nablist_race_df)
  summary(lisa_race[,1])
}
```

```{r}
spacy_function("Black")
```

```{r}
spacy_function("Hispanic")
```

```{r}
spacy_function("Multiple")
```

```{r}
spacy_function("White")
```

```{r}
save.image(file="../../data/processed_data/race_env.rds")
```

### 2023-03-18 code/analysis_code/Sex/2_sex.Rmd
more EDA and modeling

--- 2023-02-27 code/analysis_code/2_sex.Rmd
@@
-```{r}
+```{r message=FALSE, warning=FALSE}
 library(here)
@@
 h_male[cbind(match(us_nabs$StateCode, male_df$`State Abbreviation`), match(us_nabs$NeighborStateCode, male_df$`State Abbreviation`))]<-1
-nablist_male <- mat2listw(h_male)
+nablist_male <- mat2listw(h_male, style = "W")
 ```
@@
   filter(!grepl("VT", StateCode),
-         !grepl("NH", StateCode))
+         !grepl("NH", StateCode),
+         !grepl("VT", NeighborStateCode),
+         !grepl("NH", NeighborStateCode))
 
@@
 h_female[cbind(match(us_nabs_f$StateCode, female_df$`State Abbreviation`), match(us_nabs_f$NeighborStateCode, female_df$`State Abbreviation`))]<-1
-nablist_female <- mat2listw(h_female)
+nablist_female <- mat2listw(h_female, style ="W")
 

### 2023-03-18 products/manuscript/MANU.qmd
more EDA and modeling

@@
 title: "Manuscript Template for a Data Analysis Project"
-subtitle: "Exploring Factors influencing HIV-related Mortality Rates in the Georgia -- 2018 - 2020"
+subtitle: "Spatial Investigation and Exploration of Factors influencing HIV-related Mortality Rates in the United States"
 author: Hayley Hemme

### 2023-03-18 code/analysis_code/mapping.R

@@ -1,0 +1,29 @@
+library(usmap)
+
+plot_usmap(regions = "states")
+
+df<- df %>% mutate(state = tolower(State)) 
+
+
+plot_usmap(data = race_df, values = "BOX", regions = "states", include = "states") + 
+  scale_fill_continuous(low = "white", high = "blue",
+                        name = "Pc Estimates", label = scales::comma) + 
+  labs(title = "New England Region", subtitle = "Poverty Percentage Estimates for New England Counties in 2014") +
+  theme(legend.position = "right")
+
+plot_usmap(data = race_df, values = "Death Rate (estimated)",
+           regions = "states", exclude = c("HI", "AK"), color = "blue") + 
+  scale_fill_continuous(low = "white", high = "blue",
+                        name = "Death Rate (estimated)s", 
+                        label = scales::comma) + 
+  labs(title = "HIV Death Rates in the United States") +
+  theme(legend.position = "right") 
+
+plot_usmap(data = race_df, values = "Ratio Observed/Expected", 
+           regions = "states", 
+           exclude = c("HI", "AK"), color = "steelblue") +
+  scale_fill_continuous(low = "white", high = "steelblue", 
+                        name = "Death Rate (estimated)s", 
+                        label = scales::comma) + 
+  labs(title = "HIV Death Rates in the United States") +
+  theme(legend.position = "right") 

### 2023-04-08 code/analysis_code/Race/p1_EDA_race.Rmd
added more to project

@@
 autoplot(lm(race_df$`Death Rate (estimated)` ~ 1))
-autoplot(lm(race_df$`Ratio Observed/Expected` ~ 1))
+autoplot(lm(race_df$`Ratio Observed/ Expected` ~ 1))
 ```
@@
   ggplot() +
-  geom_histogram(aes(x = `Ratio Observed/Expected`, fill = Race), bins = 10)
-shapiro.test(race_df$`Ratio Observed/Expected`)
+  geom_histogram(aes(x = `Ratio Observed/ Expected`, fill = Race), bins = 10)
+shapiro.test(race_df$`Ratio Observed/ Expected`)
 ```
-Both histograms show a positively skewed distribution. ROE appears to be the slightest bit more normal.
+Both histograms show a positively skewed distribution.
 
@@
   ggplot() +
-  geom_histogram(aes(x = log(`Ratio Observed/Expected` + 1), fill = Race) , bins = 10)
-shapiro.test(log(race_df$`Ratio Observed/Expected` + 1))
+  geom_histogram(aes(x = log(`Ratio Observed/ Expected` + 1), fill = Race) , bins = 10)
+shapiro.test(log(race_df$`Ratio Observed/ Expected` + 1))
 ```
@@
 race_df_ROE <- race_df %>% 
-  filter(`Ratio Observed/Expected` > 0)
+  filter(`Ratio Observed/ Expected` > 0)
 
-b_ROE <-  boxcox(lm(race_df_ROE$`Ratio Observed/Expected` ~ 1))
+b_ROE <-  boxcox(lm(race_df_ROE$`Ratio Observed/ Expected` ~ 1))
 lambda_ROE <- b_ROE$x[which.max(b_ROE$y)]
+lambda_ROE 
 ```
@@
 race_df_ROE %>%
-  mutate(BOX = box_cox(`Ratio Observed/Expected`, lambda_ROE)) %>%
-  ggplot() + geom_histogram(aes(x= BOX, fill = Race), bins = 10) + facet_wrap(. ~ Race)
+  mutate(BOX = box_cox(`Ratio Observed/ Expected`, lambda_ROE)) %>%
+  ggplot() +
+  geom_histogram(aes(x= BOX, fill = Race), bins = 10) +
+  facet_wrap(. ~ Race) +
+  theme_classic() +
+  theme(legend.position = "none") +
+  labs(x = "Box Cox Transformation of Observed/ Expected HIV Mortality Ratio", y = "Count")
 
+ggsave(here("results/box_hist.png"), units="in", width=10, height=7, dpi=300)
+```
+
+```{r}
 race_df_ROE <- race_df_ROE %>%
-  mutate(BOX = box_cox(`Ratio Observed/Expected`, lambda_ROE))
+  mutate(BOX = box_cox(`Ratio Observed/ Expected`, lambda_ROE))
 
@@
   shapiro.test()
+## Fail to reject normality
 
@@
   shapiro.test()
+## Reject normality
 
@@
   shapiro.test() 
+## Fail to reject normality
 
@@
 plot(lm(race_df_ROE$BOX ~ 1))
+##I saved the plot above manually because I did not know how to save all three plots in one image aside from exporting.
 ```
 
-Let's save the dataframe for ROE since after the Box Cox transformation for all races was better with ROE than DR.
+The BOX transformation of ROE was better than that of DR, so let's use that one... but before wrapping up, let's go back and pull the observations that were dropped from the Box Cox transformation. 
 ```{r}
-saveRDS(race_df_ROE, here("data/processed_data/race_new.rds"))
+race_drop <- race_df %>% 
+  filter(`Ratio Observed/ Expected` == 0)
+
+race_new <- full_join(race_df_ROE, race_drop)
+
+##We need to replace the NA with a value in order to run spatial tests later. We will replace with 0. 
+race_new <-  race_new %>%
+  mutate(BOX = replace(BOX, is.na(BOX),0))
 ```
+
+```{r}
+saveRDS(race_new, here("data/processed_data/race_new.rds"))
+```

### 2023-04-08 code/analysis_code/Race/p2_PCA_race.Rmd
revisions based on feedback
added more to project

@@
 
+CREATING PCA FUNCTION
 ```{r}
@@
 
+#Selecting variables
 num_race <- df[,c(9:10, 16:25)]
@@
              gradient.cols = c("darkblue", "orange", "green"),
-             repel = TRUE))
+             repel = TRUE, title = race))
 # Variables in green are a good representation of the variable on that component.
@@
 pca_function("Black")
+ggsave(here("results/pca_black.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
 pca_function("Hispanic")
+ggsave(here("results/pca_hispanic.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
 pca_function("Multiple")
+ggsave(here("results/pca_multiple.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
 pca_function("White")
+ggsave(here("results/pca_white.png"), units="in", width=10, height=7, dpi=300)
 ```

### 2023-04-08 code/analysis_code/Race/p3_model_race.Rmd
revisions based on feedback
added more to project

@@
 library(tidyverse)
+library(MASS)
+library(spatialreg)
 library(spdep)
@@
 
-  car_race <- spautolm(log_rate ~ `Percent Living in Poverty` + 
-                             `Percent High School Education`,
+  car_race <- spautolm(`Death Cases` ~ `Median Household Income` +
+                              `Percent High School Education` + 
+                              offset(log(`Population`)),
                            data = df,
@@
 ```{r}
+sink(here("results/pois_summary_black.txt"))
+pois_function("Black")
+sink()
+```
+
+```{r}
 pois_function("Hispanic")
+```
+
+```{r}
+sink(here("results/pois_summary_hispanic.txt"))
+pois_function("Hispanic")
+sink()
 ```
@@
 ```{r}
+sink(here("results/pois_summary_multiple.txt"))
+pois_function("Multiple")
+sink()
+```
+
+```{r}
 pois_function("White")
+```
+
+```{r}
+sink(here("results/pois_summary_white.txt"))
+pois_function("White")
+sink()
 ```
@@
 ```
+```{r}
+sink(here("results/nb_summary_black.txt"))
+nb_function("Black")
+sink()
+```
 
@@
 nb_function("Hispanic")
+```
+```{r}
+sink(here("results/nb_summary_hispanic.txt"))
+nb_function("Hispanic")
+sink()
 ```
@@
 ```{r}
+sink(here("results/nb_summary_multiple.txt"))
+nb_function("Multiple")
+sink()
+```
+
+```{r}
 nb_function("White")
 ```
-The NB model accounting for overdispersion fit better than the poisson model.
+
+```{r}
+sink(here("results/nb_summary_white.txt"))
+nb_function("White")
+sink()
+```
+
+The NB model accounting for overdispersion fits better than the poisson model.
 
@@
 ```{r}
+sink(here("results/car_summary_black.txt"))
+car_function("Black")
+sink()
+```
+
+```{r}
 car_function("Hispanic")
+```
+
+```{r}
+sink(here("results/car_summary_hispanic.txt"))
+car_function("Hispanic")
+sink()
 ```
@@
 ```
+```{r}
+sink(here("results/car_summary_multiple.txt"))
+car_function("Multiple")
+sink()
+```
 
@@
 car_function("White")
+```
+```{r}
+sink(here("results/car_summary_white.txt"))
+car_function("White")
+sink()
 ```
@@
  # plot Moran's I
-  par(mfrow=c(1,2))
   print(moran.test(df$BOX, nablist_race_df, zero.policy = TRUE, randomisation = FALSE))
@@
   # calculate and print LISA statistics
-  lisa_race <- localmoran(df$`Ratio Observed/Expected`, nablist_race_df)
+  lisa_race <- localmoran(df$`Ratio Observed/ Expected`, nablist_race_df)
   summary(lisa_race[,1])
@@
 ```{r}
+png(here("results/moran_black.png"))
+sink(here("results/spacy_summary_black.txt"))
+spacy_function("Black")
+sink()
+dev.off()
+```
+
+```{r}
 spacy_function("Hispanic")
+```
+
+```{r}
+png(here("results/moran_hispanic.png"))
+sink(here("results/spacy_summary_hispanic.txt"))
+spacy_function("Hispanic")
+sink()
+dev.off()
 ```
@@
 ```{r}
+png(here("results/moran_multiple.png"))
+sink(here("results/spacy_summary_multiple.txt"))
+spacy_function("Multiple")
+sink()
+dev.off()
+```
+
+```{r}
 spacy_function("White")
@@
 ```{r}
-save.image(file="../../data/processed_data/race_env.rds")
+png(here("results/moran_white.png"))
+sink(here("results/spacy_summary_white.txt"))
+spacy_function("White")
+sink()
+dev.off()
 ```
+
+```{r}
+save.image(here("data/processed_data/race_env.rds"))
+```

### 2023-04-08 code/analysis_code/Race/p4_pulling_model_fit.Rmd
added more to project

```{r}
library(here)
library(tidyverse)
library(spdep)

df <- readRDS(here("data/processed_data/race_new.rds"))
us_nabs <- readRDS(here("data/processed_data/us_nabs.rds"))

#Subset to contiguous states
us_nabs <- us_nabs %>%
  filter(!grepl("AK", StateCode),
         !grepl("HI", StateCode), 
         !grepl("DC", StateCode))
```

The following code performs a CAR model, provides a summary, and pulls fitted values. I subset by each race. 
```{r}
black <- df %>% filter(Race == "Black")

h_black_df <- matrix(0, nrow(black), nrow(black))
h_black_df[cbind(match(us_nabs$StateCode, black$`State Abbreviation`),
                 match(us_nabs$NeighborStateCode, black$`State Abbreviation`))] <- 1

nablist_black_df <- mat2listw(h_black_df, style = "W")

car_black <- spautolm(`Death Cases` ~ `Percent Living in Poverty` + 
                        `Percent High School Education` + offset(Population),
                      listw = nablist_black_df, 
                      data = black,
                      family = "CAR")
print(summary(car_black))

##pull fitted values
black$fitted_values <- car_black$fit$fitted.values
```
We will use the

```{r}
##
multiple <- df[df$Race == "Multiple",]

h_multiple_df <- matrix(0, nrow(multiple), nrow(multiple))
h_multiple_df[cbind(match(us_nabs$StateCode, multiple$`State Abbreviation`),
                    match(us_nabs$NeighborStateCode, multiple$`State Abbreviation`))] <- 1

nablist_multiple_df <- mat2listw(h_multiple_df, style = "W")

car_multiple <- spautolm(`Death Cases` ~ `Percent Living in Poverty` + 
                        `Percent High School Education` + offset(Population),
                         data = multiple,
                         listw = nablist_multiple_df,
                         family = "CAR")
print(summary(car_multiple))

multiple$fitted_values <- car_multiple$fit$fitted.values
```

```{r}
hispanic <- df[df$Race == "Hispanic",]

h_hispanic_df <- matrix(0, nrow(hispanic), nrow(hispanic))
h_hispanic_df[cbind(match(us_nabs$StateCode, hispanic$`State Abbreviation`),
                    match(us_nabs$NeighborStateCode, hispanic$`State Abbreviation`))] <- 1

nablist_hispanic_df <- mat2listw(h_hispanic_df, style = "W")

car_hispanic <- spautolm(`Death Cases` ~ `Percent Living in Poverty` + 
                        `Percent High School Education` + offset(Population),
                         data = hispanic,
                         listw = nablist_hispanic_df,
                         family = "CAR")
print(summary(car_hispanic))

fitted_values_hispanic <- car_hispanic$fit$fitted.values

hispanic$fitted_values <- fitted_values_hispanic 
```

```{r}
white <- df[df$Race == "White",]

h_white_df <- matrix(0, nrow(white), nrow(white))
h_white_df[cbind(match(us_nabs$StateCode, white$`State Abbreviation`),
                 match(us_nabs$NeighborStateCode, white$`State Abbreviation`))] <- 1

nablist_white_df <- mat2listw(h_white_df, style = "W")

car_white <- spautolm(`Death Cases` ~ `Percent Living in Poverty` + 
                        `Percent High School Education` + offset(Population),
                      data = white,
                      listw = nablist_white_df,
                      family = "CAR")
print(summary(car_white))

fitted_values_white <- car_white$fit$fitted.values

white$fitted_values <- fitted_values_white

```

now let's combine then back into one dataframe.
```{r}
race_fit <- list(black, multiple, hispanic, white)

race_fit <-reduce(race_fit, full_join)

saveRDS(race_fit, here("data/processed_data/race_fit.rds"))
```

### 2023-04-08 code/analysis_code/Race/p5_mapping.Rmd
added more to project

In this section, I will be creating some maps of the data. I'll start by loading some package and the dataframes.

```{r}
library(here)
library(tidyverse)
library(spdep)

#loading a spatial object for the contiguous US.

states <- map_data("state")

df <- readRDS(here("data/processed_data/race_fit.rds"))
```

I do a bit more cleaning to the data to tidy things up...
```{r}
df <- df %>%
  mutate(region = tolower(State)) %>% 
  replace(is.na("Ratio Observed/ Expected"), 0) %>% 
  mutate(Exp_DR = Exp_Deaths/Population *1e5,
               OE_mortality_ratio = `Death Rate (estimated)`/Exp_DR,
         dykes_chi = ((`Death Cases`- Exp_Deaths)/ sqrt(Exp_Deaths)),
         predicted_observed_risk = fitted_values/`Death Cases`,
         predicted_observed_risk = ifelse(!is.finite(predicted_observed_risk), 0, predicted_observed_risk),
         observed_predicted_risk = `Death Cases`/fitted_values,
         observed_predicted_risk = ifelse(!is.finite(observed_predicted_risk), 0, observed_predicted_risk))
```

Next, I will merge my dataframe contain the variables of interest with the spatial object. 
```{r}
df_sf <- merge(states,df, by = "region")
# This next step is import to ensure that no "holes" appear in the map.
df_sf <- df_sf[order(df_sf$order), ]
```

Now I'm ready to start mapping! I'll begin by looking at the map of the raw HIV death rate across to contiguous US by selected race/ ethnicities.
```{r}
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Death Rate (estimated)`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "HIV Mortality Rate (raw) by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Deaths per 100,000") +
  scale_fill_distiller(palette = "RdYlGn")
```
The map for HIV-related death differs drastically across the United States by race/ethnicity. In most states, the highest HIV-related death were among Black person. This is especially true in the South region. Non-coastal western states consistently had lowest rates HIV-related deaths across the selected racial/ethnic groups, expect among multi-racial non-Hispanic persons. Non-Hispanic White persons were the only of the selected racial/ ethnic group that did not have a HIV-related death rate higher than ~10 per 100,000.

Next, let's map a probability map based on where we'd expect deaths if risk of HIV-related death was consistent across the US.
```{r}
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = Exp_DR)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race) +
  scale_fill_viridis_c() + theme_classic() + 
  theme(strip.text = element_text(size = 14),
    strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "NULL - Expected HIV Mortality Rate by Race/ Ethnicity
       in the United States, 2020", x = "", y = "", fill = "Rate")
```
Well.. I probably need to go back and reconsider using making a lambda based on specific race/ ethnic. I think that at this point, however, it is still fair to consider mapping an observed/ expected ratio using the expected deaths calculate from the lambda for the entire population. This will allow us to see differences across the race/ ethnicity, rather than masking disparities in rates due higher expected cases among certain high-risk populations, which would thus make the ratio closer to the null.

Let's map the observed/ expected ratio.
```{r}
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = OE_mortality_ratio)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race) +
  scale_fill_viridis_c() + theme_classic() + 
  theme(strip.text = element_text(size = 14),
    strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Observed/ Expected HIV Mortality Ratio by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Ratio")
```
We can see that non-Hispanic Black persons had ratios consistently (much) greater than 1, suggesting a high risk among this population.

Next, let's using Dykes and Unwin's method of smooth map using a chi square value. 
```{r}
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = dykes_chi)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + theme_classic() + 
  theme(strip.text = element_text(size = 14),
        strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Dyke's Chi Square by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Dykes and Unwin's Chi Square") +
  scale_fill_distiller(palette = "RdYlGn")
```
##Dykes and Unwin’s chi-square statistic
Dykes and Unwin (Dykes and Unwin 2001) technique  has the effect of creating a distribution of values closer to normal (as opposed to a Poisson distribution of rates and counts encountered thus far). We can therefore apply a diverging color scheme where green hues represent less than expected rates and red hues represent greater than expected rates.

https://mgimond.github.io/Spatial/mapping-rates-in-r.html
```{r}
E.Best <- EBest(df_sf$`Death Cases`,df_sf$Population)
df_sf$EBest <- E.Best$estmm * 100000
library(classInt)
brks1 <- classIntervals(df_sf$EBest, n = 10, style = "quantile")

library(RColorBrewer)
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Death Rate (estimated)`)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn")), 
                       breaks = round(brks1$brks)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom",
        legend.key.width = unit(1.5, "cm"), # Adjust the width of the legend
        legend.key.height = unit(0.5, "cm")) + # Adjust the height of the legend
  guides(fill = guide_colorbar(barwidth = 50, barheight = 1)) # Adjust the size of the colorbar
```

Next, let's make a map using the fitted values from the CAR model, the best performing model.
```{r}
##Smoothed map using fitted values.
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = fitted_values)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") 
```

Next, let's look at the map for the ratio of predicted to observed (P/O) values. 
```{r}
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = predicted_observed_risk)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") 
```

Interesting... While the AIC suggest that the CAR model performs well, the model has negative predicted values. This is problematic and suggests that it is not well-suited to the data.
Regardless, a higher P/O ratio (more red values) indicates that the model overestimated the observed values, while a lower P/O ratio (more green values) indicates that the model underestimates the observed values.

### 2023-04-08 code/analysis_code/archieved/Sex/1_EDA_sex.Rmd
added more to project

--- 2023-02-27 code/analysis_code/1_EDA_sex.Rmd
@@
 
-Visual assessment and the sharpiro test are in agreement that the data is not normally distributed. Let's see if log (x +1) transformation can help resolve this issue.
+Visual assessment and the shapiro test are in agreement that the data is not normally distributed. Let's see if log (x +1) transformation can help resolve this issue.
 
@@
 I'll try the Box-Cox transformation first.
-A limitation of this is that Box Cox cannot values of zero (or negative numbers).
+A limitation of this is that Box Cox cannot have values of zero (or negative numbers).
 We will have to subset observations greater than zero to continue with the transformation.
@@
 ```{r}
+sex_df_sub_zero <- sex_df %>% 
+  filter(`Death Rate (estimated)` <= 0)
+
 sex_df_DR <- sex_df %>% 
@@
 
-We will perform Yeo Johnson transformation on `Death Rate (estimated)` and `Standardized Mortality Ratio`.We will use `yeojohnson` from `BestNormalize` package to extract lambda and `yeo_johnson` to compute the transformation for each observation.
+We will perform Yeo Johnson transformation on `Death Rate (estimated)` and `Standardized Mortality Ratio`.
+We will use `yeojohnson` from `BestNormalize` package to extract lambda and `yeo_johnson` to compute the transformation for each observation.
 Code for `yeo_johnson` obtained from [GitHub](https://stats.stackexchange.com/questions/1444/how-should-i-transform-non-negative-data-including-zeros)
@@
 ```
+
+Let's merge back the observations we had to exclude due to limitations of the Box Cox transformation... these will be necessary to carry out spatial analyses..
+```{r}
+sex_df_ROE <- full_join(sex_df_ROE,sex_df_sub_zero)
+
+```
+
 Let's save the dataframe for ROE since this one was able to be normalized after Box Cox transformation. 

### 2023-04-08 code/analysis_code/archieved/Sex/2_sex.Rmd
added more to project

```{r message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(spdep)
```

```{r}
sex_df <- readRDS(here("data/processed_data/sex_new.rds"))
us_nabs <- readRDS(here("data/processed_data/us_nabs.rds"))

us_nabs <- us_nabs %>%
  filter(!grepl("AK", StateCode),
         !grepl("HI", StateCode), 
         !grepl("DC", StateCode))
```

```{r}
male_df <- sex_df %>%
  filter(!grepl("Alaska", State),
         !grepl("Hawaii", State),
         Sex == "Male")

#Making empty matrix
h_male <- matrix(0, nrow(male_df), nrow(male_df))

#filling matrix with '1' for neighbors
h_male[cbind(match(us_nabs$StateCode, male_df$`State Abbreviation`), 
             match(us_nabs$NeighborStateCode,
                   male_df$`State Abbreviation`))] <- 1

nablist_male <- mat2listw(h_male, style = "W")
```

```{r}
#carrying out Moran's test - global
par(mfrow=c(1,2))
moran.test(male_df$BOX, 
           nablist_male,
           zero.policy = TRUE, 
           randomisation = FALSE)
moran.plot(male_df$BOX, 
           nablist_male, 
           zero.policy = TRUE)

#carrying out local Moran's test 
lisa_male <-localmoran(male_df$`Ratio Observed/Expected`,  nablist_male)
summary(lisa_male[,1])
```

Doing the same for females...
```{r}
female_df <- sex_df %>%
  filter(!grepl("Alaska", State),
         !grepl("Hawaii", State),
         Sex == "Female") %>% 
  mutate(BOX = ifelse(is.na(BOX),0, BOX)) # replacing NA with zeros so that we can run the Moran's test.

h_female <- matrix(0, nrow(female_df), nrow(female_df))
h_female[cbind(match(us_nabs$StateCode, 
                     female_df$`State Abbreviation`), 
               match(us_nabs$NeighborStateCode, 
                     female_df$`State Abbreviation`))] <-1
nablist_female <- mat2listw(h_female, style ="W")

par(mfrow=c(1,2))
moran.test(female_df$BOX,
           nablist_female,
           zero.policy = TRUE, 
           randomisation = FALSE)

moran.plot(female_df$BOX, 
           nablist_female,
           zero.policy = TRUE)

lisa_female <-localmoran(female_df$`Ratio Observed/Expected`,  nablist_female)
summary(lisa_female[,1])
```

### 2023-04-08 code/processing_code/2a_cleaning_sex_df.Rmd
added more to project

--- 2023-02-27 code/processing_code/2_cleaning_sex_df.Rmd
@@
   mutate(Male_Rate = round(`Male Death Cases`/pop_male*1e5),
-         Female_Rate = round(`Female Death Cases`/pop_female*1e5)) %>% 
-  select(!`Year`)
+         Female_Rate = round(`Female Death Cases`/pop_female*1e5))
 ```

### 2023-04-08 code/processing_code/2b_cleaning_race_df.Rmd
added more to project

```{r, echo=FALSE, message=FALSE}
library(knitr)
library(tidyverse)
library(here)
```

```{r}
race_df <- readRDS(here("data/processed_data/race_df.rds"))
race_df <-  race_df %>%  replace(is.na("Black Death Cases"), 0) %>% 
   replace(is.na("Hispanic Death Cases"), 0) %>% 
   replace(is.na("White Death Cases"), 0) %>% 
   replace(is.na("Multiple Death Cases"), 0)

```
Let's find the HIV death rate by race
```{r}
race_df <- race_df %>% 
  mutate(black_death_rate = round(`Black Death Cases`/Black_pop*1e5),
         hispanic_death_rate = round(`Hispanic Death Cases`/Hispanic_pop*1e5),
         white_death_rate = round(`White Death Cases`/White_pop*1e5),
         multiple_death_rate = round(`Multiple Race Death Cases`/Multiple_pop*1e5))
```

Next, let's find the expected number of HIV deaths. First, we'll calculate lambda. 
```{r}
lambda_black <- race_df %>%
  drop_na(`Black Death Cases`) %>% 
  summarize(lambda_black = sum(`Black Death Cases`)/sum(Black_pop)) %>%
  as.numeric()

lambda_hispanic <- race_df %>%
  drop_na(`Hispanic Death Cases`) %>% 
  summarize(lambda_hispanic = sum(`Hispanic Death Cases`)/sum(Hispanic_pop)) %>%
  as.numeric()

lambda_white <- race_df %>%
  drop_na(`White Death Cases`) %>% 
  summarize(lambda_white = sum(`White Death Cases`)/sum(White_pop)) %>%
  as.numeric()

lambda_multiple <- race_df %>%
  drop_na(`Multiple Race Death Cases`) %>% 
  summarize(lambda_multipe = sum(`Multiple Race Death Cases`)/sum(Multiple_pop)) %>%
  as.numeric()

race_df %>%
  mutate(black_exp_deaths = round(Black_pop * lambda_black),
        white_exp_deaths = round(White_pop * lambda_white),
        hispanic_exp_deaths = round(Hispanic_pop *  lambda_hispanic),
        multiple_exp_deaths = round(Multiple_pop *  lambda_multiple))
```

NOTE: I realized after running some more analyses that calculating lambda for each race is masking disparities in HIV-related deaths between races. I believe it would more appropriate calculate a single value lambda for entire population.

```{r}
black_df <- race_df %>% select(1:3, contains("Black"), 50:62)

black_df <- black_df %>% 
  mutate(Race = "Black") %>% 
 rename("Death Rate (AIDSVu)" = `Black Death Rate`,
        "Population" = Black_pop,
        "Death Cases" = "Black Death Cases",
        "New Diagnoses Rate" = "New Diagnoses Black Rate",
        "New Diagnoses Cases" = "New Diagnoses Black Cases",
        "Percentage Diagnosed Late" = `Percent of Black Persons Diagnosed Late`,
        "Number of Persons Diagnosed Late" = `Number of Black Persons Diagnosed Late`,
        "Percentage Received Care" = `Percent of Black Persons Living with HIV who Received HIV Care`,
        "Percentage Virally Suppressed" = `Percent of Black Persons Living with HIV who were Virally Suppressed`,
        "HIV Prevalence Rate" = "Black Rate",
        "Prevalent HIV Cases" = "Black Cases",
        "Death Rate (estimated)" = black_death_rate)

hispanic_df <- race_df %>% select(1:3, contains("Hispanic"), 50:62)

hispanic_df <- hispanic_df %>% 
  mutate(Race = "Hispanic") %>% 
 rename("Death Rate (AIDSVu)" = `Hispanic Death Rate`,
        "Population" = Hispanic_pop,
        "Death Cases" = "Hispanic Death Cases",
        "New Diagnoses Rate" = "New Diagnoses Hispanic Rate",
        "New Diagnoses Cases" = "New Diagnoses Hispanic Cases",
        "Percentage Diagnosed Late" = `Percent of Hispanic Persons Diagnosed Late`,
        "Number of Persons Diagnosed Late" = `Number of Hispanic Persons Diagnosed Late`,
        "Percentage Received Care" = `Percent of Hispanic Persons Living with HIV who Received HIV Care`,
        "Percentage Virally Suppressed" = `Percent of Hispanic Persons Living with HIV who were Virally Suppressed`,
        "HIV Prevalence Rate" = "Hispanic Rate",
        "Prevalent HIV Cases" = "Hispanic Cases",
        "Death Rate (estimated)" = hispanic_death_rate)

white_df <- race_df %>% select(1:3, contains("White"), 50:62)

white_df <- white_df %>% 
  mutate(Race = "White") %>% 
 rename("Death Rate (AIDSVu)" = `White Death Rate`,
        "Population" = White_pop,
        "Death Cases" = "White Death Cases",
        "New Diagnoses Rate" = "New Diagnoses White Rate",
        "New Diagnoses Cases" = "New Diagnoses White Cases",
        "Percentage Diagnosed Late" = `Percent of White Persons Diagnosed Late`,
        "Number of Persons Diagnosed Late" = `Number of White Persons Diagnosed Late`,
        "Percentage Received Care" = `Percent of White Persons Living with HIV who Received HIV Care`,
        "Percentage Virally Suppressed" = `Percent of White Persons Living with HIV who were Virally Suppressed`,
        "HIV Prevalence Rate" = "White Rate",
        "Prevalent HIV Cases" = "White Cases",
        "Death Rate (estimated)" = white_death_rate)

multiple_df <- race_df %>% select(1:3, contains("Multiple"), 50:62)

multiple_df <- multiple_df %>% 
  mutate(Race = "Multiple") %>% 
 rename("Death Rate (AIDSVu)" = "Multiple Race Death Rate",
        "Population" = Multiple_pop,
        "Death Cases" = "Multiple Race Death Cases",
        "New Diagnoses Rate" = "New Diagnoses Multiple Race Rate",
        "New Diagnoses Cases" = "New Diagnoses Multiple Race Cases",
        "Percentage Received Care" = `Percent of Multiple Race Persons Living with HIV who Received HIV Care`,
        "Percentage Virally Suppressed" = `Percent of Multiple Race Persons Living with HIV who were Virally Suppressed`,
        "HIV Prevalence Rate" = "Multiple Races Rate",
        "Prevalent HIV Cases" = "Multiple Races Cases",
        "Death Rate (estimated)" = multiple_death_rate)

race_df <- list(black_df, hispanic_df, white_df, multiple_df)

race_df <- reduce(race_df, full_join)
```

Now, let's calculate lambda.
```{r}
lambda <- race_df %>%
  drop_na("Death Cases") %>% 
  summarize(lambda = sum(`Death Cases`)/sum(Population)) %>%
  as.numeric()

race_df <- race_df %>% 
  mutate(Exp_Deaths = Population * lambda,
         `Ratio Observed/ Expected` = `Death Cases`/Exp_Deaths)
```

Save the file and now we can move on to some exploratory analysis!
```{r}
saveRDS(race_df, here("data/processed_data/race_df_cleaned.rds"))
```

### 2023-04-08 code/processing_code/3_cleaning_us_nabs.Rmd
added more to project

--- 2023-02-27 code/processing_code/4_cleaning_us_nabs.Rmd
@@
 After some 'investigation' (troubleshooting code), this file is not exactly what I need to create the neighbor list.
-What is achieved by the following code took me a few seconds to do in Excel with copy and paste.
-But for the sake of reproducibility, I decided it's well-worth it some spend a little bit of time do this in R.
+What is achieved by the following code took me a few seconds to do in Excel with copy and paste...
+but for the sake of reproducibility, I decided it's well-worth it to some spend a little bit of time do this in R.
+I'm essentially replicating the dataframe, but crossing over the values on the replicated rows such that both variables contain the same information.
 

### 2023-04-08 products/manuscript/MANU.qmd
added more to project

@@
 ---
-
-The structure below is one possible setup for a data analysis project (including the course project).
-For a manuscript, adjust as needed.
-You don't need to have exactly these sections, but the content covering those sections should be addressed.
-
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
 
@@
 
-Question: Can the HIV death rate from 2020 in the United States be modeled by social determinants of health like poverty and/ or by HIV related outcomes (new diagnoses, prevalence PrEP use, etc?) Explain in separate analysis: Race/ ethnicity (Black, Hispanic, White, Multiple) and Sex (Male, Female).
+Question: Can the HIV death rate from 2020 in the United States be modeled by social determinants of health like poverty and/ or by HIV related outcomes (new diagnoses, prevalence PrEP use, etc?) Are these factors correlated with disparities in HIV-related mortality across different racial and ethnic groups?
 Can a spatial model explain variation in HIV death-rates at the state-level and/or regional-level after accounting for variables associated with sociodemographic, SDOH, and HIV-related outcomes?
@@
 H0 : There is not a spatial correlation HIV-related deaths and SDOH or HIV-related outcomes.
+
 Ha : There is a spatial correlation between HIV-related deaths and SDOH or HIV-related outcomes.
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a]
 
@@
 
-*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*
+I used a number of different methods throughout the various stages of this analysis.
+To ensure that data importation was simple and reproducible, I used `library(here)`.
+The vast majority of data cleaning processes were achieved using functions within the `tidyverse`.
+Some of these processes included renaming, pivoting, selecting, and mutating variables.
+Additionally, most of the data visualization and maps were obtained using ggplot2, with a few exceptions being packages with functions that are not compatible with the `tidyverse` and simple model diagnostics that base R.
+During the data visualization process, it became apparent that the dependent variable would require transformation.
+Data transformation methods of increasing complexity were performed.
+Ultimately, a Box Cox transformation was able normalize out dependent variable.
+A small value was added to each observation to ensure that all were greater than 0.
+To help assess relationships between variables, I performed a correlation analysis, most using principle component analysis (PCA).
+Data analysis methods were performed in order of increasing complexity, beginning with simple linear regression, then Poisson regression, negative-binomial regression, and finally a conditional autoregessive model (CAR) to account for spatial dependence.
+Models were judged for goodness of fit using AIC.
 
@@
 
-*Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along.* Loading the datasets from
-
-Clear the environment and start fresh.
+All data used in this analysis were obtained from publicly available sources, as note above.
 
@@
 
-Explore central tendency across demographic groups. Fit a poisson regression model predicting death-to-case ratio from percentage with x SDOH variables and percentage of x demographic group.
+Spatial dependence assessed using Moran's I.
 

### 2023-04-08 readme.qmd
added more to project

---
title: "Read Me"
format: html
editor: visual
---

This respository contains data and code to reproduce my research. Raw data is store in the `data` subfolder `raw_data`, while processed data produced by wrangling is stored under `processed_data`.

Steps to reproduce this project are:

1.  Open the R project `dataanalysis-template`
2.  Open the folder code and start by running files in `processing_code`. The files are numbered sequentially, but `1_cleaning_2020` is the only file in this folder that must be run first. The others may be run in any order.
3.  Open the folder code and start by running files in `analysis_code` , which contains sub-folders for running the analysis for race/ethnicity `race` and `sex` separately. `race` and `sex` may be run independently of each other, but contain numbers to indicate the order which these files should be run.
4.  Data files produced by these files are store in the `data` subfolder `processed_data`

### 2023-04-08 code/analysis_code/mapping.R

@@ -1,29 +1,0 @@
-library(usmap)
-
-plot_usmap(regions = "states")
-
-df<- df %>% mutate(state = tolower(State)) 
-
-
-plot_usmap(data = race_df, values = "BOX", regions = "states", include = "states") + 
-  scale_fill_continuous(low = "white", high = "blue",
-                        name = "Pc Estimates", label = scales::comma) + 
-  labs(title = "New England Region", subtitle = "Poverty Percentage Estimates for New England Counties in 2014") +
-  theme(legend.position = "right")
-
-plot_usmap(data = race_df, values = "Death Rate (estimated)",
-           regions = "states", exclude = c("HI", "AK"), color = "blue") + 
-  scale_fill_continuous(low = "white", high = "blue",
-                        name = "Death Rate (estimated)s", 
-                        label = scales::comma) + 
-  labs(title = "HIV Death Rates in the United States") +
-  theme(legend.position = "right") 
-
-plot_usmap(data = race_df, values = "Ratio Observed/Expected", 
-           regions = "states", 
-           exclude = c("HI", "AK"), color = "steelblue") +
-  scale_fill_continuous(low = "white", high = "steelblue", 
-                        name = "Death Rate (estimated)s", 
-                        label = scales::comma) + 
-  labs(title = "HIV Death Rates in the United States") +
-  theme(legend.position = "right") 

### 2023-04-08 code/analysis_code/Race/p1_EDA_race.Rmd
fixed rendering

@@
 shapiro.test(race_df$`Ratio Observed/ Expected`)
+ggsave(here("results/raw_roe_hist.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
 
+ggsave(here("results/box_hist_facet.png"), units="in", width=10, height=7, dpi=300)
+
+race_df_ROE %>%
+  mutate(BOX = box_cox(`Ratio Observed/ Expected`, lambda_ROE)) %>%
+  ggplot() +
+  geom_histogram(aes(x= BOX, fill = Race), bins = 10) +
+  theme_classic() +
+  theme(legend.position = "none") +
+  labs(x = "Box Cox Transformation of Observed/ Expected HIV Mortality Ratio", y = "Count")
+
 ggsave(here("results/box_hist.png"), units="in", width=10, height=7, dpi=300)

### 2023-04-08 code/analysis_code/Race/p4_pulling_model_fit.Rmd
fixed rendering

@@
 library(spdep)
+library(spatialreg)
 
@@
 
-car_black <- spautolm(`Death Cases` ~ `Percent Living in Poverty` + 
-                        `Percent High School Education` + offset(Population),
+car_black <- spautolm(BOX ~ `Percent Living in Poverty` + 
+                        `Percent High School Education`,
                       listw = nablist_black_df, 
@@
 
-car_multiple <- spautolm(`Death Cases` ~ `Percent Living in Poverty` + 
-                        `Percent High School Education` + offset(Population),
+car_multiple <- spautolm(BOX ~ `Percent Living in Poverty` + 
+                        `Percent High School Education`,
                          data = multiple,
@@
 
-car_hispanic <- spautolm(`Death Cases` ~ `Percent Living in Poverty` + 
-                        `Percent High School Education` + offset(Population),
+car_hispanic <- spautolm(BOX ~ `Percent Living in Poverty` + 
+                        `Percent High School Education`,
                          data = hispanic,
@@
 
-car_white <- spautolm(`Death Cases` ~ `Percent Living in Poverty` + 
-                        `Percent High School Education` + offset(Population),
+car_white <- spautolm(BOX ~ `Percent Living in Poverty` + 
+                        `Percent High School Education`,
                       data = white,

### 2023-04-08 code/analysis_code/Race/p5_mapping.Rmd
fixed rendering

@@
 
-Now I'm ready to start mapping! I'll begin by looking at the map of the raw HIV death rate across to contiguous US by selected race/ ethnicities.
+Now I'm ready to start mapping! I'll begin by looking at the map of the raw HIV death rate across to contiguous US by selected race/ ethnicity.
 ```{r}
@@
         legend.position = "bottom") +
-  labs(title = "HIV Mortality Rate (raw) by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Deaths per 100,000") +
-  scale_fill_distiller(palette = "RdYlGn")
+  labs(title = "HIV Mortality Rate (raw) by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Death rate per 100,000") +
+  scale_fill_distiller(palette = "RdYlGn") +
+  theme(plot.title= element_text(hjust = 0.5))
+
+ggsave(here("results/raw_rate_map.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
 
+Let's map the observed/ expected ratio.
+```{r}
+ggplot(df_sf, aes(long, lat)) +
+  geom_polygon(aes(group = group, fill = OE_mortality_ratio)) +
+  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
+  facet_wrap(. ~ Race) +
+  scale_fill_distiller(palette = "RdYlGn") + theme_void() + 
+  theme(strip.text = element_text(size = 14),
+    strip.background = element_blank(),
+        axis.ticks = element_blank(),
+        axis.line = element_blank(),
+        axis.text = element_blank(),
+        legend.position = "bottom") +
+  labs(title = "Observed/ Expected HIV Mortality Ratio by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Ratio") +
+  theme(plot.title= element_text(hjust = 0.5))
+
+ggsave(here("results/roe_map.png"), units="in", width=10, height=7, dpi=300)
+```
+We can see that non-Hispanic Black persons had ratios consistently (much) greater than 1, suggesting a high risk among this population.
+
+Next, let's map after the the Box_Cox transformation of the ratio of observed/ expected cases
+```{r}
+ggplot(df_sf, aes(long, lat)) +
+  geom_polygon(aes(group = group, fill = BOX)) +
+  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
+  facet_wrap(. ~ Race) +
+  scale_fill_distiller(palette = "RdYlGn") + theme_void() + 
+  theme(strip.text = element_text(size = 14),
+    strip.background = element_blank(),
+        axis.ticks = element_blank(),
+        axis.line = element_blank(),
+        axis.text = element_blank(),
+        legend.position = "bottom") +
+  labs(title = "Ratio of Observed/ Expected Cases (Box Cox transformed) by Race/ Ethnicity
+in the United States, 2020", x = "", y = "", fill = "Ratio") +
+  theme(plot.title= element_text(hjust = 0.5))
+ggsave(here("results/roe_box_map.png"), units="in", width=10, height=7, dpi=300)
+```
+
 Next, let's map a probability map based on where we'd expect deaths if risk of HIV-related death was consistent across the US.
@@
   facet_wrap(. ~ Race) +
-  scale_fill_viridis_c() + theme_classic() + 
+  scale_fill_viridis_c() + theme_void() + 
   theme(strip.text = element_text(size = 14),
@@
   labs(title = "NULL - Expected HIV Mortality Rate by Race/ Ethnicity
-       in the United States, 2020", x = "", y = "", fill = "Rate")
+       in the United States, 2020", x = "", y = "", fill = "Rate") +
+  theme(plot.title= element_text(hjust = 0.5))
 ```
@@
 
-Let's map the observed/ expected ratio.
+Next, let's using Dykes and Unwin's method of smooth map using a chi square value. 
 ```{r}
 ggplot(df_sf, aes(long, lat)) +
-  geom_polygon(aes(group = group, fill = OE_mortality_ratio)) +
+  geom_polygon(aes(group = group, fill = dykes_chi)) +
   coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
-  facet_wrap(. ~ Race) +
-  scale_fill_viridis_c() + theme_classic() + 
+  facet_wrap(. ~ Race)  + theme_void() + 
   theme(strip.text = element_text(size = 14),
@@
         legend.position = "bottom") +
-  labs(title = "Observed/ Expected HIV Mortality Ratio by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Ratio")
-```
-We can see that non-Hispanic Black persons had ratios consistently (much) greater than 1, suggesting a high risk among this population.
+  labs(title = "Dykes and Unwin's Chi Square by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Chi Square") +
+  scale_fill_distiller(palette = "RdYlGn") +
+  theme(plot.title= element_text(hjust = 0.5))
 
-Next, let's using Dykes and Unwin's method of smooth map using a chi square value. 
-```{r}
-ggplot(df_sf, aes(long, lat)) +
-  geom_polygon(aes(group = group, fill = dykes_chi)) +
-  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
-  facet_wrap(. ~ Race)  + theme_classic() + 
-  theme(strip.text = element_text(size = 14),
-        strip.background = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        axis.text = element_blank(),
-        legend.position = "bottom") +
-  labs(title = "Dyke's Chi Square by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Dykes and Unwin's Chi Square") +
-  scale_fill_distiller(palette = "RdYlGn")
+ggsave(here("results/chi_map.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
 We can therefore apply a diverging color scheme where green hues represent less than expected rates and red hues represent greater than expected rates.
-
-https://mgimond.github.io/Spatial/mapping-rates-in-r.html
-```{r}
-E.Best <- EBest(df_sf$`Death Cases`,df_sf$Population)
-df_sf$EBest <- E.Best$estmm * 100000
-library(classInt)
-brks1 <- classIntervals(df_sf$EBest, n = 10, style = "quantile")
-
-library(RColorBrewer)
-ggplot(df_sf, aes(long, lat)) +
-  geom_polygon(aes(group = group, fill =`Death Rate (estimated)`)) +
-  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn")), 
-                       breaks = round(brks1$brks)) +
-  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
-  facet_wrap(. ~ Race)  + 
-  theme_void() + 
-  theme(strip.text = element_text(size = 14),
-        legend.position = "bottom",
-        legend.key.width = unit(1.5, "cm"), # Adjust the width of the legend
-        legend.key.height = unit(0.5, "cm")) + # Adjust the height of the legend
-  guides(fill = guide_colorbar(barwidth = 50, barheight = 1)) # Adjust the size of the colorbar
-```
 
@@
   theme(strip.text = element_text(size = 14),
-        legend.position = "bottom") 
+        legend.position = "bottom") +
+  labs(title = "Smoothed Map of Square by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Ratio") +
+  theme(plot.title= element_text(hjust = 0.5))
+
+ggsave(here("results/fitted_map.png"), units="in", width=10, height=7, dpi=300)
+
 ```
@@
   theme(strip.text = element_text(size = 14),
-        legend.position = "bottom") 
+        legend.position = "bottom") +
+  labs(title = "Map of Ratio of Predicted to Observed Cases by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Ratio") +
+  theme(plot.title= element_text(hjust = 0.5))
 ``` 
-
-Interesting... While the AIC suggest that the CAR model performs well, the model has negative predicted values. This is problematic and suggests that it is not well-suited to the data.
-Regardless, a higher P/O ratio (more red values) indicates that the model overestimated the observed values, while a lower P/O ratio (more green values) indicates that the model underestimates the observed values.
+We can see that the model fitted pretty well.
+A higher P/O ratio (more red values) indicates that the model overestimated the observed values, while a lower P/O ratio (more green values) indicates that the model underestimates the observed values.
+We can see that the model underestimated rates primarily in the non-coast north-western states and north central states for persons of Hispanic, non-Hispanic White, and non-Hispanic multiple racial/ ethnic backgrounds.

### 2023-04-08 products/manuscript/MANU.qmd
fixed rendering

@@
 ---
-title: "Manuscript Template for a Data Analysis Project"
-subtitle: "Spatial Investigation and Exploration of Factors influencing HIV-related Mortality Rates in the United States"
+title: "Spatial Investigation and Exploration of Factors influencing HIV-related Mortality"
+subtitle: "United States, 2020"
 author: Hayley Hemme
@@
     highlight-style: github
-bibliography: ../dataanalysis_template_references.bib
+bibliography: ../dataanalysis_references.bib
 csl: ../apa.csl
@@
 
-## Data aquisition
+## Data acquisition
 
@@
 
-*Use a combination of text/tables/figures to explore and describe your data.
+```{r}
+
+```
+
+![Histogram of the distribution of the ratio for observed/ expected cases(raw)](images/raw_roe_hist.png){width="300"}
+
+![Histogram of the distribution of ratio of observed/ expected mortality after Box-Cox transformation](images/box_hist-01.png){width="300"}
+
+![Diagnostics for Box Cox Transformed Ratio Observed/ Expected](images/box_diag.png){width="300"}*Use a combination of text/tables/figures to explore and describe your data.
 Show the most important descriptive results here.
@@
 ## Basic statistical analysis
+
+|                                                     |                                                     |
+|-----------------------|------------------------|
+| ![Black](images/moran_black-01.png){width="200"}    | ![Multiple](images/moran_multiple.png){width="200"} |
+| ![Hispanic](images/moran_hispanic.png){width="200"} | ![White](images/moran_white.png){width="200"}       |
 
@@
 Then you load them here.*
+
+|                                              |                                         |
+|-----------------------------|------------------------------|
+| ![](images/raw_rate_map-02.png){width="250"} | ![](images/roe_map-03.png){width="250"} |
+| ![](images/roe_box_map-01.png){width="250"}  | ![](images/chi_map-02.png){width="250"} |
+| ![](images/fitted_map-02.png){width="250"}   |                                         |
 

### 2023-04-21 code/analysis_code/Race/p1_EDA_race.Rmd
more stuff

--- 2023-03-18 code/analysis_code/Race/p1_EDA_race.Rmd
@@
 library(tidyverse)
+library(broom)
+library(knitr)
+library(ggpubr)
 ```
@@
 
+Let's start by looking at the overall HIV-related death rate by race in the United States.
 ```{r}
-autoplot(lm(race_df$`Death Rate (AIDSVu)` ~ 1))
-autoplot(lm(race_df$`Death Rate (estimated)` ~ 1))
-autoplot(lm(race_df$`Ratio Observed/Expected` ~ 1))
+race_df %>% 
+  group_by(Race) %>% 
+  summarize(tot_deaths = sum(`Death Cases`),
+            tot_population = sum(`Population`),
+            overall_rate = round(tot_deaths/ tot_population *1e5,1)) %>% 
+  ungroup() %>% 
+  mutate(percent_by_race = round(tot_deaths/sum(tot_deaths)*100, 1),
+         ratio = round(overall_rate/overall_rate[Race =="White"],1)) %>%
+  kable(col.names = c("Race/ Ethnicity",
+                      "HIV-Related Deaths (n)",
+                      "Population (>12 years old)",
+                      "Death Rate (per 100,000)",
+                      "Percent",
+                      "Ratio"))
+# using white persons as reference group
+```
+
+We can seen that Black persons accounts for the 43.5% of HIV-related deaths in United States, and the rate of HIV-related deaths among this population was 7.2 times high than the rate of HIV-related deaths among White persons.
+
+Next, let's take a look at our response variable
+Univariate Analysis
+
+Let's take a look at some box plots of death rate by races and state. 
+We'll start by looking at death rate given by the AIDSVu dataset. 
+```{r}
+race_df %>% 
+  ggplot() +
+  geom_boxplot(aes(y = `Death Rate (AIDSVu)`, x = Race)) + 
+  labs(title = "HIV-Related Death Rate (AIDSVu) by Race/ Ethnicity, United States, 2020", y = "Rate (per 100,000)", x = "Race/ Ethnicity") + theme(legend.position = "none")
+```
+It seems like something might be strange about the death rates for persons of multiple races. Let's compare to the death rate that we estimated.
+
+```{r}
+race_df %>% 
+  ggplot() +
+  geom_boxplot(aes(y = `Death Rate (estimated)`, x = Race)) + 
+  labs(title = "HIV-Related Death Rate (calculated) by Race/ Ethnicity, United States, 2020", y = "Rate (per 100,000)", x = "Race/ Ethnicity") + theme(legend.position = "none")
+```
+The plots differ drastically for persons of multiple race. It is difficult to tell what might have caused this drastic change, but my inituition tells me there is something strange going on.
+
+Unfortunately, for the sake of this analysis I believe it might be best to remove persons of multiple race/ ethnicities. 
+```{r}
+race_df <- race_df %>% 
+  filter(Race != "Multiple")
 ```
@@
   ggplot() +
-  geom_histogram(aes(x = `Death Rate (estimated)`, fill = Race), bins = 10)
-shapiro.test(race_df$`Death Rate (estimated)`)
+  geom_boxplot(aes(y = `Death Rate (AIDSVu)`, x = Race)) + 
+  labs(title = "HIV-Related Death Rate (AIDSVu) by Race/ Ethnicity, United States, 2020", y = "Rate (per 100,000)", x = "Race/ Ethnicity") + theme(legend.position = "none")
 
@@
   ggplot() +
-  geom_histogram(aes(x = `Ratio Observed/Expected`, fill = Race), bins = 10)
-shapiro.test(race_df$`Ratio Observed/Expected`)
+  geom_boxplot(aes(y = `Death Rate (estimated)`, x = Race)) + 
+  labs(title = "HIV-Related Death Rate (calculated) by Race/ Ethnicity, United States, 2020", y = "Rate (per 100,000)", x = "Race/ Ethnicity") +
+  theme(legend.position = "none")
+
+ggsave(here("results/dr_boxplot.png"), units="in", width=10, height=7, dpi=300)
 ```
-Both histograms show a positively skewed distribution. ROE appears to be the slightest bit more normal.
+These plots are quite similar, so I feel good about using the counts and population data in our later analyses.
 
+Let's take a look at the distribution of HIV-related death rate. For the sake of comparison, we will use the estimated HIV death rate as our response variable. 
 ```{r}
@@
   ggplot() +
-  geom_histogram(aes(x = log(`Death Rate (estimated)` + 1), fill = Race) , bins = 10)
-shapiro.test(log(race_df$`Death Rate (estimated)` + 1))
-
-race_df %>% 
-  ggplot() +
-  geom_histogram(aes(x = log(`Ratio Observed/Expected` + 1), fill = Race) , bins = 10)
-shapiro.test(log(race_df$`Ratio Observed/Expected` + 1))
+  geom_histogram(aes(x = `Death Rate (estimated)`, group = Race), bins = 10) +
+  facet_grid(. ~ Race) + 
+  theme(legend.position = "none") +
+  labs(title = "Histogram of Raw Death Rate (estimated)", y = "n", x = "Rate (per 100,000)")
+ggsave(here("results/dr_hist.png"), units="in", width=10, height=7, dpi=300)
 ```
-The log transformation of Death Rate better fit, but there is still sufficient evidence to reject normality.
+The distribution for Hispanic and White persons is definitely not normally distributed. The distribution for Black persons seems to be roughly normal. Let's subset to test.
 
 ```{r}
-race_df_DR <- race_df %>% 
-  filter(`Death Rate (estimated)` > 0)
-b_DR <-  boxcox(lm(race_df_DR$`Death Rate (estimated)` ~ 1))
-lambda_DR <- b_DR$x[which.max(b_DR$y)]
+black_df <- race_df %>% 
+  filter(Race == "Black")
 
-race_df_ROE <- race_df %>% 
-  filter(`Ratio Observed/Expected` > 0)
+hispanic_df <- race_df %>% 
+  filter(Race == "Hispanic")
 
-b_ROE <-  boxcox(lm(race_df_ROE$`Ratio Observed/Expected` ~ 1))
-lambda_ROE <- b_ROE$x[which.max(b_ROE$y)]
+white_df <- race_df %>% 
+  filter(Race == "White")
+```
+
+```{r}
+shapiro.test(black_df$`Death Rate (estimated)`)
+```
+We fail to reject normality. No transformation of death rate for Black persons is needed.
+
+```{r}
+shapiro.test(hispanic_df$`Death Rate (estimated)`)
+# reject normality,
+
+#let's try some transformations
+shapiro.test(log(hispanic_df$`Death Rate (estimated)` + 1))
+
+hispanic_df  <- hispanic_df %>% 
+  mutate(`Death Rate (log transformed)` = log(`Death Rate (estimated)` + 1))
+```
+Log transformation was able to normalize the death rate distribution for Hispanic persons.
+
+Moving on to the transformation for the death rate of White persons.
+```{r}
+shapiro.test(log(white_df$`Death Rate (estimated)` + 1))
+```
+Log transformation was not sufficient. We will have to try a more powerful transformation method.
+
+Let's try a Box Cox transformation.
+Box Cox requires variables to be greater than 0, so we'll have to subset our dataframe to proceed with this analysis.
+We will have to add these observations back later.
+```{r}
+b_DR <-  boxcox(lm(white_df$`Death Rate (estimated)` ~ 1))
+
+lambda_DR <- b_DR$x[which.max(b_DR$y)] #pulling lambda
 ```
@@
 ```{r}
-race_df_DR %>%
-  mutate(BOX = box_cox(`Death Rate (estimated)`, lambda_DR)) %>%
-  ggplot() + geom_histogram(aes(x= BOX, fill = Race), bins = 10) + facet_wrap(. ~ Race)
+white_df <- white_df %>%
+  mutate(`Death Rate (BOX transformed)` = box_cox(`Death Rate (estimated)`, lambda_DR))
 
-race_df_DR <- race_df_DR %>%
-  mutate(BOX = box_cox(`Death Rate (estimated)`, lambda_DR))
+white_df %>% 
+  pull(`Death Rate (BOX transformed)`) %>% 
+  shapiro.test()
+```
+There is still sufficient efficient to reject normality. Let's see if a more powerful transformation works...
 
-race_df_DR %>%
-  pull(BOX) %>% 
-  shapiro.test()
-##Reject normality
+We'll try a Yeo-Johnson transformation.
+```{r}
+yeo_johnson <- function(x, lambda) {
 
-race_df_DR %>%
-  filter(Race == "Black") %>% 
-  pull(BOX) %>% 
-  shapiro.test()
-##Fail to reject normality
+    eps <- .000001
+    not_neg <- which(x >= 0)
+    is_neg  <- which(x < 0)
 
-race_df_DR %>%
-  filter(Race == "White") %>% 
-  pull(BOX) %>% 
-  shapiro.test()
-##Reject normality
+    not_neg_trans <- function(x, lambda) {
+        if (abs(lambda) < eps) log(x + 1)
+        else ((x + 1) ^ lambda - 1) / lambda
+    }
 
-race_df_DR %>%
-  filter(Race == "Hispanic") %>% 
-  pull(BOX) %>% 
-  shapiro.test()
-##Fail to reject normality
+    neg_trans <- function(x, lambda) {
+        if (abs(lambda - 2) < eps) - log(-x + 1)
+        else - ((-x + 1) ^ (2 - lambda) - 1) / (2 - lambda)
+    }
 
-race_df_DR %>%
-  filter(Race == "Multiple") %>% 
-  pull(BOX) %>% 
-  shapiro.test()
-##Fail to reject normality
+    x[not_neg] <- not_neg_trans(x[not_neg], lambda)
 
-par(mfrow=c(1,3))
-plot(lm(race_df_DR$BOX ~ 1))
-```
-Successful transformation of death rate among of Black, Hispanic, and Multiple racial backgrounds persons.
+    x[is_neg] <- neg_trans(x[is_neg], lambda)
 
-Box Cox transformation of ROE and results of Shapiro test.
-```{r}
-race_df_ROE %>%
-  mutate(BOX = box_cox(`Ratio Observed/Expected`, lambda_ROE)) %>%
-  ggplot() + geom_histogram(aes(x= BOX, fill = Race), bins = 10) + facet_wrap(. ~ Race)
+    return(x)
 
-race_df_ROE <- race_df_ROE %>%
-  mutate(BOX = box_cox(`Ratio Observed/Expected`, lambda_ROE))
+}
 
-race_df_ROE %>%
-  pull(BOX) %>% 
-  shapiro.test()
-
-race_df_ROE %>%
-  filter(Race == "Black") %>% 
-  pull(BOX) %>% 
-  shapiro.test()
-
-race_df_ROE %>%
-  filter(Race == "White") %>% 
-  pull(BOX) %>% 
-  shapiro.test()
-
-race_df_ROE %>%
-  filter(Race == "Hispanic") %>% 
-  pull(BOX) %>% 
-  shapiro.test()
-
-race_df_ROE %>%
-  filter(Race == "Multiple") %>% 
-  pull(BOX) %>% 
-  shapiro.test()
-
-par(mfrow=c(1,3))
-plot(lm(race_df_ROE$BOX ~ 1))
 ```
 
-Let's save the dataframe for ROE since after the Box Cox transformation for all races was better with ROE than DR.
 ```{r}
-saveRDS(race_df_ROE, here("data/processed_data/race_new.rds"))
+white_df <- white_df %>%
+  mutate(`Death Rate (YJ transformed)` = yeo_johnson(`Death Rate (estimated)`, lambda_DR))
+
+white_df %>% 
+  pull(`Death Rate (YJ transformed)`) %>% 
+  shapiro.test()
 ```
+Still unable to normalize. We will have to utilize methods that are not sensitive to non-normal distributions for White persons.
+
+```{r}
+df <- list(black_df, hispanic_df, white_df)
+
+df <- reduce(df, full_join)
+
+df <- df %>% 
+  mutate(outcome_DR = case_when(
+    Race == "Black" ~ `Death Rate (estimated)`,
+    Race == "White" ~ `Death Rate (estimated)`,
+    Race == "Hispanic" ~ `Death Rate (log transformed)`
+  ))
+```
+
+Let's move on to
+Bivariate Analysis
+```{r}
+ggplot(df, aes(y = outcome_DR, x = `Percent Living in Poverty`)) +
+  geom_smooth(method = "lm") + 
+  geom_point() + 
+  labs(title = "HIV Death Rate vs. Percent Living in Poverty") + 
+  stat_cor(method = "pearson") +
+  facet_wrap(Race ~ . , scales = "free_y")
+ggsave(here("results/bi_pov.png"), units="in", width=10, height=7, dpi=300)
+#did not "pass"
+
+ggplot(df, aes(y = outcome_DR, x = `Percent Unemployed`)) +
+  geom_smooth(method = "lm") +
+  geom_point() + 
+  labs(title = "HIV Death Rate vs. Percent Unemployed") + 
+  stat_cor(method = "pearson") +
+  facet_wrap(Race ~ . , scales = "free_y")
+ggsave(here("results/bi_unemployed.png"), units="in", width=10, height=7, dpi=300)
+#"passed"
+
+ggplot(df, aes(y = outcome_DR, x = `Gini Coefficient`)) +
+  geom_smooth(method = "lm") +
+  geom_point() + 
+  labs(title = "HIV Death Rate vs. Gini Coefficient") +
+  facet_wrap(Race ~ . , scales = "free_y") + 
+  stat_cor(method="pearson") 
+ggsave(here("results/bi_gini.png"), units="in", width=10, height=7, dpi=300)
+#"passed"
+
+ggplot(df, aes(y = outcome_DR, x = `Percent High School Education`)) +
+  geom_smooth(method = "lm") + 
+  geom_point() + 
+  labs(title = "HIV Death Rate vs. Percent High School Education") +
+  facet_wrap(Race ~ . , scales = "free_y") +
+  stat_cor(method="pearson")
+ggsave(here("results/bi_ed.png"), units="in", width=10, height=7, dpi=300)
+#"passed"
+
+ggplot(df, aes(y = outcome_DR, x = `Percent Living with Severe Housing Cost Burden`)) +
+  geom_smooth(method = "lm") +
+  geom_point() + 
+  labs(title = "HIV Death Rate  vs. Percent Living with Severe Housing Cost Burden") +
+  facet_wrap(Race ~ . , scales = "free_y")+ 
+  stat_cor(method="pearson")
+#"passed"
+ggsave(here("results/bi_housing.png"), units="in", width=10, height=7, dpi=300)
+
+ggplot(df, aes(y = outcome_DR, x = `Broadband of any type`)) +
+  geom_smooth(method = "lm") +
+  geom_point() + 
+  labs(title = "HIV Death Rate vs. Broadband of any type") +
+  facet_wrap(Race ~ . , scales = "free_y") + 
+  stat_cor(method="pearson")
+# did not "pass"
+```
+
+```{r}
+ggplot(df, aes(y = outcome_DR, x = `Percent Uninsured Under 65 Years Old`)) +
+  geom_smooth(method = "lm") +
+  geom_point() + 
+  labs(title = "HIV Death Rate vs. Percent Uninsured Under 65 Years Old") +
+  facet_wrap(Race ~ . , scales = "free_y") + 
+  stat_cor(method="pearson")
+# did not "pass"
+
+ggplot(df, aes(y = outcome_DR, x = `Percent Living with Food Insecurity`)) +
+  geom_smooth(method = "lm") +
+  geom_point() + 
+  labs(title = "HIV Death Rate vs. Percent Living with Food Insecurity") +
+  facet_wrap(Race ~ . , scales = "free_y") + 
+  stat_cor(method="pearson")
+# did not "pass"
+
+ggplot(df, aes(y = outcome_DR, x = `Broadband of any type`)) +
+  geom_smooth(method = "lm") +
+  geom_point() + 
+  labs(title = "HIV Death Rate vs. Broadband of any type") +
+  facet_wrap(Race ~ . , scales = "free_y") + 
+  stat_cor(method="pearson")
+# did not "pass"
+
+ggplot(df, aes(y = outcome_DR, x = `Median Household Income`)) +
+  geom_smooth(method = "lm") +
+  geom_point() + 
+  labs(title = "HIV Death Rate vs. Median Household Income") +
+  facet_wrap(Race ~ . , scales = "free_y") + 
+  stat_cor(method="pearson")
+# did not "pass"
+
+ggplot(df, aes(y = outcome_DR, x = `Percentage Diagnosed Late`)) +
+  geom_smooth(method = "lm") +
+  geom_point() + 
+  labs(title = "HIV Death Rate vs. Percentage Diagnosed Late") +
+  facet_wrap(Race ~ . , scales = "free_y") + 
+  stat_cor(method="pearson")
+
+ggplot(df, aes(y = outcome_DR, x = `Percentage Virally Suppressed`)) +
+  geom_smooth(method = "lm") +
+  geom_point() + 
+  labs(title = "HIV Death Rate vs. Percentage Virally Suppressed") +
+  facet_wrap(Race ~ . , scales = "free_y") + 
+  stat_cor(method="pearson")
+```
+Now that we have an idea of the relationship between our outcome and potential predictors, next I need to account for the fact that these variables a likely have some degree of multicollinearity.
+In the next step we will be using PCA to scale our predictors.
+
+```{r}
+saveRDS(df, here("data/processed_data/race_new.rds"))
+```

### 2023-04-21 code/analysis_code/Race/p2_PCA_race.Rmd
more stuff

--- 2023-03-18 code/analysis_code/Race/p2_PCA_race.Rmd
@@
 library(factoextra)
-library(spatialreg)
-library(MASS)
 library(here)
+library(knitr)
 library(tidyverse)
+library(broom)
 
@@
 
+PCA
 ```{r}
-pca_function <- function(race) {   
-  # filter race_df by the selected race/ethnicity
-  df <- df[df$Race == race,]
+df <- df %>%
+    select_if(~ !any(is.na(.)))
 
-num_race <- df[,c(9:10, 16:25)]
+  #separating the y variables
+df_response <- df %>%
+    select(contains(c("State", "Race", "Death", "Population", "BOX", "YJ", "Ratio", "outcome_DR")))
 
-#normalization
-norm_race <- scale(num_race)
-print(head(norm_race))
+  # separating x variables and scaling
+df_predictor <- df %>% 
+    select(contains(c("State", "Race", "Percent", "Gini", "Median", "Broadband"))) %>% 
+    #select(!contains(c("Virally", "Received", "Late"))) %>% 
+    mutate(across(where(is.numeric), scale))
+
+#Setting up correlation matrix
+corr_matrix <- df_predictor %>% 
+    select(where(is.numeric)) %>% 
+    cor()
 
   #setting up correlation matrix 
-corr_matrix_race <- cor(norm_race)
-print(ggcorrplot(corr_matrix_race))
-
-#applying pca
-pca_race <- princomp(corr_matrix_race)
-print(summary(pca_race))
-#pca loadings
-print(pca_race$loadings[, 1:2])
-
-#Scree plot
-print(fviz_eig(pca_race, addlabels = TRUE))
-# Biplot of the attributes
-print(fviz_pca_var(pca_race, col.var = "darkblue"))
-
-#Contribution of each variable 
-print(fviz_cos2(pca_race, choice = "var", axes = 1:2))
-
-# Biplot of the attributes
-print(fviz_pca_var(pca_race, col.var = "cos2",
-             gradient.cols = c("darkblue", "orange", "green"),
-             repel = TRUE))
-# Variables in green are a good representation of the variable on that component.
-}
+print(ggcorrplot(corr_matrix))
+ggsave(here("results/corr_mat.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
 ```{r}
-pca_function("Black")
+#applying pca
+pca <- princomp(corr_matrix)
+print(summary(pca))
+
+#pca loadings of principal components 1 and 2
+print(pca$loadings[, 1:2])
+
+#Scree plot
+print(fviz_eig(pca, addlabels = TRUE))
+ggsave(here("results/scree_plot.png"), units="in", width=10, height=7, dpi=300)
+```
+Since we originally had 11 variables, we should only look at components that explain more than 9.1% of variation.
+We can see from the scree plot that principal component 1 explained 63.8% of variation, and components 1 and 2 explained 84.7% of the variation, and components 1 - 3 explained 94.8%.
+Component 3 just  barely passed the threshold with 10.1%.
+All other principal components explains less that 6.2% of variation, so they will excluded.
+
+```{r}
+#Contribution of each variable 
+print(fviz_cos2(pca, choice = "var", axes = 1:2))
+
+# Biplot of the attributes
+print(fviz_pca_var(pca, col.var = "cos2",
+                   gradient.cols = c("darkblue", "orange", "green"),
+                   repel = TRUE, title = "PCA"
+                   ))
+# Variables in green are a good representation of the variable on that component.
+
+#Creating object to store scaled dataframes
+df_scaled <- full_join(df_response, df_predictor, by = join_by("State", "State Abbreviation", "States with HIV-Specific Criminal Laws/Statutes", "States That Have Prosecuted PLWH", "Race"))
 ```
@@
 ```{r}
-pca_function("Hispanic")
+saveRDS(df_scaled, here("data/processed_data/df_scaled.rds"))
 ```
-
-```{r}
-pca_function("Multiple")
-```
-
-```{r}
-pca_function("White")
-```

### 2023-04-21 code/analysis_code/Race/p3_model_race.Rmd
more stuff

--- 2023-03-18 code/analysis_code/Race/p3_model_race.Rmd
@@
 library(here)
+library(MASS)
+library(spatialreg)
+library(spdep)
 library(tidyverse)
-library(spdep)
 ```
@@
 ```{r}
-df <- readRDS(here("data/processed_data/race_new.rds"))
+df <- readRDS(here("data/processed_data/df_scaled.rds"))
 us_nabs <- readRDS(here("data/processed_data/us_nabs.rds"))
@@
   filter(!grepl("Alaska", State),
-         !grepl("Hawaii", State)) %>% 
-  mutate(log_rate = log(`Death Rate (estimated)`+ 1))
+         !grepl("Hawaii", State))
+```
+
+subset by race/ethnicity
+```{r}
+black_df <- df %>% 
+  filter(Race == "Black")
+
+hispanic_df <- df %>% 
+  filter(Race == "Hispanic")
+
+white_df <- df %>% 
+  filter(Race == "White")
 ```
@@
 Setting up functions for models
+
+```{r}
+ #multiple linear regression
+mult_function <- function(df) {
+  mult_race <- glm(outcome_DR ~ 
+                        `Gini Coefficient` + # estimate income inequality
+                        `Percent Unemployed` +
+                        `Percent Living in Poverty` +
+                        `Percent High School Education` +  #estimate education access
+                 #       `Broadband of any type` + # estimate built 
+                   #     `Percent Uninsured Under 65 Years Old`+ #estimate access to healthcare
+                        `Percent Living with Severe Housing Cost Burden`, #economic stability,
+                     data = df,
+                     family = "gaussian")
+  print(summary(mult_race))
+    ##pull fitted values
+df$mult_fit <- mult_race$fitted.values
+return(df)
+}
+
+## function to fit GLM and pull AIC
+mult_aic <- function(df) {
+  mult_race <- glm(outcome_DR ~ 
+                        `Gini Coefficient` + # estimate income inequality
+                        `Percent Unemployed` +
+                     `Percent Living in Poverty` +
+                        `Percent High School Education` +  #estimate education access
+                    #     `Broadband of any type` + # estimate built 
+                     #   `Percent Uninsured Under 65 Years Old`+ #estimate access to healthcare
+                        `Percent Living with Severe Housing Cost Burden`, #economic stability,
+                     data = df,
+                     family = "gaussian")
+aic <- AIC(mult_race)
+
+return(aic)
+}
+```
+
 ```{r}
   #poisson
-pois_function <- function(race) {
-  pois_race <- glm(`Death Cases` ~ `Median Household Income` +
-                       `Percent High School Education` +
+pois_function <- function(df) {
+  pois_race <- glm(`Death Cases`~ 
+                        `Gini Coefficient` + # estimate income inequality
+                        `Percent Unemployed` +
+                     `Percent Living in Poverty` +
+                        `Percent High School Education` +  #estimate education access
+                      #   `Broadband of any type` + # estimate built 
+                     #   `Percent Uninsured Under 65 Years Old` + #estimate access to healthcare
+                        `Percent Living with Severe Housing Cost Burden` + #economic stability
                     offset(log(`Population`)),
-                     data = df[df$Race == race,],
-                     family = poisson)
+                     data = df,
+                     family = "poisson")
   print(summary(pois_race))
+
+      ##pull fitted values
+df$pois_fit <- pois_race$fitted.values
+return(df)
 }
 
+## function to fit GLM and pull AIC
+pois_aic <- function(df) {
+  pois_race <- glm(`Death Cases`~ 
+                        `Gini Coefficient` + # estimate income inequality
+                        `Percent Unemployed` +
+                     `Percent Living in Poverty` +
+                        `Percent High School Education` +  #estimate education access
+                     #    `Broadband of any type` + # estimate built 
+                    #    `Percent Uninsured Under 65 Years Old` + #estimate access to healthcare
+                        `Percent Living with Severe Housing Cost Burden` + #economic stability
+                    offset(log(`Population`)),
+                     data = df,
+                     family = "poisson")
+aic <- AIC(pois_race)
+return(aic)
+}
+```
+
+```{r}
   #Negative binomial
-nb_function  <- function(race) {
-  neg_bin_race <- glm.nb(`Death Cases` ~ `Median Household Income` +
-                              `Percent High School Education` + 
+nb_function  <- function(df) {
+  neg_bin_race <- glm.nb(`Death Cases` ~ 
+                        `Gini Coefficient` + # estimate income inequality
+                        `Percent Unemployed` +
+                          `Percent Living in Poverty` +
+                        `Percent High School Education` +  #estimate education access
+                     #   `Broadband of any type` + # estimate built 
+                      #  `Percent Uninsured Under 65 Years Old` + #estimate access to healthcare
+                        `Percent Living with Severe Housing Cost Burden` + #economic stability
                      offset(log(`Population`)),
-                            data = df[df$Race == race,])
+                            data = df)
   print(summary(neg_bin_race))
+
+      ##pull fitted values
+df$nb_fit <- neg_bin_race$fitted.values
+
+return(df)
+
 }
 
+## function to fit GLM and pull AIC
+nb_aic <- function(df) {
+  neg_bin_race <- glm.nb(`Death Cases` ~ 
+                        `Gini Coefficient` + # estimate income inequality
+                        `Percent Unemployed` +
+                          `Percent Living in Poverty` +
+                        `Percent High School Education` +  #estimate education access
+                      #   `Broadband of any type` + # estimate built 
+                   #    `Percent Uninsured Under 65 Years Old` + #estimate access to healthcare
+                        `Percent Living with Severe Housing Cost Burden` + #economic stability
+                     offset(log(`Population`)),
+                            data = df)
+aic <- AIC(neg_bin_race)
+
+return(aic)
+
+}  
+```
+
+```{r}
 #Fitting CAR model (requires neighbors)
-car_function <- function(race) {   
-  # filter race_df by the selected race/ethnicity
-  df <- df[df$Race == race,]
-
+car_function <- function(df) {   
   h_race_df <- matrix(0, nrow(df), nrow(df))
@@
 
-  car_race <- spautolm(log_rate ~ `Percent Living in Poverty` + 
-                             `Percent High School Education`,
+  car_race <- spautolm(outcome_DR ~                      
+                        `Gini Coefficient` + # estimate income inequality
+                        `Percent Unemployed` +
+                         `Percent Living in Poverty` +
+                        `Percent High School Education` +  #estimate education access
+                       #  `Broadband of any type` + # estimate built 
+                     #   `Percent Uninsured Under 65 Years Old`+ #estimate access to healthcare
+                        `Percent Living with Severe Housing Cost Burden`,#economic stability,
                            data = df,
@@
   print(summary(car_race))
+
+  ##pull fitted values
+df$car_fit <- car_race$fit$fitted.values
+
+return(df)
+
 }
+
+car_aic <- function(df) {
+  h_race_df <- matrix(0, nrow(df), nrow(df))
+  h_race_df[cbind(match(us_nabs$StateCode, df$`State Abbreviation`),
+                  match(us_nabs$NeighborStateCode, df$`State Abbreviation`))] <- 1
+
+  nablist_race_df <- mat2listw(h_race_df, style = "W")
+
+  car_race <- spautolm(outcome_DR ~                      
+                        `Gini Coefficient` + # estimate income inequality
+                        `Percent Unemployed` +
+                         `Percent Living in Poverty` +
+                        `Percent High School Education` +  #estimate education access
+                     #    `Broadband of any type` + # estimate built 
+                    #    `Percent Uninsured Under 65 Years Old`+ #estimate access to healthcare
+                        `Percent Living with Severe Housing Cost Burden`,#economic stability,
+                           data = df,
+                           listw = nablist_race_df,
+                           family = "CAR")
+aic <- AIC(car_race)
+return(aic)
+
+} 
+```
+
+Making a function to pull AIC and organize into new dataframe
+```{r}
+aic_df <- function(x, mod_type, race) {
+  x <- tibble(x)
+  x <- x %>% 
+    mutate(Race = race, 
+           Model = mod_type,
+           AIC = round(x,2)) %>% 
+    select(Race, Model, AIC)
+}
+```
+
+Fitting multiple linear regression models
+```{r}
+black_df <- mult_function(black_df)
+
+black_mult <- mult_aic(black_df)
+black_mult <- aic_df(black_mult, "GLM", "Black")
+```
+
+```{r}
+hispanic_df <- mult_function(hispanic_df)
+
+hispanic_mult <- mult_aic(hispanic_df)
+hispanic_mult <- aic_df(hispanic_mult, "GLM", "Hispanic")
+```
+
+```{r}
+white_df <- mult_function(white_df)
+## SINCE THE OUTCOME IN NOT NORMALLY DISTRIBUTED, WE CANNOT USE GLM 
+white_df <- white_df %>% mutate(mult_fit = NA)
+
+white_mult <- mult_aic(white_df)
+white_mult <- aic_df(white_mult, "GLM", "White")
+
+white_mult <- white_mult %>% mutate(AIC = NA)
+```
+
+```{r}
+mult_mod <- list(black_mult, hispanic_mult, white_mult)
+
+mult_mod <- reduce(mult_mod, full_join)
 ```
@@
 ```{r}
-pois_function("Black")
+black_df <- pois_function(black_df)
+
+black_pois <- pois_aic(black_df)
+black_pois <- aic_df(black_pois, "Poisson", "Black")
 ```
@@
 ```{r}
-pois_function("Hispanic")
+hispanic_df <- pois_function(hispanic_df)
+
+hispanic_pois <- pois_aic(hispanic_df)
+hispanic_pois <- aic_df(hispanic_pois, "Poisson", "Hispanic")
 ```
@@
 ```{r}
-pois_function("Multiple")
+white_df <- pois_function(white_df)
+
+white_pois <- pois_aic(white_df)
+white_pois <- aic_df(white_pois, "Poisson", "White")
 ```
 
+Storing Poisson AIC
 ```{r}
-pois_function("White")
+pois_mod <- list(black_pois, hispanic_pois, white_pois)
+
+pois_mod <- reduce(pois_mod, full_join)
 ```
@@
 ```{r}
-nb_function("Black")
+black_df <- nb_function(black_df)
+
+black_nb <- nb_aic(black_df)
+black_nb <- aic_df(black_nb, "Negative Binomial", "Black")
 ```
@@
 ```{r}
-nb_function("Hispanic")
+hispanic_df <- nb_function(hispanic_df)
+
+hispanic_nb <- nb_aic(hispanic_df)
+hispanic_nb <- aic_df(hispanic_nb, "Negative Binomial", "Hispanic")
 ```
@@
 ```{r}
-nb_function("Multiple")
+white_df <- nb_function(white_df)
+
+white_nb <- nb_aic(white_df)
+white_nb <- aic_df(white_nb, "Negative Binomial", "White")
 ```
@@
 ```{r}
-nb_function("White")
+nb_mod <- list(black_nb, hispanic_nb, white_nb)
+
+nb_mod <- reduce(nb_mod, full_join)
 ```
-The NB model accounting for overdispersion fit better than the poisson model.
+The NB model accounting for overdispersion fits better than the poisson model.
 
@@
 ```{r}
-car_function("Black")
+black_df <- car_function(black_df)
+
+black_car <- car_aic(black_df)
+black_car <- aic_df(black_car, "CAR", "Black")
 ```
@@
 ```{r}
-car_function("Hispanic")
+hispanic_df <- car_function(hispanic_df)
+
+hispanic_car <- car_aic(hispanic_df)
+hispanic_car <- aic_df(hispanic_car, "CAR", "Hispanic")
 ```
@@
 ```{r}
-car_function("Multiple")
+white_df <- car_function(white_df)
+## SINCE THE OUTCOME IN NOT NORMALLY DISTRIBUTED, WE CANNOT USE CAR
+white_df <- white_df %>% mutate(car_fit = NA)
+
+white_car <- car_aic(white_df)
+white_car <- aic_df(white_car, "CAR", "White")
+white_car <- white_car %>% mutate(AIC = NA)
 ```
@@
 ```{r}
-car_function("White")
+car_mod <- list(black_car, hispanic_car, white_car)
+
+car_mod <- reduce(car_mod, full_join)
+```
+
+```{r}
+mod_aic <- list(mult_mod, pois_mod, nb_mod, car_mod)
+
+mod_aic <- reduce(mod_aic, full_join)
+```
+
+Join the data frames with all the fits.
+```{r}
+df <- list(black_df, hispanic_df, white_df)
+
+df <- reduce(df, full_join)
+
+saveRDS(df, here("data/processed_data/df_fit.rds"))
+```
+save the AIC dataframe
+```{r}
+saveRDS(mod_aic, here("data/processed_data/aic.rds"))
 ```
@@
 ```{r}
-spacy_function <- function(race) {
-  # filter race_df by the selected race/ethnicity
-  df <- df[df$Race == race,]
+spacy_function <- function(df) {
 
@@
  # plot Moran's I
-  par(mfrow=c(1,2))
-  print(moran.test(df$BOX, nablist_race_df, zero.policy = TRUE, randomisation = FALSE))
-  moran.plot(df$BOX, nablist_race_df, zero.policy = TRUE)
+  print(moran.test(df$outcome_DR, nablist_race_df, zero.policy = TRUE, randomisation = FALSE))
+  moran.plot(df$outcome_DR, nablist_race_df, zero.policy = TRUE)
 
   # calculate and print LISA statistics
-  lisa_race <- localmoran(df$`Ratio Observed/Expected`, nablist_race_df)
+  lisa_race <- localmoran(df$outcome_DR, nablist_race_df)
   summary(lisa_race[,1])
@@
 ```{r}
-spacy_function("Black")
+spacy_function(black_df)
 ```
@@
 ```{r}
-spacy_function("Hispanic")
+png(here("results/moran_black.png"))
+#sink(here("results/spacy_summary_black.txt"))
+spacy_function(black_df)
+sink()
+dev.off()
 ```
@@
 ```{r}
-spacy_function("Multiple")
+spacy_function(hispanic_df)
 ```
@@
 ```{r}
-spacy_function("White")
+png(here("results/moran_hispanic.png"))
+#sink(here("results/spacy_summary_hispanic.txt"))
+spacy_function(hispanic_df)
+sink()
+dev.off()
 ```
@@
 ```{r}
-save.image(file="../../data/processed_data/race_env.rds")
+spacy_function(white_df)
 ```
+
+```{r}
+png(here("results/moran_white.png"))
+#sink(here("results/spacy_summary_white.txt"))
+spacy_function(white_df)
+sink()
+dev.off()
+```
+
+```{r}
+save.image(here("data/processed_data/race_env.rds"))
+```

### 2023-04-21 code/analysis_code/Race/p4_mapping.Rmd
more stuff

In this section, I will be creating some maps of the data. I'll start by loading some packages and the dataframes.

```{r}
library(here)
library(tidyverse)
library(spdep)
library(RColorBrewer)

#loading a spatial object for the contiguous US.

states <- map_data("state")

df_scaled <- readRDS(here("data/processed_data/df_fit.rds"))

df_raw <- readRDS(here("data/processed_data/race_new.rds"))

mod_aic <-  readRDS(here("data/processed_data/aic.rds"))
```

I do a bit more cleaning to the data to tidy things up and create some new variables
```{r}
df_scaled <- df_scaled %>%
  mutate(region = tolower(State)) %>% 
  replace(is.na("Ratio Observed/ Expected"), 0) %>% 
  mutate(Exp_DR = Exp_Deaths/Population *1e5,
               OE_mortality_ratio = `Death Rate (estimated)`/Exp_DR,
         dykes_chi = ((`Death Cases`- Exp_Deaths)/ sqrt(Exp_Deaths)),

         ## GLM
         predicted_observed_mult = mult_fit/`Death Cases`,
         predicted_observed_mult = ifelse(!is.finite(predicted_observed_mult), 0, predicted_observed_mult),
         observed_predicted_mult = `Death Cases`/mult_fit,
         observed_predicted_mult = ifelse(!is.finite(observed_predicted_mult), 0, observed_predicted_mult),
         ## CAR
         predicted_observed_car = car_fit/`Death Cases`,
         predicted_observed_car = ifelse(!is.finite(predicted_observed_car), 0, predicted_observed_car),
         observed_predicted_car = `Death Cases`/car_fit,
         observed_predicted_car = ifelse(!is.finite(observed_predicted_car), 0, observed_predicted_car),
         ## POISSON
         predicted_observed_pois = pois_fit/`Death Cases`,
         predicted_observed_pois = ifelse(!is.finite(predicted_observed_pois), 0, predicted_observed_pois),
         observed_predicted_pois = `Death Cases`/pois_fit,
         observed_predicted_pois = ifelse(!is.finite(observed_predicted_pois), 0, observed_predicted_pois),
         ##N NEGATIVE BINOMIAL
         predicted_observed_nb = nb_fit/`Death Cases`,
         predicted_observed_nb = ifelse(!is.finite(predicted_observed_nb), 0, predicted_observed_nb),
         observed_predicted_nb = `Death Cases`/nb_fit,
         observed_predicted_nb = ifelse(!is.finite(observed_predicted_nb), 0, observed_predicted_nb))

#Best model  
best_mod_aic <- mod_aic %>%
  group_by(Race) %>%
  drop_na() %>% 
  filter(AIC == min(AIC))

df_scaled <- full_join(df_scaled, best_mod_aic, by = join_by("Race"))

##Pulling fitted values based on best fitting model
df_scaled <- df_scaled %>% 
  mutate(best_fitted = case_when(Model == "CAR" ~ car_fit,
                                 Model == "GLM" ~ mult_fit,
                                 Model ==  "Poisson" ~ pois_fit,
                                 Model == "Negative Binomial" ~ nb_fit))

df_raw <- df_raw %>%
  mutate(region = tolower(State)) 
```

Next, I will merge my dataframe contain the variables of interest with the spatial object. 
```{r}
df_sf_scale <- merge(states,df_scaled, by = "region")
# This next step is import to ensure that no "holes" appear in the map.
df_sf_scale <- df_sf_scale[order(df_sf_scale$order), ]

df_sf_raw <- merge(states,df_raw, by = "region")
# This next step is import to ensure that no "holes" appear in the map.
df_sf_raw <- df_sf_raw[order(df_sf_raw$order), ]
```

Now I'm ready to start mapping! I'll begin by looking at the map of the raw HIV death rate across to contiguous US by selected race/ ethnicity.
```{r}
ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Death Rate (estimated)`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "HIV Mortality Rate (raw) by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Death rate per 100,000") +
  scale_fill_distiller(palette = "RdYlGn") +
  theme(plot.title= element_text(hjust = 0.5))

ggsave(here("results/raw_rate_map.png"), units="in", width=10, height=4, dpi=300)
```

```{r}
ggplot(df_sf_raw, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Gini Coefficient`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Gini Coefficient, 2020", x = "", y = "", fill = "Gini coefficient (higher values represent greater income inequality)")  +
  theme(plot.title= element_text(hjust = 0.5))
ggsave(here("results/gini_map.png"), units="in", width=10, height=7, dpi=300)

ggplot(df_sf_raw, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent High School Education`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent High School Education, 2020", x = "", y = "", fill = "%") +
  theme(plot.title= element_text(hjust = 0.5)) 
ggsave(here("results/ed_map.png"), units="in", width=10, height=7, dpi=300)

ggplot(df_sf_raw, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Living with Severe Housing Cost Burden`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Living with Severe Housing Cost Burden, 2020", x = "", y = "", fill = "%") +
  theme(plot.title= element_text(hjust = 0.5)) 
ggsave(here("results/housing_map.png"), units="in", width=10, height=7, dpi=300)

ggplot(df_sf_raw, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Living in Poverty` )) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Living in Poverty, 2020", x = "", y = "", fill = "%") +
  theme(plot.title= element_text(hjust = 0.5)) 
ggsave(here("results/poverty_map.png"), units="in", width=10, height=7, dpi=300)

ggplot(df_sf_raw, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Unemployed` )) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Unemployed, 2020", x = "", y = "", fill = "%") +
  theme(plot.title= element_text(hjust = 0.5)) 
ggsave(here("results/unemployed_map.png"), units="in", width=10, height=7, dpi=300)
```
The map for HIV-related death differs drastically across the United States by race/ethnicity. In most states, the highest HIV-related death were among Black person. This is especially true in the South region. Non-coastal western states consistently had lowest rates HIV-related deaths across the selected racial/ethnic groups, expect among multi-racial non-Hispanic persons. Non-Hispanic White persons were the only of the selected racial/ ethnic group that did not have a HIV-related death rate higher than ~10 per 100,000.

Let's map the observed/ expected ratio.
```{r}
ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = OE_mortality_ratio)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race) +
  scale_fill_distiller(palette = "RdYlGn") + theme_void() + 
  theme(strip.text = element_text(size = 14),
    strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Observed/ Expected HIV Mortality Ratio by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))

ggsave(here("results/roe_map.png"), units="in", width=10, height=4, dpi=300)
```
We can see that in much of the United States non-Hispanic Black persons had ratios consistently (much) greater than 1, suggesting a higher risk among this population.

Next, let's map after the the Box_Cox transformation of the death rate
```{r}
ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = BOX)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race) +
  scale_fill_distiller(palette = "RdYlGn") + theme_void() + 
  theme(strip.text = element_text(size = 14),
    strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Death Rates (Box Cox transformed) by Race/ Ethnicity
in the United States, 2020", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))
ggsave(here("results/roe_box_map.png"), units="in", width=10, height=4, dpi=300)
```

Next, let's map a probability map based on where we'd expect deaths if risk of HIV-related death was consistent across the US.
```{r}
ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = Exp_DR)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race) +
  scale_fill_viridis_c() + theme_void() + 
  theme(strip.text = element_text(size = 14),
    strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "NULL - Expected HIV Mortality Rate by Race/ Ethnicity
       in the United States, 2020", x = "", y = "", fill = "Rate") +
  theme(plot.title= element_text(hjust = 0.5))
```
Well.. I probably need to go back and reconsider using making a lambda based on specific race/ ethnic. I think that at this point, however, it is still fair to consider mapping an observed/ expected ratio using the expected deaths calculate from the lambda for the entire population. This will allow us to see differences across the race/ ethnicity, rather than masking disparities in rates due higher expected cases among certain high-risk populations, which would thus make the ratio closer to the null.

Next, let's using Dykes and Unwin's method of smooth map using a chi square value. 
```{r}
ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = dykes_chi)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + theme_void() + 
  theme(strip.text = element_text(size = 14),
        strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Dykes and Unwin's Chi Square by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Chi Square") +
  scale_fill_distiller(palette = "RdYlGn") +
  theme(plot.title= element_text(hjust = 0.5))

ggsave(here("results/chi_map.png"), units="in", width=10, height=4, dpi=300)
```
##Dykes and Unwin’s chi-square statistic
Dykes and Unwin (Dykes and Unwin 2001) technique  has the effect of creating a distribution of values closer to normal (as opposed to a Poisson distribution of rates and counts encountered thus far). We can therefore apply a diverging color scheme where green hues represent less than expected rates and red hues represent greater than expected rates.

Next, let's make a map using the fitted values from the CAR model, the best performing model.
```{r}
##Smoothed map using fitted values from best models.
ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = best_fitted)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Smoothed Map (Fitted) of HIV-Related Death Rate by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "") +
  theme(plot.title= element_text(hjust = 0.5))

ggsave(here("results/fitted_map.png"), units="in", width=10, height=4, dpi=300)

```

Next, let's look at the map for the ratio of predicted to observed (P/O) values. 
```{r}
ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = predicted_observed_mult)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Map of Ratio of Predicted to Observed Cases (GLM) by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))

ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = predicted_observed_car)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Map of Ratio of Predicted to Observed Cases (CAR) by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))
``` 
We can see that the model fitted pretty well. A higher P/O ratio (more red values) indicates that the model overestimated the observed values, while a lower P/O ratio (more green values) indicates that the model underestimates the observed values. We can see that the model underestimated rates primarily in the non-coast north-western states and north central states for persons of Hispanic, non-Hispanic White, and non-Hispanic multiple racial/ ethnic backgrounds.

### 2023-04-21 code/processing_code/1_cleaning_2020.Rmd
added more to project
more stuff

@@
 us_race_pop <- read_xlsx(here("data/raw_data/DECENNIALPL2020.P2-2023-02-24T124235.xlsx"), sheet = 2)
-us_sex_pop <-read_csv(here("data/raw_data/raw_data.csv"), skip = 2)
+us_sex_pop <-read_csv(here("data/raw_data/raw_data.csv"), skip = 2) # not used in this analysis
 kid_pop <- read_xlsx(here("data/raw_data/Child population by age group.xlsx"))
@@
 v_sup  <- read_xlsx(here("data/raw_data/AIDSVu_State_ViralSuppression_2020.xlsx"), skip = 3)
+internet <- read_xlsx(here("data/raw_data/ACSST5Y2020.S2801-2023-04-17T232857.xlsx"), skip = 3, sheet = 2)
 ```
@@
 ```
+
 Some rates are -9 due to missingness. Let's convert these to NA
@@
 
-Some quick cleaning of the social determinant sof health file, and merging this with our two results dataframes `sex_df` and `race_df`.
+Some quick cleaning of the social determinants of health file, and merging this with our two results dataframes `sex_df` and `race_df`.
 ```{r}
@@
 
+```{r}
+internet <- internet %>% 
+  select(c(1, "Broadband of any type")) %>% 
+  slice(c(-1:-7, -48:-52,-263:-267)) %>% 
+  filter(!grepl(",", `Broadband of any type`))
+```
+
+This dataframe has some issues. Let's fill in the state names...
+```{r}
+# create a vector of state names
+states <- state.name
+
+# repeat each state name five times
+states <- states %>%
+  rep(each = 4)
+
+# sort the repeated state names in alphabetical order
+states <- sort(states)
+
+# print the resulting vector
+states
+
+internet$State<- states
+
+internet <- internet %>% 
+  select(State, `Broadband of any type`) %>% 
+  drop_na()
+
+internet$`Broadband of any type` <- gsub("%", "", internet$`Broadband of any type`)
+
+internet <- internet %>% mutate(`Broadband of any type` = as.numeric(`Broadband of any type`))
+```
+
+```{r}
+race_df <- full_join(race_df, internet)
+sex_df <- full_join(sex_df, internet)
+```
+
 We have some more cleaning to, but let's save these files for now. 
@@
 ```
-
-```{r}
-rm(deaths, df_pop, kid_pop, pop_sex, pop_less13, us_sex_pop, us_race_pop, new_pop, deaths_sex, race_pop, kid_race, total_pop, deaths_race, late_dx, new_dx, sex_dx, prev, prev_sex, prev_race, v_sup, rec_care, sdoh)
-```

### 2023-04-21 code/processing_code/2b_cleaning_race_df.Rmd
more stuff

@@
 ```{r}
-black_df <- race_df %>% select(1:3, contains("Black"), 50:62)
+black_df <- race_df %>% select(1:3, contains("Black"), 50:63)
 
@@
 
-hispanic_df <- race_df %>% select(1:3, contains("Hispanic"), 50:62)
+hispanic_df <- race_df %>% select(1:3, contains("Hispanic"), 50:63)
 
@@
 
-white_df <- race_df %>% select(1:3, contains("White"), 50:62)
+white_df <- race_df %>% select(1:3, contains("White"), 50:63)
 
@@
 
-multiple_df <- race_df %>% select(1:3, contains("Multiple"), 50:62)
+multiple_df <- race_df %>% select(1:3, contains("Multiple"), 50:63)
 

### 2023-04-21 outputs_code.Rmd
more stuff

#Table
library(kableExtra)
```{r}
df <- readRDS(here("data/processed_data/race_new.rds"))

df %>% 
  group_by(Race) %>% 
  summarize(tot_deaths = sum(`Death Cases`),
            tot_population = sum(`Population`),
            overall_rate = round(tot_deaths/ tot_population *1e5,1)) %>% 
  ungroup() %>% 
  mutate(percent_by_race = round(tot_deaths/sum(tot_deaths)*100, 1),
         ratio = round(overall_rate/overall_rate[Race =="White"],1)) %>%
  kable(col.names = c("Race/ Ethnicity",
                      "HIV-Related Deaths (n)",
                      "Population \n (>12 years old)",
                      "Death Rate \n (per 100,000)",
                      "Percent",
                      "Ratio"),
        caption = "HIV-Related Deaths Summary by Race/ Ethncity, United States, 2020",
        align = "c") %>%
  kable_minimal() %>%
  kable_styling(font_size = 50) %>% 
  save_kable(here("results/test.png"))
# using white persons as reference group
```

### 2023-04-21 products/manuscript/manuscript.qmd
more stuff

---
title: "Spatial Investigation and Exploration of Factors influencing HIV-related Mortality"
subtitle: "United States, 2020"
author: Hayley Hemme
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis_references.bib
csl: ../apa.csl
---

```{r, echo=FALSE, warning = FALSE, message=FALSE}
# load a few R packages
library(knitr)
library(readxl)
library(lubridate)
library(stringr)
library(tidyverse)
library(here)
library(kableExtra)
```

# Summary/Abstract

*Write a summary of your project.*

{{< pagebreak >}}

# Introduction

HIV has a disproportionate impact on marginalized groups, resulting in significant disparities in rate of new HIV diagnoses and poor HIV-related health outcomes, such as late HIV diagnosis, progression to AIDS, and HIV-related death. Previous study have found direct links to indicators of social determinants of health and poorer health outcomes for persons with HIV

Other studies have found an association between different SDOH measures and COVID-19 mortality that varied across racial and ethnic groups and community types

## Question:

Can the HIV death rate from 2020 in the United States be modeled by social determinants of health and/ or by HIV-related outcomes (late diagnoses, viral suppression) Are these factors correlated with disparities in HIV-related mortality across different racial and ethnic groups? Can a spatial model explain variation in HIV death-rates at the state-level and/or regional-level after accounting for variables associated with sociodemographic, SDOH, and HIV-related outcomes?

H0 : There is not a spatial correlation HIV-related deaths and SDOH or HIV-related outcomes.

Ha : There is a spatial correlation between HIV-related deaths and SDOH or HIV-related outcomes.

# Background Information

Since the emergence of the HIV epidemic in 1981 in the United States, more than 1.2 million people in the United States have been diagnosed with HIV, resulting in more than 700,000 HIV-related deaths \[cite\]. Medical advancements have led to effective antiretroviral therapies that not only treat HIV infection and reduce adverse HIV-related health outcomes, such as progression to AIDS, but have also led to use of these medications to prevent infection before and after potential exposure for people who are HIV negative Even still, more than 30,000 people in the United States continue to be diagnosed each year \[cite\]. While initiatives such the Ending the Epidemic may provide xxx, this is followed by nearly one decade of steadily maintained HIV diagnoses rates, suggesting that existence of effective medical therapy alone will not end the HIV epidemic in the United States given the complex barriers existence among populations who are most vulnerable for HIV infection. The goal of this project is to gain a between understanding of factors associated with HIV-related mortality in the United States in

# Materials and Methods

## Data Source

The data used in this project was obtained from multiple publicly available source.

### AIDSVu

Data was obtained from AIDSVu \cite{AIDSVuDatasetsIncludes} on counts and rates of HIV-related mortality by state in 2020. Social determinants of health (SDOH), were also obtained from AIDSVu by state from 2020. SDOH included a number of variable represented as a percentage particularly pertinent to health persons with HIV. AIDSVu is a public resource presented by Emory University's Rollins School of Public Health in partnership with Gilead Sciences, Inc. and the Center for AIDS Research at Emory University (CFAR) and contains HIV surveillance data and other pertinent data to monitoring HIV-related outcomes in the United States.

### US Census Bureau

Data on 2020 demographic composition by race/ethnicity and age by state was obtained from the [U.S. Census Bureau](https://data.census.gov/) \cite{USCensusBureau}. A five-year estimate of percent with access to broadband internet by state (released in 2020) was also obtained from the US Census Bureau.

### Kids Count

Data on the percentage distribution of children under 14 years old was obtained from [Kids Count](https://datacenter.kidscount.org/), an online resource that provides high-quality data on the well-being of children in the United States.

### Spatial data

Spatial object of the contiguous United States was obtained from library(maps) in R. The list of neighboring contiguous United States was obtained from a public repository on GitHub. https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state

## Data processing

### Cleaning

Data was cleaned to adjust variable format, merge dataframes, subset by race/ ethnicity, and for feature engineering. Several observations in response variables were suppressed due to small counts/ rates were represented by `-9`, and were re-coded to NA. To removed multicollinearity and to assess the relationship between predictors, principal component analysis (PCA). was used. Variables that had correlation greater that 0.3 in the first two principal components were deemed important. Scaled predictors of the the first three principal components were then used in various regression models multiple linear regression, Poisson regression, and negative binomial. Non-spatial and spatial methods were used.

### Feature Engineering

The primary response variable in this analysis was HIV-related death rate. Feature engineering of `death rate (estimated)` was performed using counts of HIV related deaths divided by the estimated population based on US Census data multiplied by 100,000 persons. Initially, persons of multiple racial backgrounds were considered in the analysis, however, significant discrepancies between the death rates provided by AIDSVu and the calculated death rate based on counts and estimated population meant that I could not reliably interpret these data. The decision to used `death rate (estimated)` instead of `death rate (AIDSVu)` was made to help ensure that differences across various models that used rates versus counts was not 'artifact'. Transformation of this variable was performed as necessary, and is discussed further in later sections.

### Variables considered

Predictor variables used in this analysis were selected to serve as proxy for SDOH. The Gini coefficient was used measure income inequality; percent living in poverty was used to estimate economic stability; percent with high school education was used to estimate access to education; percent uninsured under 65 was used to measure health care access; and percent with access to broadband internet was used to estimate the neighborhood and built environment.

## Model Selection

The relative goodness of fit of the various models used in this analysis was assessed using the Akaike information criterion (AIC).

# Analysis

![](images/test-01.png){fig-align="center"}

## Univariate analysis

![](images/dr_boxplot.png){fig-align="center" width="700"}

The distribution of death rate was assessed for each race both visually and using the Shapiro test for normality. Below is the histogram of the distributed of the raw `death rate (estimated).`

![](images/dr_hist-01.png){fig-align="center" width="700"}

The death rate for Black persons did not require transformation and the death rate for Hispanic persons was normalized using a log + 1 transformation. Despite utilizing transformations of increasing complexity and power, including Box Cox and Yeo Johnson transformations, I was not able to normalize the distribution of death rate for White persons. As a result, I was unable to include White persons in the model for assessing for spatial correlation. While there are techniques to handle spatial data that is not normally distributed, such as geographically weighted Poisson regression and geographically weight negative binomial regression, my knowledge of the implementation of these techniques limited my ability to uses these methods in my analysis.

![](images/raw_rate_map-05.png){fig-align="center"}

![](images/roe_map-05.png)

![](images/chi_map-05.png){fig-align="center"}

|                                             |                                          |
|---------------------------------------------|------------------------------------------|
| ![](images/gini_map.png){width="300"}       | ![](images/ed_map-01.png){width="300"}   |
| ![](images/housing_map.png){width="300"}    | ![](images/poverty_map.png){width="300"} |
| ![](images/unemployed_map.png){width="300"} |                                          |

```{r}

```

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

## Bivariate Analysis

![](images/bi_gini-01.png){fig-align="center" width="429"}

![](images/bi_ed-01.png){fig-align="center" width="429"}

![](images/bi_housing-01.png){fig-align="center" width="429"}

![](images/bi_unemployed.png){fig-align="center" width="429"}

![](images/bi_pov-01.png){fig-align="center" width="429"}

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

## Full analysis

![](images/corr_mat.png){fig-align="center" width="429"}

![](images/scree_plot.png){fig-align="center" width="429"}

![](images/fitted_map-03.png){fig-align="center"}

*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here.*

{{< pagebreak >}}

# Discussion

## Summary and Interpretation

*Summarize what you did, what you found and what it means.*

## Strengths and Limitations

*Discuss what you perceive as strengths and limitations of your analysis.*

# Conclusions

*What are the main take-home messages?*

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

{{< pagebreak >}}

\## Software

```{r}
sessionInfo()
```

{{< pagebreak >}}

# References

### 2023-04-25 code/analysis_code/1_EDA_alt.Rmd
modified outcome variable to MRR

```{r message=FALSE, warning=FALSE}
library(here)
library(spdep)
library(GGally)
library(skimr)
library(ggfortify)
library(MASS)
library(ggcorrplot)
library("FactoMineR")
library(tidyverse)
library(broom)
library(knitr)
library(ggpubr)
```

https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state
List-of-neighboring-states-for-each-US-state

```{r}
race_sub <- readRDS(here("data/processed_data/race_df_alt.rds"))
us_nabs <- readRDS(here("data/processed_data/us_nabs.rds"))

#Subsetting to only contiguous states.
race_sub <- race_sub %>%
  filter(!grepl("Alaska", State),
         !grepl("Hawaii", State))

us_nabs <- us_nabs %>%
  filter(!grepl("AK", StateCode),
         !grepl("HI", StateCode), 
         !grepl("DC", StateCode))
```

Let's create a variable for mortality rate ratio of Black/ White
```{r}
race_sub <- race_sub %>%
  group_by(State) %>%
  mutate(MRR = `Death Rate (estimated)`[Race == "Black"] / `Death Rate (estimated)`[Race == "White"])
```

Let's start by looking at the overall HIV-related death rate by race in the United States.
```{r}
race_sub %>% 
  group_by(Race) %>% 
  summarize(tot_deaths = sum(`Death Cases`),
            tot_population = sum(`Population`),
            overall_rate = round(tot_deaths/ tot_population *1e5,1)) %>% 
  ungroup() %>% 
  mutate(percent_by_race = round(tot_deaths/sum(tot_deaths)*100, 1),
         ratio = round(overall_rate/overall_rate[Race =="White"],1)) %>%
  kable(col.names = c("Race/ Ethnicity",
                      "HIV-Related Deaths (n)",
                      "Population (>12 years old)",
                      "Death Rate (per 100,000)",
                      "Percent",
                      "Ratio"))
# using white persons as reference group
```

We can seen that Black persons accounts for the 43.5% of HIV-related deaths in United States, and the rate of HIV-related deaths among this population was 7.2 times high than the rate of HIV-related deaths among White persons.

Next, let's take a look at our response variable
Univariate Analysis

Let's take a look at some box plots of death rate by races and state. 
We'll start by looking at death rate given by the AIDSVu dataset. 
```{r}
race_sub %>% 
  ggplot() +
  geom_boxplot(aes(y = `Death Rate (estimated)`, x = Race, group = Race)) + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio, United States, 2020", y = "Ratio", x = "Race/ Ethnicity") + theme(legend.position = "none") 
```

Histogram
```{r}
race_sub %>% 
  filter(Race == "Black") %>% 
  ggplot() +
  geom_histogram(aes(x = `MRR`), bins = 10) + 
  theme(legend.position = "none") +
  labs(title = "Histogram of Black-to-White HIV-Related Mortality Rate Ratio, United States, 2020", y = "n", x = "Ratio") + theme_classic()
ggsave(here("results/MRR_hist.png"), units="in", width=10, height=7, dpi=300)

race_sub %>% 
   filter(Race == "Black") %>% 
  pull(`MRR`) %>% 
  shapiro.test()

race_sub %>% 
  ggplot() +
  geom_boxplot(aes(y = `Death Rate (estimated)`, x = Race, group = Race)) + 
  theme(legend.position = "none") +
  labs(title = "HIV-Related Death Rate, by Race, United States, 2020", y = "Rate (per 100,000)", x = "Race") + theme_classic()
ggsave(here("results/dr_race_bp.png"), units="in", width=10, height=5, dpi=300)

race_sub %>% 
  ggplot() +
  geom_bar(aes(y = `Death Rate (estimated)`, x = `State Abbreviation`, group = `State Abbreviation`, fill = Race),
           stat = "identity", position = "dodge") + theme_classic() + 
  theme(legend.position = "bottom") +
  labs(title = "HIV-Related Death Rate of Black and White Persons, by State, United States, 2020", y = "Rate (per 100,000)", x = "State")
ggsave(here("results/dr_state.png"), units="in", width=12, height=5, dpi=300)

```

Let's move on to
Bivariate Analysis
```{r}
ggplot(race_sub, aes(y = MRR, x = `Percent Living in Poverty`)) +
  geom_smooth(method = "lm") + 
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Percent Living in Poverty") + 
  stat_cor(method = "pearson") 
#  pass
ggsave(here("results/bi_pov.png"), units="in", width=10, height=7, dpi=300)
```

```{r}
ggplot(race_sub, aes(y = MRR, x = `Percent Unemployed`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Percent Unemployed") + 
  stat_cor(method = "pearson") 
ggsave(here("results/bi_unemployed.png"), units="in", width=10, height=7, dpi=300)
#  did not pass 
```

```{r}
ggplot(race_sub, aes(y = MRR, x = `Gini Coefficient`)) +
    geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Gini Coefficient") +
  stat_cor(method = "pearson") 
ggsave(here("results/bi_gini.png"), units="in", width=10, height=7, dpi=300)
# did not "pass"
```

```{r}
ggplot(race_sub, aes(y = MRR, x = `Percent Less than High School Education`)) +
  geom_smooth(method = "lm") + 
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Percent Less than High School Education") +
  stat_cor(method="pearson")
#  did not pass
```

```{r}
ggplot(race_sub, aes(y = MRR, x = `Percent Living with Severe Housing Cost Burden`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Percent Living with Severe Housing Cost Burden") + 
  stat_cor(method="pearson")
#  did not pass
ggsave(here("results/bi_housing.png"), units="in", width=10, height=7, dpi=300)
```

```{r}
ggplot(race_sub, aes(y = MRR, x = `Percent Without an Internet subscription`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Percent Without an Internet Subscription") + 
  stat_cor(method="pearson")
# "pass"
```

```{r}
ggplot(race_sub, aes(y = MRR, x = `Percent Uninsured Under 65 Years Old`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Percent Uninsured Under 65 Years Old") + 
  stat_cor(method="pearson")
# "pass"
ggsave(here("results/bi_uninsured.png"), units="in", width=10, height=7, dpi=300)
```

```{r}
ggplot(race_sub, aes(y = MRR, x = `Percent Living with Food Insecurity`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Percent Living with Food Insecurity") + 
  stat_cor(method="pearson")
# pass"
ggsave(here("results/bi_foodinsecure.png"), units="in", width=10, height=7, dpi=300)
```

```{r}
ggplot(race_sub, aes(y = MRR, x = `Median Household Income`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Median Household Income") + 
  stat_cor(method="pearson")
# "pass"
```

```{r}
ggplot(race_sub, aes(y = MRR, x = `Percentage Diagnosed Late`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Percentage Diagnosed Late") + 
  stat_cor(method="pearson")
# "pass"
ggsave(here("results/bi_latedx.png"), units="in", width=10, height=7, dpi=300)
```

```{r}
ggplot(race_sub, aes(y = MRR, x = `Percentage Virally Suppressed`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Percentage Virally Suppressed") + 
  stat_cor(method="pearson")
#did not pass
```
Now that we have an idea of the relationship between our outcome and potential predictors, next I need to account for the fact that these variables a likely have some degree of multicollinearity. In the next step we will be using PCA to scale our predictors.

```{r}
pos_predictors <- race_sub %>% 
  select(c("State", 
           "Race",
           "Percent Living in Poverty",
           "Percent Without an Internet subscription",
           "Percent Uninsured Under 65 Years Old",
           "Percent Living with Food Insecurity",
           "Percent Living with Severe Housing Cost Burden",
           "Median Household Income",
           "Percentage Diagnosed Late"))
```

```{r}
saveRDS(pos_predictors, here("data/processed_data/pos_predictors.rds"))
saveRDS(race_sub, here("data/processed_data/race_alt.rds"))
```

### 2023-04-25 code/analysis_code/2_pca_alt.Rmd
modified outcome variable to MRR

```{r message=FALSE, warning=FALSE}
library(ggcorrplot)
library(FactoMineR)
library(factoextra)
library(here)
library(knitr)
library(tidyverse)
library(broom)

race_sub <- readRDS(here("data/processed_data/race_alt.rds"))
pos_predictors <- readRDS(here("data/processed_data/pos_predictors.rds"))

```

PCA
```{r}
race_sub <- race_sub %>%
    select_if(~ !any(is.na(.))) %>% 
  #filter(Race == "Black") %>% 
  ungroup()

  #separating the y variables
df_response <- race_sub %>%
    select(contains(c("State", "Ratio", "MRR", "Race", "Death", "Population")))

  # pulling variables that were correlated with MRR, and thenp scaling
pos_predictors_scaled <- pos_predictors %>% 
    ungroup() %>% 
    mutate(across(where(is.numeric), scale))

#Setting up correlation matrix
corr_matrix <- pos_predictors_scaled %>% 
    select(c(3:9)) %>% 
    cor()

  #setting up correlation matrix 
print(ggcorrplot(corr_matrix))
ggsave(here("results/corr_mat.png"), units="in", width=10, height=7, dpi=300)
```

```{r}
#applying pca
pca <- princomp(corr_matrix)
print(summary(pca))

#pca loadings of principal components 1 and 2
print(pca$loadings[, 1:2])

#Scree plot
print(fviz_eig(pca, addlabels = TRUE))
ggsave(here("results/scree_plot.png"), units="in", width=10, height=7, dpi=300)
```
Since we originally had 11 variables, we should only look at components that explain more than 9.1% of variation. We can see from the scree plot that principal component 1 explained 73.3% of variation, and components 1 and 2 explained 94.9% of the variation, and components 1 - 2 explained 91.1%. All other principal components explains less that 8.9% of variation, so they will excluded.

```{r}
#Contribution of each variable 
print(fviz_cos2(pca, choice = "var", axes = 1:2))
ggsave(here("results/contribution_plot.png"), units="in", width=10, height=7, dpi=300)

# Biplot of the attributes
print(fviz_pca_var(pca, col.var = "cos2",
                   gradient.cols = c("darkblue", "orange", "green"),
                   repel = TRUE, title = "PCA"
                   ))
ggsave(here("results/biplot.png"), units="in", width=10, height=7, dpi=300)
# Variables in green are a good representation of the variable on that component.

#Creating object to store scaled dataframes
df_scaled <- full_join(df_response, pos_predictors_scaled, by = join_by("State", "Race"))
```

```{r}
saveRDS(df_scaled, here("data/processed_data/df_scaled_alt.rds"))
```

### 2023-04-25 code/analysis_code/3_model_alt.Rmd
modified outcome variable to MRR

```{r message=FALSE, warning=FALSE}
library(here)
library(MASS)
library(spatialreg)
library(spdep)
library(tidyverse)
```

```{r}
df <- readRDS(here("data/processed_data/df_scaled_alt.rds"))
us_nabs <- readRDS(here("data/processed_data/us_nabs.rds"))

#Subset to contiguous states
us_nabs <- us_nabs %>%
  filter(!grepl("AK", StateCode),
         !grepl("HI", StateCode), 
         !grepl("DC", StateCode))

```

```{r}
states <- data.frame(state.name)
states$region <- case_when(
  states$state.name %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania") ~ "Northeast",
  states$state.name %in% c("Illinois", "Indiana", "Iowa", "Kansas", "Michigan", "Minnesota", "Missouri", "Nebraska", "North Dakota", "Ohio", "South Dakota", "Wisconsin") ~ "Midwest",
  states$state.name %in% c("Alabama", "Arkansas", "Delaware", "Florida", "Georgia", "Kentucky", "Louisiana", "Maryland", "Mississippi", "North Carolina", "Oklahoma", "South Carolina", "Tennessee", "Texas", "Virginia", "West Virginia") ~ "South",
  states$state.name %in% c("Arizona", "California", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Oregon", "Utah", "Washington", "Wyoming") ~ "West"
)

df <- full_join(df, states, by = join_by(State == state.name))

df <- df %>% 
  filter(!grepl("Alaska", State),
         !grepl("Hawaii", State))
```

Setting up functions for models
```{r}
 #multiple linear regression
mult_function <- function(df) {
  mult_race <- glm(MRR ~ 
                        `Percent Living in Poverty` +
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                        `Median Household Income`  +
                     #   `Percent Living with Severe Housing Cost Burden` +
                        `Percent Living with Food Insecurity`, 
                     data = df,
                     family = "gaussian")
  print(summary(mult_race))
    ##pull fitted values
df$mult_fit <- mult_race$fitted.values
return(df)
}

## function to fit GLM and pull AIC
mult_aic <- function(df) {
  mult_race <- glm(MRR ~ 
                        `Percent Living in Poverty` +
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                        `Median Household Income`  +
                       # `Percent Living with Severe Housing Cost Burden` +
                        `Percent Living with Food Insecurity`,
                     data = df,
                     family = "gaussian")
aic <- AIC(mult_race)

return(aic)
}
```

```{r}
df <- mult_function(df)

df_mult <- mult_aic(df)
```

```{r}
  #poisson
pois_function <- function(df) {

pois_race <- glm(`Death Cases` ~ Race +
                        `Percent Living in Poverty` +
                       # `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                        `Median Household Income`  +
                        `Percent Living with Food Insecurity` +
                        #`Percent Living with Severe Housing Cost Burden` +
                     offset(log(`Population`)),
                     data = df,
                     family = poisson(link = "log"))
  print(summary(pois_race))

      ##pull fitted values
df$pois_fit <- pois_race$fitted.values

return(df)

}

## function to fit GLM and pull AIC
pois_aic <- function(df) {
  pois_race <- glm(`Death Cases` ~  Race +
                        `Percent Living in Poverty` +
                      #  `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                        `Median Household Income`  +
                        `Percent Living with Food Insecurity` +
                       # `Percent Living with Severe Housing Cost Burden` +
                     offset(log(`Population`)),
                     data = df,
                     family = poisson(link = "log"))
aic <- AIC(pois_race)
return(aic)
}
```

```{r}
df <- pois_function(df)

df_pois <- pois_aic(df)
```

```{r}
  #Negative binomial
nb_function  <- function(df) {
  neg_bin_race <- glm.nb(`Death Cases` ~ Race +
                        `Percent Living in Poverty` +
                       # `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                        `Median Household Income`  +
                        `Percent Living with Food Insecurity`  +
                       # `Percent Living with Severe Housing Cost Burden` +
                     offset(log(`Population`)),
                            data = df)
  print(summary(neg_bin_race))

      ##pull fitted values
df$nb_fit <- neg_bin_race$fitted.values

return(df)

}

## function to fit GLM and pull AIC
nb_aic <- function(df) {
  neg_bin_race <- glm.nb(`Death Cases` ~ Race +
                        `Percent Living in Poverty` +
                       # `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                        `Median Household Income`  +
                        `Percent Living with Food Insecurity`  +
                        #`Percent Living with Severe Housing Cost Burden` +
                     offset(log(`Population`)),
                            data = df)
aic <- AIC(neg_bin_race)

return(aic)

}  
```

```{r}
df <- nb_function(df)

df_nb <- nb_aic(df)
```

```{r}
#Fitting CAR model (requires neighbors)
car_function <- function(df) {   
  h_race_df <- matrix(0, nrow(df), nrow(df))
  h_race_df[cbind(match(us_nabs$StateCode, df$`State Abbreviation`),
                  match(us_nabs$NeighborStateCode, df$`State Abbreviation`))] <- 1

  nablist_race_df <- mat2listw(h_race_df, style = "W")

  car_race <- spautolm(MRR ~                      
                        `Percent Living in Poverty` +
                       `Median Household Income` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                     #   `Percent Unemployed`  +
                     #   `Percent Living with Severe Housing Cost Burden` +
                        `Percent Living with Food Insecurity`, 
                           data = df,
                           listw = nablist_race_df,
                           family = "CAR")
  print(summary(car_race))

  ##pull fitted values
df$car_fit <- car_race$fit$fitted.values

return(df)

}

car_aic <- function(df) {
  h_race_df <- matrix(0, nrow(df), nrow(df))
  h_race_df[cbind(match(us_nabs$StateCode, df$`State Abbreviation`),
                  match(us_nabs$NeighborStateCode, df$`State Abbreviation`))] <- 1

  nablist_race_df <- mat2listw(h_race_df, style = "W")

  car_race <- spautolm(MRR ~                      
                        `Percent Living in Poverty` +
                       `Median Household Income` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                   #     `Percent Unemployed`  +
                      #  `Percent Living with Severe Housing Cost Burden` +
                        `Percent Living with Food Insecurity`, 
                           data = df,
                           listw = nablist_race_df,
                           family = "CAR")
aic <- AIC(car_race)
return(aic)

} 
```

```{r}
df <- car_function(df)

df_car <- car_aic(df)
```

Organizing mod_AIC into new dataframe
```{r}
aic_df <- data.frame(AIC = c(df_mult, df_pois, df_nb, df_car),
                     Model = c("GLM", "Poisson", "Negative Binomial", "CAR"),
                     Outcome = c("Ratio", "Count", "Count", "Ratio"))
aic_df <- aic_df %>% mutate(AIC  = round(AIC, 2))
```

```{r}
saveRDS(df, here("data/processed_data/df_fit_alt.rds"))

saveRDS(aic_df, here("data/processed_data/aic_alt.rds"))
```

### 2023-04-25 code/analysis_code/archieved/Race/p1_EDA_race.Rmd
modified outcome variable to MRR

--- 2023-04-21 code/analysis_code/Race/p1_EDA_race.Rmd
@@
 # using white persons as reference group
+
 ```
@@
 ```
-The plots differ drastically for persons of multiple race. It is difficult to tell what might have caused this drastic change, but my inituition tells me there is something strange going on.
+The plots differ drastically for persons of multiple race. It is difficult to tell what might have caused this drastic change, but my intuition tells me there is something strange going on.
 
@@
 
-ggplot(df, aes(y = outcome_DR, x = `Percent High School Education`)) +
+ggplot(df, aes(y = outcome_DR, x = `Percent Less than High School Education`)) +
   geom_smooth(method = "lm") + 
   geom_point() + 
-  labs(title = "HIV Death Rate vs. Percent High School Education") +
+  labs(title = "HIV Death Rate vs. Percent Less than High School Education") +
   facet_wrap(Race ~ . , scales = "free_y") +
@@
 
-ggplot(df, aes(y = outcome_DR, x = `Broadband of any type`)) +
+ggplot(df, aes(y = outcome_DR, x = `Percent Without an Internet subscription`)) +
   geom_smooth(method = "lm") +
   geom_point() + 
-  labs(title = "HIV Death Rate vs. Broadband of any type") +
+  labs(title = "HIV Death Rate vs. Percent Without an Internet Subscription") +
   facet_wrap(Race ~ . , scales = "free_y") + 
@@
   labs(title = "HIV Death Rate vs. Percent Living with Food Insecurity") +
-  facet_wrap(Race ~ . , scales = "free_y") + 
-  stat_cor(method="pearson")
-# did not "pass"
-
-ggplot(df, aes(y = outcome_DR, x = `Broadband of any type`)) +
-  geom_smooth(method = "lm") +
-  geom_point() + 
-  labs(title = "HIV Death Rate vs. Broadband of any type") +
   facet_wrap(Race ~ . , scales = "free_y") + 

### 2023-04-25 code/analysis_code/archieved/Race/p2_PCA_race.Rmd
modified outcome variable to MRR

--- 2023-04-21 code/analysis_code/Race/p2_PCA_race.Rmd
@@
 df_predictor <- df %>% 
-    select(contains(c("State", "Race", "Percent", "Gini", "Median", "Broadband"))) %>% 
-    #select(!contains(c("Virally", "Received", "Late"))) %>% 
+    select(contains(c("State", "Race", "Percent", "Gini", "Median"))) %>% 
+    select(!contains(c("Virally", "Received", "Late", "Percent High"))) %>% 
     mutate(across(where(is.numeric), scale))
@@
 Since we originally had 11 variables, we should only look at components that explain more than 9.1% of variation.
-We can see from the scree plot that principal component 1 explained 63.8% of variation, and components 1 and 2 explained 84.7% of the variation, and components 1 - 3 explained 94.8%.
-Component 3 just  barely passed the threshold with 10.1%.
-All other principal components explains less that 6.2% of variation, so they will excluded.
+We can see from the scree plot that principal component 1 explained 72.9% of variation, and components 1 and 2 explained 94.9% of the variation, and components 1 - 3 explained 94.8%.
+All other principal components explains less that 5.1% of variation, so they will excluded.
 

### 2023-04-25 code/analysis_code/archieved/Race/p3_model_race.Rmd
modified outcome variable to MRR

--- 2023-04-21 code/analysis_code/Race/p3_model_race.Rmd
@@
 
+```{r}
+states <- data.frame(state.name)
+states$region <- case_when(
+  states$state.name %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania") ~ "Northeast",
+  states$state.name %in% c("Illinois", "Indiana", "Iowa", "Kansas", "Michigan", "Minnesota", "Missouri", "Nebraska", "North Dakota", "Ohio", "South Dakota", "Wisconsin") ~ "Midwest",
+  states$state.name %in% c("Alabama", "Arkansas", "Delaware", "Florida", "Georgia", "Kentucky", "Louisiana", "Maryland", "Mississippi", "North Carolina", "Oklahoma", "South Carolina", "Tennessee", "Texas", "Virginia", "West Virginia") ~ "South",
+  states$state.name %in% c("Arizona", "California", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Oregon", "Utah", "Washington", "Wyoming") ~ "West"
+)
+
+df <- full_join(df, states, by = join_by(State == state.name))
+```
+
 subset by race/ethnicity
@@
 Setting up functions for models
-
 ```{r}
@@
                         `Percent Unemployed` +
-                        `Percent Living in Poverty` +
-                        `Percent High School Education` +  #estimate education access
-                 #       `Broadband of any type` + # estimate built 
-                   #     `Percent Uninsured Under 65 Years Old`+ #estimate access to healthcare
+                        `Percent Less than High School Education` +  #estimate education access
+                        `Percent HIV Prevalence` +
                         `Percent Living with Severe Housing Cost Burden`, #economic stability,
@@
                         `Percent Unemployed` +
-                     `Percent Living in Poverty` +
-                        `Percent High School Education` +  #estimate education access
-                    #     `Broadband of any type` + # estimate built 
-                     #   `Percent Uninsured Under 65 Years Old`+ #estimate access to healthcare
+                        `Percent Less than High School Education` +  #estimate education access
+                        `Percent HIV Prevalence` +
                         `Percent Living with Severe Housing Cost Burden`, #economic stability,
@@
                         `Percent Unemployed` +
-                     `Percent Living in Poverty` +
-                        `Percent High School Education` +  #estimate education access
-                      #   `Broadband of any type` + # estimate built 
-                     #   `Percent Uninsured Under 65 Years Old` + #estimate access to healthcare
+                        `Percent Less than High School Education` +  #estimate education access
+                        `Percent HIV Prevalence` +
                         `Percent Living with Severe Housing Cost Burden` + #economic stability
@@
                          `Percent Unemployed` +
-                     `Percent Living in Poverty` +
-                        `Percent High School Education` +  #estimate education access
-                     #    `Broadband of any type` + # estimate built 
-                    #    `Percent Uninsured Under 65 Years Old` + #estimate access to healthcare
+                         `Percent Less than High School Education` +  #estimate education access
+                        `Percent HIV Prevalence` +
                         `Percent Living with Severe Housing Cost Burden` + #economic stability
@@
                          `Percent Unemployed` +
-                          `Percent Living in Poverty` +
-                        `Percent High School Education` +  #estimate education access
-                     #   `Broadband of any type` + # estimate built 
-                      #  `Percent Uninsured Under 65 Years Old` + #estimate access to healthcare
+                        `Percent Less than High School Education` +  #estimate education access
+                        `Percent HIV Prevalence` +
                         `Percent Living with Severe Housing Cost Burden` + #economic stability
@@
                         `Percent Unemployed` +
-                          `Percent Living in Poverty` +
-                        `Percent High School Education` +  #estimate education access
-                      #   `Broadband of any type` + # estimate built 
-                   #    `Percent Uninsured Under 65 Years Old` + #estimate access to healthcare
+                        `Percent Less than High School Education` +  #estimate education access
+                        `Percent HIV Prevalence` +
                         `Percent Living with Severe Housing Cost Burden` + #economic stability
@@
                         `Percent Unemployed` +
-                         `Percent Living in Poverty` +
-                        `Percent High School Education` +  #estimate education access
-                       #  `Broadband of any type` + # estimate built 
-                     #   `Percent Uninsured Under 65 Years Old`+ #estimate access to healthcare
+                        `Percent Less than High School Education` +  #estimate education access
+                        `Percent HIV Prevalence` +
                         `Percent Living with Severe Housing Cost Burden`,#economic stability,
@@
                         `Percent Unemployed` +
-                         `Percent Living in Poverty` +
-                        `Percent High School Education` +  #estimate education access
-                     #    `Broadband of any type` + # estimate built 
-                    #    `Percent Uninsured Under 65 Years Old`+ #estimate access to healthcare
+                        `Percent Less than High School Education` +  #estimate education access
+                        `Percent HIV Prevalence` +
                         `Percent Living with Severe Housing Cost Burden`,#economic stability,
@@
 png(here("results/moran_black.png"))
-#sink(here("results/spacy_summary_black.txt"))
+sink(here("results/spacy_summary_black.txt"))
 spacy_function(black_df)
@@
 png(here("results/moran_hispanic.png"))
-#sink(here("results/spacy_summary_hispanic.txt"))
+sink(here("results/spacy_summary_hispanic.txt"))
 spacy_function(hispanic_df)
-sink()
-dev.off()
-```
-
-```{r}
-spacy_function(white_df)
-```
-
-```{r}
-png(here("results/moran_white.png"))
-#sink(here("results/spacy_summary_white.txt"))
-spacy_function(white_df)
 sink()

### 2023-04-25 code/analysis_code/archieved/Race/p4_mapping.Rmd
modified outcome variable to MRR

--- 2023-04-21 code/analysis_code/Race/p4_mapping.Rmd
@@
 ggplot(df_sf_raw, aes(long, lat)) +
-  geom_polygon(aes(group = group, fill =`Percent High School Education`)) +
+  geom_polygon(aes(group = group, fill =`Percent Less than High School Education` )) +
   coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
@@
         legend.position = "bottom") +
-  labs(title = "Percent High School Education, 2020", x = "", y = "", fill = "%") +
+  labs(title = "Percent of Adults with Less than High School Education, 2020", x = "", y = "", fill = "%") +
   theme(plot.title= element_text(hjust = 0.5)) 
@@
 ggsave(here("results/housing_map.png"), units="in", width=10, height=7, dpi=300)
-
-ggplot(df_sf_raw, aes(long, lat)) +
-  geom_polygon(aes(group = group, fill =`Percent Living in Poverty` )) +
-  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
-  theme(strip.text = element_text(size = 14),
-        legend.position = "bottom") +
-  labs(title = "Percent Living in Poverty, 2020", x = "", y = "", fill = "%") +
-  theme(plot.title= element_text(hjust = 0.5)) 
-ggsave(here("results/poverty_map.png"), units="in", width=10, height=7, dpi=300)
 
@@
 Non-Hispanic White persons were the only of the selected racial/ ethnic group that did not have a HIV-related death rate higher than ~10 per 100,000.
-
-Let's map the observed/ expected ratio.
-```{r}
-ggplot(df_sf_scale, aes(long, lat)) +
-  geom_polygon(aes(group = group, fill = OE_mortality_ratio)) +
-  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
-  facet_wrap(. ~ Race) +
-  scale_fill_distiller(palette = "RdYlGn") + theme_void() + 
-  theme(strip.text = element_text(size = 14),
-    strip.background = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        axis.text = element_blank(),
-        legend.position = "bottom") +
-  labs(title = "Observed/ Expected HIV Mortality Ratio by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Ratio") +
-  theme(plot.title= element_text(hjust = 0.5))
-
-ggsave(here("results/roe_map.png"), units="in", width=10, height=4, dpi=300)
-```
-We can see that in much of the United States non-Hispanic Black persons had ratios consistently (much) greater than 1, suggesting a higher risk among this population.
 
@@
 ggsave(here("results/fitted_map.png"), units="in", width=10, height=4, dpi=300)
-
 ```

### 2023-04-25 code/analysis_code/p4_mapping.Rmd
modified outcome variable to MRR

--- 2023-04-25 code/analysis_code/archieved/Race/p4_mapping.Rmd
@@
 
-df_scaled <- readRDS(here("data/processed_data/df_fit.rds"))
-
-df_raw <- readRDS(here("data/processed_data/race_new.rds"))
-
-mod_aic <-  readRDS(here("data/processed_data/aic.rds"))
+df_scaled <- readRDS(here("data/processed_data/df_fit_alt.rds"))
+df_raw <- readRDS(here("data/processed_data/race_alt.rds"))
+mod_aic <-  readRDS(here("data/processed_data/aic_alt.rds"))
 ```
@@
 
-#Best model  
-best_mod_aic <- mod_aic %>%
-  group_by(Race) %>%
-  drop_na() %>% 
-  filter(AIC == min(AIC))
-
-df_scaled <- full_join(df_scaled, best_mod_aic, by = join_by("Race"))
-
-##Pulling fitted values based on best fitting model
-df_scaled <- df_scaled %>% 
-  mutate(best_fitted = case_when(Model == "CAR" ~ car_fit,
-                                 Model == "GLM" ~ mult_fit,
-                                 Model ==  "Poisson" ~ pois_fit,
-                                 Model == "Negative Binomial" ~ nb_fit))
-
 df_raw <- df_raw %>%
@@
 ```{r}
+ggplot(df_sf_scale, aes(long, lat)) +
+  geom_polygon(aes(group = group, fill =`MRR`)) +
+  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
+  theme(strip.text = element_text(size = 14),
+        legend.position = "bottom") +
+  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio in the United States, 2020", x = "", y = "", fill = "Death rate per 100,000") +
+  scale_fill_distiller(palette = "RdYlGn") +
+  theme(plot.title= element_text(hjust = 0.5))
+
+ggsave(here("results/mrr_map.png"), units="in", width=10, height=4, dpi=300)
+```
+
+```{r}
 ggplot(df_sf_raw, aes(long, lat)) +
@@
 ggsave(here("results/gini_map.png"), units="in", width=10, height=7, dpi=300)
+```
 
+```{r}
 ggplot(df_sf_raw, aes(long, lat)) +
-  geom_polygon(aes(group = group, fill =`Percent Less than High School Education` )) +
+  geom_polygon(aes(group = group, fill =`Percent Living in Poverty`)) +
   coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
@@
         legend.position = "bottom") +
-  labs(title = "Percent of Adults with Less than High School Education, 2020", x = "", y = "", fill = "%") +
+  labs(title = "Percent Living in Poverty, 2020", x = "", y = "", fill = "%") +
   theme(plot.title= element_text(hjust = 0.5)) 
-ggsave(here("results/ed_map.png"), units="in", width=10, height=7, dpi=300)
+ggsave(here("results/poverty_map.png"), units="in", width=10, height=7, dpi=300)
+```
 
+```{r}
+ggplot(df_sf_raw, aes(long, lat)) +
+  geom_polygon(aes(group = group, fill =`Percent Uninsured Under 65 Years Old`)) +
+  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
+  theme(strip.text = element_text(size = 14),
+        legend.position = "bottom") +
+  labs(title = "Percent Uninsured Under 65 Years Old, 2020", x = "", y = "", fill = "%") +
+  theme(plot.title= element_text(hjust = 0.5)) 
+ggsave(here("results/uninsured_map.png"), units="in", width=10, height=7, dpi=300)
+```
+
+```{r}
 ggplot(df_sf_raw, aes(long, lat)) +
@@
         legend.position = "bottom") +
-  labs(title = "Percent Living with Severe Housing Cost Burden, 2020", x = "", y = "", fill = "%") +
+  labs(title = "Percent Living with Severe Housing Cost Burden 2020", x = "", y = "", fill = "%") +
   theme(plot.title= element_text(hjust = 0.5)) 
 ggsave(here("results/housing_map.png"), units="in", width=10, height=7, dpi=300)
+```
 
+```{r}
+ggplot(df_sf_raw, aes(long, lat)) +
+  geom_polygon(aes(group = group, fill =`Percent Living with Food Insecurity`)) +
+  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
+  theme(strip.text = element_text(size = 14),
+        legend.position = "bottom") +
+  labs(title = "Percent Living with Food Insecurity, 2020", x = "", y = "", fill = "%") +
+  theme(plot.title= element_text(hjust = 0.5)) 
+ggsave(here("results/foodinsecure_map.png"), units="in", width=10, height=7, dpi=300)
+```
+
+```{r}
+ggplot(df_sf_raw, aes(long, lat)) +
+  geom_polygon(aes(group = group, fill =`Percent Without an Internet subscription`)) +
+  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
+  theme(strip.text = element_text(size = 14),
+        legend.position = "bottom") +
+  labs(title = "Percent Without an Internet Subscription, 2020", x = "", y = "", fill = "%") +
+  theme(plot.title= element_text(hjust = 0.5)) 
+ggsave(here("results/no_internet_map.png"), units="in", width=10, height=7, dpi=300)
+```
+
+```{r}
 ggplot(df_sf_raw, aes(long, lat)) +
@@
 Non-Hispanic White persons were the only of the selected racial/ ethnic group that did not have a HIV-related death rate higher than ~10 per 100,000.
-
-Next, let's map after the the Box_Cox transformation of the death rate
-```{r}
-ggplot(df_sf_scale, aes(long, lat)) +
-  geom_polygon(aes(group = group, fill = BOX)) +
-  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
-  facet_wrap(. ~ Race) +
-  scale_fill_distiller(palette = "RdYlGn") + theme_void() + 
-  theme(strip.text = element_text(size = 14),
-    strip.background = element_blank(),
-        axis.ticks = element_blank(),
-        axis.line = element_blank(),
-        axis.text = element_blank(),
-        legend.position = "bottom") +
-  labs(title = "Death Rates (Box Cox transformed) by Race/ Ethnicity
-in the United States, 2020", x = "", y = "", fill = "Ratio") +
-  theme(plot.title= element_text(hjust = 0.5))
-ggsave(here("results/roe_box_map.png"), units="in", width=10, height=4, dpi=300)
-```
+`
 
@@
 ```
-Well..
-I probably need to go back and reconsider using making a lambda based on specific race/ ethnic.
-I think that at this point, however, it is still fair to consider mapping an observed/ expected ratio using the expected deaths calculate from the lambda for the entire population.
-This will allow us to see differences across the race/ ethnicity, rather than masking disparities in rates due higher expected cases among certain high-risk populations, which would thus make the ratio closer to the null.
 
@@
 
-Next, let's make a map using the fitted values from the CAR model, the best performing model.
 ```{r}
@@
 ggplot(df_sf_scale, aes(long, lat)) +
-  geom_polygon(aes(group = group, fill = best_fitted)) +
+  geom_polygon(aes(group = group, fill = car_fit)) +
   scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
   coord_map("albers",  lat0 = 45.5, lat1 = 29.5)  + 
-  facet_wrap(. ~ Race)  + 
   theme_void() + 

### 2023-04-25 code/processing_code/1_cleaning_2020.Rmd
modified outcome variable to MRR

@@
 v_sup  <- read_xlsx(here("data/raw_data/AIDSVu_State_ViralSuppression_2020.xlsx"), skip = 3)
-internet <- read_xlsx(here("data/raw_data/ACSST5Y2020.S2801-2023-04-17T232857.xlsx"), skip = 3, sheet = 2)
+internet <- read_xlsx(here("data/raw_data/ACSST5Y2020.S2801-2023-04-17T232857.xlsx"), skip = 2, sheet = 2)
 ```
@@
   filter(!grepl("DC", `State Abbreviation`)) %>% 
-   filter(!grepl("PR", `State Abbreviation`))
+   filter(!grepl("PR", `State Abbreviation`)) %>% 
+  mutate("Percent Less than High School Education" = 100 - `Percent High School Education`)
 
@@
 ```{r}
-internet <- internet %>% 
-  select(c(1, "Broadband of any type")) %>% 
-  slice(c(-1:-7, -48:-52,-263:-267)) %>% 
-  filter(!grepl(",", `Broadband of any type`))
+no_internet <- internet %>% 
+  select(c(1, "Without an Internet subscription")) %>% 
+  slice(c(-1:-8, -48:-52,-263:-267)) %>% 
+  filter(!grepl(",", `Without an Internet subscription`)) %>% 
+  rename(`Percent Without an Internet subscription` = `Without an Internet subscription`)
 ```
@@
 
-# print the resulting vector
-states
+no_internet$State<- states
 
-internet$State<- states
-
-internet <- internet %>% 
-  select(State, `Broadband of any type`) %>% 
+no_internet <- no_internet %>% 
+  select(State, `Percent Without an Internet subscription`) %>% 
   drop_na()
 
-internet$`Broadband of any type` <- gsub("%", "", internet$`Broadband of any type`)
+no_internet$`Percent Without an Internet subscription` <- gsub("%", "", no_internet$`Percent Without an Internet subscription`)
 
-internet <- internet %>% mutate(`Broadband of any type` = as.numeric(`Broadband of any type`))
+no_internet <- no_internet %>%
+  mutate(`Percent Without an Internet subscription` = as.numeric(`Percent Without an Internet subscription`))
 ```
@@
 ```{r}
-race_df <- full_join(race_df, internet)
-sex_df <- full_join(sex_df, internet)
+race_df <- full_join(race_df, no_internet)
+sex_df <- full_join(sex_df, no_internet)
 ```

### 2023-04-25 code/processing_code/2b_cleaning_race.Rmd
modified outcome variable to MRR

```{r, echo=FALSE, message=FALSE}
library(knitr)
library(tidyverse)
library(here)
```

```{r}
race_df <- readRDS(here("data/processed_data/race_df.rds"))
race_df <-  race_df %>%  replace(is.na("Black Death Cases"), 0) %>% 
   replace(is.na("White Death Cases"), 0)

```
Let's find the HIV death rate by race
```{r}
race_df <- race_df %>% 
  mutate(black_death_rate = round(`Black Death Cases`/Black_pop*1e5,1),
         white_death_rate = round(`White Death Cases`/White_pop*1e5,1))
```

Next, let's find the expected number of HIV deaths. First, we'll calculate lambda. 
```{r}
lambda_black <- race_df %>%
  drop_na(`Black Death Cases`) %>% 
  summarize(lambda_black = sum(`Black Death Cases`)/sum(Black_pop)) %>%
  as.numeric()

lambda_white <- race_df %>%
  drop_na(`White Death Cases`) %>% 
  summarize(lambda_white = sum(`White Death Cases`)/sum(White_pop)) %>%
  as.numeric()
```

NOTE: I realized after running some more analyses that calculating lambda for each race is masking disparities in HIV-related deaths between races. I believe it would more appropriate calculate a single value lambda for entire population.

```{r}
black_df <- race_df %>% select(1:3, contains("Black"), 50:66)

black_df <- black_df %>% 
  select(!contains("white")) %>% 
 rename("Death Rate (AIDSVu)" = `Black Death Rate`,
        "Population" = Black_pop,
        "Death Cases" = "Black Death Cases",
        "New Diagnoses Rate" = "New Diagnoses Black Rate",
        "New Diagnoses Cases" = "New Diagnoses Black Cases",
        "Percentage Diagnosed Late" = `Percent of Black Persons Diagnosed Late`,
        "Number of Persons Diagnosed Late" = `Number of Black Persons Diagnosed Late`,
        "Percentage Received Care" = `Percent of Black Persons Living with HIV who Received HIV Care`,
        "Percentage Virally Suppressed" = `Percent of Black Persons Living with HIV who were Virally Suppressed`,
        "HIV Prevalence Rate" = "Black Rate",
        "Prevalent HIV Cases" = "Black Cases",
        "Death Rate (estimated)" = black_death_rate) %>% 
  mutate(Race = "Black",
        "Expected Death Cases" = round(lambda_black * Population))

white_df <- race_df %>% select(1:3, contains("White"), 50:66)

white_df <- white_df %>% 
  select(!contains("black")) %>% 
 rename("Death Rate (AIDSVu)" = `White Death Rate`,
        "Population" = White_pop,
        "Death Cases" = "White Death Cases",
        "New Diagnoses Rate" = "New Diagnoses White Rate",
        "New Diagnoses Cases" = "New Diagnoses White Cases",
        "Percentage Diagnosed Late" = `Percent of White Persons Diagnosed Late`,
        "Number of Persons Diagnosed Late" = `Number of White Persons Diagnosed Late`,
        "Percentage Received Care" = `Percent of White Persons Living with HIV who Received HIV Care`,
        "Percentage Virally Suppressed" = `Percent of White Persons Living with HIV who were Virally Suppressed`,
        "HIV Prevalence Rate" = "White Rate",
        "Prevalent HIV Cases" = "White Cases",
        "Death Rate (estimated)" = white_death_rate) %>% 
  mutate(Race = "White",
        "Expected Death Cases" = round(lambda_white * Population))

race_df <- list(black_df, white_df)

race_df <- reduce(race_df, full_join)

race_df <- race_df %>% 
  mutate(`Percent HIV Prevalence` = round(`Prevalent HIV Cases`/ `Population`*100,2))
```

Now, let's calculate lambda - null, which would assume that there were no difference in Expected deaths by race.
```{r}
lambda <- race_df %>%
  drop_na("Death Cases") %>% 
  summarize(lambda = sum(`Death Cases`)/sum(Population)) %>%
  as.numeric()

race_df <- race_df %>% 
  mutate(`Expected Death Cases (Null)` = Population * lambda,
         `Ratio Observed/ Expected (Null)` = `Death Cases`/`Expected Death Cases (Null)`,
         `Ratio Observed/ Expected (Race)` = `Death Cases`/`Expected Death Cases`)
```

Save the file and now we can move on to some exploratory analysis!
```{r}
saveRDS(race_df, here("data/processed_data/race_df_alt.rds"))
```

### 2023-04-25 code/processing_code/archieved/2b_cleaning_race_df.Rmd
modified outcome variable to MRR

--- 2023-04-21 code/processing_code/2b_cleaning_race_df.Rmd
@@
 ```{r}
-black_df <- race_df %>% select(1:3, contains("Black"), 50:63)
+black_df <- race_df %>% select(1:3, contains("Black"), 50:64)
 
@@
 
-hispanic_df <- race_df %>% select(1:3, contains("Hispanic"), 50:63)
+hispanic_df <- race_df %>% select(1:3, contains("Hispanic"), 50:64)
 
@@
 
-white_df <- race_df %>% select(1:3, contains("White"), 50:63)
+white_df <- race_df %>% select(1:3, contains("White"), 50:64)
 
@@
 
-multiple_df <- race_df %>% select(1:3, contains("Multiple"), 50:63)
+multiple_df <- race_df %>% select(1:3, contains("Multiple"), 50:64)
 
@@
 race_df <- reduce(race_df, full_join)
+
+race_df <- race_df %>% 
+  mutate(`Percent HIV Prevalence` = round(`Prevalent HIV Cases`/ `Population`*100,2))
 ```

### 2023-04-25 outputs_code.Rmd
modified outcome variable to MRR

@@
   ungroup() %>% 
-  mutate(percent_by_race = round(tot_deaths/sum(tot_deaths)*100, 1),
-         ratio = round(overall_rate/overall_rate[Race =="White"],1)) %>%
+  mutate(ratio = round(overall_rate/overall_rate[Race =="White"],1)) %>%
   kable(col.names = c("Race/ Ethnicity",
@@
                       "Death Rate \n (per 100,000)",
-                      "Percent",
-                      "Ratio"),
-        caption = "HIV-Related Deaths Summary by Race/ Ethncity, United States, 2020",
+                      "Mortality Rate Ratio"),
+        caption = "HIV-Related Deaths Summary by Race/ Ethnicity, United States, 2020",
         align = "c") %>%
   kable_minimal() %>%
-  kable_styling(font_size = 50) %>% 
-  save_kable(here("results/test.png"))
+  kable_styling(font_size = 30)%>%
+gsub("font-size: initial !important;",
+"font-size: 30pt !important;",
+.)  %>% 
+  landscape() %>% 
+  save_kable(here("results/table1.png"))
 # using white persons as reference group

### 2023-04-25 products/manuscript/manuscript.qmd
modified outcome variable to MRR

@@
 HIV has a disproportionate impact on marginalized groups, resulting in significant disparities in rate of new HIV diagnoses and poor HIV-related health outcomes, such as late HIV diagnosis, progression to AIDS, and HIV-related death.
-Previous study have found direct links to indicators of social determinants of health and poorer health outcomes for persons with HIV
+In the United States, these disparities are most apparent when comparing HIV-related outcomes between Black and White persons.
+Previous study have found direct links to indicators of social determinants of health and poorer health outcomes for persons with HIV.
 
-Other studies have found an association between different SDOH measures and COVID-19 mortality that varied across racial and ethnic groups and community types
+Other studies have found an association between different SDOH measures and COVID-19 mortality that varied across racial and ethnic groups and community types.
+
+The goal of this project is to gain an understanding of factors associated the disparity of HIV-related death rates between Black and White persons in the United States.
 
@@
 
-Can the HIV death rate from 2020 in the United States be modeled by social determinants of health and/ or by HIV-related outcomes (late diagnoses, viral suppression) Are these factors correlated with disparities in HIV-related mortality across different racial and ethnic groups?
-Can a spatial model explain variation in HIV death-rates at the state-level and/or regional-level after accounting for variables associated with sociodemographic, SDOH, and HIV-related outcomes?
+Can the ratio of Black-to-White HIV-related mortality rate (MRR) from 2020 in the United States be modeled by social determinants of health or by HIV-related outcomes (late diagnoses, viral suppression).
+Can a spatial model explain variation in MMR at the state-level and regional-level after accounting for variables associated SDOH or HIV-related outcomes?
 
-H0 : There is not a spatial correlation HIV-related deaths and SDOH or HIV-related outcomes.
+H0 : There is no spatial correlation between Black-to-White HIV-related MRR and SDOH or HIV-related outcomes.
 
-Ha : There is a spatial correlation between HIV-related deaths and SDOH or HIV-related outcomes.
+Ha : There is a spatial correlation between Black-to-White HIV-related MRR and SDOH or HIV-related outcomes.
 
@@
 While initiatives such the Ending the Epidemic may provide xxx, this is followed by nearly one decade of steadily maintained HIV diagnoses rates, suggesting that existence of effective medical therapy alone will not end the HIV epidemic in the United States given the complex barriers existence among populations who are most vulnerable for HIV infection.
-The goal of this project is to gain a between understanding of factors associated with HIV-related mortality in the United States in
 
@@
 
-### Feature Engineering
+### Outcome Variables
 
-The primary response variable in this analysis was HIV-related death rate.
-Feature engineering of `death rate (estimated)` was performed using counts of HIV related deaths divided by the estimated population based on US Census data multiplied by 100,000 persons.
-Initially, persons of multiple racial backgrounds were considered in the analysis, however, significant discrepancies between the death rates provided by AIDSVu and the calculated death rate based on counts and estimated population meant that I could not reliably interpret these data.
-The decision to used `death rate (estimated)` instead of `death rate (AIDSVu)` was made to help ensure that differences across various models that used rates versus counts was not 'artifact'.
-Transformation of this variable was performed as necessary, and is discussed further in later sections.
+The primary response variable in this analysis is the ratio of Black-to-White HIV-related death rate (MRR).
+To compute this variable, feature engineering was performed to create `death rate (estimated)` stratified by race (Black and White) using counts of HIV related deaths divided by the estimated population based on US Census data multiplied by 100,000 persons.
+The variable `MRR` was then computed for each state by dividing the estimate death rates for Black persons divided by the estimate death for White persons.
+When using methods that required count data,
 
-### Variables considered
+Social Determinants of Health
 
-Predictor variables used in this analysis were selected to serve as proxy for SDOH.
-The Gini coefficient was used measure income inequality; percent living in poverty was used to estimate economic stability; percent with high school education was used to estimate access to education; percent uninsured under 65 was used to measure health care access; and percent with access to broadband internet was used to estimate the neighborhood and built environment.
+Predictors used in this analysis were selected to serve as proxy for SDOH.
+A number of variables were initially considered, but ultimately five predictors were selected: the Gini coefficient was used measure income inequality; percent living in poverty was used to estimate economic stability; percent uninsured under 65 was used to measure health care access; and percent without internet subscription was used to estimate the neighborhood and built environment; and percent living with food insecurity.
 
@@
 
-![](images/test-01.png){fig-align="center"}
-
 ## Univariate analysis
 
-![](images/dr_boxplot.png){fig-align="center" width="700"}
+Univariate analysis of HIV-related death rate
 
-The distribution of death rate was assessed for each race both visually and using the Shapiro test for normality. Below is the histogram of the distributed of the raw `death rate (estimated).`
+![](images/MRR_hist.png){fig-align="center"}
 
-![](images/dr_hist-01.png){fig-align="center" width="700"}
+![](images/dr_state.png){fig-align="center"}
 
-The death rate for Black persons did not require transformation and the death rate for Hispanic persons was normalized using a log + 1 transformation.
-Despite utilizing transformations of increasing complexity and power, including Box Cox and Yeo Johnson transformations, I was not able to normalize the distribution of death rate for White persons.
-As a result, I was unable to include White persons in the model for assessing for spatial correlation.
-While there are techniques to handle spatial data that is not normally distributed, such as geographically weighted Poisson regression and geographically weight negative binomial regression, my knowledge of the implementation of these techniques limited my ability to uses these methods in my analysis.
+![](images/table1.png){fig-align="center"}
 
-![](images/raw_rate_map-05.png){fig-align="center"}
-
-![](images/roe_map-05.png)
-
-![](images/chi_map-05.png){fig-align="center"}
+The distribution of death rate was assessed for each race both visually and using the Shapiro test for normality.
+Based on the Shapiro test, there was insufficient evidence to reject normality with a p-value of 0.63.
+No normalizing transformation was needed.
+Below is the histogram of the our primary outcome variable, MRR.
 
 |     |     |
-|---------------------------------------------|------------------------------------------|
-| ![](images/gini_map.png){width="300"}       | ![](images/ed_map-01.png){width="300"}   |
-| ![](images/housing_map.png){width="300"}    | ![](images/poverty_map.png){width="300"} |
-| ![](images/unemployed_map.png){width="300"} |                                          |
+|-----|-----|
+|     |     |
+|     |     |
+|     |     |
 
@@
 
-![](images/bi_gini-01.png){fig-align="center" width="429"}
+Ultimately
 
-![](images/bi_ed-01.png){fig-align="center" width="429"}
-
-![](images/bi_housing-01.png){fig-align="center" width="429"}
-
-![](images/bi_unemployed.png){fig-align="center" width="429"}
-
-![](images/bi_pov-01.png){fig-align="center" width="429"}
+|                                    |                                 |
+|------------------------------------|---------------------------------|
+| ![](images/bi_pov-03.png)          | ![](images/bi_gini-04.png)      |
+| ![](images/bi_unemployed-02.png)   | ![](images/bi_uninsured-01.png) |
+| ![](images/bi_foodinsecure-01.png) |                                 |
 
@@
 
-![](images/corr_mat.png){fig-align="center" width="429"}
+![](images/corr_mat-01.png){fig-align="center"}
 
-![](images/scree_plot.png){fig-align="center" width="429"}
-
-![](images/fitted_map-03.png){fig-align="center"}
+![](images/contriution_plot.png){fig-align="center"}
 

### 2023-04-25 code/analysis_code/1_EDA_alt.Rmd
more changes

@@
 library(knitr)
+library(kableExtra)
 library(ggpubr)
@@
 
+```{r}
+states <- data.frame(state.name)
+states$Region <- case_when(
+  states$state.name %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania") ~ "Northeast",
+  states$state.name %in% c("Illinois", "Indiana", "Iowa", "Kansas", "Michigan", "Minnesota", "Missouri", "Nebraska", "North Dakota", "Ohio", "South Dakota", "Wisconsin") ~ "Midwest",
+  states$state.name %in% c("Alabama", "Arkansas", "Delaware", "Florida", "Georgia", "Kentucky", "Louisiana", "Maryland", "Mississippi", "North Carolina", "Oklahoma", "South Carolina", "Tennessee", "Texas", "Virginia", "West Virginia") ~ "South",
+  states$state.name %in% c("Arizona", "California", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Oregon", "Utah", "Washington", "Wyoming") ~ "West"
+)
+
+race_sub <- full_join(race_sub, states, by = join_by(State == state.name))
+
+race_sub <- race_sub %>% 
+  filter(!grepl("Alaska", State),
+         !grepl("Hawaii", State))
+```
+
 Let's create a variable for mortality rate ratio of Black/ White
@@
 ```{r}
-race_sub %>% 
+table1 <- race_sub %>% 
   group_by(Race) %>% 
@@
   ungroup() %>% 
-  mutate(percent_by_race = round(tot_deaths/sum(tot_deaths)*100, 1),
-         ratio = round(overall_rate/overall_rate[Race =="White"],1)) %>%
+  mutate(ratio = round(overall_rate/overall_rate[Race =="White"],1)) %>%
   kable(col.names = c("Race/ Ethnicity",
                       "HIV-Related Deaths (n)",
-                      "Population (>12 years old)",
-                      "Death Rate (per 100,000)",
-                      "Percent",
-                      "Ratio"))
+                      "Population \n (>12 years old)",
+                      "Death Rate \n (per 100,000)",
+                      "Mortality Rate Ratio"),
+        caption = "<center>HIV-Related Deaths by Race/ Ethnicity, United States, 2020</center>",
+        align = "c") %>%
+  kable_minimal() %>%
+  kable_styling(font_size = 30)%>%
+gsub("font-size: initial !important;",
+"font-size: 30pt !important;",
+.)  %>% 
+  landscape()
+
+table1
+
+table1 %>%  
+  save_kable(here("results/table1.png"))
+# using white persons as reference group
+```
+
+```{r}
+region_table <- race_sub %>% 
+  group_by(Region, Race) %>% 
+  summarize(tot_deaths = sum(`Death Cases`),
+            tot_population = sum(`Population`),
+            overall_rate = round(tot_deaths/ tot_population *1e5,1)) %>% 
+  ungroup() %>% 
+  group_by(Region) %>% 
+  mutate(ratio = round(overall_rate/overall_rate[Race =="White"],1)) %>%
+  kable(col.names = c("Region",
+                      "Race/ Ethnicity",
+                      "HIV-Related Deaths (n)",
+                      "Population \n (>12 years old)",
+                      "Death Rate \n (per 100,000)",
+                      "Mortality Rate Ratio"),
+        caption = "<center>HIV-Related Deaths by Region, United States, 2020</center>") %>% 
+  kable_minimal() %>%
+  kable_styling(font_size = 30) %>%
+gsub("font-size: initial !important;",
+"font-size: 30pt !important;",.)  %>% 
+  landscape()
+
+region_table
+
+region_table %>%  
+  save_kable(here("results/region_table.png"))
 # using white persons as reference group
@@
 
+race_sub %>% 
+  ggplot() +
+  geom_boxplot(aes(y = `Death Rate (estimated)`, x = `Region`, group = `Region`)) + theme_classic() + 
+  theme(legend.position = "bottom") +
+  labs(title = "HIV-Related Death Rate of Black and White Persons, by Region, United States, 2020", y = "Rate (per 100,000)", x = "Region") + facet_wrap(.~ Race)
+ggsave(here("results/dr_region.png"), units="in", width=12, height=5, dpi=300)
 ```
@@
 Bivariate Analysis 
+We will deem variables with a correlation coeffiencient of |0.15| as important, and will include these in PCA
 ```{r}
@@
   stat_cor(method = "pearson") 
+#"pass"
 ggsave(here("results/bi_gini.png"), units="in", width=10, height=7, dpi=300)
-# did not "pass"
 ```
@@
 # "pass"
+ggsave(here("results/bi_no_internet.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
   stat_cor(method="pearson")
-# "pass"
+# did not "pass"
 ```
@@
   stat_cor(method="pearson")
-# "pass"
-ggsave(here("results/bi_latedx.png"), units="in", width=10, height=7, dpi=300)
+# did not"pass"
 ```
@@
            "Race",
+           "Gini Coefficient",
            "Percent Living in Poverty",
@@
            "Percent Uninsured Under 65 Years Old",
-           "Percent Living with Food Insecurity",
-           "Percent Living with Severe Housing Cost Burden",
-           "Median Household Income",
-           "Percentage Diagnosed Late"))
+           "Percent Living with Food Insecurity"))
 ```

### 2023-04-25 code/analysis_code/2_pca_alt.Rmd
more changes

@@
 df_response <- race_sub %>%
-    select(contains(c("State", "Ratio", "MRR", "Race", "Death", "Population")))
+    select(contains(c("State","Region", "Ratio", "MRR", "Race", "Death", "Population")))
 
-  # pulling variables that were correlated with MRR, and thenp scaling
+  # pulling variables that were correlated with MRR, and then scaling
 pos_predictors_scaled <- pos_predictors %>% 
@@
 corr_matrix <- pos_predictors_scaled %>% 
-    select(c(3:9)) %>% 
+    select(c(3:7)) %>% 
     cor()
@@
 #pca loadings of principal components 1 and 2
-print(pca$loadings[, 1:2])
+print(pca$loadings[, 1:3])
 
@@
 ```
-Since we originally had 11 variables, we should only look at components that explain more than 9.1% of variation.
-We can see from the scree plot that principal component 1 explained 73.3% of variation, and components 1 and 2 explained 94.9% of the variation, and components 1 - 2 explained 91.1%.
+Since we originally had 5 variables, we should only look at components that explain more than 20% of variation.
+This is includes components 1 through 3.
 All other principal components explains less that 8.9% of variation, so they will excluded.

### 2023-04-25 code/analysis_code/3_model_alt.Rmd
more changes

@@
   mult_race <- glm(MRR ~ 
-                        `Percent Living in Poverty` +
+                       # `Percent Living in Poverty` +
+                        `Gini Coefficient` + 
                         `Percent Without an Internet subscription` +
-                        `Percent Uninsured Under 65 Years Old` +
-                        `Median Household Income`  +
-                     #   `Percent Living with Severe Housing Cost Burden` +
-                        `Percent Living with Food Insecurity`, 
+                        `Percent Uninsured Under 65 Years Old`,
+                      #  `Percent Living with Food Insecurity`,
                      data = df,
@@
   mult_race <- glm(MRR ~ 
-                        `Percent Living in Poverty` +
+                       # `Percent Living in Poverty` +
+                        `Gini Coefficient` + 
                         `Percent Without an Internet subscription` +
-                        `Percent Uninsured Under 65 Years Old` +
-                        `Median Household Income`  +
-                       # `Percent Living with Severe Housing Cost Burden` +
-                        `Percent Living with Food Insecurity`,
+                        `Percent Uninsured Under 65 Years Old`,
+                      #  `Percent Living with Food Insecurity`,
                      data = df,
@@
 
-pois_race <- glm(`Death Cases` ~ Race +
-                        `Percent Living in Poverty` +
-                       # `Gini Coefficient` + 
+pois_race <- glm(`Death Cases` ~ Race + region +
+                       # `Percent Living in Poverty` +
+                        `Gini Coefficient` + 
                         `Percent Without an Internet subscription` +
                         `Percent Uninsured Under 65 Years Old` +
-                        `Median Household Income`  +
-                        `Percent Living with Food Insecurity` +
-                        #`Percent Living with Severe Housing Cost Burden` +
+                      #  `Percent Living with Food Insecurity`  +
                      offset(log(`Population`)),
@@
 pois_aic <- function(df) {
-  pois_race <- glm(`Death Cases` ~  Race +
-                        `Percent Living in Poverty` +
-                      #  `Gini Coefficient` + 
+  pois_race <- glm(`Death Cases` ~  Race + region +
+                       # `Percent Living in Poverty` +
+                        `Gini Coefficient` + 
                         `Percent Without an Internet subscription` +
                         `Percent Uninsured Under 65 Years Old` +
-                        `Median Household Income`  +
-                        `Percent Living with Food Insecurity` +
-                       # `Percent Living with Severe Housing Cost Burden` +
+                      #  `Percent Living with Food Insecurity`  +
                      offset(log(`Population`)),
@@
 nb_function  <- function(df) {
-  neg_bin_race <- glm.nb(`Death Cases` ~ Race +
-                        `Percent Living in Poverty` +
-                       # `Gini Coefficient` + 
+  neg_bin_race <- glm.nb(`Death Cases` ~ Race + region +
+                       # `Percent Living in Poverty` +
+                        `Gini Coefficient` + 
                         `Percent Without an Internet subscription` +
                         `Percent Uninsured Under 65 Years Old` +
-                        `Median Household Income`  +
-                        `Percent Living with Food Insecurity`  +
-                       # `Percent Living with Severe Housing Cost Burden` +
+                      #  `Percent Living with Food Insecurity`  +
                      offset(log(`Population`)),
@@
 nb_aic <- function(df) {
-  neg_bin_race <- glm.nb(`Death Cases` ~ Race +
-                        `Percent Living in Poverty` +
-                       # `Gini Coefficient` + 
+  neg_bin_race <- glm.nb(`Death Cases` ~ Race + region +
+                       # `Percent Living in Poverty` +
+                        `Gini Coefficient` + 
                         `Percent Without an Internet subscription` +
                         `Percent Uninsured Under 65 Years Old` +
-                        `Median Household Income`  +
-                        `Percent Living with Food Insecurity`  +
-                        #`Percent Living with Severe Housing Cost Burden` +
+                      #  `Percent Living with Food Insecurity`  +
                      offset(log(`Population`)),
@@
   car_race <- spautolm(MRR ~                      
-                        `Percent Living in Poverty` +
-                       `Median Household Income` + 
+                       # `Percent Living in Poverty` +
+                        `Gini Coefficient` + 
                         `Percent Without an Internet subscription` +
-                        `Percent Uninsured Under 65 Years Old` +
-                     #   `Percent Unemployed`  +
-                     #   `Percent Living with Severe Housing Cost Burden` +
-                        `Percent Living with Food Insecurity`, 
+                        `Percent Uninsured Under 65 Years Old`,
+                      #  `Percent Living with Food Insecurity`,
                            data = df,
@@
   car_race <- spautolm(MRR ~                      
-                        `Percent Living in Poverty` +
-                       `Median Household Income` + 
+                       # `Percent Living in Poverty` +
+                        `Gini Coefficient` + 
                         `Percent Without an Internet subscription` +
-                        `Percent Uninsured Under 65 Years Old` +
-                   #     `Percent Unemployed`  +
-                      #  `Percent Living with Severe Housing Cost Burden` +
-                        `Percent Living with Food Insecurity`, 
+                        `Percent Uninsured Under 65 Years Old`,
+                      #  `Percent Living with Food Insecurity`,
                            data = df,
@@
 aic_df <- data.frame(AIC = c(df_mult, df_pois, df_nb, df_car),
-                     Model = c("GLM", "Poisson", "Negative Binomial", "CAR"),
+                     Model = c("Multiple Linear Regression", "Poisson", "Negative Binomial", "Conditional Autoregressive"),
                      Outcome = c("Ratio", "Count", "Count", "Ratio"))
 aic_df <- aic_df %>% mutate(AIC  = round(AIC, 2))
+
+```
+
+```{r}
+library(kableExtra)
+library(knitr)
+table_fit <- aic_df %>%
+  select(c(Model, Outcome, AIC)) %>% 
+  kable(caption = "<center>Model Fit</center>",
+        align = "c") %>%
+  kable_minimal() %>%
+  kable_styling(font_size = 30) %>%
+gsub("font-size: initial !important;",
+"font-size: 30pt !important;",
+.)  %>% 
+  landscape()
+
+table_fit %>%  
+  save_kable(here("results/table_fit.png"))
+
 ```
@@
 saveRDS(df, here("data/processed_data/df_fit_alt.rds"))
+```
 
-saveRDS(aic_df, here("data/processed_data/aic_alt.rds"))
+Creating function that subsets by specified race, and compute/ plots Moran's I and lisa.
+```{r}
+spacy_function <- function(df) {
+
+  h_race_df <- matrix(0, nrow(df), nrow(df))
+  h_race_df[cbind(match(us_nabs$StateCode, df$`State Abbreviation`),
+  match(us_nabs$NeighborStateCode, df$`State Abbreviation`))] <- 1
+
+  nablist_race_df <- mat2listw(h_race_df, style = "W")
+
+ # plot Moran's I
+  print(moran.test(df$MRR, nablist_race_df, zero.policy = TRUE, randomisation = FALSE))
+  moran.plot(df$MRR, nablist_race_df, zero.policy = TRUE)
+
+  # calculate and print LISA statistics
+  lisa_race <- localmoran(df$MRR, nablist_race_df)
+  summary(lisa_race[,1])
+}
 ```
+
+```{r}
+spacy_function(df)
+```
+
+```{r}
+png(here("results/moran_mrr.png"))
+spacy_function(df)
+dev.off()
+```

### 2023-04-25 code/analysis_code/p4_mapping.Rmd
more changes

@@
 df_raw <- readRDS(here("data/processed_data/race_alt.rds"))
-mod_aic <-  readRDS(here("data/processed_data/aic_alt.rds"))
 ```
@@
   replace(is.na("Ratio Observed/ Expected"), 0) %>% 
-  mutate(Exp_DR = Exp_Deaths/Population *1e5,
+  mutate(Exp_DR = `Expected Death Cases`/Population *1e5,
                OE_mortality_ratio = `Death Rate (estimated)`/Exp_DR,
-         dykes_chi = ((`Death Cases`- Exp_Deaths)/ sqrt(Exp_Deaths)),
+         dykes_chi = ((`Death Cases`- `Expected Death Cases`)/ sqrt(`Expected Death Cases`)),
 
          ## GLM
-         predicted_observed_mult = mult_fit/`Death Cases`,
+         predicted_observed_mult = mult_fit/`MRR`,
          predicted_observed_mult = ifelse(!is.finite(predicted_observed_mult), 0, predicted_observed_mult),
@@
          ## CAR
-         predicted_observed_car = car_fit/`Death Cases`,
+         predicted_observed_car = car_fit/`MRR`,
          predicted_observed_car = ifelse(!is.finite(predicted_observed_car), 0, predicted_observed_car),
@@
          ## POISSON
-         predicted_observed_pois = pois_fit/`Death Cases`,
+         predicted_observed_pois = pois_fit/`MRR`,
          predicted_observed_pois = ifelse(!is.finite(predicted_observed_pois), 0, predicted_observed_pois),
@@
          ##N NEGATIVE BINOMIAL
-         predicted_observed_nb = nb_fit/`Death Cases`,
+         predicted_observed_nb = nb_fit/`MRR`,
          predicted_observed_nb = ifelse(!is.finite(predicted_observed_nb), 0, predicted_observed_nb),
@@
         legend.position = "bottom") +
-  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio in the United States, 2020", x = "", y = "", fill = "Death rate per 100,000") +
+  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio in the United States, 2020", x = "", y = "", fill = "Ratio") +
   scale_fill_distiller(palette = "RdYlGn") +
@@
 ggplot(df_sf_raw, aes(long, lat)) +
-  geom_polygon(aes(group = group, fill =`Percent Living in Poverty`)) +
-  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
-  theme(strip.text = element_text(size = 14),
-        legend.position = "bottom") +
-  labs(title = "Percent Living in Poverty, 2020", x = "", y = "", fill = "%") +
-  theme(plot.title= element_text(hjust = 0.5)) 
-ggsave(here("results/poverty_map.png"), units="in", width=10, height=7, dpi=300)
-```
-
-```{r}
-ggplot(df_sf_raw, aes(long, lat)) +
   geom_polygon(aes(group = group, fill =`Percent Uninsured Under 65 Years Old`)) +
@@
 ggsave(here("results/uninsured_map.png"), units="in", width=10, height=7, dpi=300)
-```
-
-```{r}
-ggplot(df_sf_raw, aes(long, lat)) +
-  geom_polygon(aes(group = group, fill =`Percent Living with Severe Housing Cost Burden`)) +
-  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
-  theme(strip.text = element_text(size = 14),
-        legend.position = "bottom") +
-  labs(title = "Percent Living with Severe Housing Cost Burden 2020", x = "", y = "", fill = "%") +
-  theme(plot.title= element_text(hjust = 0.5)) 
-ggsave(here("results/housing_map.png"), units="in", width=10, height=7, dpi=300)
-```
-
-```{r}
-ggplot(df_sf_raw, aes(long, lat)) +
-  geom_polygon(aes(group = group, fill =`Percent Living with Food Insecurity`)) +
-  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
-  theme(strip.text = element_text(size = 14),
-        legend.position = "bottom") +
-  labs(title = "Percent Living with Food Insecurity, 2020", x = "", y = "", fill = "%") +
-  theme(plot.title= element_text(hjust = 0.5)) 
-ggsave(here("results/foodinsecure_map.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
 
-```{r}
-ggplot(df_sf_raw, aes(long, lat)) +
-  geom_polygon(aes(group = group, fill =`Percent Unemployed` )) +
-  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
-  theme(strip.text = element_text(size = 14),
-        legend.position = "bottom") +
-  labs(title = "Percent Unemployed, 2020", x = "", y = "", fill = "%") +
-  theme(plot.title= element_text(hjust = 0.5)) 
-ggsave(here("results/unemployed_map.png"), units="in", width=10, height=7, dpi=300)
-```
 The map for HIV-related death differs drastically across the United States by race/ethnicity.
@@
   labs(title = "NULL - Expected HIV Mortality Rate by Race/ Ethnicity
-       in the United States, 2020", x = "", y = "", fill = "Rate") +
+       in the United States, 2020", x = "", y = "", fill = "Death Rate (per 100,000)") +
   theme(plot.title= element_text(hjust = 0.5))
@@
 ```{r}
-##Smoothed map using fitted values from best models.
+##Smoothed map using fitted values from best model (CAR)
 ggplot(df_sf_scale, aes(long, lat)) +
@@
         legend.position = "bottom") +
-  labs(title = "Smoothed Map (Fitted) of HIV-Related Death Rate by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "") +
+  labs(title = "Smoothed Map of Black-to-White HIV-Related Mortality Rate Ratio,
+United States, 2020 (CAR)", x = "Ratio", y = "", fill = "") +
   theme(plot.title= element_text(hjust = 0.5))
 
-ggsave(here("results/fitted_map.png"), units="in", width=10, height=4, dpi=300)
+ggsave(here("results/car_fit_map.png"), units="in", width=10, height=4, dpi=300)
+```
+
+```{r}
+##Smoothed map using fitted values from MULT (only marginally worse than CAR)
+ggplot(df_sf_scale, aes(long, lat)) +
+  geom_polygon(aes(group = group, fill = mult_fit)) +
+  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
+  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)  + 
+  theme_void() + 
+  theme(strip.text = element_text(size = 14),
+        legend.position = "bottom") +
+  labs(title = "Smoothed Map of Black-to-White HIV-Related Mortality Rate Ratio,
+United States, 2020 (GLM)", x = "Ratio ", y = "", fill = "") +
+  theme(plot.title= element_text(hjust = 0.5))
+
+ggsave(here("results/mult_fit_map.png"), units="in", width=10, height=4, dpi=300)
 ```
@@
   coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + 
-  facet_wrap(. ~ Race)  + 
   theme_void() + 
@@
         legend.position = "bottom") +
-  labs(title = "Map of Ratio of Predicted to Observed Cases (GLM) by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Ratio") +
+  labs(title = "Map of Predicted Over Observed Black-to-White HIV-Related Mortality Rate Ratio,
+United States, 2020 (GLM)", x = "", y = "", fill = "Ratio") +
   theme(plot.title= element_text(hjust = 0.5))
+ggsave(here("results/mult_pred_map.png"), units="in", width=7, height=4, dpi=300)
 
@@
   coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + 
-  facet_wrap(. ~ Race)  + 
   theme_void() + 
@@
         legend.position = "bottom") +
-  labs(title = "Map of Ratio of Predicted to Observed Cases (CAR) by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Ratio") +
+  labs(title = "Map of Predicted Over Observed Black-to-White HIV-Related Mortality Rate Ratio,
+United States, 2020 (CAR)", x = "", y = "", fill = "Ratio") +
   theme(plot.title= element_text(hjust = 0.5))
+ggsave(here("results/car_pred_map.png"), units="in", width=7, height=4, dpi=300)
 ``` 

### 2023-04-25 outputs_code.Rmd
more changes

--- 2023-04-21 outputs_code.Rmd
@@
 ```{r}
-df <- readRDS(here("data/processed_data/race_new.rds"))
+df <- readRDS(here("data/processed_data/race_alt.rds"))
 
@@
                       "Death Rate \n (per 100,000)",
-                      "Percent",
-                      "Ratio"),
-        caption = "HIV-Related Deaths Summary by Race/ Ethncity, United States, 2020",
+                      "Mortality Rate Ratio"),
+        caption = "<center>HIV-Related Deaths by Race/ Ethnicity, United States, 2020</center>",
         align = "c") %>%
   kable_minimal() %>%
-  kable_styling(font_size = 50) %>% 
-  save_kable(here("results/test.png"))
+  kable_styling(font_size = 30)%>%
+gsub("font-size: initial !important;",
+"font-size: 30pt !important;",
+.)  %>% 
+  landscape() %>% 
+  save_kable(here("results/table1.png"))
 # using white persons as reference group

### 2023-04-25 products/manuscript/manuscript.qmd
more changes

@@
 
-Spatial object of the contiguous United States was obtained from library(maps) in R.
-The list of neighboring contiguous United States was obtained from a public repository on GitHub.
+In order to assess spatial associations, this analysis was limited to the contiguous United States.
+Spatial object of the contiguous United States was obtained from library(maps) in R and a list of neighboring contiguous United States was obtained from a public repository on GitHub.
 https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state
@@
 
-## Model Selection
+## Statistical Analysis
 
-The relative goodness of fit of the various models used in this analysis was assessed using the Akaike information criterion (AIC).
+The goal of this analysis was to assess the spatial distribution of the Black-to-White HIV-related MRR.
+A number of predictors were analyzed using simple linear regression to predict MRR.
+Variables with a correlation coefficient of \|-0.15\| or more were selected as possible predictors.
+To address the possibility of multicollinearity between predictors, principal component analysis was then performed on these variable to aid in variable reduction.
+
+Both non-spatial and spatial models were fitted to data as appropriate.
+Analysis of the full models began with multiple linear regressions model before moving on to models of increasing complexity, such as Poisson regression and the conditional autoregressive (CAR) model.
+Models that were flexible to continuous data used `MRR` as the primary outcome, while models that required count data used the number of deaths, `Death Cases` , and used `log(Population)` as the offset, and additionally included variables `Race` and `Region` in the models.
+
+Model comparison and relative goodness of fit of the various models used in this analysis was assessed using the Akaike information criterion (AIC).
+
+Moran's *I* was computed to test the null hypothesis of spatial independence.
 
@@
 
-Univariate analysis of HIV-related death rate
+Our primary outcome of interest was the Black-to-White HIV-related MRR, which was computed as described above.
+Table 1 shows the overall number of HIV-related deaths, population size older than 12 years, death rate and the MRR in the United States in 2020.
+The rate of HIV-related deaths among this Black persons was 7.2 times high than the rate of HIV-related deaths among White persons.
 
-![](images/MRR_hist.png){fig-align="center"}
+![Table 1](images/table1-01.png){fig-align="center" width="600"}
 
-![](images/dr_state.png){fig-align="center"}
+![Figure 1](images/dr_race_bp.png){fig-align="center" width="600"}
 
-![](images/table1.png){fig-align="center"}
-
-The distribution of death rate was assessed for each race both visually and using the Shapiro test for normality.
-Based on the Shapiro test, there was insufficient evidence to reject normality with a p-value of 0.63.
+Visual assessment (Figure 1) and Shapiro test were congruent that the distribution of Black-to-White HIV-related MRR in the US in 2020 followed a normal distribution (*p* = 0.63).
 No normalizing transformation was needed.
@@
 
-|     |     |
-|-----|-----|
-|     |     |
-|     |     |
-|     |     |
+After verifying that the Black-to-White HIV-related MRR met assumptions of normality, I tested for spatial association of the Black-to-White MRR by computing Moran's *I* (*`I`* `= 0.1507`).
+For uncorrelated data, the expected value of Moran's *I* was `E(I) = - 0.0213` with a variance `var(I) = 0.0095`.
+The data showed statistically significant evidence of spatial dependence `(z =1.768; p = 0.039)`.
+Figure 2 shows a Moran's scatter plot of spatially lagged Black-to-White HIV-related MRR and Black-to-White HIV-related MRR.
+The positive slope indicates that there is a degree of spatial clustering of the Black-to-White HIV-related MRR.
 
-```{r}
+![Figure 2](images/moran_mrr.png){fig-align="center"}
 
-```
+Figure 3 and 4 show the raw HIV-related death rate by Race and State.
 
-*Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project.*
+![Figure 3](images/dr_state-01.png){fig-align="center" width="600"}
 
-## Bivariate Analysis
+![Figure 4](images/raw_rate_map-06.png){fig-align="center"}
 
-Ultimately
+Figure 5 compares the raw HIV-related death rate by Race and Region.
+These plots again show distinct disparities between rate of HIV-related death by race.
+Additionally, there is evidence of disparities in HIV-related death rate by region for Black persons.
+
+![Figure 5](images/dr_region.png){fig-align="center" width="600"}
+
+Table 2 shows maps of selected predictors in the United States in 2020.
+We can see that the `Gini Coefficient` was generally higher in Southern half of the country, indicating a greater degree of income inequality.
+`Percent Uninsured Under 65 Years Old` was highest in Texas, with the South and non-coastal Western states also appearing to have higher rates of uninsured.
+`Percent Without an Internet Subscription` was generally low through the United States, however, Wyoming had the highest percent without Internet subscription.
+
+Table 2.
+
+|                             |                               |                                    |
+|-----------------------------|-------------------------------|------------------------------------|
+| ![](images/gini_map-02.png) | ![](images/uninsured_map.png) | ![](images/no_internet_map-01.png) |
+
+## Bivariate Analysis and Principal Component Analysis
+
+To narrow down potential predictors to include in later models, I first started by performing simple linear regression models of MRR vs one of the possible predictors.
+Variables that had a correlation coefficient of \|0.15\| or more were then selected as possible variables for analysis.
+These possible predictors were then scaled and principal component analysis (PCA) was performed.
 
 |                                              |                                                 |
-|------------------------------------|---------------------------------|
-| ![](images/bi_pov-03.png)          | ![](images/bi_gini-04.png)      |
-| ![](images/bi_unemployed-02.png)   | ![](images/bi_uninsured-01.png) |
-| ![](images/bi_foodinsecure-01.png) |                                 |
+|----------------------------------------------|-------------------------------------------------|
+| ![](images/bi_gini-05.png){width="300"}      |                                                 |
+| ![](images/bi_no_internet.png){width="300"}  | ![](images/bi_pov-04.png){width="300"}          |
+| ![](images/bi_uninsured-02.png){width="300"} | ![](images/bi_foodinsecure-02.png){width="300"} |
 
-*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
+Table 2. shows the plot from the bivariate analysis of predictors that were scaled and included in the PCA. Table 3 shows the results of PCA.
+
+|                                          |                                                |
+|------------------------------------------|------------------------------------------------|
+| ![](images/corr_mat-02.png){width="300"} | ![](images/scree_plot-01.png){width="300"}     |
+| ![](images/biplot.png){width="300"}      | ![](images/contribution_plot.png){width="300"} |
+
+Results of the PCA shows that the first three principal components contributed for x % of variation between predictors.
+The variables that best represent the components were `Gini Coefficient` , `Percent Uninsured Under 65 Years Old`, `Percent Without Internet Subscription` .
+Ultimately, these variables were included as predictors in the full model.
 
@@
 
-![](images/corr_mat-01.png){fig-align="center"}
+![](images/car_fit_map-01.png){fig-align="center"}
 
-![](images/contriution_plot.png){fig-align="center"}
+![](images/mult_fit_map.png)
+
+|                                 |                                  |
+|---------------------------------|----------------------------------|
+| ![](images/car_pred_map-03.png) | ![](images/mult_pred_map-02.png) |
+
+The CAR model accounting for spatial variation only performed marginally better than the GLM model that did not account for spatial variation.
 

### 2023-04-25 products/manuscript/manuscript.qmd
updates

@@
 
-*Write a summary of your project.*
+This project aims to identify factors associated with racial disparities in HIV-related death rates in the United States.
+To assess this, we look at the ratio of Black-to-White HIV-related mortality rate (MRR) from 2020 in the United States.
+Additionally, we assess whether social determinants of health (SDOH) can explain variation in MMR at the state-level and regional-level.
+Both spatial and non-spatial methods were used.
 
@@
 
-HIV has a disproportionate impact on marginalized groups, resulting in significant disparities in rate of new HIV diagnoses and poor HIV-related health outcomes, such as late HIV diagnosis, progression to AIDS, and HIV-related death.
-In the United States, these disparities are most apparent when comparing HIV-related outcomes between Black and White persons.
-Previous study have found direct links to indicators of social determinants of health and poorer health outcomes for persons with HIV.
-
-Other studies have found an association between different SDOH measures and COVID-19 mortality that varied across racial and ethnic groups and community types.
-
-The goal of this project is to gain an understanding of factors associated the disparity of HIV-related death rates between Black and White persons in the United States.
+HIV continues to have a significant impact on marginalized communities, leading to significant disparities in the rates of new HIV diagnoses and HIV-related health outcomes, including late HIV diagnosis, progression to AIDS, and HIV-related death.
+The extent of these disparities are particularly pronounced in the United States when comparing HIV-related outcomes among Black and White persons.
+Studies have revealed a direct correlation between social determinants of health indicators and poor health outcomes for people with HIV.
+In this project, our objective is to identify the factors contributing to the disparities in HIV-related death rates between Black and White individuals in the United States.
 
@@
 
-Since the emergence of the HIV epidemic in 1981 in the United States, more than 1.2 million people in the United States have been diagnosed with HIV, resulting in more than 700,000 HIV-related deaths \[cite\].
-Medical advancements have led to effective antiretroviral therapies that not only treat HIV infection and reduce adverse HIV-related health outcomes, such as progression to AIDS, but have also led to use of these medications to prevent infection before and after potential exposure for people who are HIV negative Even still, more than 30,000 people in the United States continue to be diagnosed each year \[cite\].
-While initiatives such the Ending the Epidemic may provide xxx, this is followed by nearly one decade of steadily maintained HIV diagnoses rates, suggesting that existence of effective medical therapy alone will not end the HIV epidemic in the United States given the complex barriers existence among populations who are most vulnerable for HIV infection.
+Since the emergence of the HIV epidemic in 1981 in the United States, more than 1.2 million people in the United States have been diagnosed with HIV, resulting in more than 700,000 HIV-related deaths \cite{jun07HIVAIDSEpidemic2021}.
+Medical advancements have led to effective antiretroviral therapies that not only treat HIV infection and reduce adverse HIV-related health outcomes, such as progression to AIDS, but have also led to use of these medications to prevent infection before and after potential exposure for people who are HIV negative Even still, more than 30,000 people in the United States continue to be diagnosed each year \cite{jun07HIVAIDSEpidemic2021}.
+While the Ending the Epidemic initiative set up a goal of 90% reduction in new HIV infections by 2030 by upscaling efforts in HIV diagnoses treatment, and prevention, this is followed by nearly one decade of steady HIV diagnoses rates.
+Developing targeted intervention programs is essential to achieve this goal.
+To accomplish this, we need to gain a deeper understanding of the impact or contribution of complex factors such as social determinants of health or other barriers to poor HIV-related outcomes and disparities.
 
@@
 In order to assess spatial associations, this analysis was limited to the contiguous United States.
-Spatial object of the contiguous United States was obtained from library(maps) in R and a list of neighboring contiguous United States was obtained from a public repository on GitHub.
-https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state
+Spatial object of the contiguous United States was obtained from library(maps) in R and a list of neighboring contiguous United States was obtained from a public repository on [GitHub](https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state).
+\cite{ubikuityListofneighboringstatesforeachUSstate2022}
 
@@
 The variable `MRR` was then computed for each state by dividing the estimate death rates for Black persons divided by the estimate death for White persons.
-When using methods that required count data,
 
-Social Determinants of Health
+### Social Determinants of Health
 
@@
 |                                          |                                            |                                                 |
-|-----------------------------|-------------------------------|------------------------------------|
-| ![](images/gini_map-02.png) | ![](images/uninsured_map.png) | ![](images/no_internet_map-01.png) |
+|----------------------|-----------------------|---------------------------|
+| ![](images/gini_map-02.png){width="200"} | ![](images/uninsured_map.png){width="200"} | ![](images/no_internet_map-01.png){width="200"} |
 
@@
 |                                              |                                                 |
-|----------------------------------------------|-------------------------------------------------|
+|-----------------------------------|-------------------------------------|
 | ![](images/bi_gini-05.png){width="300"}      |                                                 |
@@
 |                                          |                                                |
-|------------------------------------------|------------------------------------------------|
+|----------------------------------|--------------------------------------|
 | ![](images/corr_mat-02.png){width="300"} | ![](images/scree_plot-01.png){width="300"}     |
@@
 
-![](images/car_fit_map-01.png){fig-align="center"}
+![](images/car_fit_map-01.png){fig-align="center" width="600"}
 
-![](images/mult_fit_map.png)
+![](images/mult_fit_map.png){width="600"}
 
 |                                              |                                               |
-|---------------------------------|----------------------------------|
-| ![](images/car_pred_map-03.png) | ![](images/mult_pred_map-02.png) |
+|----------------------------------------------|-----------------------------------------------|
+| ![](images/car_pred_map-03.png){width="350"} | ![](images/mult_pred_map-02.png){width="350"} |
 
 The CAR model accounting for spatial variation only performed marginally better than the GLM model that did not account for spatial variation.
-
-*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here.*
 

### 2023-04-26 products/manuscript/manuscript.qmd
improved manuscript

@@
 To assess this, we look at the ratio of Black-to-White HIV-related mortality rate (MRR) from 2020 in the United States.
-Additionally, we assess whether social determinants of health (SDOH) can explain variation in MMR at the state-level and regional-level.
+Additionally, we assess whether social determinants of health (SDOH) can explain variation in MRR at the state-level and regional-level.
 Both spatial and non-spatial methods were used.
@@
 
-Since the emergence of the HIV epidemic in 1981 in the United States, more than 1.2 million people in the United States have been diagnosed with HIV, resulting in more than 700,000 HIV-related deaths \cite{jun07HIVAIDSEpidemic2021}.
-Medical advancements have led to effective antiretroviral therapies that not only treat HIV infection and reduce adverse HIV-related health outcomes, such as progression to AIDS, but have also led to use of these medications to prevent infection before and after potential exposure for people who are HIV negative Even still, more than 30,000 people in the United States continue to be diagnosed each year \cite{jun07HIVAIDSEpidemic2021}.
+Since the emergence of the HIV epidemic in 1981 in the United States, more than 1.2 million people in the United States have been diagnosed with HIV, resulting in more than 700,000 HIV-related deaths [@jun07HIVAIDSEpidemic2021].
+Medical advancements have led to effective antiretroviral therapies that not only treat HIV infection and reduce adverse HIV-related health outcomes, such as progression to AIDS, but have also led to use of these medications to prevent infection before and after potential exposure for people who are HIV negative Even still, more than 30,000 people in the United States continue to be diagnosed each year [@jun07HIVAIDSEpidemic2021].
 While the Ending the Epidemic initiative set up a goal of 90% reduction in new HIV infections by 2030 by upscaling efforts in HIV diagnoses treatment, and prevention, this is followed by nearly one decade of steady HIV diagnoses rates.
@@
 
-Data was obtained from AIDSVu \cite{AIDSVuDatasetsIncludes} on counts and rates of HIV-related mortality by state in 2020.
+Data was obtained from AIDSVu [@AIDSVuDatasetsIncludes] on counts and rates of HIV-related mortality by state in 2020.
 Social determinants of health (SDOH), were also obtained from AIDSVu by state from 2020.
@@
 Data on 2020 demographic composition by race/ethnicity and age by state was obtained from the [U.S.
-Census Bureau](https://data.census.gov/) \cite{USCensusBureau}.
+Census Bureau](#0) [@USCensusBureau].
 A five-year estimate of percent with access to broadband internet by state (released in 2020) was also obtained from the US Census Bureau.
@@
 
-Data on the percentage distribution of children under 14 years old was obtained from [Kids Count](https://datacenter.kidscount.org/), an online resource that provides high-quality data on the well-being of children in the United States.
+Data on the percentage distribution of children under 14 years old was obtained from [Kids Count](https://datacenter.kidscount.org/) [@KIDSCOUNTData], an online resource that provides high-quality data on the well-being of children in the United States.
 
@@
 Spatial object of the contiguous United States was obtained from library(maps) in R and a list of neighboring contiguous United States was obtained from a public repository on [GitHub](https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state).
-\cite{ubikuityListofneighboringstatesforeachUSstate2022}
+[@ubikuity2022]
 
@@
 
-![Table 1](images/table1-01.png){fig-align="center" width="600"}
+```{r, echo=FALSE, message=FALSE, fig.cap="Table 1."}
+include_graphics(here("results/table1.png"))
+```
 
-![Figure 1](images/dr_race_bp.png){fig-align="center" width="600"}
+```{r, echo=FALSE, message=FALSE, fig.cap="Figure 1."}
+include_graphics(here("results/MRR_hist.png"))
+```
 
@@
 
-![Figure 2](images/moran_mrr.png){fig-align="center"}
+```{r, echo=FALSE, message=FALSE, fig.width = 4, fig.cap="Figure 2. Moran's Scatter Plot"}
+include_graphics(here("results/moran_mrr.png"))
+```
 
@@
 
-![Figure 4](images/raw_rate_map-06.png){fig-align="center"}
+```{r, echo=FALSE, message=FALSE, fig.cap="Figure 3. Bargraph of Raw HIV-Related Death Rates by Race by State, 2020, United States"}
+include_graphics(here("results/dr_state.png"))
+```
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Figure 4. Boxplot of HIV-Related Death by Race by Region, 2020, United States"}
+include_graphics(here("results/dr_region.png"))
+```
 
@@
 Additionally, there is evidence of disparities in HIV-related death rate by region for Black persons.
-
-![Figure 5](images/dr_region.png){fig-align="center" width="600"}
 
@@
 
-Table 2.
-
-|                                          |                                            |                                                 |
-|----------------------|-----------------------|---------------------------|
-| ![](images/gini_map-02.png){width="200"} | ![](images/uninsured_map.png){width="200"} | ![](images/no_internet_map-01.png){width="200"} |
+```{r, echo=FALSE, message=FALSE, fig.cap="Table 2", fig.align = "default"}
+include_graphics(here("results/grid_map.png"))
+```
 
@@
 
-|                                              |                                                 |
-|-----------------------------------|-------------------------------------|
-| ![](images/bi_gini-05.png){width="300"}      |                                                 |
-| ![](images/bi_no_internet.png){width="300"}  | ![](images/bi_pov-04.png){width="300"}          |
-| ![](images/bi_uninsured-02.png){width="300"} | ![](images/bi_foodinsecure-02.png){width="300"} |
+```{r, echo=FALSE, message=FALSE, fig.cap="Table 2. Simple Linear Regression between Outcome and Individual Predictors", fig.align = "default"}
+include_graphics(here("results/grid_bivar.png"))
+```
 
-Table 2. shows the plot from the bivariate analysis of predictors that were scaled and included in the PCA. Table 3 shows the results of PCA.
+Table 3. shows the plot from the bivariate analysis of predictors that were scaled and included in the PCA. Table 3 shows the results of PCA.
 
-|                                          |                                                |
-|----------------------------------|--------------------------------------|
-| ![](images/corr_mat-02.png){width="300"} | ![](images/scree_plot-01.png){width="300"}     |
-| ![](images/biplot.png){width="300"}      | ![](images/contribution_plot.png){width="300"} |
+```{r, echo=FALSE, message=FALSE, fig.cap="Table 3", fig.align = "default"}
+include_graphics(here("results/grid_pca.png"))
+```
 
-Results of the PCA shows that the first three principal components contributed for x % of variation between predictors.
-The variables that best represent the components were `Gini Coefficient` , `Percent Uninsured Under 65 Years Old`, `Percent Without Internet Subscription` .
-Ultimately, these variables were included as predictors in the full model.
+Results of the PCA shows that the first principal component contributed for 91.7 % of variation between predictors.
+The variables `Percent Uninsured Under 65 Years Old` explained the least amountof variation.
+Ultimately, only `Median Household Income`, `Percent Living in Poverty`, `Percent Living with Food Insecurity`, and `Percent Without an Internet Subscription` were included as predictors in the full model.
 
@@
 
-![](images/car_fit_map-01.png){fig-align="center" width="600"}
+```{r, echo=FALSE, message=FALSE, fig.cap="Figure.4"}
+include_graphics(here("results/car_fit_map.png"))
+```
 
-![](images/mult_fit_map.png){width="600"}
+```{r, echo=FALSE, message=FALSE, fig.cap="Figure.5"}
+include_graphics(here("results/mult_fit_map.png"))
+```
 
-|                                              |                                               |
-|----------------------------------------------|-----------------------------------------------|
-| ![](images/car_pred_map-03.png){width="350"} | ![](images/mult_pred_map-02.png){width="350"} |
+```{r, echo=FALSE, message=FALSE, fig.cap="Figure.6"}
+include_graphics(here("results/grid_pred.png"))
+```
 
@@
 
-\## Software
-
+## Software
 ```{r}
@@
 ```
-
 {{< pagebreak >}}

### 2023-04-30 code/analysis_code/3_model_alt.Rmd
fixed pca and models

--- 2023-04-25 code/analysis_code/3_model_alt.Rmd
@@
 ```{r}
-df <- readRDS(here("data/processed_data/df_scaled_alt.rds"))
+df <- readRDS(here("data/processed_data/race_alt.rds"))
+prin_comp1 <- readRDS(here("data/processed_data/prin_comp1.rds"))
 us_nabs <- readRDS(here("data/processed_data/us_nabs.rds"))
@@
   mult_race <- glm(MRR ~ 
-                        `Percent Living in Poverty` +
-                        `Percent Without an Internet subscription` +
-                        `Percent Uninsured Under 65 Years Old` +
-                        `Median Household Income`  +
-                     #   `Percent Living with Severe Housing Cost Burden` +
-                        `Percent Living with Food Insecurity`, 
+                        prin_comp1,
                      data = df,
@@
   mult_race <- glm(MRR ~ 
-                        `Percent Living in Poverty` +
-                        `Percent Without an Internet subscription` +
-                        `Percent Uninsured Under 65 Years Old` +
-                        `Median Household Income`  +
-                       # `Percent Living with Severe Housing Cost Burden` +
-                        `Percent Living with Food Insecurity`,
+                        prin_comp1,
                      data = df,
@@
 
-pois_race <- glm(`Death Cases` ~ Race +
-                        `Percent Living in Poverty` +
-                       # `Gini Coefficient` + 
-                        `Percent Without an Internet subscription` +
-                        `Percent Uninsured Under 65 Years Old` +
-                        `Median Household Income`  +
-                        `Percent Living with Food Insecurity` +
-                        #`Percent Living with Severe Housing Cost Burden` +
+pois_race <- glm(`Death Cases` ~ Race + region +
+                        prin_comp1 +
                      offset(log(`Population`)),
@@
 
-## function to fit GLM and pull AIC
+## function to fit Poisson and pull AIC
 pois_aic <- function(df) {
-  pois_race <- glm(`Death Cases` ~  Race +
-                        `Percent Living in Poverty` +
-                      #  `Gini Coefficient` + 
-                        `Percent Without an Internet subscription` +
-                        `Percent Uninsured Under 65 Years Old` +
-                        `Median Household Income`  +
-                        `Percent Living with Food Insecurity` +
-                       # `Percent Living with Severe Housing Cost Burden` +
+  pois_race <- glm(`Death Cases` ~  Race + region +
+                         prin_comp1  +
                      offset(log(`Population`)),
@@
 nb_function  <- function(df) {
-  neg_bin_race <- glm.nb(`Death Cases` ~ Race +
-                        `Percent Living in Poverty` +
-                       # `Gini Coefficient` + 
-                        `Percent Without an Internet subscription` +
-                        `Percent Uninsured Under 65 Years Old` +
-                        `Median Household Income`  +
-                        `Percent Living with Food Insecurity`  +
-                       # `Percent Living with Severe Housing Cost Burden` +
+  neg_bin_race <- glm.nb(`Death Cases` ~  Race + region +
+                         prin_comp1  +
                      offset(log(`Population`)),
@@
 
-## function to fit GLM and pull AIC
+## function to fit nb and pull AIC
 nb_aic <- function(df) {
-  neg_bin_race <- glm.nb(`Death Cases` ~ Race +
-                        `Percent Living in Poverty` +
-                       # `Gini Coefficient` + 
-                        `Percent Without an Internet subscription` +
-                        `Percent Uninsured Under 65 Years Old` +
-                        `Median Household Income`  +
-                        `Percent Living with Food Insecurity`  +
-                        #`Percent Living with Severe Housing Cost Burden` +
+  neg_bin_race <- glm.nb(`Death Cases` ~ Race + region +
+                         prin_comp1  +
                      offset(log(`Population`)),
@@
 
-  car_race <- spautolm(MRR ~                      
-                        `Percent Living in Poverty` +
-                       `Median Household Income` + 
-                        `Percent Without an Internet subscription` +
-                        `Percent Uninsured Under 65 Years Old` +
-                     #   `Percent Unemployed`  +
-                     #   `Percent Living with Severe Housing Cost Burden` +
-                        `Percent Living with Food Insecurity`, 
+  car_race <- spautolm(MRR ~   prin_comp1, 
                            data = df,
@@
 
+##pulling AIC from CAR model
 car_aic <- function(df) {
@@
   car_race <- spautolm(MRR ~                      
-                        `Percent Living in Poverty` +
-                       `Median Household Income` + 
-                        `Percent Without an Internet subscription` +
-                        `Percent Uninsured Under 65 Years Old` +
-                   #     `Percent Unemployed`  +
-                      #  `Percent Living with Severe Housing Cost Burden` +
-                        `Percent Living with Food Insecurity`, 
+                        prin_comp1,
                            data = df,
@@
 aic_df <- data.frame(AIC = c(df_mult, df_pois, df_nb, df_car),
-                     Model = c("GLM", "Poisson", "Negative Binomial", "CAR"),
+                     Model = c("Multiple Linear Regression", "Poisson", "Negative Binomial", "Conditional Autoregressive"),
                      Outcome = c("Ratio", "Count", "Count", "Ratio"))
-aic_df <- aic_df %>% mutate(AIC  = round(AIC, 2))
+
+aic_df <- aic_df %>%
+  mutate(AIC  = round(AIC, 2))
+
+```
+
+```{r}
+##Making kable table of model results
+library(kableExtra)
+library(knitr)
+table_fit <- aic_df %>%
+  select(c(Model, Outcome, AIC)) %>% 
+  kable(caption = "<center>Model Fit</center>",
+        align = "c") %>%
+  kable_minimal() %>%
+  kable_styling(font_size = 30) %>%
+gsub("font-size: initial !important;",
+"font-size: 30pt !important;",
+.)  %>% 
+  landscape()
+
+table_fit %>%  
+  save_kable(here("results/table_fit.png"))
+
 ```
@@
 saveRDS(df, here("data/processed_data/df_fit_alt.rds"))
+```
 
-saveRDS(aic_df, here("data/processed_data/aic_alt.rds"))
+Testing for spatial association. 
+```{r}
+#setting up matrix
+  h_race_df <- matrix(0, nrow(df), nrow(df))
+  h_race_df[cbind(match(us_nabs$StateCode, df$`State Abbreviation`),
+  match(us_nabs$NeighborStateCode, df$`State Abbreviation`))] <- 1
+
+#creating neighbor list 
+  nablist_race_df <- mat2listw(h_race_df, style = "W")
+
+# Using testing for spatial association using Moran's test (global)
+  print(moran.test(df$MRR, nablist_race_df, zero.policy = TRUE, randomisation = FALSE))
+# plot Moran's I
+  moran.plot(df$MRR, nablist_race_df, zero.policy = TRUE)
+  #save plot
+png(here("results/moran_mrr.png"))
+  moran.plot(df$MRR, nablist_race_df, zero.policy = TRUE)
+dev.off()
+
+  # calculate and print LISA statistics (local)
+lisa_race <- localmoran(df$MRR, nablist_race_df, zero.policy=TRUE)
+  summary(lisa_race[,1]) 
 ```

### 2023-04-30 code/analysis_code/1_EDA_alt.Rmd
Auto stash before revert of "fixed pca and models"

@@
 ```{r}
-ggplot(race_sub, aes(y = MRR, x = `Percent Living in Poverty`)) +
+bi_1 <- ggplot(race_sub, aes(y = MRR, x = `Percent Living in Poverty`)) +
   geom_smooth(method = "lm") + 
   geom_point() + 
-  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
-       Percent Living in Poverty") + 
+  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR)
+  vs. Percent Living in Poverty") + 
   stat_cor(method = "pearson") 
 #  pass 
-ggsave(here("results/bi_pov.png"), units="in", width=10, height=7, dpi=300)
+#ggsave(here("results/bi_pov.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
   geom_point() + 
-  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
-       Percent Unemployed") + 
+  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR)
+  vs. Percent Unemployed") + 
   stat_cor(method = "pearson") 
-ggsave(here("results/bi_unemployed.png"), units="in", width=10, height=7, dpi=300)
 #  did not pass 
@@
   geom_point() + 
-  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
-       Gini Coefficient") +
+  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR)
+  vs. Gini Coefficient") +
   stat_cor(method = "pearson") 
-#"pass"
-ggsave(here("results/bi_gini.png"), units="in", width=10, height=7, dpi=300)
+# did not "pass"
+#ggsave(here("results/bi_gini.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
 #  did not pass
-ggsave(here("results/bi_housing.png"), units="in", width=10, height=7, dpi=300)
+#ggsave(here("results/bi_housing.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
 ```{r}
-ggplot(race_sub, aes(y = MRR, x = `Percent Without an Internet subscription`)) +
+bi_3 <- ggplot(race_sub, aes(y = MRR, x = `Percent Without an Internet subscription`)) +
   geom_smooth(method = "lm") +
   geom_point() + 
-  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
-       Percent Without an Internet Subscription") + 
+  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR)
+  vs.  Percent Without an Internet Subscription") + 
   stat_cor(method="pearson")
 # "pass"
-ggsave(here("results/bi_no_internet.png"), units="in", width=10, height=7, dpi=300)
+#ggsave(here("results/bi_no_internet.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
 ```{r}
-ggplot(race_sub, aes(y = MRR, x = `Percent Uninsured Under 65 Years Old`)) +
+bi_4 <- ggplot(race_sub, aes(y = MRR, x = `Percent Uninsured Under 65 Years Old`)) +
   geom_smooth(method = "lm") +
   geom_point() + 
-  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
-       Percent Uninsured Under 65 Years Old") + 
+  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) 
+  vs.  Percent Uninsured Under 65 Years Old") + 
   stat_cor(method="pearson")
 # "pass"
-ggsave(here("results/bi_uninsured.png"), units="in", width=10, height=7, dpi=300)
+#ggsave(here("results/bi_uninsured.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
 ```{r}
-ggplot(race_sub, aes(y = MRR, x = `Percent Living with Food Insecurity`)) +
+bi_5 <- ggplot(race_sub, aes(y = MRR, x = `Percent Living with Food Insecurity`)) +
   geom_smooth(method = "lm") +
   geom_point() + 
-  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
-       Percent Living with Food Insecurity") + 
+  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR)
+  vs. Percent Living with Food Insecurity") + 
   stat_cor(method="pearson")
 # pass"
-ggsave(here("results/bi_foodinsecure.png"), units="in", width=10, height=7, dpi=300)
+#ggsave(here("results/bi_foodinsecure.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
 ```{r}
-ggplot(race_sub, aes(y = MRR, x = `Median Household Income`)) +
+bi_2 <- ggplot(race_sub, aes(y = MRR, x = `Median Household Income`)) +
   geom_smooth(method = "lm") +
   geom_point() + 
-  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
-       Median Household Income") + 
+  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR)
+  vs. Median Household Income") + 
   stat_cor(method="pearson")
@@
 ```{r}
-ggplot(race_sub, aes(y = MRR, x = `Percentage Diagnosed Late`)) +
+bi_6 <-ggplot(race_sub, aes(y = MRR, x = `Percentage Diagnosed Late`)) +
   geom_smooth(method = "lm") +
   geom_point() + 
-  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
-       Percentage Diagnosed Late") + 
+  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR)
+  vs.  Percentage Diagnosed Late") + 
   stat_cor(method="pearson")
-# did not"pass"
+# "pass"
 ```
@@
   geom_point() + 
-  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
-       Percentage Virally Suppressed") + 
+  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) 
+  vs.    Percentage Virally Suppressed") + 
   stat_cor(method="pearson")
@@
 ```
+
+```{r}
+library(gridExtra)
+bi_g <- grid.arrange(bi_1, bi_2, bi_4, bi_5, bi_3, bi_6, ncol = 2, nrow = 3)
+
+ggsave(here("results/grid_bivar.png"), bi_g, units="in", width=10, height=10, dpi=300) 
+```
+
 Now that we have an idea of the relationship between our outcome and potential predictors, next I need to account for the fact that these variables a likely have some degree of multicollinearity.
@@
 pos_predictors <- race_sub %>% 
-  select(c("State", 
-           "Race",
-           "Gini Coefficient",
+  select(c("Median Household Income",
            "Percent Living in Poverty",
@@
            "Percent Uninsured Under 65 Years Old",
-           "Percent Living with Food Insecurity"))
+           "Percent Living with Food Insecurity",
+           "Percentage Diagnosed Late")) %>% 
+  names()
 ```

### 2023-04-30 code/analysis_code/2_pca_alt.Rmd
Auto stash before revert of "fixed pca and models"

--- 2023-04-25 code/analysis_code/2_pca_alt.Rmd
@@
 library(broom)
+library(pls)
 
@@
 pos_predictors <- readRDS(here("data/processed_data/pos_predictors.rds"))
-
+sub <-race_sub %>% 
+  select(contains(c("State", "Race", "Ratio")), matches(pos_predictors))
 ```
@@
 ```{r}
-race_sub <- race_sub %>%
-    select_if(~ !any(is.na(.))) %>% 
-  #filter(Race == "Black") %>% 
-  ungroup()
+#Setting up correlation matrix
+pca_1 <- race_sub %>% 
+  select(all_of(pos_predictors)) %>% 
+  select(-contains("Late")) %>%  ##poorly correlated
+  cor() %>% 
+  ggcorrplot()
 
-  #separating the y variables
-df_response <- race_sub %>%
-    select(contains(c("State", "Ratio", "MRR", "Race", "Death", "Population")))
+# selects variables of interest and scale
+pca <- race_sub %>% 
+  select(matches(pos_predictors)) %>% 
+  select(-contains("Late")) %>% 
+  princomp(cor = TRUE)
 
-  # pulling variables that were correlated with MRR, and thenp scaling
-pos_predictors_scaled <- pos_predictors %>% 
-    ungroup() %>% 
-    mutate(across(where(is.numeric), scale))
-
-#Setting up correlation matrix
-corr_matrix <- pos_predictors_scaled %>% 
-    select(c(3:9)) %>% 
-    cor()
-
-  #setting up correlation matrix 
-print(ggcorrplot(corr_matrix))
-ggsave(here("results/corr_mat.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
 ```{r}
-#applying pca
-pca <- princomp(corr_matrix)
 print(summary(pca))
+
+#Scree plot
+pca_2 <-print(fviz_eig(pca, addlabels = TRUE))
 
@@
 
-#Scree plot
-print(fviz_eig(pca, addlabels = TRUE))
-ggsave(here("results/scree_plot.png"), units="in", width=10, height=7, dpi=300)
+#Storing PCA scores to use for fitting model
+prin_comp1 <- pca$scores[,1]
+
+## save
+saveRDS(prin_comp1, here("data/processed_data/prin_comp1.rds"))
 ```
-Since we originally had 11 variables, we should only look at components that explain more than 9.1% of variation.
-We can see from the scree plot that principal component 1 explained 73.3% of variation, and components 1 and 2 explained 94.9% of the variation, and components 1 - 2 explained 91.1%.
-All other principal components explains less that 8.9% of variation, so they will excluded.
+
+Since we originally had 6 variables, we should only look at components that (individually) explain more than 16.67% of variation.
+This is includes principal component 1 and 2.
+All other principal components will be excluded.
 
@@
 #Contribution of each variable 
-print(fviz_cos2(pca, choice = "var", axes = 1:2))
-ggsave(here("results/contribution_plot.png"), units="in", width=10, height=7, dpi=300)
+pca_3 <-print(fviz_cos2(pca, choice = "var", axes = 1:2))
 
 # Biplot of the attributes
-print(fviz_pca_var(pca, col.var = "cos2",
+pca_4 <- print(fviz_pca_var(pca, col.var = "cos2",
                    gradient.cols = c("darkblue", "orange", "green"),
@@
                    ))
-ggsave(here("results/biplot.png"), units="in", width=10, height=7, dpi=300)
 # Variables in green are a good representation of the variable on that component.
-
-#Creating object to store scaled dataframes
-df_scaled <- full_join(df_response, pos_predictors_scaled, by = join_by("State", "Race"))
 ```
@@
 ```{r}
-saveRDS(df_scaled, here("data/processed_data/df_scaled_alt.rds"))
+library(gridExtra)
+grid_pca <- grid.arrange(pca_1, pca_2, pca_3, pca_4, ncol = 2, nrow = 2)
+
+ggsave(here("results/grid_pca.png"), grid_pca, units="in", width=10, height=10, dpi=300)
 ```

### 2023-04-30 code/analysis_code/p4_mapping.Rmd
Auto stash before revert of "fixed pca and models"

@@
 
-df_scaled <- readRDS(here("data/processed_data/df_fit_alt.rds"))
+df_fit <- readRDS(here("data/processed_data/df_fit_alt.rds"))
 df_raw <- readRDS(here("data/processed_data/race_alt.rds"))
@@
 ```{r}
-df_scaled <- df_scaled %>%
+df_fit <- df_fit %>%
   mutate(region = tolower(State)) %>% 
@@
          observed_predicted_nb = ifelse(!is.finite(observed_predicted_nb), 0, observed_predicted_nb))
-
-df_raw <- df_raw %>%
-  mutate(region = tolower(State))
 ```
@@
 ```{r}
-df_sf_scale <- merge(states,df_scaled, by = "region")
+df_sf_fit<- merge(states,df_fit, by = "region")
 # This next step is import to ensure that no "holes" appear in the map.
-df_sf_scale <- df_sf_scale[order(df_sf_scale$order), ]
-
-df_sf_raw <- merge(states,df_raw, by = "region")
-# This next step is import to ensure that no "holes" appear in the map.
-df_sf_raw <- df_sf_raw[order(df_sf_raw$order), ]
+df_sf_fit <- df_sf_fit[order(df_sf_fit$order), ]
 ```
@@
 ```{r}
-ggplot(df_sf_scale, aes(long, lat)) +
+ggplot(df_sf_fit, aes(long, lat)) +
   geom_polygon(aes(group = group, fill =`Death Rate (estimated)`)) +
@@
 ```{r}
-ggplot(df_sf_scale, aes(long, lat)) +
+ggplot(df_sf_fit, aes(long, lat)) +
   geom_polygon(aes(group = group, fill =`MRR`)) +
@@
 ```{r}
-ggplot(df_sf_raw, aes(long, lat)) +
-  geom_polygon(aes(group = group, fill =`Gini Coefficient`)) +
+map_1 <- ggplot(df_sf_fit, aes(long, lat)) +
+  geom_polygon(aes(group = group, fill =`Percent Living in Poverty`)) +
   coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
@@
         legend.position = "bottom") +
-  labs(title = "Gini Coefficient, 2020", x = "", y = "", fill = "Gini coefficient (higher values represent greater income inequality)")  +
+  labs(title = "Percent Living in Poverty, 2020", x = "", y = "", fill = "Percent")  +
   theme(plot.title= element_text(hjust = 0.5))
-ggsave(here("results/gini_map.png"), units="in", width=10, height=7, dpi=300)
+
 ```
@@
 ```{r}
-ggplot(df_sf_raw, aes(long, lat)) +
+map_2 <-ggplot(df_sf_fit, aes(long, lat)) +
+  geom_polygon(aes(group = group, fill =`Median Household Income`/1e4)) +
+  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
+  theme(strip.text = element_text(size = 14),
+        legend.position = "bottom") +
+  labs(title = "Median Household Income, 2020", x = "", y = "", fill = "Per $10,000")  +
+  theme(plot.title= element_text(hjust = 0.5))
+
+```
+
+```{r}
+map_3 <- ggplot(df_sf_fit, aes(long, lat)) +
   geom_polygon(aes(group = group, fill =`Percent Uninsured Under 65 Years Old`)) +
@@
         legend.position = "bottom") +
-  labs(title = "Percent Uninsured Under 65 Years Old, 2020", x = "", y = "", fill = "%") +
+  labs(title = "Percent Uninsured Under 65 Years Old, 2020", x = "", y = "", fill = "Percent") +
   theme(plot.title= element_text(hjust = 0.5))
-ggsave(here("results/uninsured_map.png"), units="in", width=10, height=7, dpi=300)
+
 ```
@@
 ```{r}
-ggplot(df_sf_raw, aes(long, lat)) +
+map_4 <- ggplot(df_sf_fit, aes(long, lat)) +
   geom_polygon(aes(group = group, fill =`Percent Without an Internet subscription`)) +
@@
         legend.position = "bottom") +
-  labs(title = "Percent Without an Internet Subscription, 2020", x = "", y = "", fill = "%") +
+  labs(title = "Percent Without an Internet Subscription, 2020", x = "", y = "", fill = "Percent") +
   theme(plot.title= element_text(hjust = 0.5))
-ggsave(here("results/no_internet_map.png"), units="in", width=10, height=7, dpi=300)
+
+```
+
+```{r}
+library(gridExtra)
+grid_map<- grid.arrange(map_2, map_1, map_3, map_4, ncol = 2, nrow = 2)
+
+ggsave(here("results/grid_map.png"), grid_map, units="in", width=11, height=10, dpi=300)
 ```
@@
 ```{r}
-ggplot(df_sf_scale, aes(long, lat)) +
+ggplot(df_sf_fit, aes(long, lat)) +
   geom_polygon(aes(group = group, fill = Exp_DR)) +
@@
 ```{r}
-ggplot(df_sf_scale, aes(long, lat)) +
+ggplot(df_sf_fit, aes(long, lat)) +
   geom_polygon(aes(group = group, fill = dykes_chi)) +
@@
 ##Smoothed map using fitted values from best model (CAR)
-ggplot(df_sf_scale, aes(long, lat)) +
+car_smooth <- ggplot(df_sf_fit, aes(long, lat)) +
   geom_polygon(aes(group = group, fill = car_fit)) +
@@
         legend.position = "bottom") +
-  labs(title = "Smoothed Map of Black-to-White HIV-Related Mortality Rate Ratio,
-United States, 2020 (CAR)", x = "Ratio", y = "", fill = "") +
+  labs(title = "Smoothed Map of 
+Black-to-White HIV-Related MRR (CAR)
+United States, 2020", x = "Ratio", y = "", fill = "") +
   theme(plot.title= element_text(hjust = 0.5))
 
-ggsave(here("results/car_fit_map.png"), units="in", width=10, height=4, dpi=300)
 ```
@@
 ##Smoothed map using fitted values from MULT (only marginally worse than CAR)
-ggplot(df_sf_scale, aes(long, lat)) +
+mult_smooth <-ggplot(df_sf_fit, aes(long, lat)) +
   geom_polygon(aes(group = group, fill = mult_fit)) +
@@
         legend.position = "bottom") +
-  labs(title = "Smoothed Map of Black-to-White HIV-Related Mortality Rate Ratio,
-United States, 2020 (GLM)", x = "Ratio ", y = "", fill = "") +
+  labs(title = "Smoothed Map of 
+Black-to-White HIV-Related MRR (GLM)
+United States, 2020", x = "Ratio ", y = "", fill = "") +
   theme(plot.title= element_text(hjust = 0.5))
 
-ggsave(here("results/mult_fit_map.png"), units="in", width=10, height=4, dpi=300)
+grid_smooth <- grid.arrange(car_smooth, mult_smooth, ncol = 2, nrow = 1)
+ggsave(here("results/grid_smooth.png"), grid_smooth, units="in", width=10, height=5, dpi=300)
 ```
@@
 ```{r}
-ggplot(df_sf_scale, aes(long, lat)) +
+pred_mult <- ggplot(df_sf_fit, aes(long, lat)) +
   geom_polygon(aes(group = group, fill = predicted_observed_mult)) +
@@
         legend.position = "bottom") +
-  labs(title = "Map of Predicted Over Observed Black-to-White HIV-Related Mortality Rate Ratio,
+  labs(title = "Predicted Over Observed
+Black-to-White HIV-Related MRR
 United States, 2020 (GLM)", x = "", y = "", fill = "Ratio") +
   theme(plot.title= element_text(hjust = 0.5))
-ggsave(here("results/mult_pred_map.png"), units="in", width=7, height=4, dpi=300)
 
-ggplot(df_sf_scale, aes(long, lat)) +
+pred_car <-ggplot(df_sf_fit, aes(long, lat)) +
   geom_polygon(aes(group = group, fill = predicted_observed_car)) +
@@
         legend.position = "bottom") +
-  labs(title = "Map of Predicted Over Observed Black-to-White HIV-Related Mortality Rate Ratio,
+  labs(title = "Predicted Over Observed
+Black-to-White HIV-Related MRR
 United States, 2020 (CAR)", x = "", y = "", fill = "Ratio") +
   theme(plot.title= element_text(hjust = 0.5))
-ggsave(here("results/car_pred_map.png"), units="in", width=7, height=4, dpi=300)
+```
+
+```{r}
+grid_pred <- grid.arrange(pred_car, pred_mult, ncol = 2, nrow = 1)
+
+ggsave(here("results/grid_pred.png"), grid_pred, units="in", width=10, height=5, dpi=300)
 ```

### 2023-04-30 products/manuscript/manuscript.qmd
Auto stash before revert of "fixed pca and models"

@@
 In order to assess spatial associations, this analysis was limited to the contiguous United States.
-Spatial object of the contiguous United States was obtained from library(maps) in R and a list of neighboring contiguous United States was obtained from a public repository on [GitHub](https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state).
+Spatial object of the contiguous United States was obtained from `library(maps)` in R and a list of neighboring contiguous United States was obtained from a public repository on [GitHub](https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state).
 [@ubikuity2022]
@@
 Results of the PCA shows that the first principal component contributed for 91.7 % of variation between predictors.
-The variables `Percent Uninsured Under 65 Years Old` explained the least amountof variation.
+All other components were excluded as they explained less than 20% of variation.
+The variables `Percent Uninsured Under 65 Years Old` explained the least amount of variation.
 Ultimately, only `Median Household Income`, `Percent Living in Poverty`, `Percent Living with Food Insecurity`, and `Percent Without an Internet Subscription` were included as predictors in the full model.
@@
 ## Software
+
 ```{r}
@@
 ```
+
 {{< pagebreak >}}

### 2023-04-30 reviews/kailin_chen.qmd
Revert "Merge remote-tracking branch 'origin/master'"
This reverts commit 45cc95d9723fa752afab1fe3a55bc0b0a6700d26.

---
title: Project Review Template 
author: Kailin (Kai) Chen
date: "`r file.mtime(knitr::current_input())`"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---
# Overview

**Title of Project**: "Spatial Investigation and Exploration of Factors influencing HIV-related Mortality"

**Name of Project Author(s)**: Hayley Hemme

**Name of Project Reviewer**: Kailin (Kai) Chen

# Specific Project Content Evaluation

## Background, Context, and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments
Project context is well-described with a comprehensive background that includes summary of existing literature. It is well-placed in the context of existing literature and clearly hopes to provide an assessment of spatial correlation between Black-to-White HIV-related mortality rate and social determinants of health and HIV-related outcomes.

### Summary Assessment
*   Strong contextualization and motivation

## Question Description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments
The questions and hypotheses for this project are clear, and it is also clear how the questions relate to the data.

### Summary Assessment
*   Question/hypotheses fully clear

## Data Description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments
The data is described very well with multiple sources, most of them containing direct links. No other meta-information appears to be available.

### Summary Assessment
*   Source and overall structure of data well explained

## Data Wrangling and Exploratory Analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments
Overall, the data wrangling procedure was straightforward. Data cleaning and processing steps were reasonable and well-explained, and the data itself was cleaned well. Meaningful exploratory results were shown in the manuscript.

### Summary Assessment
*   Essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments
Analysis was done properly, and analysis methods were appropriate for data. Different components of the analysis were done in a suitable manner and well-explained.

### Summary Assessment
*   Strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments
Results are presented well, with tables and figures of publication-level quality.

### Summary Assessment
*   Results are very well presented

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments
This section is missing.

### Summary Assessment
*   Major parts of discussion missing or wrong

## Further Comments
I was very impressed with the quality of the earlier sections of your manuscript. Please work on Discussion/Conclusion.

# Overall Project Content Evaluation

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments
The project is well-structured overall, with pertinent files in appropriately-labeled folders that indicate workflow.

### Summary Assessment
*   Well structured

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

Across the different files for this analysis, the project is well-documented, and I was able to understand the majority of steps for the analysis.

### Summary Assessment
*   Fully and well documented

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

Portions of `1_EDA_alt.Rmd`, `2_pca_alt.rmd`, and `3_model_alt.Rmd` are not reproducible without manual intervention due to the same type of coding error.

### Summary Assessment
*   Small parts not reproducible or required manual intervention

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

The study overall had a high level of thoroughness, with multiple statistical models being considered. Questions and hypotheses were thoroughly addressed.

### Summary Assessment
*   Strong level of thoroughness

## Further Comments
A good way to check reproducibility is to completely clear the environment and run all qmd files in the way you intend them to be run.

### 2023-04-30 tes2.R

@@ -1,26 +1,0 @@
-# Load necessary libraries
-library(spdep)
-library(spatialreg)
-
-# Load example data (Columbus data from the spdep package)
-data(columbus)
-
-# Compute the correlation matrix
-columbus_cor <- cor(columbus)
-
-# Perform PCA using princomp
-pca <- princomp(columbus_cor)
-
-# Extract the first two principal components
-pc1 <- pca$scores[, 1]
-pc2 <- pca$scores[, 2]
-
-# Create a spatial weights matrix using inverse distance
-d <- dnearneigh(coords, d1 = 0, d2 = 1)
-w <- nb2listw(d, style = "W")
-
-# Fit a CAR regression model using the first two principal components
-car_model <- spautolm(CRIME ~ pc1 + pc2, data = columbus, listw = w)
-summary(car_model)
-
-pc1

### 2023-04-30 test1.R

@@ -1,34 +1,0 @@
-library(spdep)
-library(spatialreg)
-library(rgdal)
-library(here)
-df <- readRDS(here("data/processed_data/race_sub_scaled.rds"))
-prin_comp1 <- readRDS( here("data/processed_data/prin_comp1.rds"))
-
-us_nabs <- readRDS(here("data/processed_data/us_nabs.rds"))
-
-#Subset to contiguous states
-us_nabs <- us_nabs %>%
-  filter(!grepl("AK", StateCode),
-         !grepl("HI", StateCode), 
-         !grepl("DC", StateCode))
-
-
-h_race_df <- matrix(0, nrow(df), nrow(df))
-h_race_df[cbind(match(us_nabs$StateCode, df$`State Abbreviation`),
-                match(us_nabs$NeighborStateCode, df$`State Abbreviation`))] <- 1
-
-nablist_race_df <- mat2listw(h_race_df, style = "W")
-
-
-car_model <- spautolm(MRR ~ prin_comp1, 
-                      data = df,
-                      listw = nablist_race_df,
-                      family = "CAR", zero.policy = TRUE)
-
-car_race <- spautolm(MRR ~                      
-                       pc1,
-                     data = df,
-                     listw = nablist_race_df,
-                     family = "CAR")
-print(summary(car_race))

### 2023-04-30 code/analysis_code/1_EDA_alt.Rmd
hopefully this works...

@@
 pos_predictors <- race_sub %>% 
-  select(c("Median Household Income",
+  select(c("State", 
+           "Race",
+           "Median Household Income",
            "Percent Living in Poverty",
@@
            "Percent Living with Food Insecurity",
-           "Percentage Diagnosed Late")) %>% 
-  names()
+           "Percentage Diagnosed Late"))
 ```

### 2023-04-30 code/analysis_code/2_pca_alt.Rmd
hopefully this works...

--- 2023-04-25 code/analysis_code/2_pca_alt.Rmd
@@
   #setting up correlation matrix 
-print(ggcorrplot(corr_matrix))
-ggsave(here("results/corr_mat.png"), units="in", width=10, height=7, dpi=300)
+pca_1 <-print(ggcorrplot(corr_matrix))
+#ggsave(here("results/corr_mat.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
 #Scree plot
-print(fviz_eig(pca, addlabels = TRUE))
-ggsave(here("results/scree_plot.png"), units="in", width=10, height=7, dpi=300)
+pca_2 <-print(fviz_eig(pca, addlabels = TRUE))
+#ggsave(here("results/scree_plot.png"), units="in", width=10, height=7, dpi=300)
 ```
@@
 #Contribution of each variable 
-print(fviz_cos2(pca, choice = "var", axes = 1:2))
-ggsave(here("results/contribution_plot.png"), units="in", width=10, height=7, dpi=300)
+pca_3 <-print(fviz_cos2(pca, choice = "var", axes = 1:2))
+#ggsave(here("results/contribution_plot.png"), units="in", width=10, height=7, dpi=300)
 
 # Biplot of the attributes
-print(fviz_pca_var(pca, col.var = "cos2",
+pca_4 <- print(fviz_pca_var(pca, col.var = "cos2",
                    gradient.cols = c("darkblue", "orange", "green"),
@@
                    ))
-ggsave(here("results/biplot.png"), units="in", width=10, height=7, dpi=300)
+#ggsave(here("results/biplot.png"), units="in", width=10, height=7, dpi=300)
 # Variables in green are a good representation of the variable on that component.
@@
 ```{r}
+library(gridExtra)
+grid_pca <- grid.arrange(pca_1, pca_2, pca_3, pca_4, ncol = 2, nrow = 2)
+
+ggsave(here("results/grid_pca.png"), grid_pca, units="in", width=10, height=10, dpi=300)
+```
+
+```{r}
 saveRDS(df_scaled, here("data/processed_data/df_scaled_alt.rds"))

### 2023-04-30 code/analysis_code/3_model_alt.Rmd
hopefully this works...

--- 2023-04-25 code/analysis_code/3_model_alt.Rmd
@@
   mult_race <- glm(MRR ~ 
-                       # `Percent Living in Poverty` +
-                        `Gini Coefficient` + 
+                        `Median Household Income` +
+                        `Percent Living in Poverty` +
                         `Percent Without an Internet subscription` +
-                        `Percent Uninsured Under 65 Years Old`,
-                      #  `Percent Living with Food Insecurity`,
+                        `Percent Living with Food Insecurity`,
                      data = df,
@@
 pois_race <- glm(`Death Cases` ~ Race + region +
-                       # `Percent Living in Poverty` +
-                        `Gini Coefficient` + 
+                        `Median Household Income` +
+                        `Percent Living in Poverty` +
                         `Percent Without an Internet subscription` +
-                        `Percent Uninsured Under 65 Years Old` +
-                      #  `Percent Living with Food Insecurity`  +
+                        `Percent Living with Food Insecurity`  +
                      offset(log(`Population`)),
@@
   pois_race <- glm(`Death Cases` ~  Race + region +
-                       # `Percent Living in Poverty` +
-                        `Gini Coefficient` + 
+                        `Median Household Income` +
+                        `Percent Living in Poverty` +
                         `Percent Without an Internet subscription` +
-                        `Percent Uninsured Under 65 Years Old` +
-                      #  `Percent Living with Food Insecurity`  +
+                        `Percent Living with Food Insecurity`  +
                      offset(log(`Population`)),
@@
   neg_bin_race <- glm.nb(`Death Cases` ~ Race + region +
-                       # `Percent Living in Poverty` +
-                        `Gini Coefficient` + 
+                        `Median Household Income` +
+                        `Percent Living in Poverty` +
                         `Percent Without an Internet subscription` +
-                        `Percent Uninsured Under 65 Years Old` +
-                      #  `Percent Living with Food Insecurity`  +
+                        `Percent Living with Food Insecurity` +
                      offset(log(`Population`)),
@@
   car_race <- spautolm(MRR ~                      
-                       # `Percent Living in Poverty` +
-                        `Gini Coefficient` + 
+                        `Median Household Income` +
+                        `Percent Living in Poverty` +
                         `Percent Without an Internet subscription` +
-                        `Percent Uninsured Under 65 Years Old`,
-                      #  `Percent Living with Food Insecurity`,
+                        `Percent Living with Food Insecurity`,
                            data = df,
@@
   # calculate and print LISA statistics
-  lisa_race <- localmoran(df$MRR, nablist_race_df)
+  lisa_race <- localmoran(df$MRR, nablist_race_df, zero.policy=TRUE)
   summary(lisa_race[,1])

### 2023-04-30 code/analysis_code/p4_mapping.Rmd
hopefully this works...

--- 2023-04-25 code/analysis_code/p4_mapping.Rmd
@@
 ```{r}
-ggplot(df_sf_raw, aes(long, lat)) +
-  geom_polygon(aes(group = group, fill =`Gini Coefficient`)) +
+map_1 <- ggplot(df_sf_raw, aes(long, lat)) +
+  geom_polygon(aes(group = group, fill =`Percent Living in Poverty`)) +
   coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
@@
 ```{r}
+map_2 <-
+
   ggplot(df_sf_raw, aes(long, lat)) +
+  geom_polygon(aes(group = group, fill =`Median Household Income`/1e4)) +
+  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
+  theme(strip.text = element_text(size = 14),
+        legend.position = "bottom") +
+  labs(title = "Median Household Income, 2020", x = "", y = "", fill = "Per $10,000")  +
+  theme(plot.title= element_text(hjust = 0.5))
+
+```
+
+```{r}
+map_3 <- ggplot(df_sf_raw, aes(long, lat)) +
   geom_polygon(aes(group = group, fill =`Percent Uninsured Under 65 Years Old`)) +
@@
 ```{r}
-ggplot(df_sf_raw, aes(long, lat)) +
+map_4 <- ggplot(df_sf_raw, aes(long, lat)) +
   geom_polygon(aes(group = group, fill =`Percent Without an Internet subscription`)) +
@@
 ```{r}
-ggplot(df_sf_scale, aes(long, lat)) +
+pred_mult <- ggplot(df_sf_scale, aes(long, lat)) +
   geom_polygon(aes(group = group, fill = predicted_observed_mult)) +
@@
         legend.position = "bottom") +
-  labs(title = "Map of Predicted Over Observed Black-to-White HIV-Related Mortality Rate Ratio,
+  labs(title = "Predicted Over Observed
+Black-to-White HIV-Related MRR,
 United States, 2020 (GLM)", x = "", y = "", fill = "Ratio") +
   theme(plot.title= element_text(hjust = 0.5))
-ggsave(here("results/mult_pred_map.png"), units="in", width=7, height=4, dpi=300)
 
-ggplot(df_sf_scale, aes(long, lat)) +
+pred_car <-ggplot(df_sf_scale, aes(long, lat)) +
   geom_polygon(aes(group = group, fill = predicted_observed_car)) +
@@
         legend.position = "bottom") +
-  labs(title = "Map of Predicted Over Observed Black-to-White HIV-Related Mortality Rate Ratio,
+  labs(title = "Predicted Over Observed
+Black-to-White HIV-Related MRR,
 United States, 2020 (CAR)", x = "", y = "", fill = "Ratio") +
   theme(plot.title= element_text(hjust = 0.5))
-ggsave(here("results/car_pred_map.png"), units="in", width=7, height=4, dpi=300)
+```
+
+```{r}
+grid_pred <- grid.arrange(pred_car, pred_mult, ncol = 2, nrow = 1)
+
+ggsave(here("results/grid_pred.png"), grid_pred, units="in", width=10, height=5, dpi=300)
 ```

### 2023-04-30 code/analysis_code/1_EDA_alt.Rmd
hopefully this corrects things...

--- 2023-04-30 code/analysis_code/1_EDA_alt.Rmd
@@
 pos_predictors <- race_sub %>%
+  ungroup() %>% 
   select(c("Median Household Income",

### 2023-04-30 code/analysis_code/2_pca_alt.Rmd
hopefully this corrects things...

--- 2023-04-30 code/analysis_code/2_pca_alt.Rmd
@@
 race_sub <- readRDS(here("data/processed_data/race_alt.rds"))
+race_sub <- race_sub %>% ungroup()
 pos_predictors <- readRDS(here("data/processed_data/pos_predictors.rds"))
@@
 pca_1 <- race_sub %>% 
+  ungroup() %>% 
   select(all_of(pos_predictors)) %>% 

### 2023-04-30 code/analysis_code/p4_mapping.Rmd
hopefully this corrects things...

--- 2023-04-30 code/analysis_code/p4_mapping.Rmd
@@
 df_fit <- readRDS(here("data/processed_data/df_fit_alt.rds"))
-df_raw <- readRDS(here("data/processed_data/race_alt.rds"))
 ```
@@
 ```{r}
-df_sf_fit<- merge(states,df_fit, by = "region")
+df_sf_fit <- merge(states,df_fit, by = "region")
 # This next step is import to ensure that no "holes" appear in the map.

### 2023-04-30 reviews/Cavros_Project_Review.qmd
please work..

---
title: Final Project Review Hemme / Cavros 
author: Irene Cavros
date: 04/28/2023
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Spatial Investigation and Exploration of Factors influencing HIV-related Mortality

Name of project author(s): Hayley Hemme

Name of project reviewer: Irene Cavros

# Instructions

Write your comments and feedback below for each section/component of the project. The goal should be to help the author improve their project. Make comments as constructive and actionable as possible. You can provide both criticism and praise.

For each component, pick one summary statement by deleting the ones that do not apply and keeping only the one that you think most closely summarizes a given component.

Make sure your final document compiles/knits into a readable, well-formatted html document.

Delete any sections/text of this template that are not part of your final review document. (Including these instructions.)

# Specific project content evaluation
Evaluate the different parts of the project by filling in the sections below.

## Background, Context and Motivation

How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

Summary/abstract: 
- What you have written is very succinct and clear! 
- Consider adding a few sentences which start to prime us on the actual results of your research and what they mean.

Introduction Section:
- Very clear and well detailed.
- I know it is a fairly commonplace acronym but it may be good to just clarify what the HIV and AIDS abbreviations stand for at the top.

Background section: 
- Great job weaving citations into the text!
- Consider adding a paragraph that is specific to race and HIV.

### Summary assessment (PICK ONE, DELETE THE OTHERS)
* strong contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

Your research question and hypothesis in section 2.1 are clear. You also explain them more informally early on in the manuscript, limiting the possibility for any confusion.

### Summary assessment
* question/hypotheses fully clear

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

Great description of the data sources used for your research! I have no suggestions for improvement.

### Summary assessment
* source and overall structure of data well explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

- Nice description of the cleaning that was done and I liked that you specifically called out details like which variables were suppressed due to small counts.
-Consider adding a high level figure summarizing data before you go into outcome variables.

### Summary assessment
* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

- PCA in the data cleaning step was well described. 
- The explanation in the statistical analysis subsection in section 3 plus the analysis section 4 made sense to me and were very clear.

### Summary assessment
* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

- Table 1 was a nice high level overview of the problem before digging into more complex analyses. 
- I'm not 100% certain what I am looking at in figure 1. Could you explain in greater detail what you are presenting? 
- I think you may have added table 3 twice into the body of the manuscript. It's a nice visual though! 
- Table 2: change title to figure 2 since it is a map. Again, a great visual though!
- Are all 4 figures after table 2 considered as one figure? Also I recommend changing the title to this figure for clarity. 
- When we get to section 5.3 Full analysis, the narrative accompanying your visuals is a bit sparce. I recommend fleshing this out.

### Summary assessment
* results are very well presented

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

- Have you pushed your discussion and conclusions section? They appear blank for me unfortunately.

### Summary assessment
* major parts of discussion missing or wrong

## Further comments

Well done! There are some notes/questions I jotted down in my project review that became clear later in the manuscript but I wanted to keep them in this file anyways because it can give you an idea of what the reader is grasping at what point in the manuscript and what might need further clarification earlier on.

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

- The numbering of code files was a great way to ensure everything is run in the appropriate order.
- There are a few files in your project that I am unsure of their purpose: `outputs_code.Rmd`, `tes2.R`, and `tes1.R`
-Consider changing the folder label for `results` to something like visuals, figures, etc since your full results are elsewhere.

### Summary assessment
* well structured

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

- You do a great job going through your decision making process throughout the analysis section of the manuscript! 
- Some of your code files are extremely well documented like `p4_mapping.Rmd` but others could do with some additional narrative to accompany the code

### Summary assessment
* fully and well documented

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

- All processing code was fully reproducible. 
- Error in `1_EDA_alt` file: line 302 (error in select, unused argument)
- Error in `2_PCA_alt` file: line 13 (error in select, unused argument - was not able to run the rest of the chunks after this error)
- Error in `3_model_alt` file: line 13 (error in select, unused argument)

### Summary assessment
* small parts not reproducible or required manual intervention

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

Your project is very robust in terms of thoroughness! The research question was fully addressed.

### Summary assessment
* strong level of thoroughness

## Further comments

_Add any other comments regarding the overall project here. Write anything you think can help your classmate improve their project._

Wonderful job, Hayley! My comments/suggestions are very minute and perhaps nit picky but it's evident you put a lot of time and effort into this project! Overall, most of my suggestions are just to provide clarity but you really got all the hard parts down. Congratulations on all your excellent work!

### 2023-05-05 code/analysis_code/1_EDA.Rmd
submitting final changes

--- 2023-04-30 code/analysis_code/1_EDA_alt.Rmd
@@
 library(here)
-library(spdep)
-library(GGally)
-library(skimr)
 library(ggfortify)
-library(MASS)
-library(ggcorrplot)
-library("FactoMineR")
 library(tidyverse)
@@
 
+Creating variable for region
 ```{r}
@@
          !grepl("Hawaii", State))
-```
-
-Let's create a variable for mortality rate ratio of Black/ White
-```{r}
-race_sub <- race_sub %>%
-  group_by(State) %>%
-  mutate(MRR = `Death Rate (estimated)`[Race == "Black"] / `Death Rate (estimated)`[Race == "White"])
 ```
@@
         caption = "<center>HIV-Related Deaths by Race/ Ethnicity, United States, 2020</center>",
-        align = "c") %>%
+        align = "c", format.args = list(big.mark = ",")) %>%
   kable_minimal() %>%
-  kable_styling(font_size = 30)%>%
+  kable_styling(font_size = 50)%>%
 gsub("font-size: initial !important;",
-"font-size: 30pt !important;",
+"font-size: 50pt !important;",
 .)  %>% 
@@
                       "Mortality Rate Ratio"),
-        caption = "<center>HIV-Related Deaths by Region, United States, 2020</center>") %>% 
+        caption = "<center>HIV-Related Deaths by Region, United States, 2020</center>", format.args = list(big.mark = ",")) %>% 
   kable_minimal() %>%
-  kable_styling(font_size = 30) %>%
+  kable_styling(font_size = 50) %>%
 gsub("font-size: initial !important;",
-"font-size: 30pt !important;",.)  %>% 
+"font-size: 50pt !important;",.)  %>% 
   landscape()
@@
 
-Next, let's take a look at our response variable
-Univariate Analysis
-
-Let's take a look at some box plots of death rate by races and state. 
-We'll start by looking at death rate given by the AIDSVu dataset. 
-```{r}
-race_sub %>% 
-  ggplot() +
-  geom_boxplot(aes(y = `Death Rate (estimated)`, x = Race, group = Race)) + 
-  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio, United States, 2020", y = "Ratio", x = "Race/ Ethnicity") + theme(legend.position = "none") 
-```
-
-Histogram
+Next we'll look into the distribution of our response variable
 ```{r}
@@
   theme(legend.position = "none") +
-  labs(title = "Histogram of Black-to-White HIV-Related Mortality Rate Ratio, United States, 2020", y = "n", x = "Ratio") + theme_classic()
-ggsave(here("results/MRR_hist.png"), units="in", width=10, height=7, dpi=300)
+  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio, United States, 2020", y = "n", x = "Ratio") + theme_classic()
+ggsave(here("results/MRR_hist.png"), units="in", width=10, height=5, dpi=300)
 
@@
 
+```
+
+Let's also look at the distribution of HIV-related death rate
+```{r}
 race_sub %>% 
   ggplot() +
-  geom_boxplot(aes(y = `Death Rate (estimated)`, x = Race, group = Race)) + 
+  geom_histogram(aes(x = `Death Rate (estimated)`, fill = Race), bins = 20) + 
   theme(legend.position = "none") +
-  labs(title = "HIV-Related Death Rate, by Race, United States, 2020", y = "Rate (per 100,000)", x = "Race") + theme_classic()
-ggsave(here("results/dr_race_bp.png"), units="in", width=10, height=5, dpi=300)
+  labs(title = "HIV-Related Death Rate, United States, 2020", y = "n", x = "Deaths per 100,000") +
+  theme_classic()
+ggsave(here("results/DR_hist.png"), units="in", width=10, height=5, dpi=300)
+
+```
+
+Let's take a look at MRR by state and region
+```{r}
+race_sub %>% 
+  ggplot() +
+  geom_bar(aes(y = `MRR`, x = `State Abbreviation`, group = `State Abbreviation`),
+           stat = "identity", position = "dodge") +
+  theme_classic() + 
+  theme(legend.position = "bottom") +
+  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio, by State, United States, 2020", y = "Ratio", x = "State")
+ggsave(here("results/mrr_state.png"), units="in", width=12, height=5, dpi=300)
 
@@
   ggplot() +
-  geom_bar(aes(y = `Death Rate (estimated)`, x = `State Abbreviation`, group = `State Abbreviation`, fill = Race),
-           stat = "identity", position = "dodge") + theme_classic() + 
+  geom_boxplot(aes(y = `MRR`, x = `Region`, group = `Region`)) +
+  theme_classic() + 
   theme(legend.position = "bottom") +
-  labs(title = "HIV-Related Death Rate of Black and White Persons, by State, United States, 2020", y = "Rate (per 100,000)", x = "State")
-ggsave(here("results/dr_state.png"), units="in", width=12, height=5, dpi=300)
-
-race_sub %>% 
-  ggplot() +
-  geom_boxplot(aes(y = `Death Rate (estimated)`, x = `Region`, group = `Region`)) + theme_classic() + 
-  theme(legend.position = "bottom") +
-  labs(title = "HIV-Related Death Rate of Black and White Persons, by Region, United States, 2020", y = "Rate (per 100,000)", x = "Region") + facet_wrap(.~ Race)
-ggsave(here("results/dr_region.png"), units="in", width=12, height=5, dpi=300)
+  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio, by Region, United States, 2020", y = "Ratio", x = "Region")
+ggsave(here("results/mrr_region.png"), units="in", width=12, height=5, dpi=300)
 ```
@@
 Bivariate Analysis 
-We will deem variables with a correlation coeffiencient of |0.15| as important, and will include these in PCA
+We will deem variables with a correlation coefficient of |0.15| as important, and will include these in PCA
 ```{r}
@@
 pos_predictors <- race_sub %>%
-  select(c("Median Household Income",
+  dplyr::select(c("Median Household Income",
            "Percent Living in Poverty",

### 2023-05-05 code/analysis_code/2_PCA.Rmd
submitting final changes

```{r message=FALSE, warning=FALSE}
library(ggcorrplot)
library(FactoMineR)
library(factoextra)
library(here)
library(knitr)
library(tidyverse)
library(broom)

race_sub <- readRDS(here("data/processed_data/race_alt.rds"))

pos_predictors <- readRDS(here("data/processed_data/pos_predictors.rds"))
```

PCA
```{r}
#Setting up correlation matrix
pca_1 <- race_sub %>% 
  dplyr::select(all_of(pos_predictors)) %>% 
  cor() %>% 
  ggcorrplot()

# selects variables of interest and scale
pca <- race_sub %>% 
  dplyr::select(matches(pos_predictors)) %>% 
  princomp(cor = TRUE)
```

```{r}
print(summary(pca))

#Scree plot
pca_2 <-print(fviz_eig(pca, addlabels = TRUE))

#pca loadings of principal components 1 and 2
print(pca$loadings[, 1:2])

#Storing PCA scores to use for fitting model
prin_comp1 <- pca$scores[,1]
prin_comp2 <- pca$scores[,2]

## save
saveRDS(prin_comp1, here("data/processed_data/prin_comp1.rds"))
saveRDS(prin_comp2, here("data/processed_data/prin_comp2.rds"))
```

Since we originally had 6 variables, we should only look at components that (individually) explain more than 16.7% of variation. This is includes principal component 1 and 2. All other principal components will be excluded.

```{r}
#Contribution of each variable 
pca_3 <-print(fviz_cos2(pca, choice = "var", axes = 1:2))

# Biplot of the attributes
pca_4 <- print(fviz_pca_var(pca, col.var = "cos2",
                   gradient.cols = c("darkblue", "orange", "green"),
                   repel = TRUE, title = "Loading Plot"
                   ))
# Variables in green are a good representation of the variable on that component.
```

```{r}
library(gridExtra)
grid_pca <- grid.arrange(pca_1, pca_2, pca_3, pca_4, ncol = 2, nrow = 2)

ggsave(here("results/grid_pca.png"), grid_pca, units="in", width=10, height=10, dpi=300)
```

### 2023-05-05 code/analysis_code/3_model_fitting.Rmd
submitting final changes

```{r message=FALSE, warning=FALSE}
library(here)
library(mfx)
library(spatialreg)
library(spdep)
library(tidyverse)
library(AER)
```

```{r}
df <- readRDS(here("data/processed_data/race_alt.rds"))
prin_comp1 <- readRDS(here("data/processed_data/prin_comp1.rds"))
prin_comp2 <- readRDS(here("data/processed_data/prin_comp2.rds"))
us_nabs <- readRDS(here("data/processed_data/us_nabs.rds"))

#Subset to contiguous states
us_nabs <- us_nabs %>%
  filter(!grepl("AK", StateCode),
         !grepl("HI", StateCode), 
         !grepl("DC", StateCode))

```

```{r}
states <- data.frame(state.name)
states$region <- case_when(
  states$state.name %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania") ~ "Northeast",
  states$state.name %in% c("Illinois", "Indiana", "Iowa", "Kansas", "Michigan", "Minnesota", "Missouri", "Nebraska", "North Dakota", "Ohio", "South Dakota", "Wisconsin") ~ "Midwest",
  states$state.name %in% c("Alabama", "Arkansas", "Delaware", "Florida", "Georgia", "Kentucky", "Louisiana", "Maryland", "Mississippi", "North Carolina", "Oklahoma", "South Carolina", "Tennessee", "Texas", "Virginia", "West Virginia") ~ "South",
  states$state.name %in% c("Arizona", "California", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Oregon", "Utah", "Washington", "Wyoming") ~ "West"
)

df <- full_join(df, states, by = join_by(State == state.name))

df <- df %>% 
  filter(!grepl("Alaska", State),
         !grepl("Hawaii", State))

# reorder race variable to make White the reference group for NB
df <- df %>% 
  mutate(Race = factor(Race, levels = c("White","Black")))
```

# multiple linear regression
```{r}
mult_race <- glm(MRR ~ 
                        prin_comp1 + prin_comp2,
                     data = df,
                     family = "gaussian")
mult_sum <-   summary(mult_race)
mult_sum

#storing values in table and computing confidence interval
mult_table <- as.tibble(coef(mult_sum)) %>% 
  mutate(low = round(`Estimate` - 1.96 * `Std. Error`, 2),
         high = round(`Estimate` + 1.96 * `Std. Error`, 2),
        value = c("Intercept",
                   "PC1",
                   "PC2")) %>% 
  relocate(value, .before = Estimate) %>%
  relocate(c(low, high), .before = `t value`) %>% 
  unite("95% CI", c("low", "high"), sep = ", ", na.rm = TRUE)

  ##pulling and storing fitted values
df$mult_fit <- mult_race$fitted.values

## pulling AIC
aic_mult <- AIC(mult_race)
```

#poisson
```{r}
pois_race <- glm(`Death Cases` ~ Race + region +
                        prin_comp1 + prin_comp2 + 
                     offset(log(`Population`)),
                     data = df,
                     family = poisson(link = "log"))

#Testing for overdispersion
dispersiontest(pois_race, trafo=1)
#Evidence of overdisperion with c = 17.35. Will instead utilize negative binomial regression, which can account for overdispersion.
```
  #Negative binomial that outputs rate ratios
```{r}
neg_bin_race <- negbinirr(`Death Cases` ~  Race +
                     prin_comp1  + prin_comp2 +
                      offset(log(`Population`)),
                             data = df)
nb_sum <- summary(neg_bin_race$fit)

coef(nb_sum)

#storing values in table and computing confidence interval
nb_table <- as.tibble(coef(nb_sum)) %>% 
  mutate(low = `Estimate` - 1.96 * `Std. Error`,
         high = `Estimate` + 1.96 * `Std. Error`) %>% 
  mutate(across(!contains("z"), exp),
         value = c("Intercept",
                   "Race - Black", 
                   "PC1", 
                   "PC2")) %>% 
  relocate(value, .before = Estimate) %>%
  relocate(c(low, high), .before = `z value`) %>%  
  mutate(across(c("low", "high"), round, 2),
          across(where(is.numeric), ~format(., scientific = FALSE))) %>% 
  unite("95% CI", c(low, high), sep = ", ", na.rm = TRUE)

##pull fitted values to save in main df
df$nb_fit <- neg_bin_race$fit$fitted.values

## pull AIC
aic_nb <- AIC(neg_bin_race$fit)

print(confint(neg_bin_race$fit))

```
Testing for spatial association. 
```{r}
#setting up matrix
  h_race_df <- matrix(0, nrow(df), nrow(df))
  h_race_df[cbind(match(us_nabs$StateCode, df$`State Abbreviation`),
  match(us_nabs$NeighborStateCode, df$`State Abbreviation`))] <- 1

#creating neighbor list 
  nablist_race_df <- mat2listw(h_race_df, style = "W")

# Using testing for spatial association using Moran's test (global)
  print(moran.test(df$MRR, nablist_race_df, zero.policy = TRUE, randomisation = FALSE))
# plot Moran's I
  moran.plot(df$MRR, nablist_race_df, zero.policy = TRUE)
  #save plot
png(here("results/moran_mrr.png"))
  moran.plot(df$MRR, nablist_race_df, zero.policy = TRUE)
dev.off()

  # calculate and print LISA statistics (local)
lisa_race <- localmoran(df$MRR, nablist_race_df, zero.policy=TRUE)
summary(lisa_race[,1]) 
```

#Fitting CAR model (requires neighbors)
```{r}
  h_race_df <- matrix(0, nrow(df), nrow(df))
  h_race_df[cbind(match(us_nabs$StateCode, df$`State Abbreviation`),
                  match(us_nabs$NeighborStateCode, df$`State Abbreviation`))] <- 1

  nablist_race_df <- mat2listw(h_race_df, style = "W")

  car_race <- spautolm(MRR ~   prin_comp1 + prin_comp2, 
                           data = df,
                           listw = nablist_race_df,
                           family = "CAR")

car_sum <-  summary(car_race)
car_sum

##pull fitted values to save in main df
df$car_fit <- car_race$fit$fitted.values

#Pulling estimates and standard errors
car_table <- as.tibble(coef(car_sum)) %>% 
  mutate(low = round(`Estimate` - 1.96 * `Std. Error`, 2),
         high = round(`Estimate` + 1.96 * `Std. Error`, 2),
        value = c("Intercept",
                   "PC1",
                   "PC2")) %>% 
  relocate(value, .before = Estimate) %>%
  relocate(c(low, high), .before = `z value`) %>% 
  unite("95% CI", c("low", "high"), sep = ", ", na.rm = TRUE)

#lambda confidence interval
  car_race$lambda.se
  car_race$lambda + c(-2,2) * car_race$lambda.se

##pulling AIC from CAR model
aic_car <- AIC(car_race)

```

Organizing mod_AIC into new dataframe
```{r}
aic_df <- data.frame(AIC = c(aic_mult, aic_nb, aic_car),
                     Model = c("Multiple Linear Regression", "Negative Binomial", "Conditional Autoregressive"))

aic_df <- aic_df %>%
  mutate(AIC  = round(AIC, 2))

```

```{r}
##Making kable table of model results
library(kableExtra)
library(knitr)
table_fit <- aic_df %>%
  dplyr::select(c(Model, AIC)) %>% 
  kable(caption = "<center>Model Comparison</center>", col.names = c("Model Type", "AIC"),
        align = "c") %>%
  kable_minimal() %>%
  kable_styling(font_size = 50) %>%
gsub("font-size: initial !important;",
"font-size: 50pt !important;",
.)  %>% 
  landscape()

table_fit %>%  
  save_kable(here("results/table_fit.png"))

```

```{r}
##Making kable table of summary results

library(kableExtra)

mult_table %>%
  mutate_if(is.numeric, round, 2) %>%
  mutate(`Pr(>|t|)` = ifelse(`Pr(>|t|)` < 0.005, 
                             cell_spec(paste0("<", formatC(0.005, format = "f", digits = 3)), "latex"),
                `Pr(>|t|)`)) %>%
  kable(caption = "<center>Summary of Multiple Linear Regression Model</center>",
        align = "c",
        col.names = c("Coefficient", "Estimate", "Std. Error", "95% CI", "t value", "p value"), digits = 3) %>%
  kable_minimal() %>%
  kable_styling(font_size = 50) %>%
  gsub("font-size: initial !important;",
        "font-size: 50pt !important;",
        .)  %>% 
  landscape() %>%  
  save_kable(here("results/mult_table.png"))

nb_table %>% 
 # mutate(across(where(is.numeric), ~format(., scientific = FALSE))) %>%
  mutate(across(c(2:3,5:6), as.numeric)) %>%
  mutate(`Pr(>|z|)` = round(`Pr(>|z|)`, 2),
    `Pr(>|z|)` = ifelse(`Pr(>|z|)` < 0.005, 
                             cell_spec(paste0("<", formatC(0.005, format = "f", digits = 3)), "latex"),
                `Pr(>|z|)`)) %>%
  kable(caption = "<center>Summary of Negative Binomial Regression Model</center>",
        align = "c",
        col.names = c("Coefficient", "Estimate", "Std. Error", "95% CI", "z value", "p value"), digits = 2) %>%
  kable_minimal() %>%
  kable_styling(font_size = 50) %>%
  gsub("font-size: initial !important;",
        "font-size: 50pt !important;",
        .)  %>% 
  landscape() %>%  
  save_kable(here("results/nb_table.png"))

car_table %>%
  mutate_if(is.numeric, round, 2) %>%
  mutate(`Pr(>|z|)` = ifelse(`Pr(>|z|)` < 0.005, 
                             cell_spec(paste0("<", formatC(0.005, format = "f", digits = 3)), "latex"),
                `Pr(>|z|)`)) %>%
  kable(caption = "<center>Summary of Conditional Autoregressive Model</center>",
        align = "c",
        col.names = c("Coefficient", "Estimate", "Std. Error", "95% CI", "z value", "p value"), digits = 3) %>%
  kable_minimal() %>%
  kable_styling(font_size = 50) %>%
  gsub("font-size: initial !important;",
        "font-size: 50pt !important;",
        .)  %>% 
  landscape() %>%  
  save_kable(here("results/car_table.png"))

```

```{r}
saveRDS(df, here("data/processed_data/df_fit_alt.rds"))
```

### 2023-05-05 code/analysis_code/4_mapping.Rmd
submitting final changes

In this section, I will be creating some maps of the data. I'll start by loading some packages and the dataframes.

```{r}
library(here)
library(tidyverse)
library(RColorBrewer)

#loading a spatial object for the contiguous US.

states <- map_data("state")

df_fit <- readRDS(here("data/processed_data/df_fit_alt.rds"))
```

I do a bit more cleaning to the data to tidy things up and create some new variables for the best fitting model. Spatial methods were very marginally better than nonspatial methods.
```{r}
df_fit <- df_fit %>%
  mutate(region = tolower(State)) %>% 
 # replace(is.na("Ratio Observed/ Expected"), 0) %>% 
  #ungroup() %>% 
  mutate(dykes_chi = ((`Death Cases` - `Expected Death Cases`)/ sqrt(`Expected Death Cases`)),
         ## GLM
         predicted_observed_mult = mult_fit/`MRR`,
         predicted_observed_mult = ifelse(!is.finite(predicted_observed_mult), 0, predicted_observed_mult),
         observed_predicted_mult = `MRR`/mult_fit,
         observed_predicted_mult = ifelse(!is.finite(observed_predicted_mult), 0, observed_predicted_mult),
         ## CAR
         predicted_observed_car = car_fit/`MRR`,
         predicted_observed_car = ifelse(!is.finite(predicted_observed_car), 0, predicted_observed_car),
         observed_predicted_car = `MRR`/car_fit,
         observed_predicted_car = ifelse(!is.finite(observed_predicted_car), 0, observed_predicted_car))
```

Next, I will merge my dataframe contain the variables of interest with the spatial object. 
```{r}
df_sf_fit <- merge(states,df_fit, by = "region")
# This next step is import to ensure that no "holes" appear in the map.
df_sf_fit <- df_sf_fit[order(df_sf_fit$order), ]
```

Now I'm ready to start mapping! I'll begin by looking at the map of the raw HIV death rate across to contiguous US by selected race/ ethnicity.
```{r}
ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Death Rate (estimated)`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "HIV-Related Death Rates
Among Non-Hispanic Black and White Persons
United States, 2020", x = "", y = "", fill = "Death rate per 100,000") +
  scale_fill_distiller(palette = "RdYlGn") +
  theme(plot.title= element_text(hjust = 0.5))

ggsave(here("results/raw_rate_map.png"), units="in", width=10, height=4, dpi=300)
```

The map for HIV-related death differs drastically across the United States by race/ethnicity. In most states, the highest HIV-related death were among Black person. This is especially true in the South region. Non-coastal western states consistently had lowest rates HIV-related deaths across the selected racial/ethnic groups, expect among multi-racial non-Hispanic persons. Non-Hispanic White persons were the only of the selected racial/ ethnic group that did not have a HIV-related death rate higher than ~10 per 100,000.

```{r}
mrr_raw <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`MRR`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "a. Raw", x = "", y = "", fill = "Ratio") +
  scale_fill_distiller(palette = "RdYlGn", limits = c(0,16)) +
  theme(plot.title= element_text(hjust = 0.5))

ggsave(here("results/mrr_map.png"), units="in", width=10, height=4, dpi=300)
```

```{r}
map_1 <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Living in Poverty`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Living in Poverty", x = "", y = "", fill = "Percent")  +
  theme(plot.title= element_text(hjust = 0.5)) 
```

```{r}
map_2 <-ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Median Household Income`/1e4)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Median Household Income", x = "", y = "", fill = "Per $10,000")  +
  theme(plot.title= element_text(hjust = 0.5))
```

```{r}
map_3 <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Uninsured Under 65 Years Old`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Uninsured Under 65 Years Old", x = "", y = "", fill = "Percent") +
  theme(plot.title= element_text(hjust = 0.5)) 
```

```{r}
map_4 <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Without an Internet subscription`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Without an Internet Subscription", x = "", y = "", fill = "Percent") +
  theme(plot.title= element_text(hjust = 0.5)) 
```

```{r}
map_5 <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Living with Food Insecurity`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Living with Food Insecurity", x = "", y = "", fill = "Percent") +
  theme(plot.title= element_text(hjust = 0.5)) 
```

```{r}
map_6 <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percentage Diagnosed Late`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percentage Diagnosed Late", x = "", y = "", fill = "Percent") +
  theme(plot.title= element_text(hjust = 0.5)) 
```

```{r}
library(gridExtra)
library(ggpubr)
grid_map <- grid.arrange(map_2, map_1, map_3, map_4, map_5, map_6, ncol = 2, nrow = 3)
grid_map <- annotate_figure(grid_map, top = text_grob("Maps of Selected Predictors, United States, 2020", 
               color = "black", face = "bold", size = 14))

ggsave(here("results/grid_map.png"), grid_map, units="in", width=11, height=10, dpi=300)
```
`

Next, let's map a probability map based on where we'd expect deaths if risk of HIV-related death was consistent across the US.
```{r}
ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = MRR_null_loc)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  scale_fill_distiller(palette = "RdYlGn", limits = c(0,16)) + theme_void() + 
  theme(strip.text = element_text(size = 14),
    strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Expected Black-to-White MRR with Null Effect of Location, 
       United States, 2020", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))
```
Next, let's map a probability map based on where we'd expect deaths if risk of HIV-related death was consistent across the US and race.
```{r}
ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = `Null Expected Rate`/`Null Expected Rate`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  scale_fill_distiller(palette = "RdYlGn") + theme_void() + 
  theme(strip.text = element_text(size = 14),
    strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Expected Black-to-White MRR with Null Effect of Location and Race, United States, 2020", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))
```

Next, let's using Dykes and Unwin's method of smooth map using a chi square value. 
```{r}
ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = dykes_chi)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + theme_void() + 
  theme(strip.text = element_text(size = 14),
        strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Dykes and Unwin's Chi Square by Race/ Ethnicity, United States, 2020", x = "", y = "", fill = "Chi Square") +
  scale_fill_distiller(palette = "RdYlGn") +
  theme(plot.title= element_text(hjust = 0.5))

ggsave(here("results/chi_map.png"), units="in", width=10, height=4, dpi=300)
```
##Dykes and Unwin’s chi-square statistic
Dykes and Unwin (Dykes and Unwin 2001) technique  has the effect of creating a distribution of values closer to normal (as opposed to a Poisson distribution of rates and counts encountered thus far). We can therefore apply a diverging color scheme where green hues represent less than expected rates and red hues represent greater than expected rates.

```{r}
##Smoothed map using fitted values from best model (CAR)
car_smooth <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = car_fit)) +
   scale_fill_distiller(palette = "RdYlGn", limits = c(0,16)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)  + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "b. Smooth (fitted)", x = "Ratio", y = "", fill = "") +
  theme(plot.title= element_text(hjust = 0.5))
ggsave(here("results/car_smooth.png"))
```

Next, let's look at the map for the ratio of predicted to observed (P/O) values. 
```{r}
car_pred <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = predicted_observed_car)) +
  scale_fill_distiller(palette = "RdYlGn", limits = c(0,3)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "c. Predicted/ Observed", x = "", y = "", fill = "") +
  theme(plot.title= element_text(hjust = 0.5))
ggsave(here("results/car_pred.png"))
```

```{r}
grid_mrr <- grid.arrange(mrr_raw, car_smooth, car_pred, ncol = 3, nrow = 1)

grid_mrr <- annotate_figure(grid_mrr, top = text_grob("Maps of Black-to-White HIV-Related Mortality Rate Ratio, United States, 2020", 
               color = "black", face = "bold", size = 14))
ggsave(here("results/grid_mrr.png"), grid_mrr, units="in", width=10, height=4, dpi=300)
```
We can see that the model fitted pretty well. A higher P/O ratio (more red values) indicates that the model overestimated the observed values, while a lower P/O ratio (more green values) indicates that the model underestimates the observed values.

### 2023-05-05 code/processing_code/1_cleaning_2020.Rmd
submitting final changes

@@
 deaths_sex <- deaths %>%
-  select(1:4, contains("male"), -contains("and"), -ends_with("Stability"))
+  dplyr::select(1:4, contains("male"), -contains("and"), -ends_with("Stability"))
 
 deaths_race <- deaths %>%
-  select(1:4, contains(c("Black", "White", "Hispanic",  "Multiple")), -ends_with("Stability"))
+  dplyr::select(1:4, contains(c("Black", "White", "Hispanic",  "Multiple")), -ends_with("Stability"))
 
@@
   filter(!grepl("PR", `State Abbreviation`)) %>% 
-  select(1:3, contains(c("male", "Black", "White", "Hispanic", "Multiple")), -contains(c("and", "Stability")))
+  dplyr::select(1:3,
+                contains(c("male", "Black", "White", "Hispanic", "Multiple")),
+                -contains(c("and", "Stability")))
 
 sex_dx <- new_dx %>% 
-  select(1:7)
+  dplyr::select(1:7)
 
 new_dx <- new_dx %>% 
-  select(-contains(c("male", "Year")))
+  dplyr::select(-contains(c("male", "Year")))
 ```
@@
 race_pop <- race_pop %>% 
-  select(c(Location, starts_with(c("Total:", "Black or African American alone", "Hispanic", "White alone", "Population of two or more races:")))) %>% 
+  dplyr::select(c(Location, starts_with(c("Total:", "Black or African American alone", "Hispanic", "White alone", "Population of two or more races:")))) %>% 
   rename(c(Hispanic = `Hispanic or Latino`,
@@
 total_pop <- race_pop %>%
-  select(1:2)
+  dplyr::select(1:2)
 ```
@@
   filter(!grepl("PR", `State Abbreviation`)) %>% 
-  select(1:3, contains("Percent of")) %>% 
-  select(1:3, contains(c("Hispanic", "Black", "White", "Multiple","male")))
+  dplyr::select(1:3, contains("Percent of")) %>% 
+  dplyr::select(1:3, contains(c("Hispanic", "Black", "White", "Multiple","male")))
 
@@
 rec_care <- rec_care %>%
-  select(1:3, contains(c("Hispanic", "Black", "White", "Multiple")))
+  dplyr::select(1:3, contains(c("Hispanic", "Black", "White", "Multiple")))
 
@@
 v_sup <- v_sup %>%
-  select(1:3, contains(c("Hispanic", "Black", "White", "Multiple")))
+  dplyr::select(1:3, contains(c("Hispanic", "Black", "White", "Multiple")))
 race_df <- inner_join(race_df, v_sup)
@@
 prev_sex <- prev %>%
-  select(1:4, contains("male"), -contains("and"), -ends_with("Stability"))
+  dplyr::select(1:4, contains("male"),
+                -contains("and"),
+                -ends_with("Stability"))
 
@@
 prev_race <-   prev %>%
-  select(1:4, contains(c("Black", "White", "Hispanic",  "Multiple")), -contains("and"), -ends_with("Stability"))
+  dplyr::select(1:4, contains(c("Black", "White", "Hispanic",  "Multiple")),
+                -contains("and"),
+                -ends_with("Stability"))
 
@@
 no_internet <- internet %>% 
-  select(c(1, "Without an Internet subscription")) %>% 
+  dplyr::select(c(1, "Without an Internet subscription")) %>% 
   slice(c(-1:-8, -48:-52,-263:-267)) %>% 
@@
 no_internet <- no_internet %>% 
-  select(State, `Percent Without an Internet subscription`) %>% 
+  dplyr::select(State, `Percent Without an Internet subscription`) %>% 
   drop_na()
@@
 ```
-
 We have some more cleaning to, but let's save these files for now. 

### 2023-05-05 code/processing_code/2_cleaning_race.Rmd
submitting final changes

```{r, echo=FALSE, message=FALSE}
library(knitr)
library(tidyverse)
library(here)
```

```{r}
race_df <- readRDS(here("data/processed_data/race_df.rds"))
race_df <-  race_df %>%  replace(is.na("Black Death Cases"), 0) %>% 
   replace(is.na("White Death Cases"), 0) 
```

Next, let's find the expected number of HIV deaths. First, we'll calculate lambda. 
```{r}
lambda_black <- race_df %>%
  drop_na(`Black Death Cases`) %>% 
  summarize(lambda_black = sum(`Black Death Cases`)/sum(Black_pop)) %>%
  as.numeric()

lambda_white <- race_df %>%
  drop_na(`White Death Cases`) %>% 
  summarize(lambda_white = sum(`White Death Cases`)/sum(White_pop)) %>%
  as.numeric()
```

```{r}
black_df <- race_df %>% 
  dplyr::select(1:3, contains("Black"), 50:64) %>% 
  mutate(black_death_rate = round(`Black Death Cases`/Black_pop*1e5,1))

black_df <- black_df %>% 
    group_by(State) %>% 
 rename("Death Rate (AIDSVu)" = `Black Death Rate`,
        "Population" = Black_pop,
        "Death Cases" = "Black Death Cases",
        "New Diagnoses Rate" = "New Diagnoses Black Rate",
        "New Diagnoses Cases" = "New Diagnoses Black Cases",
        "Percentage Diagnosed Late" = `Percent of Black Persons Diagnosed Late`,
        "Number of Persons Diagnosed Late" = `Number of Black Persons Diagnosed Late`,
        "Percentage Received Care" = `Percent of Black Persons Living with HIV who Received HIV Care`,
        "Percentage Virally Suppressed" = `Percent of Black Persons Living with HIV who were Virally Suppressed`,
        "HIV Prevalence Rate" = "Black Rate",
        "Prevalent HIV Cases" = "Black Cases",
        "Death Rate (estimated)" = black_death_rate) %>% 
  mutate(Race = "Black",
        "Expected Death Cases" = round(lambda_black * Population),
        "Expected Death Rate (Race)" = round(`Expected Death Cases`/`Population` *1e5,1)) %>% 
  ungroup()

white_df <- race_df %>% 
  dplyr::select(1:3, contains("White"), 50:64) %>% 
   mutate(white_death_rate = round(`White Death Cases`/White_pop*1e5,1))

white_df <- white_df %>% 
  group_by(State) %>% 
 rename("Death Rate (AIDSVu)" = `White Death Rate`,
        "Population" = White_pop,
        "Death Cases" = "White Death Cases",
        "New Diagnoses Rate" = "New Diagnoses White Rate",
        "New Diagnoses Cases" = "New Diagnoses White Cases",
        "Percentage Diagnosed Late" = `Percent of White Persons Diagnosed Late`,
        "Number of Persons Diagnosed Late" = `Number of White Persons Diagnosed Late`,
        "Percentage Received Care" = `Percent of White Persons Living with HIV who Received HIV Care`,
        "Percentage Virally Suppressed" = `Percent of White Persons Living with HIV who were Virally Suppressed`,
        "HIV Prevalence Rate" = "White Rate",
        "Prevalent HIV Cases" = "White Cases",
        "Death Rate (estimated)" = white_death_rate) %>% 
  mutate(Race = "White",
        "Expected Death Cases" = round(lambda_white * Population),
        "Expected Death Rate (Race)" = round(`Expected Death Cases`/`Population` * 1e5, 1)) %>% 
  ungroup()

race_df <- list(black_df, white_df)

race_df <- reduce(race_df, full_join)

race_df <- race_df %>% 
  mutate(`Percent HIV Prevalence` = round(`Prevalent HIV Cases`/ `Population`*100,2))
```

Now, let's calculate lambda - null, which would assume that there is no risk difference in Expected deaths by race or location
```{r}
lambda_null <- race_df %>%
  drop_na("Death Cases") %>% 
  summarize(lambda = sum(`Death Cases`)/sum(Population)) %>%
  as.numeric()

race_df <- race_df %>% 
  mutate(`Expected Death Cases (Null)` = Population * lambda_null,
         `Difference Observed - Expected` = `Death Cases` - `Expected Death Cases`,
         `Ratio Observed/ Expected (Null)` = `Death Cases`/`Expected Death Cases (Null)`,
         `Ratio Observed/ Expected (Race)` = `Death Cases`/`Expected Death Cases`,
         `Null Expected Rate` = `Expected Death Cases (Null)`/ Population*1e5,
         `Percentage Diagnosed Late` = replace(`Percentage Diagnosed Late`, which(`Percentage Diagnosed Late` < 0), 0))
```

Let's create a variable for mortality rate ratio of Black/ White
```{r}
race_df <- race_df %>%
  group_by(State) %>%
  mutate(MRR = `Death Rate (estimated)`[Race == "Black"] / `Death Rate (estimated)`[Race == "White"]) %>%
  ungroup()

#Calculate MRR with null effect of LOCATION (since lambda was calculated using sum cases and population for each race)
race_df <- race_df %>%
  group_by(State) %>% 
  mutate(MRR_null_loc = `Expected Death Rate (Race)`[Race == "Black"] / `Expected Death Rate (Race)`[Race == "White"]) %>%
  ungroup()

# MRR with null effect of LOCATION and RACE would equal 1...
```

Save the file and now we can move on to some exploratory analysis!
```{r}
saveRDS(race_df, here("data/processed_data/race_df_alt.rds"))
```

### 2023-05-05 products/manuscript/manuscript.qmd
hopefully this corrects things...
submitting final changes

--- 2023-04-26 products/manuscript/manuscript.qmd
@@
 ---
-title: "Spatial Investigation and Exploration of Factors influencing HIV-related Mortality"
+title: "Investigating the Association between Social Determinants of Health and Racial Disparities in HIV-related Mortality Rates Among non-Hispanic Black and non-Hispanic White Persons"
 subtitle: "United States, 2020"
@@
 ```{r, echo=FALSE, warning = FALSE, message=FALSE}
-# load a few R packages
+#packages used
+library(here)
+library(ggfortify)
 library(knitr)
+library(kableExtra)
+library(ggpubr)
+library(ggcorrplot)
+library(FactoMineR)
+library(factoextra)
+library(broom)
+library(mfx)
+library(spatialreg)
+library(spdep)
+library(tidyverse)
+library(AER)
+library(RColorBrewer)
 library(readxl)
+library(reshape2)
 library(lubridate)
-library(stringr)
-library(tidyverse)
-library(here)
-library(kableExtra)
 ```
 
-# Summary/Abstract
+{{< pagebreak >}}
+
+# Summary
 
@@
 To assess this, we look at the ratio of Black-to-White HIV-related mortality rate (MRR) from 2020 in the United States.
-Additionally, we assess whether social determinants of health (SDOH) can explain variation in MRR at the state-level and regional-level.
-Both spatial and non-spatial methods were used.
+Additionally, we assess whether social determinants of health (SDOH) can explain variation in MRR at the state-level by employing spatial and non-spatial methods.
 
@@
 
-HIV continues to have a significant impact on marginalized communities, leading to significant disparities in the rates of new HIV diagnoses and HIV-related health outcomes, including late HIV diagnosis, progression to AIDS, and HIV-related death.
-The extent of these disparities are particularly pronounced in the United States when comparing HIV-related outcomes among Black and White persons.
-Studies have revealed a direct correlation between social determinants of health indicators and poor health outcomes for people with HIV.
-In this project, our objective is to identify the factors contributing to the disparities in HIV-related death rates between Black and White individuals in the United States.
-
-## Question:
-
-Can the ratio of Black-to-White HIV-related mortality rate (MRR) from 2020 in the United States be modeled by social determinants of health or by HIV-related outcomes (late diagnoses, viral suppression).
-Can a spatial model explain variation in MMR at the state-level and regional-level after accounting for variables associated SDOH or HIV-related outcomes?
-
-H0 : There is no spatial correlation between Black-to-White HIV-related MRR and SDOH or HIV-related outcomes.
-
-Ha : There is a spatial correlation between Black-to-White HIV-related MRR and SDOH or HIV-related outcomes.
+The impact of human immunodeficiency virus (HIV) on marginalized communities in the United States continues to be a significant public health concern.
+Of particular concern is the significant disparity in HIV-related health outcomes between non-Hispanic Black and non-Hispanic White persons, including the rates of new HIV diagnoses, late HIV diagnoses, and HIV-related death.
+Research has shown a direct correlation between social determinants of health (SDOH) indicators and poor health outcomes for individuals with HIV.
+In this project, our aim is to investigate the factors contributing to the disparities in HIV-related death rates between Black and White individuals in the US.
+Specifically, we will examine whether the Black-to-White HIV-related mortality rate (MRR) from 2020 can be modeled by SDOH and late diagnoses.
+Additionally, we will explore whether a spatial model can explain variation in MRR at the state level after accounting for variables associated with SDOH or HIV-related outcomes.
+Our null hypothesis is that there is no spatial correlation between Black-to-White HIV-related MRR and SDOH or HIV-related outcomes, while our alternative hypothesis suggests that there is indeed a spatial correlation.
 
@@
 
-Since the emergence of the HIV epidemic in 1981 in the United States, more than 1.2 million people in the United States have been diagnosed with HIV, resulting in more than 700,000 HIV-related deaths [@jun07HIVAIDSEpidemic2021].
-Medical advancements have led to effective antiretroviral therapies that not only treat HIV infection and reduce adverse HIV-related health outcomes, such as progression to AIDS, but have also led to use of these medications to prevent infection before and after potential exposure for people who are HIV negative Even still, more than 30,000 people in the United States continue to be diagnosed each year [@jun07HIVAIDSEpidemic2021].
+Since the emergence of the epidemic in 1981 in the United States, more than 1.2 million people in the country have been diagnosed with human immunodeficiency virus (HIV), resulting in more than 700,000 HIV-related deaths [@jun07HIVAIDSEpidemic2021].
+The virus encodes three enzymes that allow it to escape host immunity and results in chronic infection that, without treatment, ultimately leads to acquired immunodeficiency syndrome (AIDS).
+Medical advancements have led to development effective antiretroviral therapies, many of which disrupt the function of these key enzymes, and are capable of not only treating HIV infection by reducing adverse HIV-related health outcomes, including slowing progression to AIDS, but are also capable of preventing infection before and after potential exposure for people who are HIV negative.
+Even still, more than 30,000 people in the United States continue to be diagnosed each year [@jun07HIVAIDSEpidemic2021].
 While the Ending the Epidemic initiative set up a goal of 90% reduction in new HIV infections by 2030 by upscaling efforts in HIV diagnoses treatment, and prevention, this is followed by nearly one decade of steady HIV diagnoses rates.
@@
 Data was obtained from AIDSVu [@AIDSVuDatasetsIncludes] on counts and rates of HIV-related mortality by state in 2020.
+It is important to distinguish that these deaths are specifically the result of HIV infection, and do not represent all deaths that occurred in 2020 in the United States among persons with HIV.
 Social determinants of health (SDOH), were also obtained from AIDSVu by state from 2020.
-SDOH included a number of variable represented as a percentage particularly pertinent to health persons with HIV.
+SDOH included a number of variables represented as a percentage particularly pertinent to health persons with HIV.
 AIDSVu is a public resource presented by Emory University's Rollins School of Public Health in partnership with Gilead Sciences, Inc.
@@
 
-Data was cleaned to adjust variable format, merge dataframes, subset by race/ ethnicity, and for feature engineering.
-Several observations in response variables were suppressed due to small counts/ rates were represented by `-9`, and were re-coded to NA.
-To removed multicollinearity and to assess the relationship between predictors, principal component analysis (PCA).
-was used.
-Variables that had correlation greater that 0.3 in the first two principal components were deemed important.
-Scaled predictors of the the first three principal components were then used in various regression models multiple linear regression, Poisson regression, and negative binomial.
-Non-spatial and spatial methods were used.
+In the data cleaning process, various tasks were performed to ensure the data were ready for analysis.
+Firstly, variables were reformatted and dataframes were merged as necessary.
+Additionally, several observations in response variables were suppressed due to small counts/ rates were represented by `-9`, and were re-coded to `0`, rather than `NA`, as some analysis methods did not permit missing observations.
+Feature engineering was also employed, most notably for our primary outcome variable, the Black-to-White HIV-related death rate ratio (MRR).
 
@@
 The primary response variable in this analysis is the ratio of Black-to-White HIV-related death rate (MRR).
-To compute this variable, feature engineering was performed to create `death rate (estimated)` stratified by race (Black and White) using counts of HIV related deaths divided by the estimated population based on US Census data multiplied by 100,000 persons.
-The variable `MRR` was then computed for each state by dividing the estimate death rates for Black persons divided by the estimate death for White persons.
+To compute this variable, feature engineering was performed to create `death rate (estimated)` stratified by race (non-Hispanic Black and non-Hispanic White) using counts of HIV related deaths divided by the estimated population based on US Census data multiplied by 100,000 persons.
+Then, I computed `MRR` by obtaining the ratio of the estimated HIV-related death rate for non-Hispanic Black persons divided by the estimate HIV-related death rate for non-Hispanic White persons.
 
-### Social Determinants of Health
+### Predictors
 
-Predictors used in this analysis were selected to serve as proxy for SDOH.
-A number of variables were initially considered, but ultimately five predictors were selected: the Gini coefficient was used measure income inequality; percent living in poverty was used to estimate economic stability; percent uninsured under 65 was used to measure health care access; and percent without internet subscription was used to estimate the neighborhood and built environment; and percent living with food insecurity.
+Predictors used in this analysis were selected to serve as indicators for various aspects relating to SDOH, including access to healthcare, the built environment, and economic stability.
+A number of variables were initially considered, but ultimately five predictors were selected: the median household income (in US dollars); percent living in poverty (% of population); percent uninsured under 65 (% of population); percent without internet subscription(% of population); and percent living with food insecurity (% of population).
+Percentage diagnosed late is the percentage of persons diagnosed with HIV who progressed to AIDS within 12 months of diagnosis.
 
-## Statistical Analysis
+### Statistical Analysis
 
-The goal of this analysis was to assess the spatial distribution of the Black-to-White HIV-related MRR.
-A number of predictors were analyzed using simple linear regression to predict MRR.
-Variables with a correlation coefficient of \|-0.15\| or more were selected as possible predictors.
-To address the possibility of multicollinearity between predictors, principal component analysis was then performed on these variable to aid in variable reduction.
+The primary objective of this analysis was to examine the spatial distribution of the Black-to-White HIV-related MRR.
+Simple linear regression was conducted to explore possible predictors of MRR, selecting variables with a correlation coefficient of \|-0.15\| or more.
+Principal component analysis was used to reduce dimensionality and remove multicollinearity.
 
-Both non-spatial and spatial models were fitted to data as appropriate.
-Analysis of the full models began with multiple linear regressions model before moving on to models of increasing complexity, such as Poisson regression and the conditional autoregressive (CAR) model.
-Models that were flexible to continuous data used `MRR` as the primary outcome, while models that required count data used the number of deaths, `Death Cases` , and used `log(Population)` as the offset, and additionally included variables `Race` and `Region` in the models.
+To test for spatial correlation, Moran's I was employed.
+One assumption of using Moran's I in hypothesis testing is that the data follow a normal distribution.
+Normality was assessed visually using a histogram and verified with a Shapiro-Wilke test.
+The analysis utilized the following neighborhood structure:
 
-Model comparison and relative goodness of fit of the various models used in this analysis was assessed using the Akaike information criterion (AIC).
+hij { 1, if counties i and j share a county line; 0, if otherwise.
 
-Moran's *I* was computed to test the null hypothesis of spatial independence.
+Several regression models were employed, including multiple linear regression, negative binomial regression, and the conditional autoregressive model.
 
-# Analysis
+Multiple linear regression requires several key assumptions to be met, including linearity, normality of the response variable (MRR), and addressing the issue of multicollinearity.
+These assumptions were checked using scatterplots, visual assessment, Shapiro test, and PCA.
 
-## Univariate analysis
+Given that the case data for HIV-related mortality tends to be heavily right-skewed, a Poisson regression model was evaluated, which assumes that the variance and mean are equal.
+However, when the conditional variance is greater than the conditional mean, overdispersion occurs.
+In this case, the negative binomial regression model, which can accommodate overdispersion, was considered.
 
-Our primary outcome of interest was the Black-to-White HIV-related MRR, which was computed as described above.
+Finally, the conditional autoregressive (CAR) model, a type of linear regression model that can account for spatial association between neighbors, was employed.
+The CAR model's assumptions are conditional on the data at remaining site {*z~i~* ∶ *j* ≠*i*}, are normally distributed data Z(*s~i~*), and there are only pairwise spatial interactions.
+
+To compare models and assess relative goodness of fit, the Akaike information criterion (AIC) was used, where smaller values indicate better fit.
+
+# Results
+
+## Univariate Analysis
+
+Our primary outcome of interest was the Black-to-White HIV-related MRR, which was computed for each state by dividing the estimate death rates for non-Hispanic Black persons divided by the estimate death for White persons.
 Table 1 shows the overall number of HIV-related deaths, population size older than 12 years, death rate and the MRR in the United States in 2020.
@@
 
-```{r, echo=FALSE, message=FALSE, fig.cap="Table 1."}
+```{r, echo=FALSE, message=FALSE, fig.cap="Table 1. Population Charactertistics"}
 include_graphics(here("results/table1.png"))
@@
 
-```{r, echo=FALSE, message=FALSE, fig.cap="Figure 1."}
+To assess the distribution of Black-to-White HIV-related MRR in the US in 2020, we conducted a visual assessment ( @fig-include-MRR ) and a Shapiro-Wilks test.
+Our results showed that the distribution followed a normal distribution `(p= 0.63)`, and no normalizing transformations were needed.
+@fig-include-DR shows the histogram of the raw HIV-related death rate by race, which highlights significant racial disparities in risk of HIV-related death.
+
+```{r, echo=FALSE, message=FALSE, fig.width= 3, fig.cap="Histogram of Black-to-White HIV-Related Mortality Rate Ratio", label = fig-include-MRR}
 include_graphics(here("results/MRR_hist.png"))
@@
 
-Visual assessment (Figure 1) and Shapiro test were congruent that the distribution of Black-to-White HIV-related MRR in the US in 2020 followed a normal distribution (*p* = 0.63).
-No normalizing transformation was needed.
-Below is the histogram of the our primary outcome variable, MRR.
-
-After verifying that the Black-to-White HIV-related MRR met assumptions of normality, I tested for spatial association of the Black-to-White MRR by computing Moran's *I* (*`I`* `= 0.1507`).
-For uncorrelated data, the expected value of Moran's *I* was `E(I) = - 0.0213` with a variance `var(I) = 0.0095`.
-The data showed statistically significant evidence of spatial dependence `(z =1.768; p = 0.039)`.
-Figure 2 shows a Moran's scatter plot of spatially lagged Black-to-White HIV-related MRR and Black-to-White HIV-related MRR.
-The positive slope indicates that there is a degree of spatial clustering of the Black-to-White HIV-related MRR.
-
-```{r, echo=FALSE, message=FALSE, fig.width = 4, fig.cap="Figure 2. Moran's Scatter Plot"}
-include_graphics(here("results/moran_mrr.png"))
+```{r, echo=FALSE, message=FALSE, fig.width= 3, fig.cap="Histogram of HIV-Related Death Rates", label = fig-include-DR}
+include_graphics(here("results/DR_hist.png"))
 ```
 
-Figure 3 and 4 show the raw HIV-related death rate by Race and State.
+To further explore the racial disparities in HIV-related mortality, we compared the Black-to-White HIV-related mortality rate ratio (MRR) by state ( @fig-include-state ) and by region ( @fig-include-region ).
+Four states, namely Idaho, Montana, Wyoming, and Vermont, did not have an estimate for MRR due to small or suppressed rates.
+The results from these figures reveal distinct disparities between the rate of HIV-related death by race and region.
+States in the Northeast region had the highest median Black-to-White HIV-related MRR, with a ratio of approximately 8.5, while states in the West region had the lowest median Black-to-White HIV-related MRR, at approximately 4.0.
 
-![Figure 3](images/dr_state-01.png){fig-align="center" width="600"}
-
-```{r, echo=FALSE, message=FALSE, fig.cap="Figure 3. Bargraph of Raw HIV-Related Death Rates by Race by State, 2020, United States"}
-include_graphics(here("results/dr_state.png"))
+```{r, echo=FALSE, message=FALSE, fig.cap="Bargraph of Black-to-White HIV-Related MRR", label = fig-include-state}
+include_graphics(here("results/mrr_state.png"))
 ```
 
-```{r, echo=FALSE, message=FALSE, fig.cap="Figure 4. Boxplot of HIV-Related Death by Race by Region, 2020, United States"}
-include_graphics(here("results/dr_region.png"))
-```
-
-Figure 5 compares the raw HIV-related death rate by Race and Region.
-These plots again show distinct disparities between rate of HIV-related death by race.
-Additionally, there is evidence of disparities in HIV-related death rate by region for Black persons.
-
-Table 2 shows maps of selected predictors in the United States in 2020.
-We can see that the `Gini Coefficient` was generally higher in Southern half of the country, indicating a greater degree of income inequality.
-`Percent Uninsured Under 65 Years Old` was highest in Texas, with the South and non-coastal Western states also appearing to have higher rates of uninsured.
-`Percent Without an Internet Subscription` was generally low through the United States, however, Wyoming had the highest percent without Internet subscription.
-
-```{r, echo=FALSE, message=FALSE, fig.cap="Table 2", fig.align = "default"}
-include_graphics(here("results/grid_map.png"))
+```{r, echo=FALSE, message=FALSE, fig.cap="Boxplot of Black-to-White HIV-Related MRR, by Region", label = fig-include-region}
+include_graphics(here("results/mrr_region.png"))
 ```
@@
 
-To narrow down potential predictors to include in later models, I first started by performing simple linear regression models of MRR vs one of the possible predictors.
-Variables that had a correlation coefficient of \|0.15\| or more were then selected as possible variables for analysis.
-These possible predictors were then scaled and principal component analysis (PCA) was performed.
+Bivariate analysis and principal component analysis (PCA) were conducted to identify potential predictors associated with HIV-related MRR.
+Firstly, bivariate analysis was used to narrow down the range of predictors, where only variables with a correlation coefficient of \|0.15\| or more were included in the next step.
+@fig-include-bivar displays the results of the bivariate analysis, which includes the following predictors: `Percent Living in Poverty`, `Median Household Income`, `Percent Uninsured Under 65 Years Old`, `Percent Living with Food Insecurity`, `Percent Without an Internet Subscription`, and `Percentage Diagnosed Late`.
 
-```{r, echo=FALSE, message=FALSE, fig.cap="Table 2. Simple Linear Regression between Outcome and Individual Predictors", fig.align = "default"}
+```{r, echo=FALSE, message=FALSE, fig.cap="Simple Linear Regression of Outcome vs. Select Predictors", fig.align = "default",label = fig-include-bivar}
 include_graphics(here("results/grid_bivar.png"))
@@
 
-Table 3. shows the plot from the bivariate analysis of predictors that were scaled and included in the PCA. Table 3 shows the results of PCA.
+Following the bivariate analysis, PCA was performed on the selected predictors.
+The PCA was based on a correlation matrix and resulted in a scaled and centered matrix.
+@fig-include-pca shows the results of the PCA, where only principal components (PCs) that individually explained more than 16.7% variation were included in the analysis.
+PC1 accounted for 56.6% of variation, while PC2 accounted for 17.1% of variance.
+All other PCs were excluded from the analysis.
+PC1 had strong positive associations with median household income and negative associations with percent living in poverty, percent without an internet subscription, and percent living with food insecurity.
+PC2 showed very strong negative associations with the percentage diagnosed late.
 
-```{r, echo=FALSE, message=FALSE, fig.cap="Table 3", fig.align = "default"}
+```{r, echo=FALSE, message=FALSE, fig.cap="Results of Principal Component Analysis", fig.align = "default",label = fig-include-pca}
 include_graphics(here("results/grid_pca.png"))
@@
 
-Results of the PCA shows that the first principal component contributed for 91.7 % of variation between predictors.
-The variables `Percent Uninsured Under 65 Years Old` explained the least amountof variation.
-Ultimately, only `Median Household Income`, `Percent Living in Poverty`, `Percent Living with Food Insecurity`, and `Percent Without an Internet Subscription` were included as predictors in the full model.
+@fig-include-map provides a visual representation of the selected predictors using maps.
+The South generally had higher percentages of persons living in poverty, living with food insecurity, and persons uninsured under 65 years old, and had lower median household income.
+The percentage of persons diagnosed late was consistent throughout much of the United States.
+Interestingly, states with the highest percentage of persons with HIV who were diagnosed late were neighbored by states with the lowest rates.
+The percentage without an internet subscription was generally low throughout the United States, but was highest in Wyoming.
 
-## Full analysis
-
-```{r, echo=FALSE, message=FALSE, fig.cap="Figure.4"}
-include_graphics(here("results/car_fit_map.png"))
+```{r, echo=FALSE, message=FALSE, fig.cap="Maps of Selected Proxies for Social Determinants of Healths", fig.align = "default", label = fig-include-map}
+include_graphics(here("results/grid_map.png"))
 ```
 
-```{r, echo=FALSE, message=FALSE, fig.cap="Figure.5"}
-include_graphics(here("results/mult_fit_map.png"))
+## Full Models
+
+### Multivariate Linear Model
+
+The results of the multivariate linear model indicate a statistically significant positive association between Black-to-White HIV-related MRR and PC1 (*`p`* `< 0.005`), as well as a statistically significant negative association between MRR and PC2 (*`p`* `= 0.038`).
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Table 2. Summary of Multivariate Linear Model", fig.align = "default"}
+include_graphics(here("results/mult_table.png"))
 ```
 
-```{r, echo=FALSE, message=FALSE, fig.cap="Figure.6"}
-include_graphics(here("results/grid_pred.png"))
+### Negative Binomial Regression Model
+
+The negative binomial regression model was selected over Poisson regression model due to overdispersion (*`c`* `= 17.35`).
+In order to use this regression method, I use the number of HIV-related deaths in each state using log(Population) as an offset, and race, PC1 and PC2 as predictors, using non-Hispanic White persons as the reference group.
+Regression coefficients were exponentiated, resulting in rate ratios.
+The HIV-related MRR was 6.46 times higher among non-Hispanic Black persons than non-Hispanic White persons `(95% CI = (5.28, 7.93); z = 17.96; p` `< 0.005)`.
+The Black-to-White HIV-related MRR decreased with increasing components of PC1, while Black-to-White HIV-related MRR increased with increasing components of PC2, but these results were not statistically significant (*`p`* `= 0.13` and *`p`* `= 0.14` , respectively).
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Table 3. Summary of Negative Binomial Model (exponentiated)", fig.align = "default"}
+include_graphics(here("results/nb_table.png"))
 ```
 
-The CAR model accounting for spatial variation only performed marginally better than the GLM model that did not account for spatial variation.
+### Conditional Autoregressive Model (CAR)
+
+The conditional autoregressive model (CAR) indicated a statistically significant positive association between Black-to-White HIV-related MRR and PC1, (*`p`* `< 0.005`), after accounting for spatial dependence, while evidence of a negative association between Black-to-White HIV-related MRR and PC2 was not statistically significant (*`p`* `= 0.38`).
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Table 4. Summary of CAR Model", fig.align = "default"}
+include_graphics(here("results/car_table.png"))
+```
+
+## Results
+
+In 2020, a total of 13,173 HIV-related deaths occurred among non-Hispanic Black and non-Hispanic white populations in the United States, with a Black-to-White HIV-related MRR of 7.2 indicating significant racial disparities.
+@fig-include-rate displays the estimated HIV-related death rate for non-Hispanic Black and non-Hispanic White persons in the United States in 2020, highlighting significant regional differences, with the highest rates observed among non-Hispanic Black persons, particularly in the South region.
+On the other hand, non-coastal western states consistently exhibited the lowest rates of HIV-related deaths across the selected racial/ethnic groups.
+Additionally, no state in 2020 had an HIV-related death rate higher than \~10 per 100,000 among non-Hispanic White persons.
+
+```{r, echo=FALSE, message=FALSE, fig.cap="HIV-Related Mortality Rate, by Race/Ethnicity", label = fig-include-rate}
+include_graphics(here("results/raw_rate_map.png"))
+```
+
+To evaluate the spatial dependence of the data, Moran's test was performed on Black-to-White HIV-related MRR.
+The results showed statistically significant evidence of spatial dependence `(z = 1.768; p = 0.039)`, with a Moran's *I* value of `0.1507`, indicating a degree of clustering of the Black-to-White HIV-related MRR.
+For uncorrelated data, the expected value of Moran's *I* was *`E(I)`* `= -0.0213` with a variance *`var(I)`* `= 0.0095`.
+@fig-include-moran shows a Moran's scatter plot of spatially lagged Black-to-White HIV-related MRR and Black-to-White HIV-related MRR, depicting a positive slope indicative of clustering.
+
+```{r, echo=FALSE, message=FALSE, fig.width = 3, fig.cap="Moran's Scatter Plot", label = fig-include-moran}
+include_graphics(here("results/moran_mrr.png"))
+```
+
+In the full analysis, PC1 and PC2, which represented multiple indicators of social determinants of health, along with the percentage of persons diagnosed late, were used to assess the associated racial disparity in HIV-related death rates.
+The CAR model, accounting for spatial variation in the data, performed marginally better than the non-spatial generalized linear model, with an AIC of 503.18.
+Despite the slight differences, considering the results of Moran's test, the CAR model was selected as the best-performing model, as use of the spatial model was also supported by evidence of clustering from by the Moran's test.
+Table 5 provides a comparison of AIC across models.
+
+The CAR model showed statistically significant evidence of a positive association between Black-to-White HIV-related MRR and PC1; higher levels of component 1 are associated with higher levels of the Black-to-White HIV-related MRR.
+Since PC1 is positively associated with median household income and negatively associated with poverty, lack of internet subscription, and food insecurity, this suggests that higher levels of income and lower levels of poverty, lack of internet access, and food insecurity are associated with higher Black-to-White HIV-related MRR.
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Table 5. Relative Comparison of Goodness of Fit", label = tab}
+include_graphics(here("results/table_fit.png"))
+```
+
+@fig-include-mrr shows maps comparing the raw Black-to-White HIV-related MRR, fitted values from the CAR model, and the ratio of predicted over observed values.
+States with zero or suppressed counts for HIV-related deaths among non-Hispanic Black persons (Idaho, Montana, Wyoming, Vermont) are represented in dark green.
+@fig-include-mrr (c) shows that the model performed relatively well in the Northeast and South, as indicated by the light greens, which represent values closer to 1.
+The West coast states are represented in orange, indicating that the model over-predicted the disparity between HIV-related MRR between Black and White persons in these regions.
+
+```{r, echo=FALSE, message=FALSE, fig.cap="Maps of Black-to-White HIV-Related Mortality Rate Ratio", label = fig-include-mrr}
+include_graphics(here("results/grid_mrr.png"))
+```
 
@@
 
-## Summary and Interpretation
+There are significant disparities in HIV-related mortality rates between non-Hispanic Black and non-Hispanic White populations in the United States.
+This study found spatial associations of this disparity with five indicators of SDOH: median household income, percent living in poverty, percent uninsured under 65, percent without internet subscription, and percent living with food insecurity.
+This analysis found evidence of a spatial association with higher Black-to-White HIV-related MRR and higher median household income and lower rates of poverty, Internet subscription, and food insecurity.
+While these findings were somewhat surprising, they helped reveal that HIV-related health disparities are complex, and other factors likely contribute to high HIV-related mortality rates among non-Hispanic Black persons.
 
-*Summarize what you did, what you found and what it means.*
+This study has several limitations that should be addressed.
+Firstly, while the Black-to-White HIV-related mortality rate ratio can be used to identifying areas with significant racial disparities in HIV-related mortality rates, it may not accurately reflect areas with where HIV deaths are especially high.
+Additionally, this study is vulnerable to ecological fallacy, as the data was aggregated at the state-level, which can potential mask individual-level relationships between the predictors and outcome variable.
+Furthermore, state-level data may not capture the variation in HIV-related health disparities within states, such as differences in risk factors and outcomes between urban and rural populations or those between different racial and ethnic groups.
+Another limitation is that the neighborhood structure used in this study may not capture the true relationships between neighboring counties.
+In Georgia, this might particularly be an issue, since county sizes and shapes differ significantly.
+Including additional neighborhood structures in the analysis, such as one that defines neighbors based on those that lay within *x* distance of one another, may be a better approach.
+Finally, the analysis is limited to the 2020 and to non-Hispanic Black and non-Hispanic White persons, and may not reflect changes over time or across other racial and ethnic groups.
 
-## Strengths and Limitations
+Strengths of this study included the use of multiple predictors to assess the relationship between SDOH and the Black-to-White HIV-related MRR.
+Additionally, use of various regression models, including non-spatial and spatial methods, allowed for a more comprehensive analysis of the data.
+The use of PCA helped to address multicollinearity issues among the predictors.
 
-*Discuss what you perceive as strengths and limitations of your analysis.*
-
-# Conclusions
-
-*What are the main take-home messages?*
-
-*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.
+This analysis highlights the need for targeted interventions to address HIV-related health disparities.
+Future research would greatly benefit from analyzing individual-level data to better understand factors contributing to HIV-related racial disparities, as well as utilizing additional neighborhood structures, and a broader range of years.
 
@@
 
-## Software
+# Software
+
 ```{r}

### 2023-05-05 readme.qmd
submitting final changes

@@
 
-This respository contains data and code to reproduce my research. Raw data is store in the `data` subfolder `raw_data`, while processed data produced by wrangling is stored under `processed_data`.
+This respository contains data and code to reproduce my project. Raw data is store in the `data` subfolder `raw_data`, while processed data produced by wrangling is stored under `processed_data`.
 
@@
 2.
-Open the folder code and start by running files in `processing_code`.
-The files are numbered sequentially, but `1_cleaning_2020` is the only file in this folder that must be run first.
+Open the folder `code` and start by running files in `processing_code`.
+The files are numbered sequentially, but `1_cleaning_2020` is the only file in this folder that must be run in order.
 The others may be run in any order.
-3.
-Open the folder code and start by running files in `analysis_code` , which contains sub-folders for running the analysis for race/ethnicity `race` and `sex` separately.
-`race` and `sex` may be run independently of each other, but contain numbers to indicate the order which these files should be run.
+3.  Next, run the files in `analysis_code` . The files are numbered sequentially in the order that they should be run.
 4.  Data files produced by these files are store in the `data` subfolder `processed_data`
+5.  Images and tables produced by these files are stored in the `results` folder.
+6.   The manuscript is stored in subfolder `manuscript` within the `products` folder.
